### 2020-01-27 paper/manuscript.Rmd
Update manuscript.Rmd

---
title: The problem and promise of scale in multilayer networks
author: Quinn M.R. Webber^1^, Jack Hendrix^1^, Alec L. Robitaille^2^,  Eric Vander Wal^1,2^
header-includes:  ^1^Cognitive and Behavioural Ecology Interdisciplinary Program, Memorial University of Newfoundland^2^; Department of Biology, Memorial University of Newfoundland, Canada.

output: 
    #word_document
    github_document
---

### 2020-01-27 paper/manuscript.Rmd
add ms to rmd

@@
 ---
-title: The problem and promise of scale in multilayer networks
+title: Interisland movements of woodland caribou in Newfoundland
 author: Quinn M.R. Webber^1^, Jack Hendrix^1^, Alec L. Robitaille^2^,  Eric Vander Wal^1,2^
@@
 ---
+
+## Abstract
+
+## Introduction
+Caribou are exceptional swimmers.
+Ample evidence exists that caribou swim in streams, rivers, and lakes during migration (Leblond et al.
+2016) to avoid predators (Bergerud 1985) and access islands during calving (Bergerud et al.
+1990).
+Despite a large number of coastal and island caribou herds, limited evidence exists suggesting that caribou swim in the ocean as readily as they swim in freshwater (see Table 1).
+A key exception is Peary caribou swimming up to 2.6km between the Queen Elizabeth Islands (Miller, 1995; Miller, 2002).
+For caribou living on small islands in the ocean, the same notions for why caribou swim in freshwater may also explain why caribou swim in the ocean (Leblond et al.
+2016).
+Specifically, forage limitation (Miller, 2002) and predator avoidance (Jeffery et al.
+2007) have been proposed as potential explanations.
+
+Swimming for quadrupedal terrestrial mammals can be thought of as a set of trade-offs.
+Costs include high energy expenditure and risk of drowning.
+Specifically, even though caribou are adept swimmers, the energetic expenditure associated with swimming for quadrupedal mammals is significantly higher than walking or running (Fish 1993).
+Additionally, drowning is also possible (Miller & Gunn, 1985) although the proportion of caribou that die from drowning is unknown.
+Swimming between islands or from the mainland to nearby islands can be beneficial.
+Among the most important benefits include avoiding predators and seeking under-foraged areas.
+
+Forage limitation could explain movement between islands.
+Forage scarcity has been proposed as a potential reason why caribou move between arctic islands on the sea-ice in winter (Miller et al.
+1977).
+For very small islands, over-grazing by newly arrived caribou may rapidly limit forage (Bergerud et al.
+1990).
+Without sea ice, movement from the mainland to an island as well as between islands requires caribou to assess the trade-off associated with swimming.
+The underlying mechanism driving forage limitation and the trade-off associated with swimming could be density-dependent habitat selection governed by the Ideal Free Distribution (Morris 1987; Bradbury et al.
+2015).
+Ideal Free Distribution theory suggests that habitat selection is density-dependent and that variation in density between habitat patches leads to a fitness equilibrium (Bradbury et al.
+2015).
+Fitness in a habitat patch depends on density where, ideally, the available resources on a habitat patch can sustain a specific number of individuals.
+In a hypothetical example, when density of a given habitat patch has exceeded the fitness equilibrium, animals are predicted to relocate and settle new habitat patches to reach equilibrium.
+Islands therefore represent distinct habitat patches with clear costs and benefits associated with swimming.
+
+Woodland caribou are well known to disperse to remote locations, including islands, during calving (Bergerud and Page 1987; Cumming and Beange 1987).
+Moving to remote locations is a viable anti-predator strategy for caribou because it reduces detection and encounter rates by predators.
+For instance, the use of shoreline habitat in summer was deemed an effective anti-predator calving strategy because caribou avoided predators by using islands and peninsulas or swimming as a flight response (Bergerud et al.
+1990).
+The use of water, and by association occasionally swimming, is therefore part of the fine-scale interactions between caribou and their predators and the use of islands appears to be an effective anti-predator strategy.
+	Here, we describe visual observation of a single swimming event by a caribou in the North Atlantic Ocean.
+This observation led us to investigate the prevalence of swimming of GPS radio-collared caribou in the Fogo Island archipelago in Newfoundland, Canada.
+We posit two potential mechanisms explaining the trade-offs associated with swimming in the North Atlantic Ocean: forage limitation and the associated density-dependent habitat selection and predator avoidance.
+
+## Materials and methods
+
+### Study area
+
+Our study area was the Fogo Island, Newfoundland, Canada (49°N, 54°W).
+Fogo Island is a small (237.71 km2) island off the northeastern coast of Newfoundland and has humid climate with year-round precipitation.
+While Fogo Island is the largest island in the area, the archipelago includes at least three larger islands: Western Indian (77.6 km2), Eastern Indian (38.7 km2) and Change (XX km2) Islands, as well as numerous smaller islands, including Blundon’s (1.18 km2), North Long (1.01 km2), South Long (0.48 km2), Kate’s (1.64 km2), and Brother’s (1.59 km2) Islands (see Figure 1).
+All islands in the Fogo Island archipelago consist largely of coniferous and mixed forests of balsam fir (Abies balsamea), black spruce (Picea mariana), and white birch (Betula payrifera) as well as bogs, lakes, lichen and rocky barrens.
+
+Caribou were introduced to Fogo Island (n = 26 animals introduced between 1964–67) and Change Island (n = 5 animals introduced in 1964) from the main Island of Newfoundland as part of a series of translocations and introductions throughout the province (Bergerud and Mercer 1989).
+While caribou populations are in decline throughout Canada (Festa-Bianchet et al.
+2011), as well as in Newfoundland (Bastille-Rousseau et al.
+2013), the Fogo Island herd has not declined to the same extent and currently consists of approximately 300 animals (Newfoundland and Labrador Wildlife Division, unpublished data).
+
+In Newfoundland, wolves (Canis lupus) are extirpated, so coyotes (Canis latrans) and black bears (Ursus americanus) are the primary predators of caribou (Bastille-Rousseau, Schaefer, et al.
+2016).
+Coyotes and black bears are responsible for the majority of mortalities for neonate caribou calves (Bastille-Rousseau, Schaefer, et al.
+2016), although predation can still occur after the calving period (Lewis and Mahoney 2014).
+However, on Fogo Island, coyotes are the only predator of caribou.
+
+### Caribou observation and swimming
+
+We fitted 29 (n = 15 in 2016; n = 14 in 2017) adult female caribou on Fogo Island (n = 25) and Western Indian Island (n = 4) with GPS collars (1240g, GPS 4400M; Lotek Wireless Inc., Newmarket, Ontario, Canada).
+Collars were deployed on individual caribou for two years and collars were redeployed on the same individuals for the duration of our study.
+Collars were programmed to collect relocation fixes every 2 hours.
+All animal capture and handling procedures were consistent with the American Society of Mammologists guidelines (Sikes and Gannon 2011) and were approved by Memorial University Animal Use Protocol No.
+20152067.
+During routine fieldwork on May 30, 2017, we observed an unmarked male caribou swim between Perry and Eastern Indian Islands; a swim of at least 470m, which took ~9 minutes (~52m/minute).
+To determine the prevalence of ocean swimming for the Fogo Island caribou herd, we queried nearly 200,000 GPS relocations from 26/29 collared adult female caribou between April 2016 and March 2019.
+We determined all instances of individual caribou moving between islands.
+We restricted the GPS dataset to the ice-free months of the year (beginning of April to end of November) and we assumed caribou swam in the ocean between islands when consecutive relocation fixes, i.e., 2 hour intervals, were located on different islands.
+For each swimming event, we identified the origin island, the terminal island, and the duration of time spent on the terminal island before swimming again.
+
+### Statistical analysis
+We were interested in determining what factors might influence how long an individual remains on the terminus island.
+We therefore used linear regressions and regressed the duration of time spent on the terminal island as a function of the Julian date of arrival to the island, size of the terminal island, and the paired identity of islands (i.e., Fogo to Perry or Perry to Eastern Indian).
+All statistical analyses were conducted in R.
+
+## Discussion
+We observed one swimming event in the field as well as 86 remotely sensed swimming events from three individual caribou over a three-year period.
+In general, caribou swam more frequently to larger islands.
+Fogo, Perry, and Eastern Indian islands are the largest, and the most commonly occupied.
+While caribou often swim across streams, rivers, and lakes (Leblond et al.
+2016), few observations exist of caribou swimming in the ocean (see Table 1).
+Of the existing or inferred observations of caribou swimming in the ocean, none appear to be as prolific as the number and frequency of swimming events we observed.
+We surmise that for caribou living on small islands, the primary drivers of swimming are likely related to forage limitation and predator avoidance.
+
+### Forage limitation and density-dependence
+Forage limitation and over-grazing is a major issue for caribou populations (Schaefer et al.
+2016; Zamin et al.
+2017).
+Reduced access to foraging opportunities can result in higher conspecific competition for resources and lowered reproductive success for adult females (Schaefer et al.
+2016).
+For caribou in the Fogo Island archipelago, the role of forage limitation in animal habitat selection patterns and movement remains unknown, but it is possible that forage limitation and increased conspecific competition influences swimming between islands.
+
+Caribou were introduced to Fogo Island in the 1960s and for several decades the population did not exceed ~100 individuals (Bergerud and Mercer 1989; Newfoundland and Labrador Wildlife Division, unpublished data).
+During the 1990s, population density reached 300 animals and anecdotal evidence suggests that caribou began to occupy other islands in the archipelago.
+We surmise that, as density increased over time, competition among conspecifics and density-dependent habitat selection resulted in expansion of the population to nearby islands.
+Following density-dependent habitat selection theory, caribou should swim to new islands when the average fitness of individuals on the starting island exceeds the density-fitness equilibrium (Morris 1987).
+Animals would therefore be predicted to swim to new islands (i.e.
+settle new habitat patches) to reach a fitness equilibrium.
+While this is an ultimate explanation to a series of proximate observations, it is possible for density-dependent habitat selection to operate at such a fine-scale (Webber and Vander Wal 2018).
+Evidence from the Sable Island horse (Equus ferus caballus) system corroborates our suggestion that swimming among islands could be driven by density-dependent habitat selection (van Beest et al.
+2014).
+For horses, selection for high quality habitat was highest when population density was low, but individuals tended to settle in lower quality habitats as population density increased, following Ideal Free Distribution theory (van Beest et al.
+2014).
+Taken together, caribou may follow density-dependent habitat selection (Wittmer et al.
+2007) and as population density increased, forage was depleted and competition increased.
+
+### Predator avoidance
+Predator avoidance could contribute to swimming behaviour of caribou in our system as well as others (Miller 2002).
+Unlike elsewhere in the range of caribou, wolves are extirpated from Newfoundland, including Fogo Island.
+In Newfoundland, coyotes are the primary predator of caribou (Bastille-Rousseau et al.
+2016a).
+While predation by coyotes on adult female caribou is possible, it is relatively rare on Fogo Island (Webber and Vander Wal unpublished data).
+Therefore, any effect of predation on swimming behaviour is likely related to fine-scale encounters between coyotes and caribou which result in caribou swimming as a flight response.
+Alternatively, during calving, females may swim to small islands to give birth (Bergerud and Page 1987) or swim to islands with a calf at heel to avoid predation.
+Swimming with a calf at heel has been observed for caribou in freshwater lakes on Fogo Island (Webber and Vander Wal unpublished data), so we anticipate it may also be possible in the ocean.
+We suggest future studies assess encounter rates among caribou and coyotes to determine what proportion of encounters result in swimming behaviour.
+
+Our observations represent the most frequent and numerous swimming events for caribou in the ocean (for comparison see Table 1).
+While past observations (e.g.
+Miller 1995, 2002; Jeffery et al.
+2007; Ricca et al.
+2012) have noted the unique nature of caribou swimming in the ocean, we suggest that this behaviour is likely more common than previously thought for caribou living on oceanic islands.
+We posit forage limitation and density-dependent habitat selection is an ultimate explanation for why caribou swimming in the ocean while predator avoidance is a proximate explanation.
+Although our inference is limited to observations, we suggest that because swimming behaviour is so common in our system, that density-dependent habitat selection and predation are plausible mechanisms explaining swimming behaviour in Fogo Island caribou.
+
+## Acknowledgements
+We thank M.
+Laforge, M.
+Bonar, C.
+Hart, and R.
+Huang for help in the field.
+Logistical support was provided L.
+Bixby.
+We also thank members of the Newfoundland Wildlife Division including S.
+Moores, B.
+Adams, K.
+Lewis, C.
+Doucet, W.
+Barney, F.
+Norman, R.
+Otto, J.
+Neville, P.
+Saunders, T.
+Porter, P.
+Tremblett, S.
+Gullage, T.
+Hodder, D.
+Jennings, J.
+McGinn, M.
+Frampton, P.
+McClement, and S.
+Mahoney for data collection, management, and logistics.
+We thank T.
+Bergerud and S.
+Mahoney for their vision in initiating much of the work on caribou in Newfoundland.
+Funding for this study was provided by a Vanier Canada Graduate Scholarship to QMRW and a NSERC Discovery Grant to EVW.
+
+## References

### 2020-02-07 paper/manuscript.Rmd
updated reference formats to be consistent; changed author line; do we include indigenous land statements in the acknowledgements section? I have in the past but not sure if that's SOP here Questions: - not sure how far along the statistics are, but could we include distance to nearest island as a predictor in that model? I would think that caribou would spend a longer time on an island if the distance you'd have to swim to leave is longer - IFD assumes cost-free travel. But given the risks & energetic costs of swimming, wouldn't you be quicker to leave an island once resource availability beg

@@
 title: Interisland movements of woodland caribou in Newfoundland
-author: Quinn M.R. Webber^1^, Jack Hendrix^1^, Alec L. Robitaille^2^,  Eric Vander Wal^1,2^
+author: Quinn M.R. Webber^1^, Jack G. Hendrix^1^, Alec L. Robitaille^2^,  Eric Vander Wal^1,2^
 header-includes:  ^1^Cognitive and Behavioural Ecology Interdisciplinary Program, Memorial University of Newfoundland^2^; Department of Biology, Memorial University of Newfoundland, Canada.
@@
 Despite a large number of coastal and island caribou herds, limited evidence exists suggesting that caribou swim in the ocean as readily as they swim in freshwater (see Table 1).
-A key exception is Peary caribou swimming up to 2.6km between the Queen Elizabeth Islands (Miller, 1995; Miller, 2002).
+A key exception is Peary caribou swimming up to 2.6km between the Queen Elizabeth Islands (Miller 1995; Miller 2002).
 For caribou living on small islands in the ocean, the same notions for why caribou swim in freshwater may also explain why caribou swim in the ocean (Leblond et al.
 2016).
-Specifically, forage limitation (Miller, 2002) and predator avoidance (Jeffery et al.
+Specifically, forage limitation (Miller 2002) and predator avoidance (Jeffery et al.
 2007) have been proposed as potential explanations.
@@
 Specifically, even though caribou are adept swimmers, the energetic expenditure associated with swimming for quadrupedal mammals is significantly higher than walking or running (Fish 1993).
-Additionally, drowning is also possible (Miller & Gunn, 1985) although the proportion of caribou that die from drowning is unknown.
+Additionally, drowning is also possible (Miller & Gunn 1985) although the proportion of caribou that die from drowning is unknown.
 Swimming between islands or from the mainland to nearby islands can be beneficial.
-Among the most important benefits include avoiding predators and seeking under-foraged areas.
+Among the most important benefits are predator avoidance and seeking under-foraged areas.
 
@@
 
-Woodland caribou are well known to disperse to remote locations, including islands, during calving (Bergerud and Page 1987; Cumming and Beange 1987).
+Woodland caribou are well known to disperse to remote locations, including islands, during calving (Bergerud & Page 1987; Cumming & Beange 1987).
 Moving to remote locations is a viable anti-predator strategy for caribou because it reduces detection and encounter rates by predators.
@@
 
-Our study area was the Fogo Island, Newfoundland, Canada (49°N, 54°W).
+Our study area was the Fogo Island archipelago, Newfoundland, Canada (49°N, 54°W).
 Fogo Island is a small (237.71 km2) island off the northeastern coast of Newfoundland and has humid climate with year-round precipitation.
@@
 
-Caribou were introduced to Fogo Island (n = 26 animals introduced between 1964–67) and Change Island (n = 5 animals introduced in 1964) from the main Island of Newfoundland as part of a series of translocations and introductions throughout the province (Bergerud and Mercer 1989).
+Caribou were introduced to Fogo Island (n = 26 animals introduced between 1964–67) and Change Island (n = 5 animals introduced in 1964) from the main Island of Newfoundland as part of a series of translocations and introductions throughout the province (Bergerud & Mercer 1989).
 While caribou populations are in decline throughout Canada (Festa-Bianchet et al.
@@
 
-In Newfoundland, wolves (Canis lupus) are extirpated, so coyotes (Canis latrans) and black bears (Ursus americanus) are the primary predators of caribou (Bastille-Rousseau, Schaefer, et al.
+In Newfoundland, wolves (Canis lupus) are extirpated, so coyotes (Canis latrans) and black bears (Ursus americanus) are the primary predators of caribou (Bastille-Rousseau et al.
 2016).
-Coyotes and black bears are responsible for the majority of mortalities for neonate caribou calves (Bastille-Rousseau, Schaefer, et al.
-2016), although predation can still occur after the calving period (Lewis and Mahoney 2014).
+Coyotes and black bears are responsible for the majority of mortalities for neonate caribou calves (Bastille-Rousseau et al.
+2016), although predation can still occur after the calving period (Lewis & Mahoney 2014).
 However, on Fogo Island, coyotes are the only predator of caribou.
@@
 Collars were programmed to collect relocation fixes every 2 hours.
-All animal capture and handling procedures were consistent with the American Society of Mammologists guidelines (Sikes and Gannon 2011) and were approved by Memorial University Animal Use Protocol No.
+All animal capture and handling procedures were consistent with the American Society of Mammologists guidelines (Sikes & Gannon 2011) and were approved by Memorial University Animal Use Protocol No.
 20152067.
-During routine fieldwork on May 30, 2017, we observed an unmarked male caribou swim between Perry and Eastern Indian Islands; a swim of at least 470m, which took ~9 minutes (~52m/minute).
+During routine fieldwork on May 30, 2017, we observed an unmarked male caribou swim between Perry and Eastern Indian Islands, a swim of at least 470m which took ~9 minutes (~52m/minute).
 To determine the prevalence of ocean swimming for the Fogo Island caribou herd, we queried nearly 200,000 GPS relocations from 26/29 collared adult female caribou between April 2016 and March 2019.
@@
 ### Statistical analysis
-We were interested in determining what factors might influence how long an individual remains on the terminus island.
-We therefore used linear regressions and regressed the duration of time spent on the terminal island as a function of the Julian date of arrival to the island, size of the terminal island, and the paired identity of islands (i.e., Fogo to Perry or Perry to Eastern Indian).
+We were interested in determining what factors might influence how long an individual remains on the terminal island.
+We therefore used linear regressions to test how the Julian date of arrival to the island, size of the terminal island, and the paired identity of islands (i.e., Fogo to Perry or Perry to Eastern Indian) influenced the duration of time spent on the terminal island.
 All statistical analyses were conducted in R.
@@
 
-Caribou were introduced to Fogo Island in the 1960s and for several decades the population did not exceed ~100 individuals (Bergerud and Mercer 1989; Newfoundland and Labrador Wildlife Division, unpublished data).
+Caribou were introduced to Fogo Island in the 1960s and for several decades the population did not exceed ~100 individuals (Bergerud & Mercer 1989; Newfoundland and Labrador Wildlife Division, unpublished data).
 During the 1990s, population density reached 300 animals and anecdotal evidence suggests that caribou began to occupy other islands in the archipelago.
@@
 settle new habitat patches) to reach a fitness equilibrium.
-While this is an ultimate explanation to a series of proximate observations, it is possible for density-dependent habitat selection to operate at such a fine-scale (Webber and Vander Wal 2018).
+While this is an ultimate explanation to a series of proximate observations, it is possible for density-dependent habitat selection to operate at such a fine-scale (Webber & Vander Wal 2018).
 Evidence from the Sable Island horse (Equus ferus caballus) system corroborates our suggestion that swimming among islands could be driven by density-dependent habitat selection (van Beest et al.
@@
 2016a).
-While predation by coyotes on adult female caribou is possible, it is relatively rare on Fogo Island (Webber and Vander Wal unpublished data).
+While predation by coyotes on adult female caribou is possible, it is relatively rare on Fogo Island (Webber & Vander Wal unpublished data).
 Therefore, any effect of predation on swimming behaviour is likely related to fine-scale encounters between coyotes and caribou which result in caribou swimming as a flight response.
-Alternatively, during calving, females may swim to small islands to give birth (Bergerud and Page 1987) or swim to islands with a calf at heel to avoid predation.
-Swimming with a calf at heel has been observed for caribou in freshwater lakes on Fogo Island (Webber and Vander Wal unpublished data), so we anticipate it may also be possible in the ocean.
+Alternatively, during calving, females may swim to small islands to give birth (Bergerud & Page 1987) or swim to islands with a calf at heel to avoid predation.
+Swimming with a calf at heel has been observed for caribou in freshwater lakes on Fogo Island (Webber & Vander Wal unpublished data), so we anticipate it may also be possible in the ocean.
 We suggest future studies assess encounter rates among caribou and coyotes to determine what proportion of encounters result in swimming behaviour.
@@
 
+Land acknowledgement of Beothuk & Mi'kmaq people?
+
 ## References

### 2020-02-17 paper/manuscript.Rmd
fix yaml author

@@
 author: Quinn M.R. Webber^1^, Jack G. Hendrix^1^, Alec L. Robitaille^2^,  Eric Vander Wal^1,2^
-header-includes:  ^1^Cognitive and Behavioural Ecology Interdisciplinary Program, Memorial University of Newfoundland^2^; Department of Biology, Memorial University of Newfoundland, Canada.
-
+date:  ^1^Cognitive and Behavioural Ecology Interdisciplinary Program, Memorial University of Newfoundland; ^2^ Department of Biology, Memorial University of Newfoundland, Canada.
 output: 

### 2020-02-17 scripts/1-extractRaster.R
add script titles, authors

@@ -1,31 +1,46 @@
+### Extract OSM data ====
+# Quinn M.R. Webber, Jack G. Hendrix, Alec L. Robitaille,  Eric Vander Wal
 
 
 ### Packages ----
-libs <- c('data.table', 'ggplot2', 'dplyr',
-          'raster', 'prettymapr', 'rosm', 
-          'stars', 'fasterize')
+libs <- c('data.table',
+          'ggplot2',
+          # 'dplyr',
+          'raster',
+          # 'prettymapr',
+          'rosm',
+          'stars',
+          'fasterize')
 lapply(libs, require, character.only = TRUE)
 
-### extract islands from osm 
-coords <- makebbox(49.5194, -54.1878, 
-                   49.65, -54.3533)
+### extract islands from osm
+coords <- makebbox(49.5194,-54.1878,
+                   49.65,-54.3533)
 
-coordsOSM <- osm.raster(coords, projection=CRS('+proj=utm +zone=21 ellps=WGS84'), crop=TRUE)
+coordsOSM <-
+  osm.raster(coords,
+             projection = CRS('+proj=utm +zone=21 ellps=WGS84'),
+             crop = TRUE)
 coordsOSM2 <- coordsOSM[[1]] != 170
 coordsOSM2[coordsOSM2 == 0] <- NA
-coordsOSM3 <- st_as_stars(coordsOSM2) %>% 
-  st_as_sf(as_points = FALSE, use_integer = TRUE, merge = TRUE, na.rm = TRUE)
-
+coordsOSM3 <- st_as_sf(
+  st_as_stars(coordsOSM2),
+  as_points = FALSE,
+  use_integer = TRUE,
+  merge = TRUE,
+  na.rm = TRUE
+)
 coordsOSM3$island <- seq_along(coordsOSM3$geometry)
 
-r <- raster(ymn = st_bbox(coordsOSM3)$ymin,
-            ymx = st_bbox(coordsOSM3)$ymax, 
-            xmn = st_bbox(coordsOSM3)$xmin,
-            xmx = st_bbox(coordsOSM3)$xmax, 
-            crs = CRS('+proj=utm +zone=21 ellps=WGS84'))
+r <- raster(
+  ymn = st_bbox(coordsOSM3)$ymin,
+  ymx = st_bbox(coordsOSM3)$ymax,
+  xmn = st_bbox(coordsOSM3)$xmin,
[…]

### 2020-02-17 scripts/2-extractPoints.R

@@ -1,3 +1,5 @@
+### Extract Island Caribou ====
+# Quinn M.R. Webber, Jack G. Hendrix, Alec L. Robitaille,  Eric Vander Wal
 
 
 ### Packages ----

### 2020-02-17 scripts/3-extractArea.R

@@ -1,3 +1,5 @@
+### Island area vs ====
+# Quinn M.R. Webber, Jack G. Hendrix, Alec L. Robitaille,  Eric Vander Wal
 
 ### Packages ----
 libs <- c('data.table', 'ggplot2', 'dplyr',

### 2020-02-17 scripts/1-extractRaster.R
use sp instead

@@ -7,6 +7,7 @@ libs <- c('data.table',
           'ggplot2',
           # 'dplyr',
           'raster',
+          'sp',
           # 'prettymapr',
           'rosm',
           'stars',
@@ -14,8 +15,7 @@ libs <- c('data.table',
 lapply(libs, require, character.only = TRUE)
 
 ### extract islands from osm
-coords <- makebbox(49.5194,-54.1878,
-                   49.65,-54.3533)
+coords <- bbox(SpatialPoints(cbind(c(49.5194, 49.65), c(-54.1878, -54.3533))))
 
 coordsOSM <-
   osm.raster(coords,

### 2020-02-19 paper/manuscript.Rmd
reflow P1 few edits
reflow

@@
 
-## Abstract
+Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
+that caribou swim in streams, rivers, and lakes during migration (Leblond et al.
+2016) to avoid predators (Bergerud 1985) and access islands during calving
+(Bergerud et al. 1990). Despite a large number of coastal and island caribou
+herds, limited evidence exists suggesting that caribou swim in the ocean as
+readily as they swim in freshwater (see Table 1). A key exception is Peary
+caribou swimming up to 2.6 km between the Queen Elizabeth Islands (Miller 1995;
+Miller 2002). Swimming for quadrupedal terrestrial mammals can be thought of as
+a set of trade-offs. Costs include high energy expenditure and risk of drowning.
+Specifically, even though caribou are adept swimmers, the energetic expenditure
+associated with swimming for quadrupedal mammals is significantly higher than
+walking or running (Fish 1993). Additionally, drowning is also possible (Miller
+& Gunn 1985) although the proportion of caribou that die from drowning is
+unknown. Swimming between islands or from the mainland to nearby islands can be
+beneficial. Among the most important benefits are predator avoidance and seeking
+under-foraged areas.
 
-## Introduction
-Caribou are exceptional swimmers.
-Ample evidence exists that caribou swim in streams, rivers, and lakes during migration (Leblond et al.
-2016) to avoid predators (Bergerud 1985) and access islands during calving (Bergerud et al.
-1990).
-Despite a large number of coastal and island caribou herds, limited evidence exists suggesting that caribou swim in the ocean as readily as they swim in freshwater (see Table 1).
-A key exception is Peary caribou swimming up to 2.6km between the Queen Elizabeth Islands (Miller 1995; Miller 2002).
-For caribou living on small islands in the ocean, the same notions for why caribou swim in freshwater may also explain why caribou swim in the ocean (Leblond et al.
-2016).
-Specifically, forage limitation (Miller 2002) and predator avoidance (Jeffery et al.
-2007) have been proposed as potential explanations.
-
-Swimming for quadrupedal terrestrial mammals can be thought of as a set of trade-offs.
-Costs include high energy expenditure and risk of drowning.
-Specifically, even though caribou are adept swimmers, the energetic expenditure associated with swimming for quadrupedal mammals is significantly higher than walking or running (Fish 1993).
-Additionally, drowning is also possible (Miller & Gunn 1985) although the proportion of caribou that die from drowning is unknown.
-Swimming between islands or from the mainland to nearby islands can be beneficial.
-Among the most important benefits are predator avoidance and seeking under-foraged areas.
-
-Forage limitation could explain movement between islands.
-Forage scarcity has been proposed as a potential reason why caribou move between arctic islands on the sea-ice in winter (Miller et al.
-1977).
-For very small islands, over-grazing by newly arrived caribou may rapidly limit forage (Bergerud et al.
-1990).
-Without sea ice, movement from the mainland to an island as well as between islands requires caribou to assess the trade-off associated with swimming.
-The underlying mechanism driving forage limitation and the trade-off associated with swimming could be density-dependent habitat selection governed by the Ideal Free Distribution (Morris 1987; Bradbury et al.
-2015).
-Ideal Free Distribution theory suggests that habitat selection is density-dependent and that variation in density between habitat patches leads to a fitness equilibrium (Bradbury et al.
-2015).
-Fitness in a habitat patch depends on density where, ideally, the available resources on a habitat patch can sustain a specific number of individuals.
-In a hypothetical example, when density of a given habitat patch has exceeded the fitness equilibrium, animals are predicted to relocate and settle new habitat patches to reach equilibrium.
-Islands therefore represent distinct habitat patches with clear costs and benefits associated with swimming.
-
-Woodland caribou are well known to disperse to remote locations, including islands, during calving (Bergerud & Page 1987; Cumming & Beange 1987).
-Moving to remote locations is a viable anti-predator strategy for caribou because it reduces detection and encounter rates by predators.
-For instance, the use of shoreline habitat in summer was deemed an effective anti-predator calving strategy because caribou avoided predators by using islands and peninsulas or swimming as a flight response (Bergerud et al.
-1990).
-The use of water, and by association occasionally swimming, is therefore part of the fine-scale interactions between caribou and their predators and the use of islands appears to be an effective anti-predator strategy.
-	Here, we describe visual observation of a single swimming event by a caribou in the North Atlantic Ocean.
-This observation led us to investigate the prevalence of swimming of GPS radio-collared caribou in the Fogo Island archipelago in Newfoundland, Canada.
-We posit two potential mechanisms explaining the trade-offs associated with swimming in the North Atlantic Ocean: forage limitation and the associated density-dependent habitat selection and predator avoidance.
+We describe visual observation of a single swimming event by a caribou in the
+North Atlantic Ocean. This observation led us to investigate the prevalence of
+swimming of GPS radio-collared caribou in the Fogo Island archipelago in
+Newfoundland, Canada. We posit two potential mechanisms explaining the
+trade-offs associated with swimming in the North Atlantic Ocean: forage
+limitation and the associated density-dependent habitat selection and predator
+avoidance. For caribou living on small islands in the ocean, the same notions
+for why caribou swim in freshwater may also explain why caribou swim in the
+ocean (Leblond et al. 2016). Specifically, forage limitation (Miller 2002) and
+predator avoidance (Jeffery et al. 2007) have been proposed as potential
+explanations.
 
@@
 ## Discussion
+
+Forage limitation could explain movement between islands.
+Forage scarcity has been proposed as a potential reason why caribou move between arctic islands on the sea-ice in winter (Miller et al.
+1977).
+For very small islands, over-grazing by newly arrived caribou may rapidly limit forage (Bergerud et al.
+1990).
+Without sea ice, movement from the mainland to an island as well as between islands requires caribou to assess the trade-off associated with swimming.
+The underlying mechanism driving forage limitation and the trade-off associated with swimming could be density-dependent habitat selection governed by the Ideal Free Distribution (Morris 1987; Bradbury et al.
+2015).
+Ideal Free Distribution theory suggests that habitat selection is density-dependent and that variation in density between habitat patches leads to a fitness equilibrium (Bradbury et al.
+2015).
+Fitness in a habitat patch depends on density where, ideally, the available resources on a habitat patch can sustain a specific number of individuals.
+In a hypothetical example, when density of a given habitat patch has exceeded the fitness equilibrium, animals are predicted to relocate and settle new habitat patches to reach equilibrium.
+Islands therefore represent distinct habitat patches with clear costs and benefits associated with swimming.
+
+Woodland caribou are well known to disperse to remote locations, including islands, during calving (Bergerud & Page 1987; Cumming & Beange 1987).
+Moving to remote locations is a viable anti-predator strategy for caribou because it reduces detection and encounter rates by predators.
+For instance, the use of shoreline habitat in summer was deemed an effective anti-predator calving strategy because caribou avoided predators by using islands and peninsulas or swimming as a flight response (Bergerud et al.
+1990).
+The use of water, and by association occasionally swimming, is therefore part of the fine-scale interactions between caribou and their predators and the use of islands appears to be an effective anti-predator strategy.
+
 We observed one swimming event in the field as well as 86 remotely sensed swimming events from three individual caribou over a three-year period.

### 2020-02-19 paper/manuscript.Rmd
monster reflow

@@
 
-## Materials and methods
+We observed one swimming event in the field as well as 86 remotely sensed
+swimming events from three individual caribou over a three-year period. In
+general, caribou swam more frequently to larger islands. Fogo, Perry, and
+Eastern Indian islands are the largest, and the most commonly occupied. While
+caribou often swim across streams, rivers, and lakes (Leblond et al. 2016), few
+observations exist of caribou swimming in the ocean (see Table 1). Of the
+existing or inferred observations of caribou swimming in the ocean, none appear
+to be as prolific as the number and frequency of swimming events we observed. We
+surmise that for caribou living on small islands, the primary drivers of
+swimming are likely related to forage limitation and predator avoidance.
 
-### Study area
+The Fogo Island archipelago is located in Newfoundland, Canada (49°N, 54°W), and
+comprises approximately ten islands. While Fogo Island is the largest island
+(237.71 km^2), the archipelago includes at least three larger islands: Western
+Indian (77.6 km2), Eastern Indian (38.7 km2) and Change (XX km2) Islands, as
+well as numerous smaller islands, including Blundon’s (1.18 km2), North Long
+(1.01 km2), South Long (0.48 km2), Kate’s (1.64 km2), and Brother’s (1.59 km2)
+Islands (see Figure 1). All islands in the Fogo Island archipelago consist
+largely of coniferous and mixed forests of balsam fir (Abies balsamea), black
+spruce (Picea mariana), and white birch (Betula payrifera) as well as bogs,
+lakes, lichen and rocky barrens.
 
-Our study area was the Fogo Island archipelago, Newfoundland, Canada (49°N, 54°W).
-Fogo Island is a small (237.71 km2) island off the northeastern coast of Newfoundland and has humid climate with year-round precipitation.
-While Fogo Island is the largest island in the area, the archipelago includes at least three larger islands: Western Indian (77.6 km2), Eastern Indian (38.7 km2) and Change (XX km2) Islands, as well as numerous smaller islands, including Blundon’s (1.18 km2), North Long (1.01 km2), South Long (0.48 km2), Kate’s (1.64 km2), and Brother’s (1.59 km2) Islands (see Figure 1).
-All islands in the Fogo Island archipelago consist largely of coniferous and mixed forests of balsam fir (Abies balsamea), black spruce (Picea mariana), and white birch (Betula payrifera) as well as bogs, lakes, lichen and rocky barrens.
+Caribou were introduced to Fogo Island (n = 26 animals introduced between
+1964–67) and Change Island (n = 5 animals introduced in 1964) from the main
+Island of Newfoundland as part of a series of translocations and introductions
+throughout the province (Bergerud & Mercer 1989). While caribou populations are
+in decline throughout Canada (Festa-Bianchet et al. 2011), as well as in
+Newfoundland (Bastille-Rousseau et al. 2013), the Fogo Island herd has not
+declined to the same extent and currently consists of approximately 300 animals
+(Newfoundland and Labrador Wildlife Division, unpublished data).
 
-Caribou were introduced to Fogo Island (n = 26 animals introduced between 1964–67) and Change Island (n = 5 animals introduced in 1964) from the main Island of Newfoundland as part of a series of translocations and introductions throughout the province (Bergerud & Mercer 1989).
-While caribou populations are in decline throughout Canada (Festa-Bianchet et al.
-2011), as well as in Newfoundland (Bastille-Rousseau et al.
-2013), the Fogo Island herd has not declined to the same extent and currently consists of approximately 300 animals (Newfoundland and Labrador Wildlife Division, unpublished data).
+In Newfoundland, wolves (Canis lupus) are extirpated, so coyotes (Canis latrans)
+and black bears (Ursus americanus) are the primary predators of caribou
+(Bastille-Rousseau et al. 2016). Coyotes and black bears are responsible for the
+majority of mortalities for neonate caribou calves (Bastille-Rousseau et al.
+2016), although predation can still occur after the calving period (Lewis &
+Mahoney 2014). However, on Fogo Island, coyotes are the only predator of
+caribou.
 
-In Newfoundland, wolves (Canis lupus) are extirpated, so coyotes (Canis latrans) and black bears (Ursus americanus) are the primary predators of caribou (Bastille-Rousseau et al.
-2016).
-Coyotes and black bears are responsible for the majority of mortalities for neonate caribou calves (Bastille-Rousseau et al.
-2016), although predation can still occur after the calving period (Lewis & Mahoney 2014).
-However, on Fogo Island, coyotes are the only predator of caribou.
+We fitted 29 (n = 15 in 2016; n = 14 in 2017) adult female caribou on Fogo
+Island (n = 25) and Western Indian Island (n = 4) with GPS collars (1240g, GPS
+4400M; Lotek Wireless Inc., Newmarket, Ontario, Canada). Collars were deployed
+on individual caribou for two years and collars were redeployed on the same
+individuals for the duration of our study. Collars were programmed to collect
+relocation fixes every 2 hours. All animal capture and handling procedures were
+consistent with the American Society of Mammologists guidelines (Sikes & Gannon
+2011) and were approved by Memorial University Animal Use Protocol No. 20152067.
+During routine fieldwork on May 30, 2017, we observed an unmarked male caribou
+swim between Perry and Eastern Indian Islands, a swim of at least 470m which
+took ~9 minutes (~52m/minute). To determine the prevalence of ocean swimming for
+the Fogo Island caribou herd, we queried nearly 200,000 GPS relocations from
+26/29 collared adult female caribou between April 2016 and March 2019. We
+determined all instances of individual caribou moving between islands. We
+restricted the GPS dataset to the ice-free months of the year (beginning of
+April to end of November) and we assumed caribou swam in the ocean between
+islands when consecutive relocation fixes, i.e., 2 hour intervals, were located
+on different islands. For each swimming event, we identified the origin island,
+the terminal island, and the duration of time spent on the terminal island
+before swimming again.
 
-### Caribou observation and swimming
+Forage limitation could explain movement between islands. Forage scarcity has
+been proposed as a potential reason why caribou move between arctic islands on
+the sea-ice in winter (Miller et al. 1977). For very small islands, over-grazing
+by newly arrived caribou may rapidly limit forage (Bergerud et al. 1990).
+Without sea ice, movement from the mainland to an island as well as between
+islands requires caribou to assess the trade-off associated with swimming. The
+underlying mechanism driving forage limitation and the trade-off associated with
+swimming could be density-dependent habitat selection governed by the Ideal Free
+Distribution (Morris 1987; Bradbury et al. 2015). Ideal Free Distribution theory
+suggests that habitat selection is density-dependent and that variation in
+density between habitat patches leads to a fitness equilibrium (Bradbury et al.
+2015). Fitness in a habitat patch depends on density where, ideally, the
+available resources on a habitat patch can sustain a specific number of
+individuals. In a hypothetical example, when density of a given habitat patch
+has exceeded the fitness equilibrium, animals are predicted to relocate and
+settle new habitat patches to reach equilibrium. Islands therefore represent
+distinct habitat patches with clear costs and benefits associated with swimming.
 
-We fitted 29 (n = 15 in 2016; n = 14 in 2017) adult female caribou on Fogo Island (n = 25) and Western Indian Island (n = 4) with GPS collars (1240g, GPS 4400M; Lotek Wireless Inc., Newmarket, Ontario, Canada).
-Collars were deployed on individual caribou for two years and collars were redeployed on the same individuals for the duration of our study.
-Collars were programmed to collect relocation fixes every 2 hours.
-All animal capture and handling procedures were consistent with the American Society of Mammologists guidelines (Sikes & Gannon 2011) and were approved by Memorial University Animal Use Protocol No.
-20152067.
-During routine fieldwork on May 30, 2017, we observed an unmarked male caribou swim between Perry and Eastern Indian Islands, a swim of at least 470m which took ~9 minutes (~52m/minute).
-To determine the prevalence of ocean swimming for the Fogo Island caribou herd, we queried nearly 200,000 GPS relocations from 26/29 collared adult female caribou between April 2016 and March 2019.
-We determined all instances of individual caribou moving between islands.
-We restricted the GPS dataset to the ice-free months of the year (beginning of April to end of November) and we assumed caribou swam in the ocean between islands when consecutive relocation fixes, i.e., 2 hour intervals, were located on different islands.
-For each swimming event, we identified the origin island, the terminal island, and the duration of time spent on the terminal island before swimming again.
+Woodland caribou are well known to disperse to remote locations, including
+islands, during calving (Bergerud & Page 1987; Cumming & Beange 1987). Moving to
+remote locations is a viable anti-predator strategy for caribou because it
+reduces detection and encounter rates by predators. For instance, the use of
+shoreline habitat in summer was deemed an effective anti-predator calving
+strategy because caribou avoided predators by using islands and peninsulas or
+swimming as a flight response (Bergerud et al. 1990). The use of water, and by
+association occasionally swimming, is therefore part of the fine-scale
+interactions between caribou and their predators and the use of islands appears
+to be an effective anti-predator strategy.
 
-### Statistical analysis
-We were interested in determining what factors might influence how long an individual remains on the terminal island.
-We therefore used linear regressions to test how the Julian date of arrival to the island, size of the terminal island, and the paired identity of islands (i.e., Fogo to Perry or Perry to Eastern Indian) influenced the duration of time spent on the terminal island.
-All statistical analyses were conducted in R.
+Forage limitation and over-grazing is a major issue for caribou populations
+(Schaefer et al. 2016; Zamin et al. 2017). Reduced access to foraging
+opportunities can result in higher conspecific competition for resources and
+lowered reproductive success for adult females (Schaefer et al. 2016). For
+caribou in the Fogo Island archipelago, the role of forage limitation in animal
+habitat selection patterns and movement remains unknown, but it is possible that
+forage limitation and increased conspecific competition influences swimming
+between islands.
 
-## Discussion
+Caribou were introduced to Fogo Island in the 1960s and for several decades the
+population did not exceed ~100 individuals (Bergerud & Mercer 1989; Newfoundland
+and Labrador Wildlife Division, unpublished data). During the 1990s, population
+density reached 300 animals and anecdotal evidence suggests that caribou began
+to occupy other islands in the archipelago. We surmise that, as density
+increased over time, competition among conspecifics and density-dependent
+habitat selection resulted in expansion of the population to nearby islands.
+Following density-dependent habitat selection theory, caribou should swim to new
+islands when the average fitness of individuals on the starting island exceeds
+the density-fitness equilibrium (Morris 1987). Animals would therefore be
+predicted to swim to new islands (i.e. settle new habitat patches) to reach a
+fitness equilibrium. While this is an ultimate explanation to a series of
+proximate observations, it is possible for density-dependent habitat selection
+to operate at such a fine-scale (Webber & Vander Wal 2018). Evidence from the
+Sable Island horse (Equus ferus caballus) system corroborates our suggestion
+that swimming among islands could be driven by density-dependent habitat
+selection (van Beest et al. 2014). For horses, selection for high quality
+habitat was highest when population density was low, but individuals tended to
+settle in lower quality habitats as population density increased, following
+Ideal Free Distribution theory (van Beest et al. 2014). Taken together, caribou
+may follow density-dependent habitat selection (Wittmer et al. 2007) and as
+population density increased, forage was depleted and competition increased.
 
-Forage limitation could explain movement between islands.
-Forage scarcity has been proposed as a potential reason why caribou move between arctic islands on the sea-ice in winter (Miller et al.
-1977).
-For very small islands, over-grazing by newly arrived caribou may rapidly limit forage (Bergerud et al.
-1990).
-Without sea ice, movement from the mainland to an island as well as between islands requires caribou to assess the trade-off associated with swimming.
-The underlying mechanism driving forage limitation and the trade-off associated with swimming could be density-dependent habitat selection governed by the Ideal Free Distribution (Morris 1987; Bradbury et al.
-2015).
-Ideal Free Distribution theory suggests that habitat selection is density-dependent and that variation in density between habitat patches leads to a fitness equilibrium (Bradbury et al.
-2015).
-Fitness in a habitat patch depends on density where, ideally, the available resources on a habitat patch can sustain a specific number of individuals.
-In a hypothetical example, when density of a given habitat patch has exceeded the fitness equilibrium, animals are predicted to relocate and settle new habitat patches to reach equilibrium.
-Islands therefore represent distinct habitat patches with clear costs and benefits associated with swimming.
+Predator avoidance could contribute to swimming behaviour of caribou in our
+system as well as others (Miller 2002). Unlike elsewhere in the range of
+caribou, wolves are extirpated from Newfoundland, including Fogo Island. In
+Newfoundland, coyotes are the primary predator of caribou (Bastille-Rousseau et
+al. 2016a). While predation by coyotes on adult female caribou is possible, it
+is relatively rare on Fogo Island (Webber & Vander Wal unpublished data).
+Therefore, any effect of predation on swimming behaviour is likely related to
+fine-scale encounters between coyotes and caribou which result in caribou
+swimming as a flight response. Alternatively, during calving, females may swim
+to small islands to give birth (Bergerud & Page 1987) or swim to islands with a
+calf at heel to avoid predation. Swimming with a calf at heel has been observed
+for caribou in freshwater lakes on Fogo Island (Webber & Vander Wal unpublished
+data), so we anticipate it may also be possible in the ocean. We suggest future
+studies assess encounter rates among caribou and coyotes to determine what
+proportion of encounters result in swimming behaviour.
 
-Woodland caribou are well known to disperse to remote locations, including islands, during calving (Bergerud & Page 1987; Cumming & Beange 1987).
-Moving to remote locations is a viable anti-predator strategy for caribou because it reduces detection and encounter rates by predators.
-For instance, the use of shoreline habitat in summer was deemed an effective anti-predator calving strategy because caribou avoided predators by using islands and peninsulas or swimming as a flight response (Bergerud et al.
-1990).
-The use of water, and by association occasionally swimming, is therefore part of the fine-scale interactions between caribou and their predators and the use of islands appears to be an effective anti-predator strategy.
-
-We observed one swimming event in the field as well as 86 remotely sensed swimming events from three individual caribou over a three-year period.
-In general, caribou swam more frequently to larger islands.
-Fogo, Perry, and Eastern Indian islands are the largest, and the most commonly occupied.
-While caribou often swim across streams, rivers, and lakes (Leblond et al.
-2016), few observations exist of caribou swimming in the ocean (see Table 1).
-Of the existing or inferred observations of caribou swimming in the ocean, none appear to be as prolific as the number and frequency of swimming events we observed.
-We surmise that for caribou living on small islands, the primary drivers of swimming are likely related to forage limitation and predator avoidance.
-
-### Forage limitation and density-dependence
-Forage limitation and over-grazing is a major issue for caribou populations (Schaefer et al.
-2016; Zamin et al.
-2017).
-Reduced access to foraging opportunities can result in higher conspecific competition for resources and lowered reproductive success for adult females (Schaefer et al.
-2016).
-For caribou in the Fogo Island archipelago, the role of forage limitation in animal habitat selection patterns and movement remains unknown, but it is possible that forage limitation and increased conspecific competition influences swimming between islands.
-
-Caribou were introduced to Fogo Island in the 1960s and for several decades the population did not exceed ~100 individuals (Bergerud & Mercer 1989; Newfoundland and Labrador Wildlife Division, unpublished data).
-During the 1990s, population density reached 300 animals and anecdotal evidence suggests that caribou began to occupy other islands in the archipelago.
-We surmise that, as density increased over time, competition among conspecifics and density-dependent habitat selection resulted in expansion of the population to nearby islands.
-Following density-dependent habitat selection theory, caribou should swim to new islands when the average fitness of individuals on the starting island exceeds the density-fitness equilibrium (Morris 1987).
-Animals would therefore be predicted to swim to new islands (i.e.
-settle new habitat patches) to reach a fitness equilibrium.
-While this is an ultimate explanation to a series of proximate observations, it is possible for density-dependent habitat selection to operate at such a fine-scale (Webber & Vander Wal 2018).
-Evidence from the Sable Island horse (Equus ferus caballus) system corroborates our suggestion that swimming among islands could be driven by density-dependent habitat selection (van Beest et al.
-2014).
-For horses, selection for high quality habitat was highest when population density was low, but individuals tended to settle in lower quality habitats as population density increased, following Ideal Free Distribution theory (van Beest et al.
-2014).
-Taken together, caribou may follow density-dependent habitat selection (Wittmer et al.
-2007) and as population density increased, forage was depleted and competition increased.
-
-### Predator avoidance
-Predator avoidance could contribute to swimming behaviour of caribou in our system as well as others (Miller 2002).
-Unlike elsewhere in the range of caribou, wolves are extirpated from Newfoundland, including Fogo Island.
-In Newfoundland, coyotes are the primary predator of caribou (Bastille-Rousseau et al.
-2016a).
-While predation by coyotes on adult female caribou is possible, it is relatively rare on Fogo Island (Webber & Vander Wal unpublished data).
-Therefore, any effect of predation on swimming behaviour is likely related to fine-scale encounters between coyotes and caribou which result in caribou swimming as a flight response.
-Alternatively, during calving, females may swim to small islands to give birth (Bergerud & Page 1987) or swim to islands with a calf at heel to avoid predation.
-Swimming with a calf at heel has been observed for caribou in freshwater lakes on Fogo Island (Webber & Vander Wal unpublished data), so we anticipate it may also be possible in the ocean.
-We suggest future studies assess encounter rates among caribou and coyotes to determine what proportion of encounters result in swimming behaviour.
-
-Our observations represent the most frequent and numerous swimming events for caribou in the ocean (for comparison see Table 1).
-While past observations (e.g.
-Miller 1995, 2002; Jeffery et al.
-2007; Ricca et al.
-2012) have noted the unique nature of caribou swimming in the ocean, we suggest that this behaviour is likely more common than previously thought for caribou living on oceanic islands.
-We posit forage limitation and density-dependent habitat selection is an ultimate explanation for why caribou swimming in the ocean while predator avoidance is a proximate explanation.
-Although our inference is limited to observations, we suggest that because swimming behaviour is so common in our system, that density-dependent habitat selection and predation are plausible mechanisms explaining swimming behaviour in Fogo Island caribou.
+Our observations represent the most frequent and numerous swimming events for
+caribou in the ocean (for comparison see Table 1). While past observations (e.g.
+Miller 1995, 2002; Jeffery et al. 2007; Ricca et al. 2012) have noted the unique
+nature of caribou swimming in the ocean, we suggest that this behaviour is
+likely more common than previously thought for caribou living on oceanic
+islands. We posit forage limitation and density-dependent habitat selection is
+an ultimate explanation for why caribou swimming in the ocean while predator
+avoidance is a proximate explanation. Although our inference is limited to
+observations, we suggest that because swimming behaviour is so common in our
+system, that density-dependent habitat selection and predation are plausible
+mechanisms explaining swimming behaviour in Fogo Island caribou.
 
 ## Acknowledgements 
-We thank M.
-Laforge, M.
-Bonar, C.
-Hart, and R.
-Huang for help in the field.
-Logistical support was provided L.
-Bixby.
-We also thank members of the Newfoundland Wildlife Division including S.
-Moores, B.
-Adams, K.
-Lewis, C.
-Doucet, W.
-Barney, F.
-Norman, R.
-Otto, J.
-Neville, P.
-Saunders, T.
-Porter, P.
-Tremblett, S.
-Gullage, T.
-Hodder, D.
-Jennings, J.
-McGinn, M.
-Frampton, P.
-McClement, and S.
-Mahoney for data collection, management, and logistics.
-We thank T.
-Bergerud and S.
-Mahoney for their vision in initiating much of the work on caribou in Newfoundland.
-Funding for this study was provided by a Vanier Canada Graduate Scholarship to QMRW and a NSERC Discovery Grant to EVW.
+We thank M. Laforge, M. Bonar, C. Hart, and R. Huang for help in the field.
+Logistical support was provided L. Bixby. We also thank members of the
+Newfoundland Wildlife Division including S. Moores, B. Adams, K. Lewis, C.
+Doucet, W. Barney, F. Norman, R. Otto, J. Neville, P. Saunders, T. Porter, P.
+Tremblett, S. Gullage, T. Hodder, D. Jennings, J. McGinn, M. Frampton, P.
+McClement, and S. Mahoney for data collection, management, and logistics. We
+thank T. Bergerud and S. Mahoney for their vision in initiating much of the work
+on caribou in Newfoundland. Funding for this study was provided by a National
+Sciences and Engineering Research Council (NSERC) Vanier Canada Graduate
+Scholarship to QMRW, NSERC Canda Graduate Scholarship to JGH, and a NSERC
+Discovery Grant to EVW.
 

### 2020-02-19 paper/manuscript.Rmd
acknowledgements
Update manuscript.Rmd and manuscript.md

@@
 title: Interisland movements of woodland caribou in Newfoundland
-author: Quinn M.R. Webber^1^, Jack G. Hendrix^1^, Alec L. Robitaille^2^,  Eric Vander Wal^1,2^
-date:  ^1^Cognitive and Behavioural Ecology Interdisciplinary Program, Memorial University of Newfoundland; ^2^ Department of Biology, Memorial University of Newfoundland, Canada.
+author: Quinn M.R. Webber^1\*^, Jack G. Hendrix^1^, Alec L. Robitaille^2^,  Eric Vander Wal^1,2^
+date: |
+  | ^1^Cognitive and Behavioural Ecology Interdisciplinary Program, Memorial University of Newfoundland
+  | ^2^Department of Biology, Memorial University of Newfoundland, Canada
+  | \*Corresponding author: Quinn M.R. Webber (webber.quinn@gmail.com)
 output: 
-    #word_document
     github_document
+  # bookdown::pdf_document2:
+  #    number_sections: true
+  #    toc: false
+  #bookdown::word_document2:
+  #  reference_docx: template.docx
+#bibliography: ../references/knitcite.bib
+#csl: csl.csl
 ---
+
+\newpage
 
@@
 2016) to avoid predators (Bergerud 1985) and access islands during calving
-(Bergerud et al. 1990). Despite a large number of coastal and island caribou
-herds, limited evidence exists suggesting that caribou swim in the ocean as
-readily as they swim in freshwater (see Table 1). A key exception is Peary
+(Bergerud et al. 1990). Swimming for quadrupedal terrestrial mammals can be
+thought of as a set of trade-offs. Costs include high energy expenditure and
+risk of drowning. Specifically, even though caribou are adept swimmers, the
+energetic expenditure associated with swimming for quadrupedal mammals is
+significantly higher than walking or running (Fish 1993). Additionally, drowning
+is also possible (Miller & Gunn 1985) although the proportion of caribou that
+die from drowning is unknown. Despite a large number of coastal and island
+caribou herds, limited evidence exists suggesting that caribou swim in the ocean
+as readily as they swim in freshwater (see Table 1). A key exception is Peary
 caribou swimming up to 2.6 km between the Queen Elizabeth Islands (Miller 1995;
-Miller 2002). Swimming for quadrupedal terrestrial mammals can be thought of as
-a set of trade-offs. Costs include high energy expenditure and risk of drowning.
-Specifically, even though caribou are adept swimmers, the energetic expenditure
-associated with swimming for quadrupedal mammals is significantly higher than
-walking or running (Fish 1993). Additionally, drowning is also possible (Miller
-& Gunn 1985) although the proportion of caribou that die from drowning is
-unknown. Swimming between islands or from the mainland to nearby islands can be
-beneficial. Among the most important benefits are predator avoidance and seeking
-under-foraged areas.
+Miller 2002).
 
-We describe visual observation of a single swimming event by a caribou in the
-North Atlantic Ocean. This observation led us to investigate the prevalence of
-swimming of GPS radio-collared caribou in the Fogo Island archipelago in
-Newfoundland, Canada. We posit two potential mechanisms explaining the
+Here, we describe the visual observation of an adult male caribou swimming
+between islands in the Fogo Island archipleago in the North Atlantic Ocean. This
+observation prompted us to query a large database of GPS radio-collared caribou
+in the Fogo Island caribou herd in Newfoundland, Canada. We identified  86
+remotely sensed swimming events between XX islands from three individual caribou
+over a three-year period.
+
+We posit two potential mechanisms explaining the
 trade-offs associated with swimming in the North Atlantic Ocean: forage
@@
 
-We observed one swimming event in the field as well as 86 remotely sensed
-swimming events from three individual caribou over a three-year period. In
-general, caribou swam more frequently to larger islands. Fogo, Perry, and
-Eastern Indian islands are the largest, and the most commonly occupied. While
-caribou often swim across streams, rivers, and lakes (Leblond et al. 2016), few
-observations exist of caribou swimming in the ocean (see Table 1). Of the
-existing or inferred observations of caribou swimming in the ocean, none appear
-to be as prolific as the number and frequency of swimming events we observed. We
-surmise that for caribou living on small islands, the primary drivers of
-swimming are likely related to forage limitation and predator avoidance.
+While caribou often swim across streams, rivers, and lakes (Leblond et al.
+2016), few observations exist of caribou swimming in the ocean (see Table 1). Of
+the existing or inferred observations of caribou swimming in the ocean, none
+appear to be as prolific as the number and frequency of swimming events we
+observed. We surmise that for caribou living on small islands, the primary
+drivers of swimming are likely related to forage limitation and predator
+avoidance.
 
@@
 Scholarship to QMRW, NSERC Canda Graduate Scholarship to JGH, and a NSERC
-Discovery Grant to EVW.
+Discovery Grant to EVW. We respectfully acknowledge the territory in which data
+were collected and analyzed as the ancestral homelands of the Beothuk, and the
+island of Newfoundland as the ancestral homelands of the Mi’kmaq and Beothuk.
 
-Land acknowledgement of Beothuk & Mi'kmaq people?
-
-## References
+## Literature Cited

### 2020-02-19 paper/manuscript.Rmd
edit fogo info

@@
 Here, we describe the visual observation of an adult male caribou swimming
-between islands in the Fogo Island archipleago in the North Atlantic Ocean. This
-observation prompted us to query a large database of GPS radio-collared caribou
-in the Fogo Island caribou herd in Newfoundland, Canada. We identified  86
-remotely sensed swimming events between XX islands from three individual caribou
-over a three-year period.
+between islands in the Fogo Island archipleago in the North Atlantic Ocean.
+During routine fieldwork on May 30, 2017, we observed an unmarked adult male
+caribou swim between Perry and Eastern Indian Islands (Figure 1), a swim of at
+least 470 m which took ~9 minutes (~52 m per minute). This observation prompted
+us to query a large database of GPS radio-collared caribou in the Fogo Island
+caribou herd in Newfoundland, Canada. We identified  86 remotely sensed swimming
+events between XX islands from three individual caribou over a three-year
+period.
+
+Caribou were introduced to Fogo Island (n = 26 animals introduced between
+1964–67) and Change Island (n = 5 animals introduced in 1964) from the main
+Island of Newfoundland as part of a series of translocations and introductions
+throughout the province (Bergerud & Mercer 1989). While caribou populations are
+in decline throughout Canada (Festa-Bianchet et al. 2011), as well as in
+Newfoundland (Bastille-Rousseau et al. 2013), the Fogo Island herd has not
+declined to the same extent and currently consists of approximately 300 animals
+(Newfoundland and Labrador Wildlife Division, unpublished data). Since 2016 we
+have been studying the Fogo Island caribou herd (Bonar et al. 2017; Bonar et al.
+2018; Peignier et al. 2019). To date, we have fitted XX adult female caribou
+with GPS collars (1240g, GPS 4400M; Lotek Wireless Inc., Newmarket, Ontario,
+Canada), which were programmed to collect relocation fixes every 2 hours (see
+Bonar et al. 2018 for details).
 
@@
 
-Caribou were introduced to Fogo Island (n = 26 animals introduced between
-1964–67) and Change Island (n = 5 animals introduced in 1964) from the main
-Island of Newfoundland as part of a series of translocations and introductions
-throughout the province (Bergerud & Mercer 1989). While caribou populations are
-in decline throughout Canada (Festa-Bianchet et al. 2011), as well as in
-Newfoundland (Bastille-Rousseau et al. 2013), the Fogo Island herd has not
-declined to the same extent and currently consists of approximately 300 animals
-(Newfoundland and Labrador Wildlife Division, unpublished data).
-
 In Newfoundland, wolves (Canis lupus) are extirpated, so coyotes (Canis latrans)
@@
 caribou.
-
-We fitted 29 (n = 15 in 2016; n = 14 in 2017) adult female caribou on Fogo
-Island (n = 25) and Western Indian Island (n = 4) with GPS collars (1240g, GPS
-4400M; Lotek Wireless Inc., Newmarket, Ontario, Canada). Collars were deployed
-on individual caribou for two years and collars were redeployed on the same
-individuals for the duration of our study. Collars were programmed to collect
-relocation fixes every 2 hours. All animal capture and handling procedures were
-consistent with the American Society of Mammologists guidelines (Sikes & Gannon
-2011) and were approved by Memorial University Animal Use Protocol No. 20152067.
-During routine fieldwork on May 30, 2017, we observed an unmarked male caribou
-swim between Perry and Eastern Indian Islands, a swim of at least 470m which
-took ~9 minutes (~52m/minute). To determine the prevalence of ocean swimming for
-the Fogo Island caribou herd, we queried nearly 200,000 GPS relocations from
-26/29 collared adult female caribou between April 2016 and March 2019. We
-determined all instances of individual caribou moving between islands. We
-restricted the GPS dataset to the ice-free months of the year (beginning of
-April to end of November) and we assumed caribou swam in the ocean between
-islands when consecutive relocation fixes, i.e., 2 hour intervals, were located
-on different islands. For each swimming event, we identified the origin island,
-the terminal island, and the duration of time spent on the terminal island
-before swimming again.
 

### 2020-02-19 paper/manuscript.Rmd
Update manuscript.Rmd and manuscript.md

@@
 Here, we describe the visual observation of an adult male caribou swimming
-between islands in the Fogo Island archipleago in the North Atlantic Ocean.
-During routine fieldwork on May 30, 2017, we observed an unmarked adult male
-caribou swim between Perry and Eastern Indian Islands (Figure 1), a swim of at
-least 470 m which took ~9 minutes (~52 m per minute). This observation prompted
-us to query a large database of GPS radio-collared caribou in the Fogo Island
-caribou herd in Newfoundland, Canada. We identified  86 remotely sensed swimming
-events between XX islands from three individual caribou over a three-year
-period.
+between islands in the Fogo Island archipleago off the coast of Newfoundland,
+Canada. During routine fieldwork on May 30, 2017, we observed an unmarked adult
+male caribou swim between Perry and Eastern Indian Islands (Figure 1), a swim of
+at least 470 m which took ~9 minutes (~52 m per minute). This observation
+prompted us to query a large database of GPS radio-collared caribou. We
+identified  86 remotely sensed swimming events between XX islands from three
+individual caribou over a three-year period.
 
@@
 comprises approximately ten islands. While Fogo Island is the largest island
-(237.71 km^2), the archipelago includes at least three larger islands: Western
+(237.71 km^2^), the archipelago includes at least three larger islands: Western
 Indian (77.6 km2), Eastern Indian (38.7 km2) and Change (XX km2) Islands, as

### 2020-02-19 paper/manuscript.Rmd
refs

@@
   #  reference_docx: template.docx
-#bibliography: ../references/knitcite.bib
+bibliography: ../references/knitcite.bib
 #csl: csl.csl
@@
 that caribou swim in streams, rivers, and lakes during migration (Leblond et al.
-2016) to avoid predators (Bergerud 1985) and access islands during calving
+2016) to avoid predators [@Bergerud_1985] and access islands during calving
 (Bergerud et al. 1990). Swimming for quadrupedal terrestrial mammals can be

### 2020-02-19 Untitled.R

@@ -1,0 +1,4 @@
+
+
+### Packages ----
+library(knitcitations)

### 2020-02-19 paper/citations.R

@@ -1,0 +1,12 @@
+### Packages ----
+library(knitcitations)
+
+## only 15 citations
+
+## Bergerud 1985
+citep('10.1139/z85-199')
+
+
+### Write out bib ----
+write.bibtex(file = "references/knitcite.bib")
+

### 2020-02-19 paper/manuscript.Rmd
P1

@@
 energetic expenditure associated with swimming for quadrupedal mammals is
-significantly higher than walking or running (Fish 1993). Additionally, drowning
-is also possible (Miller & Gunn 1985) although the proportion of caribou that
-die from drowning is unknown. Despite a large number of coastal and island
-caribou herds, limited evidence exists suggesting that caribou swim in the ocean
-as readily as they swim in freshwater (see Table 1). A key exception is Peary
-caribou swimming up to 2.6 km between the Queen Elizabeth Islands (Miller 1995;
-Miller 2002).
+significantly higher than walking or running (Fish 1993), whiledrowning
+is also possible (Miller & Gunn 1985). Despite a large number of coastal and
+island caribou herds, only a handful of examples have documented caribou
+swimming in the ocean (Table 1).
 

### 2020-02-19 paper/manuscript.Rmd
P3

@@
 
+<!--DO WE NEED THIS SECTION?--> 
 Caribou were introduced to Fogo Island (n = 26 animals introduced between
@@
 
-We posit two potential mechanisms explaining the
-trade-offs associated with swimming in the North Atlantic Ocean: forage
-limitation and the associated density-dependent habitat selection and predator
-avoidance. For caribou living on small islands in the ocean, the same notions
-for why caribou swim in freshwater may also explain why caribou swim in the
-ocean (Leblond et al. 2016). Specifically, forage limitation (Miller 2002) and
+We posit two potential mechanisms explaining the trade-offs associated with
+swimming in the North Atlantic Ocean: forage limitation and predator avoidance.
+For caribou living on small islands in the ocean, the same notions for why
+caribou swim in freshwater may also explain why caribou swim in the ocean
+(Leblond et al. 2016). Specifically, forage limitation (Miller 2002) and
 predator avoidance (Jeffery et al. 2007) have been proposed as potential
 explanations.
-
-While caribou often swim across streams, rivers, and lakes (Leblond et al.
-2016), few observations exist of caribou swimming in the ocean (see Table 1). Of
-the existing or inferred observations of caribou swimming in the ocean, none
-appear to be as prolific as the number and frequency of swimming events we
-observed. We surmise that for caribou living on small islands, the primary
-drivers of swimming are likely related to forage limitation and predator
-avoidance.
 

### 2020-02-19 paper/manuscript.Rmd
edits reflow yes

@@
 
-In Newfoundland, wolves (Canis lupus) are extirpated, so coyotes (Canis latrans)
-and black bears (Ursus americanus) are the primary predators of caribou
-(Bastille-Rousseau et al. 2016). Coyotes and black bears are responsible for the
-majority of mortalities for neonate caribou calves (Bastille-Rousseau et al.
-2016), although predation can still occur after the calving period (Lewis &
-Mahoney 2014). However, on Fogo Island, coyotes are the only predator of
-caribou.
+Forage limitation and over-grazing is a major issue for caribou populations
+(Schaefer et al. 2016; Zamin et al. 2017) and could explain movement between
+islands. Reduced access to foraging opportunities can result in higher
+conspecific competition for resources and lowered reproductive success for adult
+females (Schaefer et al. 2016). Forage scarcity has been proposed as a potential
+reason why caribou move between arctic islands on the sea-ice in winter (Miller
+et al. 1977).  For very small islands, over-grazing by newly arrived caribou may
+rapidly limit forage (Bergerud et al. 1990). Without sea ice, movement from the
+mainland to an island as well as between islands requires caribou to assess the
+trade-off associated with swimming.
 
-Forage limitation could explain movement between islands. Forage scarcity has
-been proposed as a potential reason why caribou move between arctic islands on
-the sea-ice in winter (Miller et al. 1977). For very small islands, over-grazing
-by newly arrived caribou may rapidly limit forage (Bergerud et al. 1990).
-Without sea ice, movement from the mainland to an island as well as between
-islands requires caribou to assess the trade-off associated with swimming. The
-underlying mechanism driving forage limitation and the trade-off associated with
-swimming could be density-dependent habitat selection governed by the Ideal Free
-Distribution (Morris 1987; Bradbury et al. 2015). Ideal Free Distribution theory
-suggests that habitat selection is density-dependent and that variation in
-density between habitat patches leads to a fitness equilibrium (Bradbury et al.
-2015). Fitness in a habitat patch depends on density where, ideally, the
+The underlying mechanism driving forage limitation and the trade-off associated
+with swimming could be density-dependent habitat selection governed by the Ideal
+Free Distribution (Morris 1987; Bradbury et al. 2015). Ideal Free Distribution
+theory suggests that habitat selection is density-dependent and that variation
+in density between habitat patches leads to a fitness equilibrium (Bradbury et
+al. 2015). Fitness in a habitat patch depends on density where, ideally, the
 available resources on a habitat patch can sustain a specific number of
@@
 distinct habitat patches with clear costs and benefits associated with swimming.
+For caribou in the Fogo Island archipelago, the role of forage limitation in
+animal habitat selection patterns and movement remains unknown, but it is
+possible that forage limitation and increased conspecific competition influences
+swimming between islands.
 
-Woodland caribou are well known to disperse to remote locations, including
-islands, during calving (Bergerud & Page 1987; Cumming & Beange 1987). Moving to
-remote locations is a viable anti-predator strategy for caribou because it
-reduces detection and encounter rates by predators. For instance, the use of
-shoreline habitat in summer was deemed an effective anti-predator calving
-strategy because caribou avoided predators by using islands and peninsulas or
-swimming as a flight response (Bergerud et al. 1990). The use of water, and by
-association occasionally swimming, is therefore part of the fine-scale
-interactions between caribou and their predators and the use of islands appears
-to be an effective anti-predator strategy.
-
-Forage limitation and over-grazing is a major issue for caribou populations
-(Schaefer et al. 2016; Zamin et al. 2017). Reduced access to foraging
-opportunities can result in higher conspecific competition for resources and
-lowered reproductive success for adult females (Schaefer et al. 2016). For
-caribou in the Fogo Island archipelago, the role of forage limitation in animal
-habitat selection patterns and movement remains unknown, but it is possible that
-forage limitation and increased conspecific competition influences swimming
-between islands.
+Predator avoidance could contribute to swimming behaviour of caribou (Miller
+2002). Unlike elsewhere in the range of caribou, wolves (*Canis lupus*) are
+extirpated from Newfoundland, including Fogo Island, and coyotes (*Canis latrans*) and black bears
+(*Ursus americanus*) are the primary predators of caribou (Bastille-Rousseau et
+al. 2016). While predation by coyotes on adult female caribou is possible, it is
+relatively rare on Fogo Island (Webber & Vander Wal unpublished data).
+Therefore, any effect of predation on swimming behaviour is likely related to
+fine-scale encounters between coyotes and caribou which result in caribou
+swimming as a flight response. Alternatively, during calving, females may swim
+to small islands to give birth (Bergerud & Page 1987) or swim to islands with a
+calf at heel to avoid predation. Swimming with a calf at heel has been observed
+for caribou in freshwater lakes on Fogo Island (Webber & Vander Wal unpublished
+data), so we anticipate it may also be possible in the ocean. We suggest future
+studies assess encounter rates among caribou and coyotes to determine what
+proportion of encounters result in swimming behaviour.Woodland caribou are well
+known to disperse to remote locations, including islands, during calving
+(Bergerud & Page 1987; Cumming & Beange 1987). Moving to remote locations is a
+viable anti-predator strategy for caribou because it reduces detection and
+encounter rates by predators. For instance, the use of shoreline habitat in
+summer was deemed an effective anti-predator calving strategy because caribou
+avoided predators by using islands and peninsulas or swimming as a flight
+response (Bergerud et al. 1990). The use of water, and by association
+occasionally swimming, is therefore part of the fine-scale interactions between
+caribou and their predators and the use of islands appears to be an effective
+anti-predator strategy. Coyotes and black bears are responsible for the majority
+of mortalities for neonate caribou calves (Bastille-Rousseau et al. 2016),
+although predation can still occur after the calving period (Lewis & Mahoney
+2014). However, on Fogo Island, coyotes are the only predator of caribou.
 
@@
 population density increased, forage was depleted and competition increased.
-
-Predator avoidance could contribute to swimming behaviour of caribou in our
-system as well as others (Miller 2002). Unlike elsewhere in the range of
-caribou, wolves are extirpated from Newfoundland, including Fogo Island. In
-Newfoundland, coyotes are the primary predator of caribou (Bastille-Rousseau et
-al. 2016a). While predation by coyotes on adult female caribou is possible, it
-is relatively rare on Fogo Island (Webber & Vander Wal unpublished data).
-Therefore, any effect of predation on swimming behaviour is likely related to
-fine-scale encounters between coyotes and caribou which result in caribou
-swimming as a flight response. Alternatively, during calving, females may swim
-to small islands to give birth (Bergerud & Page 1987) or swim to islands with a
-calf at heel to avoid predation. Swimming with a calf at heel has been observed
-for caribou in freshwater lakes on Fogo Island (Webber & Vander Wal unpublished
-data), so we anticipate it may also be possible in the ocean. We suggest future
-studies assess encounter rates among caribou and coyotes to determine what
-proportion of encounters result in swimming behaviour.
 

### 2020-02-19 paper/manuscript.Rmd
wolves

@@
 Predator avoidance could contribute to swimming behaviour of caribou (Miller
-2002). Unlike elsewhere in the range of caribou, wolves (*Canis lupus*) are
-extirpated from Newfoundland, including Fogo Island, and coyotes (*Canis latrans*) and black bears
-(*Ursus americanus*) are the primary predators of caribou (Bastille-Rousseau et
-al. 2016). While predation by coyotes on adult female caribou is possible, it is
-relatively rare on Fogo Island (Webber & Vander Wal unpublished data).
-Therefore, any effect of predation on swimming behaviour is likely related to
-fine-scale encounters between coyotes and caribou which result in caribou
-swimming as a flight response. Alternatively, during calving, females may swim
-to small islands to give birth (Bergerud & Page 1987) or swim to islands with a
-calf at heel to avoid predation. Swimming with a calf at heel has been observed
-for caribou in freshwater lakes on Fogo Island (Webber & Vander Wal unpublished
-data), so we anticipate it may also be possible in the ocean. We suggest future
-studies assess encounter rates among caribou and coyotes to determine what
-proportion of encounters result in swimming behaviour.Woodland caribou are well
-known to disperse to remote locations, including islands, during calving
-(Bergerud & Page 1987; Cumming & Beange 1987). Moving to remote locations is a
-viable anti-predator strategy for caribou because it reduces detection and
-encounter rates by predators. For instance, the use of shoreline habitat in
-summer was deemed an effective anti-predator calving strategy because caribou
-avoided predators by using islands and peninsulas or swimming as a flight
-response (Bergerud et al. 1990). The use of water, and by association
-occasionally swimming, is therefore part of the fine-scale interactions between
-caribou and their predators and the use of islands appears to be an effective
-anti-predator strategy. Coyotes and black bears are responsible for the majority
-of mortalities for neonate caribou calves (Bastille-Rousseau et al. 2016),
-although predation can still occur after the calving period (Lewis & Mahoney
-2014). However, on Fogo Island, coyotes are the only predator of caribou.
+2002). Although caribou co-evolved with wolves (*Canis lupus*) throughout their range, wolves are extirpated from Newfoundland, including Fogo Island. Coyotes (*Canis latrans*)
+and black bears are the primary predators of
+caribou (Bastille-Rousseau et al. 2016). While predation by coyotes on adult
+female caribou is possible, it is relatively rare on Fogo Island (Webber &
+Vander Wal unpublished data). Therefore, any effect of predation on swimming
+behaviour is likely related to fine-scale encounters between coyotes and caribou
+which result in caribou swimming as a flight response. Alternatively, during
+calving, females may swim to small islands to give birth (Bergerud & Page 1987)
+or swim to islands with a calf at heel to avoid predation. Swimming with a calf
+at heel has been observed for caribou in freshwater lakes on Fogo Island (Webber
+& Vander Wal unpublished data), so we anticipate it may also be possible in the
+ocean. We suggest future studies assess encounter rates among caribou and
+coyotes to determine what proportion of encounters result in swimming
+behaviour.Woodland caribou are well known to disperse to remote locations,
+including islands, during calving (Bergerud & Page 1987; Cumming & Beange 1987).
+Moving to remote locations is a viable anti-predator strategy for caribou
+because it reduces detection and encounter rates by predators. For instance, the
+use of shoreline habitat in summer was deemed an effective anti-predator calving
+strategy because caribou avoided predators by using islands and peninsulas or
+swimming as a flight response (Bergerud et al. 1990). The use of water, and by
+association occasionally swimming, is therefore part of the fine-scale
+interactions between caribou and their predators and the use of islands appears
+to be an effective anti-predator strategy. Coyotes and black bears are
+responsible for the majority of mortalities for neonate caribou calves
+(Bastille-Rousseau et al. 2016), although predation can still occur after the
+calving period (Lewis & Mahoney 2014). However, on Fogo Island, coyotes are the
+only predator of caribou.
 

### 2020-02-19 paper/manuscript.Rmd
calving swim-bo-baggins

@@
 Predator avoidance could contribute to swimming behaviour of caribou (Miller
-2002). Although caribou co-evolved with wolves (*Canis lupus*) throughout their range, wolves are extirpated from Newfoundland, including Fogo Island. Coyotes (*Canis latrans*)
-and black bears are the primary predators of
-caribou (Bastille-Rousseau et al. 2016). While predation by coyotes on adult
-female caribou is possible, it is relatively rare on Fogo Island (Webber &
-Vander Wal unpublished data). Therefore, any effect of predation on swimming
-behaviour is likely related to fine-scale encounters between coyotes and caribou
-which result in caribou swimming as a flight response. Alternatively, during
-calving, females may swim to small islands to give birth (Bergerud & Page 1987)
-or swim to islands with a calf at heel to avoid predation. Swimming with a calf
-at heel has been observed for caribou in freshwater lakes on Fogo Island (Webber
-& Vander Wal unpublished data), so we anticipate it may also be possible in the
-ocean. We suggest future studies assess encounter rates among caribou and
-coyotes to determine what proportion of encounters result in swimming
-behaviour.Woodland caribou are well known to disperse to remote locations,
+2002). Although caribou co-evolved with wolves (*Canis lupus*) throughout their
+range, wolves are extirpated from Newfoundland, including Fogo Island. Coyotes
+(*Canis latrans*) and black bears (*Ursus americanus*) are the primary predators
+of caribou in Newfoundland (Bastille-Rousseau et al. 2016). While predation by
+coyotes on adult female caribou does occur (ref?), it is relatively rare and the
+effect of predation on swimming behaviour is more likely the result of
+encounters which result in a flight response. For example, during calving,
+females may swim to small islands to give birth (Bergerud & Page 1987) or swim
+to islands in freshwater lakes with their calves to avoid predation.
+
+Woodland caribou are also well known to disperse to remote locations,
 including islands, during calving (Bergerud & Page 1987; Cumming & Beange 1987).

### 2020-02-19 paper/manuscript.Rmd
Update manuscript.Rmd and manuscript.docx

@@
 output: 
-    github_document
+   # github_document
   # bookdown::pdf_document2:
@@
   #    toc: false
-  #bookdown::word_document2:
+  bookdown::word_document2:
   #  reference_docx: template.docx
@@
 
+## Introduction
 Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
@@
 
+<!--I don't know if we need this and if yes, where it will go - thoughts?--> 
 The Fogo Island archipelago is located in Newfoundland, Canada (49°N, 54°W), and
@@
 
+## Forage limitation
 Forage limitation and over-grazing is a major issue for caribou populations
@@
 
+## Predator avoidance
 Predator avoidance could contribute to swimming behaviour of caribou (Miller
@@
 
+## Summary and conclusion
 Caribou were introduced to Fogo Island in the 1960s and for several decades the
@@
 
-Our observations represent the most frequent and numerous swimming events for
-caribou in the ocean (for comparison see Table 1). While past observations (e.g.
-Miller 1995, 2002; Jeffery et al. 2007; Ricca et al. 2012) have noted the unique
-nature of caribou swimming in the ocean, we suggest that this behaviour is
-likely more common than previously thought for caribou living on oceanic
-islands. We posit forage limitation and density-dependent habitat selection is
-an ultimate explanation for why caribou swimming in the ocean while predator
-avoidance is a proximate explanation. Although our inference is limited to
-observations, we suggest that because swimming behaviour is so common in our
-system, that density-dependent habitat selection and predation are plausible
-mechanisms explaining swimming behaviour in Fogo Island caribou.
+Our observations add to the evidence that caribou can, and occaissionaly do,
+swim in the ocean (Table 1). While past observations (e.g. Miller 1995, 2002;
+Jeffery et al. 2007; Ricca et al. 2012) have noted the unique nature of caribou
+swimming in the ocean, we suggest that this behaviour is likely more common than
+previously thought for caribou living on oceanic islands. We posit forage
+limitation and the associated density-dependent habitat selection is an ultimate
+explanation for why caribou swimming in the ocean while predator avoidance is a
+proximate explanation. Although our inference is limited to observations, our
+observations suggest that islands represent discrete foraging patches for
+terrestrial animals that could influence fitness via increased foraging
+opportunities and safety from predators.
 

### 2020-02-19 paper/manuscript.Rmd
restructured a lot of ms: - changed up intro to explain what we did, and then discussed some hypotheses post-hoc -

@@
 2016) to avoid predators [@Bergerud_1985] and access islands during calving
-(Bergerud et al. 1990). Swimming for quadrupedal terrestrial mammals can be
-thought of as a set of trade-offs. Costs include high energy expenditure and
-risk of drowning. Specifically, even though caribou are adept swimmers, the
-energetic expenditure associated with swimming for quadrupedal mammals is
-significantly higher than walking or running (Fish 1993), whiledrowning
+(Bergerud et al. 1990). Even for adept swimmers like caribou, the
+energetic expenditure of swimming for quadrupedal mammals is
+significantly higher than walking or running (Fish 1993), while drowning
 is also possible (Miller & Gunn 1985). Despite a large number of coastal and
@@
 
-Here, we describe the visual observation of an adult male caribou swimming
-between islands in the Fogo Island archipleago off the coast of Newfoundland,
-Canada. During routine fieldwork on May 30, 2017, we observed an unmarked adult
-male caribou swim between Perry and Eastern Indian Islands (Figure 1), a swim of
-at least 470 m which took ~9 minutes (~52 m per minute). This observation
-prompted us to query a large database of GPS radio-collared caribou. We
-identified  86 remotely sensed swimming events between XX islands from three
-individual caribou over a three-year period.
+The Fogo Island archipelago, off the coast of Newfoundland, Canada (49°N, 54°W),
+is home to approximately 300 woodland caribou. During routine fieldwork on May
+30, 2017, we observed an unmarked adult male caribou swim between Perry and
+Eastern Indian Islands (Figure 1), a swim of at least 470 m which took ~9
+minutes (~52 m per minute). This observation prompted us to investigate the
+subset of GPS radio-collared caribou in the population (n = 29 adult females).
+Using these remotely sensed relocations, we determined that this observation was
+not unique and that several individuals engaged in this behaviour (details below).
+<!--Is it better to just state our results here, or give this sort of teaser?
+Initially I had "86 events by 3 individuals" here, but then we ignore that for a couple paragraphs of context before getting back into it...
+which seemed weird-->
 
-<!--DO WE NEED THIS SECTION?--> 
-Caribou were introduced to Fogo Island (n = 26 animals introduced between
-1964–67) and Change Island (n = 5 animals introduced in 1964) from the main
+Fogo Island is the largest island in the archipelago (237.71 km^2^), but there
+are at least three other large islands: Western Indian (77.6 km2), Eastern
+Indian (38.7 km2) and Change (XX km2) Islands, as well as numerous smaller
+islands, including Blundon’s (1.18 km2), North Long (1.01 km2), South Long (0.48
+km2), Kate’s (1.64 km2), and Brother’s (1.59 km2) Islands (see Figure 1).
+Habitats are similar across the archipelago, consisting largely of coniferous
+and mixed forests of balsam fir (*Abies balsamea*), black spruce (*Picea
+mariana*), and white birch (*Betula papyrifera*) as well as bogs, lakes, lichen
+and rocky barrens.
+
+Caribou were introduced to Fogo Island (n = 26 animals between
+1964–67) and Change Island (n = 5 animals in 1964) from the main
 Island of Newfoundland as part of a series of translocations and introductions
@@
 have been studying the Fogo Island caribou herd (Bonar et al. 2017; Bonar et al.
-2018; Peignier et al. 2019). To date, we have fitted XX adult female caribou
+2018; Peignier et al. 2019). To date, we have fitted 29 adult female caribou
 with GPS collars (1240g, GPS 4400M; Lotek Wireless Inc., Newmarket, Ontario,
 Canada), which were programmed to collect relocation fixes every 2 hours (see
-Bonar et al. 2018 for details).
+Bonar et al. 2018 for details). We used GPS locations from these individuals to investigate oceanic swimming events.
 
-We posit two potential mechanisms explaining the trade-offs associated with
-swimming in the North Atlantic Ocean: forage limitation and predator avoidance.
-For caribou living on small islands in the ocean, the same notions for why
-caribou swim in freshwater may also explain why caribou swim in the ocean
-(Leblond et al. 2016). Specifically, forage limitation (Miller 2002) and
-predator avoidance (Jeffery et al. 2007) have been proposed as potential
-explanations.
+We identified swimming events as two consecutive GPS locations from an
+individual on different islands. This region typically experiences pack ice
+during the winter, so that caribou may be able to travel between islands by
+walking over the ice. We restricted our GPS data to the ice-free period of the
+year (April 1 - November 25) to avoid this possibility. Using these criteria, we identified three adult female caribou that collectively engaged in 86 swimming events (Fig 1??).
 
-<!--I don't know if we need this and if yes, where it will go - thoughts?--> 
-The Fogo Island archipelago is located in Newfoundland, Canada (49°N, 54°W), and
-comprises approximately ten islands. While Fogo Island is the largest island
-(237.71 km^2^), the archipelago includes at least three larger islands: Western
-Indian (77.6 km2), Eastern Indian (38.7 km2) and Change (XX km2) Islands, as
-well as numerous smaller islands, including Blundon’s (1.18 km2), North Long
-(1.01 km2), South Long (0.48 km2), Kate’s (1.64 km2), and Brother’s (1.59 km2)
-Islands (see Figure 1). All islands in the Fogo Island archipelago consist
-largely of coniferous and mixed forests of balsam fir (Abies balsamea), black
-spruce (Picea mariana), and white birch (Betula payrifera) as well as bogs,
-lakes, lichen and rocky barrens.
+Caribou may swim between islands in the ocean for similar reasons that drive caribou swimming in freshwater (Leblond et al. 2016). We present two such
+explanations for these swimming events, forage limitation (Miller 2002) and
+predator avoidance (Jeffery et al. 2007), and discuss why these swimming events appear to be relatively commonplace for only a small subset of the population.
 
 ## Forage limitation
-Forage limitation and over-grazing is a major issue for caribou populations
+Forage limitation and over-grazing is a major concern for caribou populations
 (Schaefer et al. 2016; Zamin et al. 2017) and could explain movement between
-islands. Reduced access to foraging opportunities can result in higher
-conspecific competition for resources and lowered reproductive success for adult
-females (Schaefer et al. 2016). Forage scarcity has been proposed as a potential
-reason why caribou move between arctic islands on the sea-ice in winter (Miller
-et al. 1977).  For very small islands, over-grazing by newly arrived caribou may
-rapidly limit forage (Bergerud et al. 1990). Without sea ice, movement from the
-mainland to an island as well as between islands requires caribou to assess the
+islands. Conspecific competition for resources driven by reduced foraging opportunities can decrease reproductive success of adult females
+(Schaefer et al. 2016). Forage scarcity has been proposed as a potential reason
+for caribou moving between arctic islands on the sea-ice in winter (Miller et
+al. 1977).  For very small islands, over-grazing by newly arrived caribou may
+rapidly deplete forage (Bergerud et al. 1990). Without sea ice, movement from the
+mainland to an island, or between islands, requires caribou to assess the
 trade-off associated with swimming.
 
-The underlying mechanism driving forage limitation and the trade-off associated
-with swimming could be density-dependent habitat selection governed by the Ideal
-Free Distribution (Morris 1987; Bradbury et al. 2015). Ideal Free Distribution
-theory suggests that habitat selection is density-dependent and that variation
-in density between habitat patches leads to a fitness equilibrium (Bradbury et
-al. 2015). Fitness in a habitat patch depends on density where, ideally, the
-available resources on a habitat patch can sustain a specific number of
-individuals. In a hypothetical example, when density of a given habitat patch
-has exceeded the fitness equilibrium, animals are predicted to relocate and
-settle new habitat patches to reach equilibrium. Islands therefore represent
-distinct habitat patches with clear costs and benefits associated with swimming.
-For caribou in the Fogo Island archipelago, the role of forage limitation in
-animal habitat selection patterns and movement remains unknown, but it is
-possible that forage limitation and increased conspecific competition influences
-swimming between islands.
+This trade-off can be understood as density-dependent habitat selection governed
+by the Ideal Free Distribution (Morris 1987; Bradbury et al. 2015). Ideal Free
+Distribution theory predicts a fitness equilibrium: when the density in a given
+habitat patch has exceeded the optimum for fitness within that patch, animals
+should relocate and settle new habitat patches, so that fitness is equal across
+all patches (Bradbury et al. 2015). Islands act as discrete habitat patches in
+this case, but the costs associated with swimming create a trade-off when moving
+between patches. For caribou in the Fogo Island archipelago, the role of forage
+limitation in animal habitat selection patterns and movement remains unknown,
+but swimming between islands may in part be explained by forage limitation and
+conspecific competition.
 
 ## Predator avoidance
-Predator avoidance could contribute to swimming behaviour of caribou (Miller
-2002). Although caribou co-evolved with wolves (*Canis lupus*) throughout their
-range, wolves are extirpated from Newfoundland, including Fogo Island. Coyotes
-(*Canis latrans*) and black bears (*Ursus americanus*) are the primary predators
-of caribou in Newfoundland (Bastille-Rousseau et al. 2016). While predation by
-coyotes on adult female caribou does occur (ref?), it is relatively rare and the
-effect of predation on swimming behaviour is more likely the result of
-encounters which result in a flight response. For example, during calving,
-females may swim to small islands to give birth (Bergerud & Page 1987) or swim
-to islands in freshwater lakes with their calves to avoid predation.
+Predator avoidance could also contribute to swimming behaviour of caribou
+(Miller 2002). During calving, female caribou are known to swim to small islands
+in freshwater lakes give birth, or with their calves after birth, to avoid
+predation (Bergerud & Page 1987). The use of shoreline habitat in summer is
+considered an effective anti-predator calving strategy by allowing for swimming
+to islands or peninsulas as a predation avoidance flight response (Bergerud et
+al. 1990). The use of water, and by association occasionally swimming, is
+therefore part of the fine-scale interactions between caribou and their
+predators and the use of islands appears to be an effective anti-predator
+strategy.
 
-Woodland caribou are also well known to disperse to remote locations,
-including islands, during calving (Bergerud & Page 1987; Cumming & Beange 1987).
-Moving to remote locations is a viable anti-predator strategy for caribou
-because it reduces detection and encounter rates by predators. For instance, the
-use of shoreline habitat in summer was deemed an effective anti-predator calving
-strategy because caribou avoided predators by using islands and peninsulas or
-swimming as a flight response (Bergerud et al. 1990). The use of water, and by
-association occasionally swimming, is therefore part of the fine-scale
-interactions between caribou and their predators and the use of islands appears
-to be an effective anti-predator strategy. Coyotes and black bears are
-responsible for the majority of mortalities for neonate caribou calves
-(Bastille-Rousseau et al. 2016), although predation can still occur after the
-calving period (Lewis & Mahoney 2014). However, on Fogo Island, coyotes are the
-only predator of caribou.
+Historic predators of caribou such as wolves (*Canis lupus*) and black bears (*Ursus americanus*) are not present in the Fogo Island archipelago; coyotes (*Canis latrans*) are the top predator in this region.
+Predation by
+coyotes on adult caribou is relatively rare (ref?), so the potential predator effect on caribou swimming behaviour is likely to be a flight response to non-lethal encounters.
+Calves are susceptible to coyotes, particularly young calves, but the energetic costs to oceanic swimming are likely to be far greater for small calves.
+
+<!-- did the swimming events occur during the calving season? were these three breeders/non-breeders? did any of these females make these journeys with their calves?  -->
+
+## Individual variation 
+Three of 29 GPS-collared female caribou swam between islands in the Fogo
+archipelago. These swimming events were relatively frequent, on average occuring
+once every 24 days within our date range. However, we identified zero swimming
+events for the other 26 collared females. It remains unclear why this behaviour
+is seemingly common among 10% of our marked population, while 90% never engaged
+in it. Our initial direct observation of swimming was an adult male, but as we
+only have GPS data for females, we are unable to make conclusions on the
+prevalence of this behaviour among male caribou. It is theoretically possible
+that these other females moved rapidly back and forth between islands within the
+2 hour fix rate of our GPS collars, which we would be unable to detect. We
+consider this highly unlikely given the range of residency times for caribou on
+any given island (min - max: XX - XX).
+<!-- What is the typical residency time? Is this a true statement or am I just making this up? -->
+
+The two hypotheses presented above are generally thought to apply uniformly
+across populations, but individual caribou could evaluate the costs and benefits
+of swimming to another island differently. Potentially those three females,
+compared to the other 26, considered swimming less costly or perceived
+competition or predation to be of greater concern. The individual attributes
+that might influence how animals vary in their assessments of these trade-offs
+are unknown, and potential predictors such as age or personality traits are
+unknown for these caribou.
 
 ## Summary and conclusion
-Caribou were introduced to Fogo Island in the 1960s and for several decades the
+For several decades after their introduction to Fogo Island, the caribou
 population did not exceed ~100 individuals (Bergerud & Mercer 1989; Newfoundland
 and Labrador Wildlife Division, unpublished data). During the 1990s, population
-density reached 300 animals and anecdotal evidence suggests that caribou began
-to occupy other islands in the archipelago. We surmise that, as density
-increased over time, competition among conspecifics and density-dependent
-habitat selection resulted in expansion of the population to nearby islands.
-Following density-dependent habitat selection theory, caribou should swim to new
-islands when the average fitness of individuals on the starting island exceeds
-the density-fitness equilibrium (Morris 1987). Animals would therefore be
-predicted to swim to new islands (i.e. settle new habitat patches) to reach a
-fitness equilibrium. While this is an ultimate explanation to a series of
-proximate observations, it is possible for density-dependent habitat selection
-to operate at such a fine-scale (Webber & Vander Wal 2018). Evidence from the
-Sable Island horse (Equus ferus caballus) system corroborates our suggestion
-that swimming among islands could be driven by density-dependent habitat
-selection (van Beest et al. 2014). For horses, selection for high quality
-habitat was highest when population density was low, but individuals tended to
-settle in lower quality habitats as population density increased, following
-Ideal Free Distribution theory (van Beest et al. 2014). Taken together, caribou
-may follow density-dependent habitat selection (Wittmer et al. 2007) and as
-population density increased, forage was depleted and competition increased.
+density reached 300 animals and anecdotal evidence suggests that [only then???]
+caribou began to occupy other islands in the archipelago. We surmise that as
+competition among conspecifics increased along with population size,
+density-dependent habitat selection resulted in expansion of the population to
+nearby islands. Following Ideal Free Distribution theory, caribou should swim to
+new islands when the average fitness of individuals on the starting island
+exceeds the density-fitness equilibrium (Morris 1987). While this is an ultimate
+explanation to a series of proximate observations, it is possible for
+density-dependent habitat selection to operate at such a fine-scale (Webber &
+Vander Wal 2018). Evidence from the Sable Island horse (Equus ferus caballus)
+system corroborates this hypothesis (van Beest et al. 2014). For horses,
+selection for high quality habitat was strongest at low population densities,
+but individuals tended to settle in lower quality habitats as population density
+increased, following Ideal Free Distribution theory (van Beest et al. 2014).
+Predator avoidance can also contribute to caribou swimming behaviour, but in
+this situation it seems a less likely explanation given predation is primarily a
+concern for young calves, which are the least likely to engage in this costly
+behaviour.
+<!-- this last bit is sort of me bullshitting, not sure it's actually defensible?-->
 
-Our observations add to the evidence that caribou can, and occaissionaly do,
-swim in the ocean (Table 1). While past observations (e.g. Miller 1995, 2002;
-Jeffery et al. 2007; Ricca et al. 2012) have noted the unique nature of caribou
+Our observations add to the evidence that caribou can, and occasionally do, swim
+in the ocean (Table 1). While past observations (e.g. Miller 1995, 2002; Jeffery
+et al. 2007; Ricca et al. 2012) have noted the exceptional nature of caribou
 swimming in the ocean, we suggest that this behaviour is likely more common than
@@
 limitation and the associated density-dependent habitat selection is an ultimate
-explanation for why caribou swimming in the ocean while predator avoidance is a
-proximate explanation. Although our inference is limited to observations, our
-observations suggest that islands represent discrete foraging patches for
-terrestrial animals that could influence fitness via increased foraging
-opportunities and safety from predators.
+explanation for why caribou swimming in the ocean, with predator avoidance as a
+potential proximate explanation. Although our inference is limited to
+observations, our observations suggest that islands represent discrete foraging
+patches for terrestrial animals that could influence fitness via increased
+foraging opportunities and safety from predators. We also suggest that
+individuals vary in their evaluation of costs and benefits of movement between
+these habitat patches.
 
@@
 We thank M. Laforge, M. Bonar, C. Hart, and R. Huang for help in the field.
-Logistical support was provided L. Bixby. We also thank members of the
+Logistical support was provided by L. Bixby. We also thank members of the
 Newfoundland Wildlife Division including S. Moores, B. Adams, K. Lewis, C.
@@
 thank T. Bergerud and S. Mahoney for their vision in initiating much of the work
-on caribou in Newfoundland. Funding for this study was provided by a National
-Sciences and Engineering Research Council (NSERC) Vanier Canada Graduate
-Scholarship to QMRW, NSERC Canda Graduate Scholarship to JGH, and a NSERC
-Discovery Grant to EVW. We respectfully acknowledge the territory in which data
-were collected and analyzed as the ancestral homelands of the Beothuk, and the
-island of Newfoundland as the ancestral homelands of the Mi’kmaq and Beothuk.
+on caribou in Newfoundland. Funding for this study was provided by the National
+Sciences and Engineering Research Council (QMRW, JGH, EVW). We respectfully
+acknowledge the territory in which data were collected and analyzed as the
+ancestral homelands of the Beothuk, and the island of Newfoundland as the
+ancestral homelands of the Mi’kmaq and Beothuk.
+<!-- Simplified the funding statement, do we need to name the specific scholarships/grants? Eric opinion??-->
 

### 2020-02-19 paper/manuscript.Rmd
sorry I didn't finish my comments on the last stage -___- I also: - streamlined intro/context - explained methods - added section on individual variation (classic Jack, tbh)

@@
 Island of Newfoundland as part of a series of translocations and introductions
-throughout the province (Bergerud & Mercer 1989). While caribou populations are
-in decline throughout Canada (Festa-Bianchet et al. 2011), as well as in
-Newfoundland (Bastille-Rousseau et al. 2013), the Fogo Island herd has not
-declined to the same extent and currently consists of approximately 300 animals
-(Newfoundland and Labrador Wildlife Division, unpublished data). Since 2016 we
+throughout the province (Bergerud & Mercer 1989). Since 2016 we
 have been studying the Fogo Island caribou herd (Bonar et al. 2017; Bonar et al.

### 2020-02-20 paper/manuscript.Rmd
Update manuscript.Rmd, manuscript.docx, and ~$nuscript.docx

@@
 author: Quinn M.R. Webber^1\*^, Jack G. Hendrix^1^, Alec L. Robitaille^2^,  Eric Vander Wal^1,2^
-date: |
-  | ^1^Cognitive and Behavioural Ecology Interdisciplinary Program, Memorial University of Newfoundland
-  | ^2^Department of Biology, Memorial University of Newfoundland, Canada
-  | \*Corresponding author: Quinn M.R. Webber (webber.quinn@gmail.com)
+date: 
+   ^1^Cognitive and Behavioural Ecology Interdisciplinary Program, Memorial University of Newfoundland
+   ^2^Department of Biology, Memorial University of Newfoundland, Canada
+   #\*Corresponding author: Quinn M.R. Webber (webber.quinn@gmail.com)
 output: 
@@
 
-Caribou were introduced to Fogo Island (n = 26 animals between
-1964–67) and Change Island (n = 5 animals in 1964) from the main
-Island of Newfoundland as part of a series of translocations and introductions
-throughout the province (Bergerud & Mercer 1989). Since 2016 we
-have been studying the Fogo Island caribou herd (Bonar et al. 2017; Bonar et al.
-2018; Peignier et al. 2019). To date, we have fitted 29 adult female caribou
-with GPS collars (1240g, GPS 4400M; Lotek Wireless Inc., Newmarket, Ontario,
-Canada), which were programmed to collect relocation fixes every 2 hours (see
-Bonar et al. 2018 for details). We used GPS locations from these individuals to investigate oceanic swimming events.
+Caribou were introduced to Fogo Island (n = 26 animals between 1964–67) and
+Change Island (n = 5 animals in 1964) from the main Island of Newfoundland as
+part of a series of translocations and introductions throughout the province
+(Bergerud & Mercer 1989). Since 2016 we have been studying the Fogo Island
+caribou herd (Bonar et al. 2017; Bonar et al. 2018; Peignier et al. 2019). To
+date, we have fitted 29 adult female caribou with GPS collars (1240g, GPS 4400M;
+Lotek Wireless Inc., Newmarket, Ontario, Canada), which were programmed to
+collect relocation fixes every 2 hours (see Bonar et al. 2018 for details). We
+used GPS locations from these individuals to investigate oceanic swimming
+events.
 

### 2020-02-20 paper/manuscript.Rmd
P2
reflow

@@
 output: 
-   # github_document
+    github_document
   # bookdown::pdf_document2:
@@
   #    toc: false
-  bookdown::word_document2:
+  #bookdown::word_document2:
   #  reference_docx: template.docx
@@
 Eastern Indian Islands (Figure 1), a swim of at least 470 m which took ~9
-minutes (~52 m per minute). This observation prompted us to investigate the
-subset of GPS radio-collared caribou in the population (n = 29 adult females).
-Using these remotely sensed relocations, we determined that this observation was
-not unique and that several individuals engaged in this behaviour (details below).
+minutes (~52 m per minute). This observation prompted us to investigate whether
+GPS radio-collared caribou in the population (n = 29 adult females) also swam
+between islands. Using these remotely sensed relocations, we determined that
+this observation was not unique and that several individuals engaged in this
+behaviour (details below).
 <!--Is it better to just state our results here, or give this sort of teaser?
@@
 individual on different islands. This region typically experiences pack ice
-during the winter, so that caribou may be able to travel between islands by
+during the winter, so caribou may be able to travel between islands by
 walking over the ice. We restricted our GPS data to the ice-free period of the
-year (April 1 - November 25) to avoid this possibility. Using these criteria, we identified three adult female caribou that collectively engaged in 86 swimming events (Fig 1??).
+year (April 1 - November 25) to avoid this possibility. Using these criteria, we
+identified three adult female caribou that collectively engaged in 86 swimming
+events (Figure 2).
 
-Caribou may swim between islands in the ocean for similar reasons that drive caribou swimming in freshwater (Leblond et al. 2016). We present two such
+Caribou may swim between islands in the ocean for similar reasons that drive
+caribou swimming in freshwater (Leblond et al. 2016). We present two
 explanations for these swimming events, forage limitation (Miller 2002) and
-predator avoidance (Jeffery et al. 2007), and discuss why these swimming events appear to be relatively commonplace for only a small subset of the population.
+predator avoidance (Jeffery et al. 2007), and discuss why these swimming events
+appear to be relatively commonplace in the population.
 

### 2020-02-21 paper/manuscript.Rmd
reflow
Update citations.R, manuscript.Rmd, and manuscript.md

@@
 Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
-that caribou swim in streams, rivers, and lakes during migration (Leblond et al.
-2016) to avoid predators [@Bergerud_1985] and access islands during calving
+that caribou swim in streams, rivers, and lakes during migration [@Leblond_2016] 
+to avoid predators [@Bergerud_1985] and access islands during calving
 (Bergerud et al. 1990). Even for adept swimmers like caribou, the
@@
 (Schaefer et al. 2016; Zamin et al. 2017) and could explain movement between
-islands. Conspecific competition for resources driven by reduced foraging opportunities can decrease reproductive success of adult females
-(Schaefer et al. 2016). Forage scarcity has been proposed as a potential reason
-for caribou moving between arctic islands on the sea-ice in winter (Miller et
-al. 1977).  For very small islands, over-grazing by newly arrived caribou may
-rapidly deplete forage (Bergerud et al. 1990). Without sea ice, movement from the
+islands. Conspecific competition for resources driven by reduced foraging
+opportunities can decrease reproductive success of adult females (Schaefer et
+al. 2016). Forage scarcity has been proposed as a potential reason for caribou
+moving between arctic islands on the sea-ice in winter (Miller et al. 1977).
+For very small islands, over-grazing by newly arrived caribou may rapidly
+deplete forage (Bergerud et al. 1990). Without sea ice, movement from the
 mainland to an island, or between islands, requires caribou to assess the
@@
 
-Historic predators of caribou such as wolves (*Canis lupus*) and black bears (*Ursus americanus*) are not present in the Fogo Island archipelago; coyotes (*Canis latrans*) are the top predator in this region.
-Predation by
-coyotes on adult caribou is relatively rare (ref?), so the potential predator effect on caribou swimming behaviour is likely to be a flight response to non-lethal encounters.
-Calves are susceptible to coyotes, particularly young calves, but the energetic costs to oceanic swimming are likely to be far greater for small calves.
+Historic predators of caribou such as wolves (*Canis lupus*) and black bears
+(*Ursus americanus*) are not present in the Fogo Island archipelago; coyotes
+(*Canis latrans*) are the top predator in this region. Predation by coyotes on
+adult caribou is relatively rare (ref?), so the potential predator effect on
+caribou swimming behaviour is likely to be a flight response to non-lethal
+encounters. Calves are susceptible to coyotes, particularly young calves, but
+the energetic costs to oceanic swimming are likely to be far greater for small
+calves.
 

### 2020-02-21 paper/citations.R

@@ -6,6 +6,8 @@ library(knitcitations)
 ## Bergerud 1985
 citep('10.1139/z85-199')
 
+## Leblond et al 2016
+citep('10.1186/s40462-016-0079-4')
 
 ### Write out bib ----
 write.bibtex(file = "references/knitcite.bib")

### 2020-02-21 paper/manuscript.Rmd
P3

@@
 Eastern Indian Islands (Figure 1), a swim of at least 470 m which took ~9
-minutes (~52 m per minute). This observation prompted us to investigate whether
-GPS radio-collared caribou in the population (n = 29 adult females) also swam
-between islands. Using these remotely sensed relocations, we determined that
-this observation was not unique and that several individuals engaged in this
-behaviour (details below).
+minutes (~52 m per minute).
+This observation prompted us to investigate whether swimming behaviour was common among GPS radio-collared caribou (n = 29 adult females) in the population.
+We identified swimming events as two consecutive GPS locations from an
+individual on different islands. This region typically experiences pack ice
+during the winter, so caribou may be able to travel between islands by
+walking over the ice. We restricted our GPS data to the ice-free period of the
+year (April 1 - November 25) to avoid this possibility. Using these criteria, we
+identified three adult female caribou that collectively engaged in 86 swimming
+events (Figure 2).
 <!--Is it better to just state our results here, or give this sort of teaser?
@@
 which seemed weird-->
-
-Fogo Island is the largest island in the archipelago (237.71 km^2^), but there
-are at least three other large islands: Western Indian (77.6 km2), Eastern
-Indian (38.7 km2) and Change (XX km2) Islands, as well as numerous smaller
-islands, including Blundon’s (1.18 km2), North Long (1.01 km2), South Long (0.48
-km2), Kate’s (1.64 km2), and Brother’s (1.59 km2) Islands (see Figure 1).
-Habitats are similar across the archipelago, consisting largely of coniferous
-and mixed forests of balsam fir (*Abies balsamea*), black spruce (*Picea
-mariana*), and white birch (*Betula papyrifera*) as well as bogs, lakes, lichen
-and rocky barrens.
 
@@
 
-We identified swimming events as two consecutive GPS locations from an
-individual on different islands. This region typically experiences pack ice
-during the winter, so caribou may be able to travel between islands by
-walking over the ice. We restricted our GPS data to the ice-free period of the
-year (April 1 - November 25) to avoid this possibility. Using these criteria, we
-identified three adult female caribou that collectively engaged in 86 swimming
-events (Figure 2).
+Fogo Island is the largest island in the archipelago (237.71 km^2^), but there
+are at least three other large islands: Western Indian (77.6 km2), Eastern
+Indian (38.7 km2) and Change (XX km2) Islands, as well as numerous smaller
+islands, including Blundon’s (1.18 km2), North Long (1.01 km2), South Long (0.48
+km2), Kate’s (1.64 km2), and Brother’s (1.59 km2) Islands (see Figure 1).
+Habitats are similar across the archipelago, consisting largely of coniferous
+and mixed forests of balsam fir (*Abies balsamea*), black spruce (*Picea
+mariana*), and white birch (*Betula papyrifera*) as well as bogs, lakes, lichen
+and rocky barrens.
 

### 2020-02-21 paper/manuscript.Rmd
blah blah blah

@@
 to avoid predators [@Bergerud_1985] and access islands during calving
-(Bergerud et al. 1990). Even for adept swimmers like caribou, the
+[@Bergerud_1990]. Even for adept swimmers like caribou, the
 energetic expenditure of swimming for quadrupedal mammals is
 significantly higher than walking or running (Fish 1993), while drowning
-is also possible (Miller & Gunn 1985). Despite a large number of coastal and
+is also possible [@Miller_1986]. Despite a large number of coastal and
 island caribou herds, only a handful of examples have documented caribou
@@
 Eastern Indian Islands (Figure 1), a swim of at least 470 m which took ~9
-minutes (~52 m per minute).
-This observation prompted us to investigate whether swimming behaviour was common among GPS radio-collared caribou (n = 29 adult females) in the population.
-We identified swimming events as two consecutive GPS locations from an
-individual on different islands. This region typically experiences pack ice
-during the winter, so caribou may be able to travel between islands by
-walking over the ice. We restricted our GPS data to the ice-free period of the
-year (April 1 - November 25) to avoid this possibility. Using these criteria, we
-identified three adult female caribou that collectively engaged in 86 swimming
-events (Figure 2).
+minutes (~52 m per minute). This observation prompted us to investigate whether
+swimming behaviour was common among GPS radio-collared caribou in the population
+(n = 29 adult females: 1240g, GPS 4400M; Lotek Wireless Inc., Newmarket,
+Ontario, Canada). We identified swimming events when two consecutive GPS (2-hour
+relocation rates) locations from an individual occurred on different islands.
+This region typically experiences pack ice during the winter, so caribou may be
+able to travel between islands by walking over the ice. We restricted our GPS
+data to the ice-free period of the year (April 1 - November 25) to avoid this
+possibility. Using these criteria, we identified three adult female caribou that
+collectively engaged in 86 swimming events (Figure 2).
+
 <!--Is it better to just state our results here, or give this sort of teaser?

### 2020-02-21 paper/citations.R

@@ -6,8 +6,14 @@ library(knitcitations)
 ## Bergerud 1985
 citep('10.1139/z85-199')
 
+## Bergerud 1990
+citep('10.1016/S0003-3472(05)80882-6')
+
 ## Leblond et al 2016
 citep('10.1186/s40462-016-0079-4')
+
+## Miller and Gunn 1986 
+citep('10.14430/arctic2052')
 
 ### Write out bib ----
 write.bibtex(file = "references/knitcite.bib")

### 2020-02-21 scripts/1-extractRaster.R
add mapview

@@ -9,6 +9,7 @@ libs <- c('data.table',
           'sf',
           'rosm',
           'stars',
+          'mapview',
           'fasterize')
 lapply(libs, require, character.only = TRUE)
 

### 2020-02-21 scripts/2-extractPoints.R
Update 2-extractPoints.R

@@ -21,7 +21,7 @@ caribou <- caribou[EASTING > 690000 & EASTING < 800000 &
                      NORTHING > 5470000  & NORTHING < 5520000]
 
 ## Sub by date 
-caribou <- caribou[JDate > 90 & JDate < 330]
+caribou <- caribou[JDate > 90 & JDate < 365]
 
 ## Sub by animals that swam
 swimmers <- caribou[ANIMAL_ID == "FO2016011" |
@@ -83,5 +83,7 @@ swimmers$MoveIsland[swimmers$difference == 0] <- "Stay"
 swimmers[, counter := rowid(rleid(StartIsland))]
 
 duration <- swimmers[difference != 0]
+duration[!is.na(MoveIsland)]
+
 
 fwrite(duration, "output/duration.csv")

### 2020-02-21 paper/manuscript.Rmd
Update manuscript.Rmd

@@
 able to travel between islands by walking over the ice. We restricted our GPS
-data to the ice-free period of the year (April 1 - November 25) to avoid this
-possibility. Using these criteria, we identified three adult female caribou that
-collectively engaged in 86 swimming events (Figure 2).
+data to the ice-free period of the year (April 1 - November 25) and we
+identified three adult female caribou that collectively engaged in 86 swimming
+events (Figure 2).
 
+In total, 
 <!--Is it better to just state our results here, or give this sort of teaser?
@@
 
-Caribou were introduced to Fogo Island (n = 26 animals between 1964–67) and
+<!-- Caribou were introduced to Fogo Island (n = 26 animals between 1964–67) and
 Change Island (n = 5 animals in 1964) from the main Island of Newfoundland as
@@
 used GPS locations from these individuals to investigate oceanic swimming
-events.
+events. -->
 

### 2020-02-21 scripts/1-extractRaster.R
comment throughout, update output names

@@ -14,36 +14,43 @@ libs <- c('data.table',
 lapply(libs, require, character.only = TRUE)
 
 ### Extract islands from OSM ----
-# Set min max coords
+# Set up bounding box
 coords <- c(ymin = 49.5194,
             ymax = 49.65,
             xmin = -54.3533,
             xmax = -54.1878)
-latlon <- st_crs(4326)
 bb <- bbox(matrix(c(coords[['xmin']], coords[['xmax']], 
                     coords[['ymin']], coords[['ymax']]), 
                   nrow = 2))
+
+# Projections
+latlon <- st_crs(4326)
 utm <- CRS('+proj=utm +zone=21 ellps=WGS84')
+
+# Download osm in bbox as raster stack
 coordsOSM <- osm.raster(bb,
                         projection = utm,
                         crop = TRUE)
+layer1 <- coordsOSM[[1]]
+islands <- layer1 != 170
+islands[islands == 0] <- NA
 
-coordsOSM2 <- coordsOSM[[1]] != 170
-coordsOSM2[coordsOSM2 == 0] <- NA
-coordsOSM3 <- st_as_sf(
-  st_as_stars(coordsOSM),
+# Convert to an sf object
+sfislands <- st_as_sf(
+  st_as_stars(islands),
   as_points = FALSE,
   use_integer = TRUE,
   merge = TRUE,
   na.rm = TRUE
 )
-coordsOSM3$island <- seq_along(coordsOSM3$geometry)
+sfislands$island <- seq_along(sfislands$geometry)
 
-# zzz
+# Project to UTM
 utmbb <- projectExtent(raster(ext = extent(bb), crs = latlon$proj4string), 
                        crs = utm)
-r2 <- fasterize(coordsOSM3, utmbb, field = "island")
+out <- fasterize(sfislands, utmbb, field = "island")
 
-saveRDS(coordsOSM3, "output/coordsOSM.Rds")
-saveRDS(r2, "output/islandRaster.Rds")
+### Output ----
+saveRDS(sfislands, "output/islandsPoly.Rds")
+saveRDS(out, "output/islandsRaster.Rds")
 

### 2020-02-21 scripts/1-extractRaster.R
rm mapview

@@ -9,7 +9,6 @@ libs <- c('data.table',
           'sf',
           'rosm',
           'stars',
-          'mapview',
           'fasterize')
 lapply(libs, require, character.only = TRUE)
 

### 2020-02-21 scripts/2-extractPoints.R
update input

@@ -10,7 +10,7 @@ lapply(libs, require, character.only = TRUE)
 
 ## load data
 caribou <- fread('input/FogoCaribou.csv')
-r <- readRDS('output/islandRaster.RDS')
+r <- readRDS('output/islandsRaster.RDS')
 
 ## Loc fields
 utm21N <- '+proj=utm +zone=21 ellps=WGS84'

### 2020-02-21 scripts/3-extractArea.R

@@ -9,7 +9,7 @@ lapply(libs, require, character.only = TRUE)
 
 ## load files
 duration <- fread("output/duration.csv")
-r <- readRDS('output/coordsOSM.RDS')
+r <- readRDS('output/islandsPoly.RDS')
 
 ## subset duration data
 duration <- duration[, c("ANIMAL_ID","Year","StartIsland","counter")]

### 2020-02-21 scripts/2-extractPoints.R
.Rds not .RDS

@@ -3,14 +3,22 @@
 
 
 ### Packages ----
-libs <- c('data.table', 'ggplot2', 'dplyr',
-          'raster', 'prettymapr', 'rosm', 
-          'stars', 'fasterize', 'rgdal')
+libs <- c(
+  'data.table',
+  'ggplot2',
+  # 'dplyr',
+  'raster'
+  # 'prettymapr',
+  # 'rosm',
+  # 'stars',
+  # 'fasterize',
+  # 'rgdal'
+)
 lapply(libs, require, character.only = TRUE)
 
 ## load data
 caribou <- fread('input/FogoCaribou.csv')
-r <- readRDS('output/islandsRaster.RDS')
+r <- readRDS('output/islandsRaster.Rds')
 
 ## Loc fields
 utm21N <- '+proj=utm +zone=21 ellps=WGS84'

### 2020-02-21 scripts/2-extractPoints.R
bug risk, why explicitly selecting ids?

@@ -7,12 +7,12 @@ libs <- c(
   'data.table',
   'ggplot2',
   # 'dplyr',
-  'raster'
+  'raster',
   # 'prettymapr',
   # 'rosm',
   # 'stars',
   # 'fasterize',
-  # 'rgdal'
+  'rgdal'
 )
 lapply(libs, require, character.only = TRUE)
 
@@ -32,6 +32,7 @@ caribou <- caribou[EASTING > 690000 & EASTING < 800000 &
 caribou <- caribou[JDate > 90 & JDate < 365]
 
 ## Sub by animals that swam
+# TODO: why explicitly selecting?
 swimmers <- caribou[ANIMAL_ID == "FO2016011" |
                       ANIMAL_ID == "FO2017001" |
                       ANIMAL_ID == "FO2017013"]

### 2020-02-21 scripts/2-extractPoints.R
dont drop to three individuals

@@ -33,12 +33,12 @@ caribou <- caribou[JDate > 90 & JDate < 365]
 
 ## Sub by animals that swam
 # TODO: why explicitly selecting?
-swimmers <- caribou[ANIMAL_ID == "FO2016011" |
-                      ANIMAL_ID == "FO2017001" |
-                      ANIMAL_ID == "FO2017013"]
+# swimmers <- caribou[ANIMAL_ID == "FO2016011" |
+#                       ANIMAL_ID == "FO2017001" |
+#                       ANIMAL_ID == "FO2017013"]
 
 ## Extract points on different islands
-swimmers[, islands := extract(r, matrix(c(EASTING, NORTHING), ncol = 2))]
+caribou[, islands := extract(r, matrix(c(EASTING, NORTHING), ncol = 2))]
 
 ## check points in reference to raster map
 swimmers[, .N, by = "islands"]

### 2020-02-21 scripts/2-extractPoints.R
island numbers dont match

@@ -54,15 +54,16 @@ plot(NORTHING ~ EASTING, data = swimmers[islands == 55],
 swimmers <- swimmers[!is.na(islands),]
 
 ## rename islands
-swimmers$StartIsland[swimmers$islands == 43] <- "Fogo"
-swimmers$StartIsland[swimmers$islands == 53] <- "North Long"
-swimmers$StartIsland[swimmers$islands == 55] <- "North Long"
-swimmers$StartIsland[swimmers$islands == 58] <- "Blundon"
-swimmers$StartIsland[swimmers$islands == 67] <- "Brother"
-swimmers$StartIsland[swimmers$islands == 68] <- "W. Indian"
-swimmers$StartIsland[swimmers$islands == 70] <- "South Long"
-swimmers$StartIsland[swimmers$islands == 71] <- "E. Indian"
-swimmers$StartIsland[swimmers$islands == 74] <- "Kate"
+# TODO: update... these values dont match island numbers 
+# swimmers$StartIsland[swimmers$islands == 43] <- "Fogo"
+# swimmers$StartIsland[swimmers$islands == 53] <- "North Long"
+# swimmers$StartIsland[swimmers$islands == 55] <- "North Long"
+# swimmers$StartIsland[swimmers$islands == 58] <- "Blundon"
+# swimmers$StartIsland[swimmers$islands == 67] <- "Brother"
+# swimmers$StartIsland[swimmers$islands == 68] <- "W. Indian"
+# swimmers$StartIsland[swimmers$islands == 70] <- "South Long"
+# swimmers$StartIsland[swimmers$islands == 71] <- "E. Indian"
+# swimmers$StartIsland[swimmers$islands == 74] <- "Kate"
 
 ## determine when swimming occurred 
 swimmers[, difference := data.table::shift(islands, type = "lead") - islands, by = .(ANIMAL_ID, Year)]

### 2020-02-21 scripts/2-extractPoints.R
more than three individuals
"FO2016011" "FO2017001" "FO2017013" "FO2016001"

@@ -12,6 +12,7 @@ libs <- c(
   # 'rosm',
   # 'stars',
   # 'fasterize',
+  'mapview',
   'rgdal'
 )
 lapply(libs, require, character.only = TRUE)
@@ -41,7 +42,17 @@ caribou <- caribou[JDate > 90 & JDate < 365]
 caribou[, islands := extract(r, matrix(c(EASTING, NORTHING), ncol = 2))]
 
 ## check points in reference to raster map
-swimmers[, .N, by = "islands"]
+caribou[, .N, islands]
+
+
+mapview(
+  caribou[islands != 32280],
+  xcol = 'EASTING',
+  ycol = 'NORTHING',
+  zcol = 'ANIMAL_ID',
+  crs = utm21N
+)
+
 plot(r, xlim = c(685000, 710000), 
      ylim = c(5490000, 5502000))
 par(new = T)

### 2020-02-21 scripts/2-extractPoints.R
any reason to sub to a bbox / julian day limit?

@@ -26,11 +26,11 @@ utm21N <- '+proj=utm +zone=21 ellps=WGS84'
 caribou[, c('EASTING', 'NORTHING') := as.data.table(project(cbind(X_COORD, Y_COORD), utm21N))]
 
 ## Sub by bounding box
-caribou <- caribou[EASTING > 690000 & EASTING < 800000 &
-                     NORTHING > 5470000  & NORTHING < 5520000]
+# caribou <- caribou[EASTING > 690000 & EASTING < 800000 &
+#                      NORTHING > 5470000  & NORTHING < 5520000]
 
 ## Sub by date 
-caribou <- caribou[JDate > 90 & JDate < 365]
+# caribou <- caribou[JDate > 90 & JDate < 365]
 
 ## Sub by animals that swam
 # TODO: why explicitly selecting?

### 2020-02-21 scripts/2-extractPoints.R
rm base plot, add maps

@@ -45,22 +45,6 @@ caribou[, islands := extract(r, matrix(c(EASTING, NORTHING), ncol = 2))]
 caribou[, .N, islands]
 
 
-mapview(
-  caribou[islands != 32280],
-  xcol = 'EASTING',
-  ycol = 'NORTHING',
-  zcol = 'ANIMAL_ID',
-  crs = utm21N
-)
-
-plot(r, xlim = c(685000, 710000), 
-     ylim = c(5490000, 5502000))
-par(new = T)
-plot(NORTHING ~ EASTING, data = swimmers[islands == 55], 
-     xlim = c(685000, 710000), 
-     ylim = c(5490000, 5502000))
-
-
 ## cut points that aren't on an island
 swimmers <- swimmers[!is.na(islands),]
 
@@ -108,3 +92,13 @@ duration[!is.na(MoveIsland)]
 
 
 fwrite(duration, "output/duration.csv")
+
+
+### Maps ----
+mapview(
+  caribou[islands != 32280],
+  xcol = 'EASTING',
+  ycol = 'NORTHING',
+  zcol = 'ANIMAL_ID',
+  crs = utm21N
+)

### 2020-02-21 scripts/2-extractPoints.R
tidy comments, sections

@@ -17,36 +17,39 @@ libs <- c(
 )
 lapply(libs, require, character.only = TRUE)
 
-## load data
+
+### Load data ----
 caribou <- fread('input/FogoCaribou.csv')
 r <- readRDS('output/islandsRaster.Rds')
 
-## Loc fields
+### Prep data ----
+# Project coordinates
 utm21N <- '+proj=utm +zone=21 ellps=WGS84'
 caribou[, c('EASTING', 'NORTHING') := as.data.table(project(cbind(X_COORD, Y_COORD), utm21N))]
 
-## Sub by bounding box
+# Sub by bounding box
 # caribou <- caribou[EASTING > 690000 & EASTING < 800000 &
 #                      NORTHING > 5470000  & NORTHING < 5520000]
 
-## Sub by date 
+# Sub by date 
 # caribou <- caribou[JDate > 90 & JDate < 365]
 
-## Sub by animals that swam
+# Sub by animals that swam
 # TODO: why explicitly selecting?
 # swimmers <- caribou[ANIMAL_ID == "FO2016011" |
 #                       ANIMAL_ID == "FO2017001" |
 #                       ANIMAL_ID == "FO2017013"]
 
-## Extract points on different islands
+
+### Extract islands ----
+# Extract points on different islands
 caribou[, islands := extract(r, matrix(c(EASTING, NORTHING), ncol = 2))]
 
-## check points in reference to raster map
+# Count locs by islands
 caribou[, .N, islands]
 
-
-## cut points that aren't on an island
-swimmers <- swimmers[!is.na(islands),]
+# Cut points that aren't on an island
+swimmers <- caribou[!is.na(islands)]
 
 ## rename islands
 # TODO: update... these values dont match island numbers 
@@ -61,7 +64,8 @@ swimmers <- swimmers[!is.na(islands),]
 # swimmers$StartIsland[swimmers$islands == 74] <- "Kate"
 
 ## determine when swimming occurred 
-swimmers[, difference := data.table::shift(islands, type = "lead") - islands, by = .(ANIMAL_ID, Year)]
+swimmers[, difference := paste(islands, shift(islands, type = "lead"), sep = '-'), 
+         by = .(ANIMAL_ID, Year)]
 
 ## rename the shifted difference column to the inhabited island
 swimmers$MoveIsland[swimmers$difference == -10] <- "Blundon"

### 2020-02-21 scripts/2-extractPoints.R
rm shifted difference for shifted paste

@@ -51,7 +51,7 @@ caribou[, .N, islands]
 # Cut points that aren't on an island
 swimmers <- caribou[!is.na(islands)]
 
-## rename islands
+# rename islands
 # TODO: update... these values dont match island numbers 
 # swimmers$StartIsland[swimmers$islands == 43] <- "Fogo"
 # swimmers$StartIsland[swimmers$islands == 53] <- "North Long"
@@ -63,30 +63,10 @@ swimmers <- caribou[!is.na(islands)]
 # swimmers$StartIsland[swimmers$islands == 71] <- "E. Indian"
 # swimmers$StartIsland[swimmers$islands == 74] <- "Kate"
 
-## determine when swimming occurred 
+
+# Determine between which islands swimming occured
 swimmers[, difference := paste(islands, shift(islands, type = "lead"), sep = '-'), 
          by = .(ANIMAL_ID, Year)]
-
-## rename the shifted difference column to the inhabited island
-swimmers$MoveIsland[swimmers$difference == -10] <- "Blundon"
-swimmers$MoveIsland[swimmers$difference == 3] <- "E. Indian"
-swimmers$MoveIsland[swimmers$difference == -3] <- "W. Indian"
-swimmers$MoveIsland[swimmers$difference == -1] <- "Brother"
-swimmers$MoveIsland[swimmers$difference == 1] <- "W. Indian"
-swimmers$MoveIsland[swimmers$StartIsland == "Blundon" & 
-                    swimmers$difference == 13] <- "E. Indian"
-swimmers$MoveIsland[swimmers$StartIsland == "North Long" & 
-                    swimmers$difference == 13] <- "W. Indian"
-swimmers$MoveIsland[swimmers$difference == -13] <- "North Long"
-swimmers$MoveIsland[swimmers$difference == -25] <- "Fogo"
-swimmers$MoveIsland[swimmers$difference == 25] <- "W. Indian"
-swimmers$MoveIsland[swimmers$difference == -15] <- "North Long"
-swimmers$MoveIsland[swimmers$difference == 15] <- "W. Indian"
-swimmers$MoveIsland[swimmers$difference == -6] <- "W. Indian"
-swimmers$MoveIsland[swimmers$difference == 6] <- "Kate"
-swimmers$MoveIsland[swimmers$difference == 2] <- "South Long"
-swimmers$MoveIsland[swimmers$difference == -2] <- "W. Indian"
[…]

### 2020-02-21 scripts/2-extractPoints.R
island instead of islands

@@ -43,29 +43,31 @@ caribou[, c('EASTING', 'NORTHING') := as.data.table(project(cbind(X_COORD, Y_COO
 
 ### Extract islands ----
 # Extract points on different islands
-caribou[, islands := extract(r, matrix(c(EASTING, NORTHING), ncol = 2))]
+caribou[, island := extract(r, matrix(c(EASTING, NORTHING), ncol = 2))]
 
-# Count locs by islands
-caribou[, .N, islands]
+# Count locs by island
+caribou[, .N, island]
 
 # Cut points that aren't on an island
-swimmers <- caribou[!is.na(islands)]
+swimmers <- caribou[!is.na(island)]
 
-# rename islands
+# rename island
 # TODO: update... these values dont match island numbers 
-# swimmers$StartIsland[swimmers$islands == 43] <- "Fogo"
-# swimmers$StartIsland[swimmers$islands == 53] <- "North Long"
-# swimmers$StartIsland[swimmers$islands == 55] <- "North Long"
-# swimmers$StartIsland[swimmers$islands == 58] <- "Blundon"
-# swimmers$StartIsland[swimmers$islands == 67] <- "Brother"
-# swimmers$StartIsland[swimmers$islands == 68] <- "W. Indian"
-# swimmers$StartIsland[swimmers$islands == 70] <- "South Long"
-# swimmers$StartIsland[swimmers$islands == 71] <- "E. Indian"
-# swimmers$StartIsland[swimmers$islands == 74] <- "Kate"
+# swimmers$StartIsland[swimmers$island == 43] <- "Fogo"
+# swimmers$StartIsland[swimmers$island == 53] <- "North Long"
+# swimmers$StartIsland[swimmers$island == 55] <- "North Long"
+# swimmers$StartIsland[swimmers$island == 58] <- "Blundon"
+# swimmers$StartIsland[swimmers$island == 67] <- "Brother"
+# swimmers$StartIsland[swimmers$island == 68] <- "W. Indian"
+# swimmers$StartIsland[swimmers$island == 70] <- "South Long"
+# swimmers$StartIsland[swimmers$island == 71] <- "E. Indian"
+# swimmers$StartIsland[swimmers$island == 74] <- "Kate"
 
 
-# Determine between which islands swimming occured
-swimmers[, difference := paste(islands, shift(islands, type = "lead"), sep = '-'), 
+# Determine between which island swimming occured
+swimmers[, endisland]
[…]

### 2020-02-21 scripts/2-extractPoints.R
tidy endisland, paste shift etc

@@ -65,22 +65,30 @@ swimmers <- caribou[!is.na(island)]
 
 
 # Determine between which island swimming occured
-swimmers[, endisland]
-swimmers[, difference := paste(island, data.table::shift(island, type = "lead"), 
-                               sep = '-'), 
+swimmers[, endisland := data.table::shift(island, type = "lead")]
+swimmers[island != endisland, 
+         diff := paste(island, endisland, sep = '-'), 
          by = .(ANIMAL_ID, Year)]
 
-## count number of fixes on each island
-swimmers[, counter := rowid(rleid(StartIsland))]
+# Count number of fixes on each island
+swimmers[, counter := .N, island]
 
-duration <- swimmers[difference != 0]
-duration[!is.na(MoveIsland)]
+duration <- swimmers[island != endisland]
 
 
+### Output ----
 fwrite(duration, "output/duration.csv")
 
 
 ### Maps ----
+mapview(
+  duration],
+  xcol = 'EASTING',
+  ycol = 'NORTHING',
+  zcol = 'ANIMAL_ID',
+  crs = utm21N
+)
+
 mapview(
   caribou[island != 32280],
   xcol = 'EASTING',

### 2020-02-21 scripts/2-extractPoints.R
bbox makes sense...

@@ -28,8 +28,8 @@ utm21N <- '+proj=utm +zone=21 ellps=WGS84'
 caribou[, c('EASTING', 'NORTHING') := as.data.table(project(cbind(X_COORD, Y_COORD), utm21N))]
 
 # Sub by bounding box
-# caribou <- caribou[EASTING > 690000 & EASTING < 800000 &
-#                      NORTHING > 5470000  & NORTHING < 5520000]
+caribou <- caribou[EASTING > 690000 & EASTING < 800000 &
+                     NORTHING > 5470000  & NORTHING < 5520000]
 
 # Sub by date 
 # caribou <- caribou[JDate > 90 & JDate < 365]
@@ -82,7 +82,7 @@ fwrite(duration, "output/duration.csv")
 
 ### Maps ----
 mapview(
-  duration],
+  duration,
   xcol = 'EASTING',
   ycol = 'NORTHING',
   zcol = 'ANIMAL_ID',

### 2020-02-21 scripts/2-extractPoints.R
use between instead

@@ -28,8 +28,8 @@ utm21N <- '+proj=utm +zone=21 ellps=WGS84'
 caribou[, c('EASTING', 'NORTHING') := as.data.table(project(cbind(X_COORD, Y_COORD), utm21N))]
 
 # Sub by bounding box
-caribou <- caribou[EASTING > 690000 & EASTING < 800000 &
-                     NORTHING > 5470000  & NORTHING < 5520000]
+caribou <- caribou[EASTING %between% c(690000, 800000) &
+                     NORTHING %between% c(5470000, 5520000)]
 
 # Sub by date 
 # caribou <- caribou[JDate > 90 & JDate < 365]
@@ -72,6 +72,10 @@ swimmers[island != endisland,
 
 # Count number of fixes on each island
 swimmers[, counter := .N, island]
+# TODO: get duration on islands rowid(rleid(StartIsland))]
+
+
+
 
 duration <- swimmers[island != endisland]
 

### 2020-02-21 scripts/2-extractPoints.R
set fogonum

@@ -49,6 +49,7 @@ caribou[, island := extract(r, matrix(c(EASTING, NORTHING), ncol = 2))]
 caribou[, .N, island]
 
 # Cut points that aren't on an island
+# fogonum <- 32280
 swimmers <- caribou[!is.na(island)]
 
 # rename island

### 2020-02-21 scripts/2-extractPoints.R
trouble getting individual events

@@ -75,8 +75,18 @@ swimmers[island != endisland,
 swimmers[, counter := .N, island]
 # TODO: get duration on islands rowid(rleid(StartIsland))]
 
+# Routes
+# TODO: careful island != isnt because endisland is na
+swimmers[, i := .I]
+swimmers[island != endisland & !is.na(endisland), swim := 'start']
+swimmers[i %in% swimmers[!is.na(swim), i+1], swim := 'end']
 
+swimmers[swim == 'start', event := .I]
+swimmers[order(i)][swim == 'end', event := data.table::shift(event, 1, 'lag')]
 
+ggplot(swimmers[!is.na(swim)]) +
+  geom_line(aes(EASTING, NORTHING, color = ANIMAL_ID, )) + 
+  
 
 duration <- swimmers[island != endisland]
 
@@ -86,6 +96,13 @@ fwrite(duration, "output/duration.csv")
 
 
 ### Maps ----
+mapview(
+  swimmers[!is.na(swim)],
+  xcol = 'EASTING',
+  ycol = 'NORTHING',
+  zcol = 'ANIMAL_ID',
+  crs = utm21N
+)
 mapview(
   duration,
   xcol = 'EASTING',

### 2020-02-21 paper/manuscript.Rmd
refs refszzzzz

@@
 
-The Fogo Island archipelago, off the coast of Newfoundland, Canada (49°N, 54°W),
+The Fogo Island archipelago, off the coast of Newfoundland, Canada,
 is home to approximately 300 woodland caribou. During routine fieldwork on May
-30, 2017, we observed an unmarked adult male caribou swim between Perry and
+30, 2017, we observed an unmarked adult male caribou swim between Wester and
 Eastern Indian Islands (Figure 1), a swim of at least 470 m which took ~9
@@
 Caribou may swim between islands in the ocean for similar reasons that drive
-caribou swimming in freshwater (Leblond et al. 2016). We present two
+caribou swimming in freshwater [@Leblond_2016]. We present two
 explanations for these swimming events, forage limitation (Miller 2002) and
@@
 Forage limitation and over-grazing is a major concern for caribou populations
-(Schaefer et al. 2016; Zamin et al. 2017) and could explain movement between
+[@Schaefer_2016] and could explain movement between
 islands. Conspecific competition for resources driven by reduced foraging
-opportunities can decrease reproductive success of adult females (Schaefer et
-al. 2016). Forage scarcity has been proposed as a potential reason for caribou
-moving between arctic islands on the sea-ice in winter (Miller et al. 1977).
+opportunities can decrease reproductive success of adult females [@Schaefer_2016]. Forage scarcity has been proposed as a potential reason for caribou
+moving between arctic islands on the sea-ice in winter [@Miller_1977].
 For very small islands, over-grazing by newly arrived caribou may rapidly
-deplete forage (Bergerud et al. 1990). Without sea ice, movement from the
+deplete forage [@Bergerud_1990]. Without sea ice, movement from the
 mainland to an island, or between islands, requires caribou to assess the
@@
 This trade-off can be understood as density-dependent habitat selection governed
-by the Ideal Free Distribution (Morris 1987; Bradbury et al. 2015). Ideal Free
+by the Ideal Free Distribution [@Morris_1987; @Bradbury_2015]. Ideal Free
 Distribution theory predicts a fitness equilibrium: when the density in a given

### 2020-02-21 paper/citations.R

@@ -9,11 +9,23 @@ citep('10.1139/z85-199')
 ## Bergerud 1990
 citep('10.1016/S0003-3472(05)80882-6')
 
+## Bradbury et al 2015
+citep('10.1093/beheco/arv078')
+
 ## Leblond et al 2016
 citep('10.1186/s40462-016-0079-4')
 
+## Miller et al 1977
+citep('10.1139/z77-131')
+
 ## Miller and Gunn 1986 
 citep('10.14430/arctic2052')
+
+## Morris 1987
+citep('10.2307/2937087')
+
+## Schaefer 2016
+citep('10.1093/jmammal/gyv184')
 
 ### Write out bib ----
 write.bibtex(file = "references/knitcite.bib")

### 2020-02-21 paper/manuscript.Rmd
forage limitation

@@
 Forage limitation and over-grazing is a major concern for caribou populations
-[@Schaefer_2016] and could explain movement between
-islands. Conspecific competition for resources driven by reduced foraging
-opportunities can decrease reproductive success of adult females [@Schaefer_2016]. Forage scarcity has been proposed as a potential reason for caribou
-moving between arctic islands on the sea-ice in winter [@Miller_1977].
-For very small islands, over-grazing by newly arrived caribou may rapidly
-deplete forage [@Bergerud_1990]. Without sea ice, movement from the
-mainland to an island, or between islands, requires caribou to assess the
-trade-off associated with swimming.
+[@Schaefer_2015] and could explain movement between islands. Conspecific
+competition for resources driven by reduced foraging opportunities can decrease
+reproductive success of adult females [@Schaefer_2015]. Forage scarcity has been
+proposed as a potential reason for caribou moving between arctic islands on the
+sea-ice in winter [@Miller_1977]. For very small islands, over-grazing by newly
+arrived caribou may rapidly deplete forage [@Bergerud_1990]. Without sea ice,
+movement from the mainland to an island, or between islands, requires caribou to
+assess the trade-off associated with swimming.
 
@@
 should relocate and settle new habitat patches, so that fitness is equal across
-all patches (Bradbury et al. 2015). Islands act as discrete habitat patches in
+all patches [@Bradbury_2015]. Islands act as discrete habitat patches in
 this case, but the costs associated with swimming create a trade-off when moving
 between patches. For caribou in the Fogo Island archipelago, the role of forage
-limitation in animal habitat selection patterns and movement remains unknown,
+limitation in animal habitat selection patterns remains unknown,
 but swimming between islands may in part be explained by forage limitation and

### 2020-02-21 paper/manuscript.Rmd
predators and refs

@@
 in freshwater lakes give birth, or with their calves after birth, to avoid
-predation (Bergerud & Page 1987). The use of shoreline habitat in summer is
-considered an effective anti-predator calving strategy by allowing for swimming
-to islands or peninsulas as a predation avoidance flight response (Bergerud et
-al. 1990). The use of water, and by association occasionally swimming, is
-therefore part of the fine-scale interactions between caribou and their
-predators and the use of islands appears to be an effective anti-predator
-strategy.
+predation [@Bergerud_1987]. The use of shoreline habitat in summer is considered
+an effective anti-predator calving strategy by allowing for swimming to islands
+or peninsulas as a predation avoidance flight response [@Bergerud_1990]. The use
+of water, and by association occasionally swimming, is therefore part of the
+fine-scale interactions between caribou and their predators and the use of
+islands appears to be an effective anti-predator strategy.
 

### 2020-02-21 paper/citations.R

@@ -6,6 +6,9 @@ library(knitcitations)
 ## Bergerud 1985
 citep('10.1139/z85-199')
 
+## Bergerud and Page 1987
+citep('10.1139/z87-249')
+
 ## Bergerud 1990
 citep('10.1016/S0003-3472(05)80882-6')
 
@@ -20,6 +23,9 @@ citep('10.1139/z77-131')
 
 ## Miller and Gunn 1986 
 citep('10.14430/arctic2052')
+
+## Miller 2002
+#citep('10.2307/40512446')
 
 ## Morris 1987
 citep('10.2307/2937087')

### 2020-02-21 paper/manuscript.Rmd
more pred stuff

@@
 Predator avoidance could also contribute to swimming behaviour of caribou
-(Miller 2002). During calving, female caribou are known to swim to small islands
-in freshwater lakes give birth, or with their calves after birth, to avoid
-predation [@Bergerud_1987]. The use of shoreline habitat in summer is considered
-an effective anti-predator calving strategy by allowing for swimming to islands
-or peninsulas as a predation avoidance flight response [@Bergerud_1990]. The use
-of water, and by association occasionally swimming, is therefore part of the
-fine-scale interactions between caribou and their predators and the use of
-islands appears to be an effective anti-predator strategy.
+(Miller 2002) <!-- find doi for this article -->. During calving, female caribou
+are known to swim to small islands in freshwater lakes give birth, or with their
+calves after birth, to avoid predation [@Bergerud_1987]. The use of shoreline
+habitat in summer is considered an effective anti-predator calving strategy by
+allowing for swimming to islands or peninsulas as a predation avoidance flight
+response [@Bergerud_1990]. The use of water, and by association occasionally
+swimming, is therefore part of the fine-scale interactions between caribou and
+their predators and the use of islands appears to be an effective anti-predator
+strategy.
 
@@
 (*Canis latrans*) are the top predator in this region. Predation by coyotes on
-adult caribou is relatively rare (ref?), so the potential predator effect on
+adult caribou is relatively rare (ref?), so the potential effect of predators on
 caribou swimming behaviour is likely to be a flight response to non-lethal
 encounters. Calves are susceptible to coyotes, particularly young calves, but
-the energetic costs to oceanic swimming are likely to be far greater for small
-calves.
+the energetic costs to oceanic swimming for calves is likely to be far greater for small
+calves compared to adults.
 
@@
 ## Individual variation 
-Three of 29 GPS-collared female caribou swam between islands in the Fogo
+In total, 3 of 29 collared female caribou swam between islands in the Fogo
 archipelago. These swimming events were relatively frequent, on average occuring
-once every 24 days within our date range. However, we identified zero swimming
-events for the other 26 collared females. It remains unclear why this behaviour
-is seemingly common among 10% of our marked population, while 90% never engaged
-in it. Our initial direct observation of swimming was an adult male, but as we
-only have GPS data for females, we are unable to make conclusions on the
-prevalence of this behaviour among male caribou. It is theoretically possible
-that these other females moved rapidly back and forth between islands within the
-2 hour fix rate of our GPS collars, which we would be unable to detect. We
-consider this highly unlikely given the range of residency times for caribou on
-any given island (min - max: XX - XX).
+once every 24 <!--double check--> days during ice-free seasons. However, we
+identified zero swimming events for the other 26 collared females. It remains
+unclear why this behaviour is seemingly common among 10% of our marked
+population, while 90% never engaged in it. Our initial direct observation of
+swimming was an adult male, but as we only have GPS data for females, we are
+unable to make conclusions on the prevalence of this behaviour among male
+caribou. It is theoretically possible that these other females moved rapidly
+back and forth between islands within the 2 hour fix rate of our GPS collars,
+which we would be unable to detect. We consider this highly unlikely given the
+range of residency times for caribou on any given island (min - max: XX - XX).
 <!-- What is the typical residency time? Is this a true statement or am I just making this up? -->

### 2020-02-21 paper/manuscript.Rmd
edits

@@
 archipelago. These swimming events were relatively frequent, on average occuring
-once every 24 <!--double check--> days during ice-free seasons. However, we
-identified zero swimming events for the other 26 collared females. It remains
-unclear why this behaviour is seemingly common among 10% of our marked
-population, while 90% never engaged in it. Our initial direct observation of
+once every 24 <!--double check--> days during ice-free seasons. However, despite relatively frequent swimming events for these three individuals, we were unable to
+identify any swimming events for the other 26 collared females. It remains
+unclear why ocean swimming is relatively common among ~10% of our marked
+population, while the other 90% never swam. A potential explanation is the Our initial direct observation of
 swimming was an adult male, but as we only have GPS data for females, we are

### 2020-02-21 paper/citations.R
Update citations.R

@@ -9,6 +9,9 @@ citep('10.1139/z85-199')
 ## Bergerud and Page 1987
 citep('10.1139/z87-249')
 
+## Bergerud and Mercer 1989 - get actual doi
+#citep('10.2307/3782635')
+
 ## Bergerud 1990
 citep('10.1016/S0003-3472(05)80882-6')
 
@@ -24,7 +27,7 @@ citep('10.1139/z77-131')
 ## Miller and Gunn 1986 
 citep('10.14430/arctic2052')
 
-## Miller 2002
+## Miller 2002 -- get actual doi
 #citep('10.2307/40512446')
 
 ## Morris 1987

### 2020-02-21 paper/manuscript.Rmd
Update manuscript.Rmd

@@
 archipelago. These swimming events were relatively frequent, on average occuring
-once every 24 <!--double check--> days during ice-free seasons. However, despite relatively frequent swimming events for these three individuals, we were unable to
-identify any swimming events for the other 26 collared females. It remains
+once every 24 <!--double check--> days during ice-free seasons. However, despite
+relatively frequent swimming events for these three individuals, we were unable
+to identify any swimming events for the other 26 collared females. It remains
 unclear why ocean swimming is relatively common among ~10% of our marked
-population, while the other 90% never swam. A potential explanation is the Our initial direct observation of
-swimming was an adult male, but as we only have GPS data for females, we are
-unable to make conclusions on the prevalence of this behaviour among male
-caribou. It is theoretically possible that these other females moved rapidly
-back and forth between islands within the 2 hour fix rate of our GPS collars,
-which we would be unable to detect. We consider this highly unlikely given the
-range of residency times for caribou on any given island (min - max: XX - XX).
-<!-- What is the typical residency time? Is this a true statement or am I just making this up? -->
+population, while the other 90% never swam. Of the three collared individuals,
+two were collared on Western Indian Island and the third was collared on Fogo
+Island. It is possible that individuals born on, or that live on, Eastern and
+Western Indian Island swim are more likely to swim between islands because
+inter-island distances are shorter than between Western Indian and Fogo Islands
+(Figure 2). Furthermore, it is unclear whether swimming is more common among
+males or females. Our initial observation  was of an adult male, but our GPS
+data only includes females. It is theoretically possible that these other
+females moved rapidly back and forth between islands within the 2 hour fix rate
+of our GPS collars, which we would be unable to detect. We consider this highly
+unlikely given the range of residency times for caribou on any given island (min
+- max: XX - XX). <!--double check-->
+<!-- What is the typical residency time? Is this a true statement or am I just making this up? QW: this is something we will look into, I think. -->
 
-The two hypotheses presented above are generally thought to apply uniformly
-across populations, but individual caribou could evaluate the costs and benefits
-of swimming to another island differently. Potentially those three females,
-compared to the other 26, considered swimming less costly or perceived
-competition or predation to be of greater concern. The individual attributes
-that might influence how animals vary in their assessments of these trade-offs
-are unknown, and potential predictors such as age or personality traits are
-unknown for these caribou.
+The forage limitation and predator avoidance hypotheses are generally thought to
+apply uniformly across populations, but individual caribou could evaluate the
+costs and benefits of swimming to another island differently. Potentially those
+three females, compared to the other 26, considered swimming less costly or
+perceived competition or predation to be of greater concern. The individual
+attributes that might influence how animals vary in their assessments of these
+trade-offs are unknown, and potential predictors such as age or personality
+traits are unknown for these caribou.
 
@@
 new islands when the average fitness of individuals on the starting island
-exceeds the density-fitness equilibrium (Morris 1987). While this is an ultimate
+exceeds the density-fitness equilibrium [@Morris_1987]. While this is an ultimate
 explanation to a series of proximate observations, it is possible for

### 2020-02-21 paper/citations.R
refz

@@ -18,6 +18,9 @@ citep('10.1016/S0003-3472(05)80882-6')
 ## Bradbury et al 2015
 citep('10.1093/beheco/arv078')
 
+## Jeffrey et al 2007
+citep('10.7557/2.27.4.320')
+
 ## Leblond et al 2016
 citep('10.1186/s40462-016-0079-4')
 
@@ -35,6 +38,12 @@ citep('10.2307/2937087')
 
 ## Schaefer 2016
 citep('10.1093/jmammal/gyv184')
+
+## Webber and Vander Wal 2018
+citep('10.1111/1365-2656.12773')
+
+## van Beest et al 2014
+citep('10.1111/1365-2656.12115')
 
 ### Write out bib ----
 write.bibtex(file = "references/knitcite.bib")

### 2020-02-21 paper/manuscript.Rmd
Update manuscript.Rmd and manuscript.md

@@
 explanations for these swimming events, forage limitation (Miller 2002) and
-predator avoidance (Jeffery et al. 2007), and discuss why these swimming events
+predator avoidance [@Jeffrey_2007], and discuss why these swimming events
 appear to be relatively commonplace in the population.
@@
 males or females. Our initial observation  was of an adult male, but our GPS
-data only includes females. It is theoretically possible that these other
+data only includes females. 
+<!-- QW: I'm not sure we need to say this.. 
+It is theoretically possible that these other
 females moved rapidly back and forth between islands within the 2 hour fix rate
@@
 apply uniformly across populations, but individual caribou could evaluate the
-costs and benefits of swimming to another island differently. Potentially those
-three females, compared to the other 26, considered swimming less costly or
-perceived competition or predation to be of greater concern. The individual
-attributes that might influence how animals vary in their assessments of these
-trade-offs are unknown, and potential predictors such as age or personality
-traits are unknown for these caribou.
+costs and benefits of swimming to another island differently. It is possible
+that some individuals, in our case ~10%, considered swimming less costly or that
+they perceived competition or predation more acutely than their conspecifics.
+Intrinsic life-history, physiological, or behavioural traits that might
+influence how animals vary in their assessments of swimming trade-offs remains
+unknown. Future work could assess the role of age, stress, or an individuals
+tendency to move potential predictors of swimming behaviour.
 
@@
 new islands when the average fitness of individuals on the starting island
-exceeds the density-fitness equilibrium [@Morris_1987]. While this is an ultimate
-explanation to a series of proximate observations, it is possible for
-density-dependent habitat selection to operate at such a fine-scale (Webber &
-Vander Wal 2018). Evidence from the Sable Island horse (Equus ferus caballus)
-system corroborates this hypothesis (van Beest et al. 2014). For horses,
-selection for high quality habitat was strongest at low population densities,
-but individuals tended to settle in lower quality habitats as population density
-increased, following Ideal Free Distribution theory (van Beest et al. 2014).
-Predator avoidance can also contribute to caribou swimming behaviour, but in
-this situation it seems a less likely explanation given predation is primarily a
-concern for young calves, which are the least likely to engage in this costly
-behaviour.
-<!-- this last bit is sort of me bullshitting, not sure it's actually defensible?-->
+exceeds the density-fitness equilibrium [@Morris_1987]. While this is an
+ultimate explanation to a series of proximate observations, fine-scale
+density-dependent habitat selection is theoretically possible
+[@Webber_2017]. For horses (*Equus ferus caballus*), selection for high quality
+habitat was strongest at low population densities, but individuals tended to
+settle in lower quality habitats as population density increased, following
+Ideal Free Distribution theory [@van_Beest_2013]. Predator avoidance could also
+contribute to caribou swimming behaviour, but given the presumed low density of coyotes on Fogo Island and that predation primarily only affects young calves, we suspect predator is a possible, but less likely explanation.
+<!-- this last bit is sort of me bullshitting, not sure it's actually defensible? 
+QW: changed it a bit -->
 
 Our observations add to the evidence that caribou can, and occasionally do, swim
-in the ocean (Table 1). While past observations (e.g. Miller 1995, 2002; Jeffery
-et al. 2007; Ricca et al. 2012) have noted the exceptional nature of caribou
+in the ocean (Table 1). While past observations (e.g. @Miller_1995;
+@Jeffrey_2007; @Ricca_2012] have noted the exceptional nature of caribou
 swimming in the ocean, we suggest that this behaviour is likely more common than

### 2020-02-21 paper/manuscript.Rmd
refz

@@
 Our observations add to the evidence that caribou can, and occasionally do, swim
-in the ocean (Table 1). While past observations (e.g. @Miller_1995;
-@Jeffrey_2007; @Ricca_2012] have noted the exceptional nature of caribou
-swimming in the ocean, we suggest that this behaviour is likely more common than
-previously thought for caribou living on oceanic islands. We posit forage
-limitation and the associated density-dependent habitat selection is an ultimate
-explanation for why caribou swimming in the ocean, with predator avoidance as a
-potential proximate explanation. Although our inference is limited to
-observations, our observations suggest that islands represent discrete foraging
-patches for terrestrial animals that could influence fitness via increased
-foraging opportunities and safety from predators. We also suggest that
-individuals vary in their evaluation of costs and benefits of movement between
-these habitat patches.
+in the ocean (Table 1). While past observations (e.g.
+@Jeffery_2007; @Ricca_2012] <!-- add Miller 1995, Miller 2002 --> have noted the
+exceptional nature of caribou swimming in the ocean, we suggest that this
+behaviour is likely more common than previously thought for caribou living on
+oceanic islands. We posit forage limitation and the associated density-dependent
+habitat selection is an ultimate explanation for why caribou swimming in the
+ocean, with predator avoidance as a potential proximate explanation. Although
+our inference is limited to observations, our observations suggest that islands
+represent discrete foraging patches for terrestrial animals that could influence
+fitness via increased foraging opportunities and safety from predators. We also
+suggest that individuals vary in their evaluation of costs and benefits of
+movement between these habitat patches.
 

### 2020-02-21 paper/citations.R

@@ -39,6 +39,9 @@ citep('10.2307/2937087')
 ## Schaefer 2016
 citep('10.1093/jmammal/gyv184')
 
+## Ricca et al 2012
+citep('10.1007/s10530-012-0195-z')
+
 ## Webber and Vander Wal 2018
 citep('10.1111/1365-2656.12773')
 

### 2020-02-21 paper/citations.R
Miller ref

@@ -30,6 +30,9 @@ citep('10.1139/z77-131')
 ## Miller and Gunn 1986 
 citep('10.14430/arctic2052')
 
+## Miller 1995 -- get actual doi
+#citep('40511612')
+
 ## Miller 2002 -- get actual doi
 #citep('10.2307/40512446')
 

### 2020-02-21 paper/manuscript.Rmd
blargh

@@
 output: 
-    github_document
+  #  github_document
   # bookdown::pdf_document2:
@@
   #    toc: false
-  #bookdown::word_document2:
+   bookdown::word_document2:
   #  reference_docx: template.docx
@@
 to avoid predators [@Bergerud_1985] and access islands during calving
-[@Bergerud_1990]. Even for adept swimmers like caribou, the
-energetic expenditure of swimming for quadrupedal mammals is
-significantly higher than walking or running (Fish 1993), while drowning
-is also possible [@Miller_1986]. Despite a large number of coastal and
-island caribou herds, only a handful of examples have documented caribou
-swimming in the ocean (Table 1).
+[@Bergerud_1990]. Even for adept swimmers like caribou, the energetic
+expenditure of swimming for quadrupedal mammals is significantly higher than
+walking or running, while drowning is also possible [@Miller_1986].
+Despite a large number of coastal and island caribou herds, only a handful of
+examples have documented caribou swimming in the ocean [e.g. @Jeffery_2007;
+@Ricca_2012]. <!-- add Miller 1995/2002 or reference table 1 if we decide to
+keep this -->
 

### 2020-02-21 paper/citations.R

@@ -10,13 +10,15 @@ citep('10.1139/z85-199')
 citep('10.1139/z87-249')
 
 ## Bergerud and Mercer 1989 - get actual doi
-#citep('10.2307/3782635')
 
 ## Bergerud 1990
 citep('10.1016/S0003-3472(05)80882-6')
 
 ## Bradbury et al 2015
 citep('10.1093/beheco/arv078')
+
+## Fish 1993
+citep('10.1071/ZO9940079')
 
 ## Jeffrey et al 2007
 citep('10.7557/2.27.4.320')
@@ -31,10 +33,8 @@ citep('10.1139/z77-131')
 citep('10.14430/arctic2052')
 
 ## Miller 1995 -- get actual doi
-#citep('40511612')
 
 ## Miller 2002 -- get actual doi
-#citep('10.2307/40512446')
 
 ## Morris 1987
 citep('10.2307/2937087')

### 2020-02-21 paper/manuscript.Rmd
Update manuscript.Rmd

@@
 data to the ice-free period of the year (April 1 - November 25) and we
-identified three adult female caribou that collectively engaged in 86 swimming
+identified three adult female caribou that collectively had 86 swimming
 events (Figure 2).
 
-In total, 
-<!--Is it better to just state our results here, or give this sort of teaser?
-Initially I had "86 events by 3 individuals" here, but then we ignore that for a couple paragraphs of context before getting back into it...
-which seemed weird-->
+In total, these 86 swimming events occured in month (n = Y), month (n = Y), and
+month (n = Y), while the remaining events were in month, month, and month (n =
+Y). On average, caribou swam between islands approximately ever 24 days, but
+some individuals swam up to XX times per week/month/whatever. <!--Is it better
+to just state our results here, or give this sort of teaser? Initially I had "86
+events by 3 individuals" here, but then we ignore that for a couple paragraphs
+of context before getting back into it... which seemed weird-->
 
@@
 
-Fogo Island is the largest island in the archipelago (237.71 km^2^), but there
-are at least three other large islands: Western Indian (77.6 km2), Eastern
-Indian (38.7 km2) and Change (XX km2) Islands, as well as numerous smaller
-islands, including Blundon’s (1.18 km2), North Long (1.01 km2), South Long (0.48
-km2), Kate’s (1.64 km2), and Brother’s (1.59 km2) Islands (see Figure 1).
-Habitats are similar across the archipelago, consisting largely of coniferous
-and mixed forests of balsam fir (*Abies balsamea*), black spruce (*Picea
-mariana*), and white birch (*Betula papyrifera*) as well as bogs, lakes, lichen
-and rocky barrens.
+<!-- still not sure we need this paragraph, I wonder if the island size
+information could be included in the figure caption?--> Fogo Island is the
+largest island in the archipelago (237.71 km^2^), but there are at least three
+other large islands: Western Indian (77.6 km2), Eastern Indian (38.7 km2) and
+Change (XX km2) Islands, as well as numerous smaller islands, including
+Blundon’s (1.18 km2), North Long (1.01 km2), South Long (0.48 km2), Kate’s (1.64
+km2), and Brother’s (1.59 km2) Islands (see Figure 1). Habitats are similar
+across the archipelago, consisting largely of coniferous and mixed forests of
+balsam fir (*Abies balsamea*), black spruce (*Picea mariana*), and white birch
+(*Betula papyrifera*) as well as bogs, lakes, lichen and rocky barrens.
 

### 2020-02-21 paper/manuscript.Rmd
Update manuscript.Rmd and ~$nuscript.docx

@@
    bookdown::word_document2:
-  #  reference_docx: template.docx
+    reference_docx: template.docx
 bibliography: ../references/knitcite.bib

### 2020-02-21 paper/manuscript.Rmd
Update manuscript.Rmd, manuscript.docx, and 2 more files...

@@
    bookdown::word_document2:
-    reference_docx: template.docx
+    #reference_docx: template.docx
 bibliography: ../references/knitcite.bib
@@
 
-## Introduction
+# Introduction
 Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
-that caribou swim in streams, rivers, and lakes during migration [@Leblond_2016]
-to avoid predators [@Bergerud_1985] and access islands during calving
+that caribou swim in streams, rivers, and lakes during migration [@Leblond_2016],
+to avoid predators [@Bergerud_1985], and access islands during calving
 [@Bergerud_1990]. Even for adept swimmers like caribou, the energetic
 expenditure of swimming for quadrupedal mammals is significantly higher than
-walking or running, while drowning is also possible [@Miller_1986].
+walking or running [@Fish_1994], while drowning is also possible [@Miller_1986].
 Despite a large number of coastal and island caribou herds, only a handful of
@@
 
-The Fogo Island archipelago, off the coast of Newfoundland, Canada,
-is home to approximately 300 woodland caribou. During routine fieldwork on May
-30, 2017, we observed an unmarked adult male caribou swim between Wester and
-Eastern Indian Islands (Figure 1), a swim of at least 470 m which took ~9
-minutes (~52 m per minute). This observation prompted us to investigate whether
-swimming behaviour was common among GPS radio-collared caribou in the population
-(n = 29 adult females: 1240g, GPS 4400M; Lotek Wireless Inc., Newmarket,
-Ontario, Canada). We identified swimming events when two consecutive GPS (2-hour
-relocation rates) locations from an individual occurred on different islands.
-This region typically experiences pack ice during the winter, so caribou may be
-able to travel between islands by walking over the ice. We restricted our GPS
+The Fogo Island archipelago, off the coast of Newfoundland, Canada, is home to
+approximately 300 woodland caribou. During routine fieldwork on May 30, 2017, we
+observed an unmarked adult male caribou swim between Western and Eastern Indian
+Islands (Figure 1), a swim of at least 470 m which took ~9 minutes (~52 m per
+minute). This observation prompted us to investigate whether swimming behaviour
+was common among GPS radio-collared caribou in the population (n = 29 adult
+females; 1240g, GPS 4400M; Lotek Wireless Inc., Newmarket, Ontario, Canada). We
+identified swimming events as two consecutive GPS (2-hour relocation rates)
+locations from an individual occurred on different islands. Northeastern
+Newfoundland typically experiences pack ice during winter and caribou are known
+locally to travel between islands by walking over the ice. We restricted our GPS
 data to the ice-free period of the year (April 1 - November 25) and we
-identified three adult female caribou that collectively had 86 swimming
-events (Figure 2).
+identified three adult female caribou that collectively had 86 swimming events
+(Figure 2). In addition to our own visual and remotely sensed observations,
+residents of Fogo Island have also reported observing caribou swimming between
+islands on numerous occaisions.
 
@@
 
-## Forage limitation
+# Forage limitation
 Forage limitation and over-grazing is a major concern for caribou populations
@@
 
-## Predator avoidance
+# Predator avoidance
 Predator avoidance could also contribute to swimming behaviour of caribou
@@
 
-## Individual variation 
+# Individual variation 
 In total, 3 of 29 collared female caribou swam between islands in the Fogo
@@
 
-## Summary and conclusion
+# Summary and conclusion
 For several decades after their introduction to Fogo Island, the caribou
@@
 
-## Acknowledgements 
+# Acknowledgements 
 We thank M. Laforge, M. Bonar, C. Hart, and R. Huang for help in the field.
@@
 
-## Literature Cited
+# Literature Cited

### 2020-02-21 paper/citations.R
refs

@@ -3,6 +3,9 @@ library(knitcitations)
 
 ## only 15 citations
 
+## Bastille-Rousseau et al 2016
+citep('10.1139/cjz-2015-0127')
+
 ## Bergerud 1985
 citep('10.1139/z85-199')
 
@@ -41,6 +44,9 @@ citep('10.2307/2937087')
 
 ## Schaefer 2016
 citep('10.1093/jmammal/gyv184')
+
+## Rayl et al 2014
+citep('10.1644/13-MAMM-A-133.1')
 
 ## Ricca et al 2012
 citep('10.1007/s10530-012-0195-z')

### 2020-02-21 paper/manuscript.Rmd
Update manuscript.Rmd and ~$nuscript.docx

@@
 part of a series of translocations and introductions throughout the province
-(Bergerud & Mercer 1989). Since 2016 we have been studying the Fogo Island
-caribou herd (Bonar et al. 2017; Bonar et al. 2018; Peignier et al. 2019). To
-date, we have fitted 29 adult female caribou with GPS collars (1240g, GPS 4400M;
-Lotek Wireless Inc., Newmarket, Ontario, Canada), which were programmed to
-collect relocation fixes every 2 hours (see Bonar et al. 2018 for details). We
-used GPS locations from these individuals to investigate oceanic swimming
-events. -->
+(Bergerud & Mercer 1989).
 
@@
 Historic predators of caribou such as wolves (*Canis lupus*) and black bears
-(*Ursus americanus*) are not present in the Fogo Island archipelago; coyotes
-(*Canis latrans*) are the top predator in this region. Predation by coyotes on
-adult caribou is relatively rare (ref?), so the potential effect of predators on
-caribou swimming behaviour is likely to be a flight response to non-lethal
-encounters. Calves are susceptible to coyotes, particularly young calves, but
-the energetic costs to oceanic swimming for calves is likely to be far greater for small
-calves compared to adults.
+(*Ursus americanus*) are not present in the Fogo Island archipelago, thus
+leaving coyotes (*Canis latrans*) as the top predator. Predation by coyotes on
+adult caribou is relatively rare [@Bastille_Rousseau_2016], so the potential
+effect of predators on caribou swimming behaviour is limited to non-consumptive
+effects associated with encounters and subsequent flight response. Adult females
+with calves at heel may be an exception because calves are highly susceptible to
+coyotes [@Rayl_2014], but the energetic costs for calves to swim in the ocean is
+likely far greater than adults. As one might expect, none <!--confirm--> of the
+86 swimming events we observed occurred in the weeks after calves were born,
+suggesting swimming with in the ocean with a calf-at-heel is likely more risky
+than other predator avoidance behaviours.
 

### 2020-02-21 paper/manuscript.Rmd
Update manuscript.Rmd

@@
 males or females. Our initial observation  was of an adult male, but our GPS
-data only includes females. 
+data only includes females. <!-- QW: not sure how to finish this off -->
 <!-- QW: I'm not sure we need to say this.. 
@@
 Our observations add to the evidence that caribou can, and occasionally do, swim
-in the ocean (Table 1). While past observations (e.g.
-@Jeffery_2007; @Ricca_2012] <!-- add Miller 1995, Miller 2002 --> have noted the
-exceptional nature of caribou swimming in the ocean, we suggest that this
-behaviour is likely more common than previously thought for caribou living on
-oceanic islands. We posit forage limitation and the associated density-dependent
-habitat selection is an ultimate explanation for why caribou swimming in the
-ocean, with predator avoidance as a potential proximate explanation. Although
-our inference is limited to observations, our observations suggest that islands
-represent discrete foraging patches for terrestrial animals that could influence
-fitness via increased foraging opportunities and safety from predators. We also
-suggest that individuals vary in their evaluation of costs and benefits of
-movement between these habitat patches.
+in the ocean (e.g. @Jeffery_2007; @Ricca_2012]. <!-- add Miller 1995, Miller
+2002 -->  Taken together, we suggest that swimming is likely more common than
+previously thought for caribou living on oceanic islands. We posit forage
+limitation and the associated density-dependent habitat selection is an ultimate
+explanation for why caribou swimming in the ocean, with predator avoidance as a
+potential proximate explanation. Although our inference is limited to
+observations, our observations suggest that islands represent discrete foraging
+patches for terrestrial animals that could influence fitness via increased
+foraging opportunities and safety from predators. We also suggest that
+individuals vary in their evaluation of costs and benefits of movement between
+these habitat patches.
 

### 2020-02-21 paper/manuscript.Rmd
Update manuscript.Rmd, manuscript.docx, and ~$nuscript.docx

@@
 ---
-title: Interisland movements of woodland caribou in Newfoundland
+title: Interisland movements and oceanic swimming of woodland caribou in Newfoundland
 author: Quinn M.R. Webber^1\*^, Jack G. Hendrix^1^, Alec L. Robitaille^2^,  Eric Vander Wal^1,2^
@@
 explanations for these swimming events, forage limitation (Miller 2002) and
-predator avoidance [@Jeffrey_2007], and discuss why these swimming events
+predator avoidance [@Jeffery_2007], and discuss why these swimming events
 appear to be relatively commonplace in the population.

### 2020-02-21 paper/manuscript.Rmd
Update manuscript.Rmd

@@
 Predator avoidance could also contribute to swimming behaviour of caribou
-(Miller 2002) <!-- find doi for this article -->. During calving, female caribou
-are known to swim to small islands in freshwater lakes give birth, or with their
+(Miller 2002).<!-- find doi for this article --> During calving, female caribou
+are known to swim to small islands in freshwater lakes and give birth, or with their
 calves after birth, to avoid predation [@Bergerud_1987]. The use of shoreline

### 2020-02-22 paper/citations.R
update citations

@@ -4,55 +4,55 @@ library(knitcitations)
 ## only 15 citations
 
 ## Bastille-Rousseau et al 2016
-citep('10.1139/cjz-2015-0127')
+#citep('10.1139/cjz-2015-0127')
 
 ## Bergerud 1985
-citep('10.1139/z85-199')
+#citep('10.1139/z85-199')
 
 ## Bergerud and Page 1987
-citep('10.1139/z87-249')
+#citep('10.1139/z87-249')
 
 ## Bergerud and Mercer 1989 - get actual doi
 
 ## Bergerud 1990
-citep('10.1016/S0003-3472(05)80882-6')
+#citep('10.1016/S0003-3472(05)80882-6')
 
 ## Bradbury et al 2015
-citep('10.1093/beheco/arv078')
+#citep('10.1093/beheco/arv078')
 
 ## Fish 1993
-citep('10.1071/ZO9940079')
+#citep('10.1071/ZO9940079')
 
 ## Jeffrey et al 2007
-citep('10.7557/2.27.4.320')
+#citep('10.7557/2.27.4.320')
 
 ## Leblond et al 2016
-citep('10.1186/s40462-016-0079-4')
+#citep('10.1186/s40462-016-0079-4')
 
 ## Miller et al 1977
-citep('10.1139/z77-131')
+#citep('10.1139/z77-131')
 
 ## Miller and Gunn 1986 
-citep('10.14430/arctic2052')
+#citep('10.14430/arctic2052')
 
 ## Miller 1995 -- get actual doi
-
+# yes
 ## Miller 2002 -- get actual doi
-
+#yes
 ## Morris 1987
-citep('10.2307/2937087')
+#citep('10.2307/2937087')
 
 ## Schaefer 2016
-citep('10.1093/jmammal/gyv184')
+#citep('10.1093/jmammal/gyv184')
 
 ## Rayl et al 2014
-citep('10.1644/13-MAMM-A-133.1')
+#citep('10.1644/13-MAMM-A-133.1')
 
 ## Ricca et al 2012
-citep('10.1007/s10530-012-0195-z')
+#citep('10.1007/s10530-012-0195-z')
 
 ## Webber and Vander Wal 2018
-citep('10.1111/1365-2656.12773')
+#citep('10.1111/1365-2656.12773')
 
 ## van Beest et al 2014
 citep('10.1111/1365-2656.12115')

### 2020-02-22 paper/manuscript.Rmd
refs

@@
 output: 
-  #  github_document
+    github_document
   # bookdown::pdf_document2:
@@
   #    toc: false
-   bookdown::word_document2:
+  # bookdown::word_document2:
     #reference_docx: template.docx
-bibliography: ../references/knitcite.bib
+bibliography: ../references/swim.bib
 #csl: csl.csl
@@
 Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
-that caribou swim in streams, rivers, and lakes during migration [@Leblond_2016],
-to avoid predators [@Bergerud_1985], and access islands during calving
-[@Bergerud_1990]. Even for adept swimmers like caribou, the energetic
+that caribou swim in streams, rivers, and lakes during migration [@Leblond2016],
+to avoid predators [@Bergerud1985], and access islands during calving
+[@Bergerud1990]. Even for adept swimmers like caribou, the energetic
 expenditure of swimming for quadrupedal mammals is significantly higher than
-walking or running [@Fish_1994], while drowning is also possible [@Miller_1986].
+walking or running [@Fish1993], while drowning is also possible [@Miller1986].
 Despite a large number of coastal and island caribou herds, only a handful of
-examples have documented caribou swimming in the ocean [e.g. @Jeffery_2007;
-@Ricca_2012]. <!-- add Miller 1995/2002 or reference table 1 if we decide to
-keep this -->
+examples have documented caribou swimming in the ocean [e.g. @Miller1995; @Miller2002; @Jeffery2007; @Ricca2012].
 

### 2020-02-22 paper/manuscript.Rmd
rezfddf

@@
 expenditure of swimming for quadrupedal mammals is significantly higher than
-walking or running [@Fish1993], while drowning is also possible [@Miller1986].
+walking or running <!--[@Fish1993]-->, while drowning is also possible [@Miller1985].
 Despite a large number of coastal and island caribou herds, only a handful of
@@
 Caribou may swim between islands in the ocean for similar reasons that drive
-caribou swimming in freshwater [@Leblond_2016]. We present two
-explanations for these swimming events, forage limitation (Miller 2002) and
-predator avoidance [@Jeffery_2007], and discuss why these swimming events
+caribou swimming in freshwater [@Leblond2016]. We present two
+explanations for these swimming events, forage limitation [@Miller2002] and
+predator avoidance [@Jeffery2007], and discuss why these swimming events
 appear to be relatively commonplace in the population.
@@
 Forage limitation and over-grazing is a major concern for caribou populations
-[@Schaefer_2015] and could explain movement between islands. Conspecific
+[@Schaefer2016] and could explain movement between islands. Conspecific
 competition for resources driven by reduced foraging opportunities can decrease
-reproductive success of adult females [@Schaefer_2015]. Forage scarcity has been
+reproductive success of adult females [@Schaefer2016]. Forage scarcity has been
 proposed as a potential reason for caribou moving between arctic islands on the
-sea-ice in winter [@Miller_1977]. For very small islands, over-grazing by newly
-arrived caribou may rapidly deplete forage [@Bergerud_1990]. Without sea ice,
+sea-ice in winter [@Miller1977]. For very small islands, over-grazing by newly
+arrived caribou may rapidly deplete forage [@Bergerud1990]. Without sea ice,
 movement from the mainland to an island, or between islands, requires caribou to
@@
 This trade-off can be understood as density-dependent habitat selection governed
-by the Ideal Free Distribution [@Morris_1987; @Bradbury_2015]. Ideal Free
+by the Ideal Free Distribution [@Morris1987; @Bradbury2015]. Ideal Free
 Distribution theory predicts a fitness equilibrium: when the density in a given
@@
 should relocate and settle new habitat patches, so that fitness is equal across
-all patches [@Bradbury_2015]. Islands act as discrete habitat patches in
+all patches [@Bradbury2015]. Islands act as discrete habitat patches in
 this case, but the costs associated with swimming create a trade-off when moving
@@
 Predator avoidance could also contribute to swimming behaviour of caribou
-(Miller 2002).<!-- find doi for this article --> During calving, female caribou
-are known to swim to small islands in freshwater lakes and give birth, or with their
-calves after birth, to avoid predation [@Bergerud_1987]. The use of shoreline
-habitat in summer is considered an effective anti-predator calving strategy by
-allowing for swimming to islands or peninsulas as a predation avoidance flight
-response [@Bergerud_1990]. The use of water, and by association occasionally
-swimming, is therefore part of the fine-scale interactions between caribou and
-their predators and the use of islands appears to be an effective anti-predator
-strategy.
+[@Miller2002]. During calving, female caribou are known to swim to small islands
+in freshwater lakes and give birth, or with their calves after birth, to avoid
+predation [@Bergerud1985]. The use of shoreline habitat in summer is considered
+an effective anti-predator calving strategy by allowing for swimming to islands
+or peninsulas as a predation avoidance flight response [@Bergerud1990]. The use
+of water, and by association occasionally swimming, is therefore part of the
+fine-scale interactions between caribou and their predators and the use of
+islands appears to be an effective anti-predator strategy.
 
@@
 leaving coyotes (*Canis latrans*) as the top predator. Predation by coyotes on
-adult caribou is relatively rare [@Bastille_Rousseau_2016], so the potential
+adult caribou is relatively rare [@BastilleRousseau2016], so the potential
 effect of predators on caribou swimming behaviour is limited to non-consumptive
@@
 with calves at heel may be an exception because calves are highly susceptible to
-coyotes [@Rayl_2014], but the energetic costs for calves to swim in the ocean is
+coyotes [@Rayl2014], but the energetic costs for calves to swim in the ocean is
 likely far greater than adults. As one might expect, none <!--confirm--> of the
@@
 For several decades after their introduction to Fogo Island, the caribou
-population did not exceed ~100 individuals (Bergerud & Mercer 1989; Newfoundland
-and Labrador Wildlife Division, unpublished data). During the 1990s, population
+population did not exceed ~100 individuals [@Bergerud1989 ; Newfoundland and
+Labrador Wildlife Division, unpublished data). During the 1990s, population
 density reached 300 animals and anecdotal evidence suggests that [only then???]
@@
 new islands when the average fitness of individuals on the starting island
-exceeds the density-fitness equilibrium [@Morris_1987]. While this is an
-ultimate explanation to a series of proximate observations, fine-scale
-density-dependent habitat selection is theoretically possible
-[@Webber_2017]. For horses (*Equus ferus caballus*), selection for high quality
-habitat was strongest at low population densities, but individuals tended to
-settle in lower quality habitats as population density increased, following
-Ideal Free Distribution theory [@van_Beest_2013]. Predator avoidance could also
-contribute to caribou swimming behaviour, but given the presumed low density of coyotes on Fogo Island and that predation primarily only affects young calves, we suspect predator is a possible, but less likely explanation.
+exceeds the density-fitness equilibrium [@Morris1987]. While this is an ultimate
+explanation to a series of proximate observations, fine-scale density-dependent
+habitat selection is theoretically possible [@Webber2018]. Predator avoidance
+could also contribute to caribou swimming behaviour, but given the presumed low
+density of coyotes on Fogo Island and that predation primarily only affects
+young calves, we suspect predator is a possible, but less likely explanation.
 <!-- this last bit is sort of me bullshitting, not sure it's actually defensible? 
@@
 Our observations add to the evidence that caribou can, and occasionally do, swim
-in the ocean (e.g. @Jeffery_2007; @Ricca_2012]. <!-- add Miller 1995, Miller
-2002 -->  Taken together, we suggest that swimming is likely more common than
-previously thought for caribou living on oceanic islands. We posit forage
-limitation and the associated density-dependent habitat selection is an ultimate
-explanation for why caribou swimming in the ocean, with predator avoidance as a
-potential proximate explanation. Although our inference is limited to
-observations, our observations suggest that islands represent discrete foraging
-patches for terrestrial animals that could influence fitness via increased
-foraging opportunities and safety from predators. We also suggest that
-individuals vary in their evaluation of costs and benefits of movement between
-these habitat patches.
+in the ocean [e.g. @Miller1995; @Miller2002; @Jeffery2007; @Ricca2012]. Taken
+together, we suggest that swimming is likely more common than previously thought
+for caribou living on oceanic islands. We posit forage limitation and the
+associated density-dependent habitat selection is an ultimate explanation for
+why caribou swimming in the ocean, with predator avoidance as a potential
+proximate explanation. Although our inference is limited to observations, our
+observations suggest that islands represent discrete foraging patches for
+terrestrial animals that could influence fitness via increased foraging
+opportunities and safety from predators. We also suggest that individuals vary
+in their evaluation of costs and benefits of movement between these habitat
+patches.
 

### 2020-02-22 paper/manuscript.Rmd
fiddly refs

@@
   #    toc: false
-  # bookdown::word_document2:
-    #reference_docx: template.docx
+   bookdown::word_document2:
+    reference_docx: template.docx
 bibliography: ../references/swim.bib
@@
 For several decades after their introduction to Fogo Island, the caribou
-population did not exceed ~100 individuals [@Bergerud1989 ; Newfoundland and
-Labrador Wildlife Division, unpublished data). During the 1990s, population
+population did not exceed ~100 individuals [@Bergerud1989, Newfoundland and
+Labrador Wildlife Division, unpublished data]. During the 1990s, population
 density reached 300 animals and anecdotal evidence suggests that [only then???]
@@
 density of coyotes on Fogo Island and that predation primarily only affects
-young calves, we suspect predator is a possible, but less likely explanation.
+young calves, we suspect predator risk is a possible, but less likely
+explanation for swimming.
 <!-- this last bit is sort of me bullshitting, not sure it's actually defensible? 
-QW: changed it a bit -->
+QW: changed it a bit - how is this? -->
 

### 2020-02-22 paper/manuscript.Rmd
header

@@
 
-# Introduction
+**Introduction**
+
 Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
@@
 
-# Forage limitation
+**Forage limitation**
+
 Forage limitation and over-grazing is a major concern for caribou populations
@@
 
-# Predator avoidance
+**Predator avoidance**
+
 Predator avoidance could also contribute to swimming behaviour of caribou
@@
 leaving coyotes (*Canis latrans*) as the top predator. Predation by coyotes on
-adult caribou is relatively rare [@BastilleRousseau2016], so the potential
+adult caribou is relatively rare [@Bastille-Rousseau2016], so the potential
 effect of predators on caribou swimming behaviour is limited to non-consumptive
@@
 
-# Individual variation 
+**Individual variation**
+
 In total, 3 of 29 collared female caribou swam between islands in the Fogo
@@
 
-# Summary and conclusion
+**Summary and conclusion**
+
 For several decades after their introduction to Fogo Island, the caribou
@@
 
-# Acknowledgements 
+**Acknowledgements**
+
 We thank M. Laforge, M. Bonar, C. Hart, and R. Huang for help in the field.
 Logistical support was provided by L. Bixby. We also thank members of the
-Newfoundland Wildlife Division including S. Moores, B. Adams, K. Lewis, C.
-Doucet, W. Barney, F. Norman, R. Otto, J. Neville, P. Saunders, T. Porter, P.
-Tremblett, S. Gullage, T. Hodder, D. Jennings, J. McGinn, M. Frampton, P.
-McClement, and S. Mahoney for data collection, management, and logistics. We
-thank T. Bergerud and S. Mahoney for their vision in initiating much of the work
-on caribou in Newfoundland. Funding for this study was provided by the National
-Sciences and Engineering Research Council (QMRW, JGH, EVW). We respectfully
-acknowledge the territory in which data were collected and analyzed as the
-ancestral homelands of the Beothuk, and the island of Newfoundland as the
-ancestral homelands of the Mi’kmaq and Beothuk.
+Newfoundland Wildlife Division including S. Moores, B. Adams, C. Doucet, W.
+Barney, and J. Neville for logistical support in the field as well as help with
+data data collection and management. We thank T. Bergerud and S. Mahoney for
+their vision in initiating much of the work on caribou in Newfoundland. Funding
+for this study was provided by the National Sciences and Engineering Research
+Council (QMRW, JGH, EVW). We respectfully acknowledge the territory in which
+data were collected and analyzed as the ancestral homelands of the Beothuk, and
+the island of Newfoundland as the ancestral homelands of the Mi’kmaq and
+Beothuk.
 <!-- Simplified the funding statement, do we need to name the specific scholarships/grants? Eric opinion??-->
 
+\newpage
+
+\captionsetup[figure]{labelformat=empty}
+
+![**Figure 1**: Image of swimming unmarled adult male caribou (*Rangifer tarandus*) from Western to Eastern Indian Island taken on 30 May 2017.](../graphics/IMG_1562.JPG)
+
+\newpage
+
 # Literature Cited

### 2020-02-22 paper/manuscript.Rmd
blah blah blah blah

@@
 date: 
-   ^1^Cognitive and Behavioural Ecology Interdisciplinary Program, Memorial University of Newfoundland
+    ^1^Cognitive and Behavioural Ecology Interdisciplinary Program, Memorial
+   University of Newfoundland
     ^2^Department of Biology, Memorial University of Newfoundland, Canada
@@
   #    toc: false
-   bookdown::word_document2:
-    reference_docx: template.docx
+   #bookdown::word_document2:
+    #reference_docx: template.docx
 bibliography: ../references/swim.bib
@@
 locally to travel between islands by walking over the ice. We restricted our GPS
-data to the ice-free period of the year (April 1 - November 25) and we
-identified three adult female caribou that collectively had 86 swimming events
-(Figure 2). In addition to our own visual and remotely sensed observations,
-residents of Fogo Island have also reported observing caribou swimming between
-islands on numerous occaisions.
+data to the ice-free period of the year (April 1 - November 25) <!--confirm
+dates --> and we identified three adult female caribou that collectively had 86
+swimming events (Figure 2). In addition to our own visual and remotely sensed
+observations, residents of Fogo Island have also reported observing caribou
+swimming between islands on numerous occaisions.
 
+<!-- add numbers in here -->
 In total, these 86 swimming events occured in month (n = Y), month (n = Y), and
@@
 part of a series of translocations and introductions throughout the province
-(Bergerud & Mercer 1989).
-
-<!-- still not sure we need this paragraph, I wonder if the island size
-information could be included in the figure caption?--> Fogo Island is the
-largest island in the archipelago (237.71 km^2^), but there are at least three
-other large islands: Western Indian (77.6 km2), Eastern Indian (38.7 km2) and
-Change (XX km2) Islands, as well as numerous smaller islands, including
-Blundon’s (1.18 km2), North Long (1.01 km2), South Long (0.48 km2), Kate’s (1.64
-km2), and Brother’s (1.59 km2) Islands (see Figure 1). Habitats are similar
-across the archipelago, consisting largely of coniferous and mixed forests of
-balsam fir (*Abies balsamea*), black spruce (*Picea mariana*), and white birch
-(*Betula papyrifera*) as well as bogs, lakes, lichen and rocky barrens.
+(Bergerud & Mercer 1989). -->
 
@@
 archipelago. These swimming events were relatively frequent, on average occuring
-once every 24 <!--double check--> days during ice-free seasons. However, despite
+once every 24 <!--confirm--> days during ice-free seasons. However, despite
 relatively frequent swimming events for these three individuals, we were unable
@@
 
-# Literature Cited
+\captionsetup[figure]{labelformat=empty}
+
+![**Figure 2**: Map of the Fogo Island archipelago. Fogo Island is the largest
+island in the archipelago (237.71 km^2^), but there are at least three other
+large islands: Western Indian (77.6 km^2^), Eastern Indian (38.7 km^2^) and
+Change (XX km^2^) Islands, as well as numerous smaller islands, including
+Blundon’s (1.18 km^2^), North Long (1.01 km^2^), South Long (0.48 km^2^), Kate’s
+(1.64 km^2^), and Brother’s (1.59 km^2^) Islands. Habitats are similar across
+the archipelago, consisting largely of coniferous and mixed forests of balsam
+fir (*Abies balsamea*), black spruce (*Picea mariana*), and white birch 
+(*Betula papyrifera*) as well as bogs, lakes, lichen and rocky
+barrens.]()<!--(../graphics/IMG_1562.JPG)-->
+
+\newpage
+
+**Literature Cited**

### 2020-02-24 paper/manuscript.Rmd
comment out new fig
fix typos

@@
 observations, residents of Fogo Island have also reported observing caribou
-swimming between islands on numerous occaisions.
+swimming between islands on numerous occasions.
 
 <!-- add numbers in here -->
-In total, these 86 swimming events occured in month (n = Y), month (n = Y), and
+In total, these 86 swimming events occurred in month (n = Y), month (n = Y), and
 month (n = Y), while the remaining events were in month, month, and month (n =
@@
 In total, 3 of 29 collared female caribou swam between islands in the Fogo
-archipelago. These swimming events were relatively frequent, on average occuring
+archipelago. These swimming events were relatively frequent, on average occurring
 once every 24 <!--confirm--> days during ice-free seasons. However, despite
@@
 
-![**Figure 1**: Image of swimming unmarled adult male caribou (*Rangifer tarandus*) from Western to Eastern Indian Island taken on 30 May 2017.](../graphics/IMG_1562.JPG)
+![**Figure 1**: Image of swimming unmarked adult male caribou (*Rangifer tarandus*) from Western to Eastern Indian Island taken on 30 May 2017.](../graphics/IMG_1562.JPG)
 
@@
 
-![**Figure 2**: Map of the Fogo Island archipelago. Fogo Island is the largest
+<!--![**Figure 2**: Map of the Fogo Island archipelago. Fogo Island is the largest
 island in the archipelago (237.71 km^2^), but there are at least three other

### 2020-02-24 scripts/1-extract-islands.js
fst islands with EE

@@ -1,0 +1,41 @@
+// Extract islands with SENTINEL - 2
+// Alec Robitaille
+
+var geometry = ee.Geometry.Polygon(
+        [[[-54.45048689517963, 49.760281833675236],
+          [-54.45048689517963, 49.49787241520817],
+          [-53.984941240882755, 49.49787241520817],
+          [-53.984941240882755, 49.760281833675236]]], null, false);
+          
+var col = ee.ImageCollection('COPERNICUS/S2')
+  .filterDate('2019-07-01', '2019-10-01')
+  .filterBounds(geometry)
+  
+var value = function(img) {
+  return(img.addBands(img.divide(10000)
+                         .select('B4', 'B3', 'B2')
+                         .rgbToHsv()
+                         .select('value')))
+}
+
+var dark = function(img) {
+  return(img.addBands(ee.Image(1).divide(img.select('value')).rename(['dark'])))
+}
+
+var qual = col.map(value)
+              .map(dark)
+              .qualityMosaic('dark')
+
+var land = qual.select('B8').gt(350)
+
+Map.addLayer(col)
+Map.addLayer(qual)
+Map.addLayer(land)
+
+
+Export.image.toDrive({
+  image: land,
+  description: 'caribouswimming',
+  scale: 20,
+  fileNamePrefix: 'fogo-land-caribou-swimming'
+})

### 2020-02-24 scripts/1-extract-islands.js
add export region

@@ -1,4 +1,4 @@
-// Extract islands with SENTINEL - 2
+// Extract islands with SENTINEL2
 // Alec Robitaille
 
 var geometry = ee.Geometry.Polygon(
@@ -9,33 +9,34 @@ var geometry = ee.Geometry.Polygon(
           
 var col = ee.ImageCollection('COPERNICUS/S2')
   .filterDate('2019-07-01', '2019-10-01')
-  .filterBounds(geometry)
+  .filterBounds(geometry);
   
 var value = function(img) {
   return(img.addBands(img.divide(10000)
                          .select('B4', 'B3', 'B2')
                          .rgbToHsv()
-                         .select('value')))
-}
+                         .select('value')));
+};
 
 var dark = function(img) {
-  return(img.addBands(ee.Image(1).divide(img.select('value')).rename(['dark'])))
-}
+  return(img.addBands(ee.Image(1).divide(img.select('value')).rename(['dark'])));
+};
 
 var qual = col.map(value)
               .map(dark)
-              .qualityMosaic('dark')
+              .qualityMosaic('dark');
 
-var land = qual.select('B8').gt(350)
+var land = qual.select('B8').gt(350);
 
-Map.addLayer(col)
-Map.addLayer(qual)
-Map.addLayer(land)
+Map.addLayer(col);
+Map.addLayer(qual);
+Map.addLayer(land);
 
 
 Export.image.toDrive({
   image: land,
   description: 'caribouswimming',
   scale: 20,
-  fileNamePrefix: 'fogo-land-caribou-swimming'
-})
+  fileNamePrefix: 'fogo-land-caribou-swimming',
+  region: geometry
+});

### 2020-02-24 scripts/1-extract-islands.js
add crs export

@@ -38,5 +38,6 @@ Export.image.toDrive({
   description: 'caribouswimming',
   scale: 20,
   fileNamePrefix: 'fogo-land-caribou-swimming',
-  region: geometry
+  region: geometry,
+  crs: 'EPSG:32621'
 });

### 2020-02-24 scripts/2-extractPoints.R
read new ee raster

@@ -20,7 +20,8 @@ lapply(libs, require, character.only = TRUE)
 
 ### Load data ----
 caribou <- fread('input/FogoCaribou.csv')
-r <- readRDS('output/islandsRaster.Rds')
+r <- raster('output/fogo-land-caribou-swimming.tif')
+# r <- readRDS('output/islandsRaster.Rds')
 
 ### Prep data ----
 # Datetime

### 2020-02-24 paper/manuscript.Rmd
Update manuscript.Rmd

@@
 expenditure of swimming for quadrupedal mammals is significantly higher than
-walking or running <!--[@Fish1993]-->, while drowning is also possible [@Miller1985].
+walking or running, <!--[@Fish1993]--> while drowning is also possible [@Miller1985].
 Despite a large number of coastal and island caribou herds, only a handful of
@@
 locally to travel between islands by walking over the ice. We restricted our GPS
-data to the ice-free period of the year (April 1 - November 25) <!--confirm
-dates --> and we identified three adult female caribou that collectively had 86
-swimming events (Figure 2). In addition to our own visual and remotely sensed
-observations, residents of Fogo Island have also reported observing caribou
-swimming between islands on numerous occasions.
+data to the ice-free period of the year (April 1 - November 25)<!--confirm dates
+--> and we identified three adult female caribou that collectively had 86
+swimming events over three years (Figure 2). In addition to our own visual and
+remotely sensed observations, residents of Fogo Island have also reported
+observing caribou swimming between islands on numerous occasions.
 

### 2020-02-24 scripts/1-extract-islands.js
calc connected components in ee

@@ -32,12 +32,13 @@ Map.addLayer(col);
 Map.addLayer(qual);
 Map.addLayer(land);
 
+var conn = land.connectedComponents(ee.Kernel.square(1), 256).toDouble()
 
 Export.image.toDrive({
-  image: land,
+  image: conn,
   description: 'caribouswimming',
   scale: 20,
-  fileNamePrefix: 'fogo-land-caribou-swimming',
+  fileNamePrefix: 'fogo-components-caribou-swimming',
   region: geometry,
   crs: 'EPSG:32621'
 });

### 2020-02-24 scripts/2-extractPoints.R
calc connected components in r

@@ -24,6 +24,13 @@ r <- raster('output/fogo-land-caribou-swimming.tif')
 # r <- readRDS('output/islandsRaster.Rds')
 
 ### Prep data ----
+# Generate connected components
+conn <- clump(r)
+
+# Reduce small island/pixel noise
+fuzz <- clump(buffer(r, 10))
+
+
 # Datetime
 caribou[, c('idate', 'itime') := .(as.IDate(idate), as.ITime(itime))]
 

### 2020-02-24 scripts/2-extractPoints.R
igraph needed for clumps

@@ -6,7 +6,7 @@
 libs <- c(
   'data.table',
   'ggplot2',
-  # 'dplyr',
+  'igraph',
   'raster',
   # 'prettymapr',
   # 'rosm',

### 2020-02-24 scripts/2-extractPoints.R
need NAs to buffer

@@ -28,7 +28,9 @@ r <- raster('output/fogo-land-caribou-swimming.tif')
 conn <- clump(r)
 
 # Reduce small island/pixel noise
-fuzz <- clump(buffer(r, 10))
+rna <- r
+rna[rna != 1] <- NA
+fuzz <- clump(buffer(r == 1, 10))
 
 
 # Datetime

### 2020-03-02 paper/manuscript.Rmd
minor edits

@@
 expenditure of swimming for quadrupedal mammals is significantly higher than
-walking or running, <!--[@Fish1993]--> while drowning is also possible [@Miller1985].
+walking or running, <!--[@Fish1993]--> and drowning is also possible [@Miller1985].
 Despite a large number of coastal and island caribou herds, only a handful of
@@
 observed an unmarked adult male caribou swim between Western and Eastern Indian
-Islands (Figure 1), a swim of at least 470 m which took ~9 minutes (~52 m per
+Islands (Figure 1), a distance of at least 470 m which took ~9 minutes (~52 m per
 minute). This observation prompted us to investigate whether swimming behaviour

### 2020-03-09 paper/manuscript.Rmd
edits to discussion, removed most predator stuff, added a couple sentences on non-ice-free movements that need numbers

@@
 Caribou may swim between islands in the ocean for similar reasons that drive
-caribou swimming in freshwater [@Leblond2016]. We present two
-explanations for these swimming events, forage limitation [@Miller2002] and
-predator avoidance [@Jeffery2007], and discuss why these swimming events
-appear to be relatively commonplace in the population.
+caribou swimming in freshwater [@Leblond2016]. We propose that one such
+explanation, the forage limitation hypothesis, is likely driving these oceanic
+swimming events [@Miller2002].
+
+Predation pressure is also known to affect caribou swimming behaviour, both in
+immediate situations to escape predators [@Bergerud1990] and also pre-emptively
+to give birth or raise calves in safer areas like isolated islands
+[@Bergerud1985]. The Fogo Island archipelago does not contain historic caribou
+predators such as wolves (*Canis lupus*) or black bears (*Ursus americanus*),
+leaving coyotes (*Canis latrans*) as the top predator. Coyotes may predate
+calves [@Rayl2014] but do not pose a substantial risk to adult caribou
+[@Bastille-Rousseau2016]; predation in general does not exert as strong a force
+on Fogo Island caribou as in mainland populations. Calves that are most
+vulnerable to predation likely face a much higher energetic cost of oceanic
+swimming than do adults, and indeed none of the females observed swimming had a
+calf with them at the time.<!-- female-centric rather than time-of-year
+["swimming events occurred in the weeks after calves were born"] seems better,
+if we can confirm this--> Given the frequency of these swimming events, and that
+swimming in the ocean with a calf-at-heel is riskier than other predator
+avoidance strategies, we suspect that predation risk is not a main driver of
+oceanic swimming in Fogo Island caribou.
 
@@
 
-**Predator avoidance**
-
-Predator avoidance could also contribute to swimming behaviour of caribou
-[@Miller2002]. During calving, female caribou are known to swim to small islands
-in freshwater lakes and give birth, or with their calves after birth, to avoid
-predation [@Bergerud1985]. The use of shoreline habitat in summer is considered
-an effective anti-predator calving strategy by allowing for swimming to islands
-or peninsulas as a predation avoidance flight response [@Bergerud1990]. The use
-of water, and by association occasionally swimming, is therefore part of the
-fine-scale interactions between caribou and their predators and the use of
-islands appears to be an effective anti-predator strategy.
-
-Historic predators of caribou such as wolves (*Canis lupus*) and black bears
-(*Ursus americanus*) are not present in the Fogo Island archipelago, thus
-leaving coyotes (*Canis latrans*) as the top predator. Predation by coyotes on
-adult caribou is relatively rare [@Bastille-Rousseau2016], so the potential
-effect of predators on caribou swimming behaviour is limited to non-consumptive
-effects associated with encounters and subsequent flight response. Adult females
-with calves at heel may be an exception because calves are highly susceptible to
-coyotes [@Rayl2014], but the energetic costs for calves to swim in the ocean is
-likely far greater than adults. As one might expect, none <!--confirm--> of the
-86 swimming events we observed occurred in the weeks after calves were born,
-suggesting swimming with in the ocean with a calf-at-heel is likely more risky
-than other predator avoidance behaviours.
-
-<!-- did the swimming events occur during the calving season? were these three breeders/non-breeders? did any of these females make these journeys with their calves?  -->
-
 **Individual variation**
 
-In total, 3 of 29 collared female caribou swam between islands in the Fogo
-archipelago. These swimming events were relatively frequent, on average occurring
-once every 24 <!--confirm--> days during ice-free seasons. However, despite
-relatively frequent swimming events for these three individuals, we were unable
-to identify any swimming events for the other 26 collared females. It remains
-unclear why ocean swimming is relatively common among ~10% of our marked
-population, while the other 90% never swam. Of the three collared individuals,
-two were collared on Western Indian Island and the third was collared on Fogo
-Island. It is possible that individuals born on, or that live on, Eastern and
-Western Indian Island swim are more likely to swim between islands because
-inter-island distances are shorter than between Western Indian and Fogo Islands
-(Figure 2). Furthermore, it is unclear whether swimming is more common among
-males or females. Our initial observation  was of an adult male, but our GPS
-data only includes females. <!-- QW: not sure how to finish this off -->
-<!-- QW: I'm not sure we need to say this.. 
+In total, 12 of 29 collared female caribou swam between islands in the Fogo
+archipelago. Only 3 of these 12 had more than XX swimming events during the
+study period, so approximately one third of our marked population swam
+irregularly and infrequently, whereas for ~10% of individuals these swimming
+events were relatively common, on average occurring once every 24 <!--confirm-->
+days during ice-free seasons. We did not detect any swimming events for the
+remaining 17 individuals; it remains unclear whether a subset of the population
+does not engage in this behaviour at all, or whether these swimming events are
+extremely rare for most caribou and we happened to detect some for those nine
+individuals with very few observed swimming events. We are also unable to draw
+conclusions about the prevalence of swimming in the ocean by male caribou: our
+initial visual observation was of an adult male, but our GPS data only included
+females.
+
+<!-- "During the winter, when individuals could theoretically move between islands more easily by walking across ice, XX individuals were observed moving between islands.
+YY [(a) majority, (b) minority] of these were not detected swimming during the ice-free season and only moved between islands overwinter.
+This suggests that [(a) moving between islands is a relatively consistent behaviour within individuals, where some caribou refrain from both swimming and walking over ice] / [(b) caribou who avoided swimming between islands would still take advantage of the benefits provided by these patches, but only when travelling to them was less risky and costly] -->
+
+<!-- QW: I'm not sure we need to say the below.. 
+JGH - yeah, it's pretty extra I guess. I'm not really a fan of the above para either, it's not very convincing and doesn't add much
 It is theoretically possible that these other
@@
 
-The forage limitation and predator avoidance hypotheses are generally thought to
-apply uniformly across populations, but individual caribou could evaluate the
-costs and benefits of swimming to another island differently. It is possible
-that some individuals, in our case ~10%, considered swimming less costly or that
-they perceived competition or predation more acutely than their conspecifics.
-Intrinsic life-history, physiological, or behavioural traits that might
-influence how animals vary in their assessments of swimming trade-offs remains
-unknown. Future work could assess the role of age, stress, or an individuals
-tendency to move potential predictors of swimming behaviour.
+Regardless, oceanic swimming is relatively common among a small minority of our
+marked population. Of the three collared individuals, two were collared on
+Western Indian Island and the third was collared on Fogo Island. Inter-island
+distances between Western and Eastern Indian Islands, and the smaller islands
+around them, are shorter than those between Fogo Island and others (Figure 2).
+Individuals that were born on or spend more time on these smaller islands may
+swim between them more frequently than animals that spend most of their time on
+Fogo Island. The forage limitation hypothesis is generally thought to apply
+uniformly across populations, but individual caribou could evaluate the costs
+and benefits of swimming to another island differently. It is possible that some
+individuals, in our case ~10%, considered swimming to be less costly or
+perceived competition or predation more acutely than their conspecifics.
 
@@
 explanation to a series of proximate observations, fine-scale density-dependent
-habitat selection is theoretically possible [@Webber2018]. Predator avoidance
-could also contribute to caribou swimming behaviour, but given the presumed low
-density of coyotes on Fogo Island and that predation primarily only affects
-young calves, we suspect predator risk is a possible, but less likely
-explanation for swimming.
-<!-- this last bit is sort of me bullshitting, not sure it's actually defensible? 
-QW: changed it a bit - how is this? -->
+habitat selection is theoretically possible [@Webber2018]. Given the presumed
+low density of coyotes on Fogo Island and the minimal risk they pose to all but
+young calves, who are least likely to engage in oceanic swimming, predation risk
+does not appear to be a major driver of the behaviour in this system.
+<!-- QW: changed it a bit - how is this? 
+JGH: simplified these two para's a bit, took out most of the predator stuff.
+I'm a bit confused by the sentence above citing your 2018 paper, I'm not really sure what it's saying & how it relates to the rest of the section.
+there were comments to that effect as well, maybe expand the end a bit to make it clearer how that relates?
+-->
 
@@
 together, we suggest that swimming is likely more common than previously thought
-for caribou living on oceanic islands. We posit forage limitation and the
+for caribou living on oceanic islands, and that forage limitation and the
 associated density-dependent habitat selection is an ultimate explanation for
-why caribou swimming in the ocean, with predator avoidance as a potential
-proximate explanation. Although our inference is limited to observations, our
-observations suggest that islands represent discrete foraging patches for
-terrestrial animals that could influence fitness via increased foraging
-opportunities and safety from predators. We also suggest that individuals vary
-in their evaluation of costs and benefits of movement between these habitat
-patches.
+this phenomenon. Although our inference is limited to observations, islands
+appear to represent discrete foraging patches for terrestrial animals that can
+influence fitness via increased foraging opportunities. We also posit that
+individuals can vary in their evaluation of costs and benefits of movement
+between these habitat patches.
 

### 2020-03-09 scripts/2-extractPoints.R

@@ -1,0 +1,87 @@
+### Extract Island Caribou ====
+# Quinn M.R. Webber, Jack G. Hendrix, Alec L. Robitaille,  Eric Vander Wal
+
+
+### Packages ----
+libs <- c('data.table', 'ggplot2', 'dplyr',
+          'raster', 'prettymapr', 'rosm', 
+          'stars', 'fasterize', 'rgdal')
+lapply(libs, require, character.only = TRUE)
+
+## load data
+caribou <- fread('input/FogoCaribou.csv')
+r <- readRDS('output/islandRaster.RDS')
+
+## Loc fields
+utm21N <- '+proj=utm +zone=21 ellps=WGS84'
+caribou[, c('EASTING', 'NORTHING') := as.data.table(project(cbind(X_COORD, Y_COORD), utm21N))]
+
+## Sub by bounding box
+caribou <- caribou[EASTING > 690000 & EASTING < 800000 &
+                     NORTHING > 5470000  & NORTHING < 5520000]
+
+## Sub by date 
+caribou <- caribou[JDate > 90 & JDate < 330]
+
+## Sub by animals that swam
+swimmers <- caribou[ANIMAL_ID == "FO2016011" |
+                      ANIMAL_ID == "FO2017001" |
+                      ANIMAL_ID == "FO2017013"]
+
+## Extract points on different islands
+swimmers[, islands := extract(r, matrix(c(EASTING, NORTHING), ncol = 2))]
+
+## check points in reference to raster map
+swimmers[, .N, by = "islands"]
+plot(r, xlim = c(685000, 710000), 
+     ylim = c(5490000, 5502000))
+par(new = T)
+plot(NORTHING ~ EASTING, data = swimmers[islands == 55], 
+     xlim = c(685000, 710000), 
+     ylim = c(5490000, 5502000))
+
+
+## cut points that aren't on an island
+swimmers <- swimmers[!is.na(islands),]
+
+## rename islands
+swimmers$StartIsland[swimmers$islands == 43] <- "Fogo"
+swimmers$StartIsland[swimmers$islands == 53] <- "North Long"
+swimmers$StartIsland[swimmers$islands == 55] <- "North Long"
+swimmers$StartIsland[swimmers$islands == 58] <- "Blundon"
+swimmers$StartIsland[swimmers$islands == 67] <- "Brother"
+swimmers$StartIsland[swimmers$islands == 68] <- "W. Indian"
[…]

### 2020-03-17 paper/manuscript.Rmd
minor eds

@@
 <!-- add numbers in here -->
-In total, these 86 swimming events occurred in month (n = Y), month (n = Y), and
-month (n = Y), while the remaining events were in month, month, and month (n =
-Y). On average, caribou swam between islands approximately ever 24 days, but
+In total, these XX swimming events occurred throughout the ice-free season, with a peak in November. On average, caribou swam between islands approximately ever XX days, but
 some individuals swam up to XX times per week/month/whatever. <!--Is it better
@@
 caribou swimming in freshwater [@Leblond2016]. We propose that one such
-explanation, the forage limitation hypothesis, is likely driving these oceanic
+explanation, the forage limitation hypothesis, is likely driving oceanic
 swimming events [@Miller2002].
@@
 oceanic swimming in Fogo Island caribou.
-
-**Forage limitation**
 

### 2020-03-17 paper/manuscript.Rmd
insert fogo deets

@@
 
-<!-- Caribou were introduced to Fogo Island (n = 26 animals between 1964–67) and
-Change Island (n = 5 animals in 1964) from the main Island of Newfoundland as
-part of a series of translocations and introductions throughout the province
-(Bergerud & Mercer 1989). -->
-
 Caribou may swim between islands in the ocean for similar reasons that drive
@@
 competition for resources driven by reduced foraging opportunities can decrease
-reproductive success of adult females [@Schaefer2016]. Forage scarcity has been
+reproductive success of adult females [@Schaefer2016].Forage scarcity has been
 proposed as a potential reason for caribou moving between arctic islands on the
-sea-ice in winter [@Miller1977]. For very small islands, over-grazing by newly
-arrived caribou may rapidly deplete forage [@Bergerud1990]. Without sea ice,
+sea-ice in winter [@Miller1977]. Caribou were introduced to Fogo Island (n = 26
+animals between 1964–67) and Change Island (n = 5 animals in 1964) from
+Newfoundland as part of a series of translocations and introductions throughout
+the province (Bergerud & Mercer 1989). It is unknown when caribou began
+colonizing nearby islands, but given that some are small, it is likely forage is
+depleted relatively quickly. While forage depletion is possible for Fogo Island,
+For very small islands, over-grazing by newly arrived caribou may rapidly
+deplete forage [@Bergerud1990].
+
+Without sea ice,
 movement from the mainland to an island, or between islands, requires caribou to

### 2020-03-17 paper/manuscript.Rmd
intro edits

@@
 to avoid predators [@Bergerud1985], and access islands during calving
-[@Bergerud1990]. Even for adept swimmers like caribou, the energetic
-expenditure of swimming for quadrupedal mammals is significantly higher than
-walking or running, <!--[@Fish1993]--> and drowning is also possible [@Miller1985].
-Despite a large number of coastal and island caribou herds, only a handful of
-examples have documented caribou swimming in the ocean [e.g. @Miller1995; @Miller2002; @Jeffery2007; @Ricca2012].
+[@Bergerud1990]. Even for adept swimmers like caribou, the energetic expenditure
+of swimming for quadrupedal mammals is significantly higher than walking or
+running, <!--[@Fish1993]--> and drowning is also possible [@Miller1985]. Despite
+a large number of coastal and island caribou herds, only a handful of examples
+have documented caribou swimming in the ocean [e.g. @Miller1995; @Miller2002;
+@Jeffery2007; @Ricca2012].
 
@@
 locally to travel between islands by walking over the ice. We restricted our GPS
-data to the ice-free period of the year (April 1 - November 25)<!--confirm dates
+data to the ice-free period of the year (April 1 - December 31) <!--confirm dates
 --> and we identified three adult female caribou that collectively had 86
@@
 <!-- add numbers in here -->
-In total, these XX swimming events occurred throughout the ice-free season, with a peak in November. On average, caribou swam between islands approximately ever XX days, but
-some individuals swam up to XX times per week/month/whatever. <!--Is it better
+In total, these XX swimming events occurred throughout the ice-free season, with
+a peak in November. On average, caribou swam between islands approximately ever
+XX days, but some individuals swam up to XX times per week/month/whatever.
+<!--Is it better
 to just state our results here, or give this sort of teaser? Initially I had "86
@@
 
+<!-- DO WE WANT TO CUT THIS? -->
 Predation pressure is also known to affect caribou swimming behaviour, both in
@@
 the province (Bergerud & Mercer 1989). It is unknown when caribou began
-colonizing nearby islands, but given that some are small, it is likely forage is
-depleted relatively quickly. While forage depletion is possible for Fogo Island,
-For very small islands, over-grazing by newly arrived caribou may rapidly
-deplete forage [@Bergerud1990].
+colonizing nearby islands, but given that some are small, over-grazing by newly
+arrived caribou may rapidly deplete forage [@Bergerud1990]. We posit that some caribou swim to nearby islands and forage
 

### 2020-03-17 paper/manuscript.Rmd
respond to JGH comment
edits

@@
 events by 3 individuals" here, but then we ignore that for a couple paragraphs
-of context before getting back into it... which seemed weird-->
+of context before getting back into it... which seemed weird
+QW: I agree. I've added some Fogo context-->
 
@@
 colonizing nearby islands, but given that some are small, over-grazing by newly
-arrived caribou may rapidly deplete forage [@Bergerud1990]. We posit that some caribou swim to nearby islands and forage
+arrived caribou may rapidly deplete forage [@Bergerud1990]. Although
+unconfirmed, it is also possible that Fogo Island has experienced forage
+depletion since the introduction of caribou in the 1960s. More likely, however,
+is fine-scale competition for foraging resources that drives caribou to swim to
+new islands.
 
-Without sea ice,
-movement from the mainland to an island, or between islands, requires caribou to
-assess the trade-off associated with swimming.
-
-This trade-off can be understood as density-dependent habitat selection governed
-by the Ideal Free Distribution [@Morris1987; @Bradbury2015]. Ideal Free
-Distribution theory predicts a fitness equilibrium: when the density in a given
-habitat patch has exceeded the optimum for fitness within that patch, animals
-should relocate and settle new habitat patches, so that fitness is equal across
-all patches [@Bradbury2015]. Islands act as discrete habitat patches in
+Movement between islands requires caribou to assess the trade-off associated
+with swimming. This trade-off can be understood as density-dependent habitat
+selection governed by the Ideal Free Distribution [@Morris1987; @Bradbury2015].
+Ideal Free Distribution theory predicts a fitness equilibrium: when the density
+in a given habitat patch has exceeded the optimum for fitness within that patch,
+animals should relocate and settle new habitat patches, so that fitness is equal
+across all patches [@Bradbury2015]. Islands act as discrete habitat patches in
 this case, but the costs associated with swimming create a trade-off when moving
 between patches. For caribou in the Fogo Island archipelago, the role of forage
-limitation in animal habitat selection patterns remains unknown,
-but swimming between islands may in part be explained by forage limitation and
-conspecific competition.
+limitation in animal habitat selection patterns remains unknown, but swimming
+between islands may in part be explained by forage limitation and conspecific
+competition.
 

### 2020-03-17 paper/manuscript.Rmd
align
add knitr opts chunk
add pkgs chunk
add input chunk
basic summary stats as vars

@@
 ---
+
+```{r knitropts, include = FALSE}
+knitr::opts_chunk$set(echo = FALSE,
+                      eval = TRUE,
+                      fig.align = 'center')
+```
+
+```{r pkgs}
+library(spatsoc)
+library(data.table)
+```
+
+```{r input}
+edges <- readRDS('output/island-edges.Rds')
+icefree <- edges[season == 'icefree']
+```
+
+```{r summary-stats}
+## Only icefree
+# Number of events
+nevents <- icefree[, .N]
+
+# Number of individuals
+nids <- icefree[, uniqueN(ANIMAL_ID)]
+
+# Number of events by individual
+neventid <- icefree[, .N, ANIMAL_ID]
+
+## Full
+# Number of events by individual + icefree season
+nfull <- edges[, .N, .(ANIMAL_ID, season)]
+```
 
@@
 does not appear to be a major driver of the behaviour in this system.
-<!-- QW: changed it a bit - how is this? 
-JGH: simplified these two para's a bit, took out most of the predator stuff.
-I'm a bit confused by the sentence above citing your 2018 paper, I'm not really sure what it's saying & how it relates to the rest of the section.
-there were comments to that effect as well, maybe expand the end a bit to make it clearer how that relates?
--->
+<!-- QW: changed it a bit - how is this? JGH: simplified these two para's a bit,
+took out most of the predator stuff. I'm a bit confused by the sentence above
+citing your 2018 paper, I'm not really sure what it's saying & how it relates to
+the rest of the section. there were comments to that effect as well, maybe
+expand the end a bit to make it clearer how that relates? -->
 
@@
 Beothuk.
-<!-- Simplified the funding statement, do we need to name the specific scholarships/grants? Eric opinion??-->
+<!-- Simplified the funding statement, do we need to name the specific
+scholarships/grants? Eric opinion??-->
 

### 2020-03-17 paper/manuscript.Rmd
remove predation chunk

--- 2020-03-17 paper/manuscript.Rmd
@@
 swimming events [@Miller2002].
-
-<!-- DO WE WANT TO CUT THIS? -->
-Predation pressure is also known to affect caribou swimming behaviour, both in
-immediate situations to escape predators [@Bergerud1990] and also pre-emptively
-to give birth or raise calves in safer areas like isolated islands
-[@Bergerud1985]. The Fogo Island archipelago does not contain historic caribou
-predators such as wolves (*Canis lupus*) or black bears (*Ursus americanus*),
-leaving coyotes (*Canis latrans*) as the top predator. Coyotes may predate
-calves [@Rayl2014] but do not pose a substantial risk to adult caribou
-[@Bastille-Rousseau2016]; predation in general does not exert as strong a force
-on Fogo Island caribou as in mainland populations. Calves that are most
-vulnerable to predation likely face a much higher energetic cost of oceanic
-swimming than do adults, and indeed none of the females observed swimming had a
-calf with them at the time.<!-- female-centric rather than time-of-year
-["swimming events occurred in the weeks after calves were born"] seems better,
-if we can confirm this--> Given the frequency of these swimming events, and that
-swimming in the ocean with a calf-at-heel is riskier than other predator
-avoidance strategies, we suspect that predation risk is not a main driver of
-oceanic swimming in Fogo Island caribou.
 

### 2020-03-17 paper/manuscript.Rmd
Update manuscript.Rmd and manuscript.md
Update manuscript.Rmd

@@
 XX days, but some individuals swam up to XX times per week/month/whatever.
+Caribou may swim between islands in the ocean for similar reasons that drive
+caribou swimming in freshwater [@Leblond2016]. We propose that one such
+explanation, the forage limitation hypothesis, is likely driving oceanic
+swimming events [@Miller2002].
 <!--Is it better
@@
 QW: I agree. I've added some Fogo context-->
-
-Caribou may swim between islands in the ocean for similar reasons that drive
-caribou swimming in freshwater [@Leblond2016]. We propose that one such
-explanation, the forage limitation hypothesis, is likely driving oceanic
-swimming events [@Miller2002].
 
@@
 Labrador Wildlife Division, unpublished data]. During the 1990s, population
-density reached 300 animals and anecdotal evidence suggests that [only then???]
+density reached 300 animals and anecdotal evidence suggests that
 caribou began to occupy other islands in the archipelago. We surmise that as

### 2020-03-17 paper/manuscript.Rmd
Update manuscript.Rmd

@@
 Labrador Wildlife Division, unpublished data]. During the 1990s, population
-density reached 300 animals and anecdotal evidence suggests that
-caribou began to occupy other islands in the archipelago. We surmise that as
+density reached 300 animals and it is unknown when caribou began to occupy other
+islands, anecdotal evidence suggests it was during the 1990s. We surmise that as
 competition among conspecifics increased along with population size,
@@
 habitat selection is theoretically possible [@Webber2018]. Given the presumed
-low density of coyotes on Fogo Island and the minimal risk they pose to all but
-young calves, who are least likely to engage in oceanic swimming, predation risk
-does not appear to be a major driver of the behaviour in this system.
+low density of coyotes on Fogo Island and the minimal risk they pose, predation risk
+does not appear to be a major driver of swimming.
+
 <!-- QW: changed it a bit - how is this? JGH: simplified these two para's a bit,

### 2020-03-17 paper/manuscript.Rmd
Update manuscript.Rmd

@@
 exceeds the density-fitness equilibrium [@Morris1987]. While this is an ultimate
-explanation to a series of proximate observations, fine-scale density-dependent
-habitat selection is theoretically possible [@Webber2018]. Given the presumed
-low density of coyotes on Fogo Island and the minimal risk they pose, predation risk
-does not appear to be a major driver of swimming.
+explanation, our data suggest caribou only remain on smaller islands for a few
+days at a time, so the fitness equilibrium remains a theoretical construct. More
+likely, however, is that fine-scale competition, density-dependent habitat
+selection, and forage depletion drives individuals to peridiodically swim
+between islands. Given the presumed low density of coyotes on Fogo Island and
+the minimal risk they pose, predation risk does not appear to be a major driver
+of swimming.
 

### 2020-03-18 paper/manuscript.Rmd
numbers

@@
 data to the ice-free period of the year (April 1 - December 31) <!--confirm dates
---> and we identified three adult female caribou that collectively had 86
+--> and we identified 12 adult female caribou that collectively had XX
 swimming events over three years (Figure 2). In addition to our own visual and

### 2020-03-18 paper/manuscript.Rmd
edits

@@
 a peak in November. On average, caribou swam between islands approximately ever
-XX days, but some individuals swam up to XX times per week/month/whatever.
-Caribou may swim between islands in the ocean for similar reasons that drive
-caribou swimming in freshwater [@Leblond2016]. We propose that one such
-explanation, the forage limitation hypothesis, is likely driving oceanic
-swimming events [@Miller2002].
+XX days, but some individuals swam up to XX times per year (range XX-XX).
+Caribou may swim between islands in the ocean for similar reasons they swim in
+freshwater [@Leblond2016]. We propose that one such explanation, the forage
+limitation hypothesis, is likely driving oceanic swimming events [@Miller2002].
 <!--Is it better

### 2020-03-18 paper/manuscript.Rmd
Update manuscript.Rmd and manuscript.md

@@
 freshwater [@Leblond2016]. We propose that one such explanation, the forage
-limitation hypothesis, is likely driving oceanic swimming events [@Miller2002].
+limitation hypothesis, is a likely explanation for oceanic swimming for caribou that live on islands in the Fogo Island archipelago.
 <!--Is it better

### 2020-03-18 paper/manuscript.Rmd
Update manuscript.Rmd and manuscript.md
minor eds

@@
 freshwater [@Leblond2016]. We propose that one such explanation, the forage
-limitation hypothesis, is a likely explanation for oceanic swimming for caribou that live on islands in the Fogo Island archipelago.
+limitation hypothesis, is a likely explanation for oceanic swimming for caribou
+that live on islands in the Fogo Island archipelago.
 <!--Is it better
@@
 competition for resources driven by reduced foraging opportunities can decrease
-reproductive success of adult females [@Schaefer2016].Forage scarcity has been
+reproductive success of adult females [@Schaefer2016]. Forage scarcity has been
 proposed as a potential reason for caribou moving between arctic islands on the
@@
 colonizing nearby islands, but given that some are small, over-grazing by newly
-arrived caribou may rapidly deplete forage [@Bergerud1990]. Although
+arrived caribou may rapidly deplete forage on an annual cycle [@Bergerud1990]. Although
 unconfirmed, it is also possible that Fogo Island has experienced forage

### 2020-03-18 paper/manuscript.Rmd
Update manuscript.Rmd

@@
 limitation in animal habitat selection patterns remains unknown, but swimming
-between islands may in part be explained by forage limitation and conspecific
-competition.
-
-**Individual variation**
+between islands may be at least partially governed by density-dependent habitat selection.
 

### 2020-03-18 paper/manuscript.Rmd
Update manuscript.Rmd

@@
 archipelago. Only 3 of these 12 had more than XX swimming events during the
-study period, so approximately one third of our marked population swam
-irregularly and infrequently, whereas for ~10% of individuals these swimming
-events were relatively common, on average occurring once every 24 <!--confirm-->
-days during ice-free seasons. We did not detect any swimming events for the
-remaining 17 individuals; it remains unclear whether a subset of the population
-does not engage in this behaviour at all, or whether these swimming events are
-extremely rare for most caribou and we happened to detect some for those nine
-individuals with very few observed swimming events. We are also unable to draw
-conclusions about the prevalence of swimming in the ocean by male caribou: our
-initial visual observation was of an adult male, but our GPS data only included
-females.
+study period; ~30% of the marked population swam at least XX times, whereas for
+~10% of individuals these swimming events were relatively common, on average
+occurring once every XX  days during ice-free seasons. We did not
+detect any swimming events for the remaining 17 individuals. It remains unclear
+whether these individuals do not swim, or whether swimming is rare for most caribou. 
+We are also unable to draw conclusions about the prevalence of
+swimming in the ocean by male caribou: our initial visual observation was of an
+adult male, but our GPS data only included females.
 

### 2020-03-18 paper/manuscript.Rmd
delete header
Update manuscript.Rmd

@@
 
-Regardless, oceanic swimming is relatively common among a small minority of our
-marked population. Of the three collared individuals, two were collared on
+Regardless, oceanic swimming is relatively common among a minority of
+marked caribou. Of the three collared individuals that swam the most, two were collared on
 Western Indian Island and the third was collared on Fogo Island. Inter-island
@@
 perceived competition or predation more acutely than their conspecifics.
-
-**Summary and conclusion**
 
@@
 
-<!-- QW: changed it a bit - how is this? JGH: simplified these two para's a bit,
+<!-- QW: changed it a bit - how is this? 
+JGH: simplified these two para's a bit,
 took out most of the predator stuff. I'm a bit confused by the sentence above
@@
 Our observations add to the evidence that caribou can, and occasionally do, swim
-in the ocean [e.g. @Miller1995; @Miller2002; @Jeffery2007; @Ricca2012]. Taken
-together, we suggest that swimming is likely more common than previously thought
+in the ocean [e.g. @Miller1995; @Miller2002; @Jeffery2007; @Ricca2012]. 
+We suggest that swimming is likely more common than previously thought
 for caribou living on oceanic islands, and that forage limitation and the

### 2020-03-18 paper/manuscript.Rmd
r nids

@@
 observed an unmarked adult male caribou swim between Western and Eastern Indian
-Islands (Figure 1), a distance of at least 470 m which took ~9 minutes (~52 m per
-minute). This observation prompted us to investigate whether swimming behaviour
-was common among GPS radio-collared caribou in the population (n = 29 adult
-females; 1240g, GPS 4400M; Lotek Wireless Inc., Newmarket, Ontario, Canada). We
-identified swimming events as two consecutive GPS (2-hour relocation rates)
-locations from an individual occurred on different islands. Northeastern
+Islands (Figure 1), a distance of at least 470 m which took ~9 minutes (~52 m
+per minute). This observation prompted us to investigate whether swimming
+behaviour was common among GPS radio-collared caribou in the population (n = 29
+adult females; 1240g, GPS 4400M; Lotek Wireless Inc., Newmarket, Ontario,
+Canada). We identified swimming events as two consecutive GPS (2-hour relocation
+rates) locations from an individual occurred on different islands. Northeastern
 Newfoundland typically experiences pack ice during winter and caribou are known
 locally to travel between islands by walking over the ice. We restricted our GPS
-data to the ice-free period of the year (April 1 - December 31) <!--confirm dates
---> and we identified 12 adult female caribou that collectively had XX
-swimming events over three years (Figure 2). In addition to our own visual and
-remotely sensed observations, residents of Fogo Island have also reported
-observing caribou swimming between islands on numerous occasions.
+data to the ice-free period of the year (April 1 - December 31) and we
+identified  adult female caribou that collectively had `r nids` swimming events
+over three years (Figure 2). In addition to our own visual and remotely sensed
+observations, residents of Fogo Island have also reported observing caribou
+swimming between islands on numerous occasions.
 

### 2020-03-18 paper/manuscript.Rmd
Update manuscript.Rmd

@@
 data to the ice-free period of the year (April 1 - December 31) and we
-identified  adult female caribou that collectively had `r nids` swimming events
+identified  adult female caribou that collectively had ``` `r nids` swimming events
 over three years (Figure 2). In addition to our own visual and remotely sensed
@@
 <!-- add numbers in here -->
-In total, these XX swimming events occurred throughout the ice-free season, with
+In total, `r nevents` swimming events occurred throughout the ice-free season, with
 a peak in November. On average, caribou swam between islands approximately ever

### 2020-03-18 paper/manuscript.Rmd
edit wd - thanks alec for all the help it's been really great thanks for being so helpful
add number

@@
 ```{r input}
-edges <- readRDS('output/island-edges.Rds')
+edges <- readRDS('../output/island-edges.Rds')
 icefree <- edges[season == 'icefree']
@@
 neventid <- icefree[, .N, ANIMAL_ID]
+
+# Max per individual
+neventidMax <- max(neventid$N)
+
+# Mean per individual
+neventidMean <- round(mean(neventid$N))
+
+# Min per individual
+neventidMin <- min(neventid$N)
+
+# Duration of time per island
+dur <- round(icefree[, mean(islandlen)], digits = 0)
+durMin <- round(icefree[, min(islandlen)], digits = 0)
+durMax <- round(icefree[, max(islandlen)], digits = 0)
 
@@
 data to the ice-free period of the year (April 1 - December 31) and we
-identified  adult female caribou that collectively had ``` `r nids` swimming events
+identified  adult female caribou that collectively had `r nids` swimming events
 over three years (Figure 2). In addition to our own visual and remotely sensed
@@
 <!-- add numbers in here -->
-In total, `r nevents` swimming events occurred throughout the ice-free season, with
-a peak in November. On average, caribou swam between islands approximately ever
-XX days, but some individuals swam up to XX times per year (range XX-XX).
-Caribou may swim between islands in the ocean for similar reasons they swim in
-freshwater [@Leblond2016]. We propose that one such explanation, the forage
-limitation hypothesis, is a likely explanation for oceanic swimming for caribou
-that live on islands in the Fogo Island archipelago.
+In total, `r nevents` swimming events occurred throughout the ice-free season,
+with a peak in November. On average, caribou swam approximately `r neventidMean`
+(range = `r neventidMin`-`r neventidMax`) per year and remained on those islands
+for approximately `r dur` (range = `r durMin`-`r durMax`) days before swimming again. Caribou may swim between
+islands in the ocean for similar reasons they swim in freshwater [@Leblond2016].
+We propose that one such explanation, the forage limitation hypothesis, is a
+likely explanation for oceanic swimming for caribou that live on islands in the
+Fogo Island archipelago.
 <!--Is it better

### 2020-03-18 paper/manuscript.Rmd
n ids/events

@@
 data to the ice-free period of the year (April 1 - December 31) and we
-identified  adult female caribou that collectively had `r nids` swimming events
+identified `r nids` adult female caribou that collectively had `r nevents` swimming events
 over three years (Figure 2). In addition to our own visual and remotely sensed
@@
 archipelago. Only 3 of these 12 had more than XX swimming events during the
-study period; ~30% of the marked population swam at least XX times, whereas for
+study period; ~30% of the marked population swam at least once, whereas for
 ~10% of individuals these swimming events were relatively common, on average

### 2020-03-18 scripts/3-generate-edges.R
Update 3-generate-edges.R

@@ -51,6 +51,6 @@ net <- graph_from_data_frame(
                    island]
 )
 
-### Output ----
+
 saveRDS(edges, 'output/island-edges.Rds')
 saveRDS(net, 'output/island-network.Rds')

### 2020-03-18 paper/manuscript.Rmd
Update manuscript.Rmd and manuscript.md

@@
 
+# IDs > 10 swimming
+n10 <- length(icefree[, .N, ANIMAL_ID][N > 10]$N)
+
 # Duration of time per island
@@
 durMax <- round(icefree[, max(islandlen)], digits = 0)
+
+ind <- edges[ANIMAL_ID == "FO2016011" | ANIMAL_ID == "FO2017001" | ANIMAL_ID == "FO2017013"]
+indMean <- round(ind[, mean(islandlen)], digits = 0)
 
@@
 In total, 12 of 29 collared female caribou swam between islands in the Fogo
-archipelago. Only 3 of these 12 had more than XX swimming events during the
+archipelago. Only `r n10` of these 12 had more than 10 swimming events during the
 study period; ~30% of the marked population swam at least once, whereas for
 ~10% of individuals these swimming events were relatively common, on average
-occurring once every XX  days during ice-free seasons. We did not
+occurring approximately every `r indMean` days during ice-free seasons. We did not
 detect any swimming events for the remaining 17 individuals. It remains unclear

### 2020-03-18 paper/manuscript.Rmd
Update manuscript.Rmd and manuscript.md

--- 2020-03-18 paper/manuscript.Rmd
@@
 durMax <- round(icefree[, max(islandlen)], digits = 0)
+
+# Regular swimmers
+ind <- edges[ANIMAL_ID == "FO2016011" | ANIMAL_ID == "FO2017001" | ANIMAL_ID == "FO2017013"]
+indMean <- round(ind[, mean(islandlen)], digits = 0)
+
+# Irregular swimmers
+ind2 <- edges[ANIMAL_ID != "FO2016011" | ANIMAL_ID != "FO2017001" | ANIMAL_ID != "FO2017013"]
+ind2Mean <- round(ind2[, mean(islandlen)], digits = 0)
 
@@
 In total, 12 of 29 collared female caribou swam between islands in the Fogo
-archipelago. Only 3 of these 12 had more than XX swimming events during the
-study period; ~30% of the marked population swam at least once, whereas for
-~10% of individuals these swimming events were relatively common, on average
-occurring once every XX  days during ice-free seasons. We did not
-detect any swimming events for the remaining 17 individuals. It remains unclear
-whether these individuals do not swim, or whether swimming is rare for most caribou. 
-We are also unable to draw conclusions about the prevalence of
-swimming in the ocean by male caribou: our initial visual observation was of an
-adult male, but our GPS data only included females.
+archipelago. Only `r n10` of these 12 individuals had more than 10 swimming
+events, suggesting that for ~10% of narjed individuals swimming is relatively
+common, on average occurring approximately every `r indMean` days during
+ice-free seasons. We did not detect any swimming events for the remaining 17
+individuals. It remains unclear whether these individuals do not swim, or
+whether swimming is rare for most caribou. Overall, there appears to be three
+groups of individuals: those that swim often (~`r indMean` days), those that
+swim occasionally (~`r ind2Mean` days), and those that never swam. We are also
+unable to draw conclusions about the prevalence of swimming in the ocean by male
+caribou: our initial visual observation was of an adult male, but our GPS data
+only included females.
 

### 2020-03-18 paper/manuscript.Rmd
swim-bo-baggins

@@
 whether swimming is rare for most caribou. Overall, there appears to be three
-groups of individuals: those that swim often (~`r indMean` days), those that
-swim occasionally (~`r ind2Mean` days), and those that never swam. We are also
+groups of individuals: those that swim often (n = 3, every ~`r indMean` days), those that
+swim occasionally (n = 9, every ~`r ind2Mean` days), and those that never swam (n = 17). We are also
 unable to draw conclusions about the prevalence of swimming in the ocean by male

### 2020-03-18 paper/manuscript.Rmd
Update manuscript.Rmd, manuscript.docx, and manuscript.md

@@
   #    toc: false
-   #bookdown::word_document2:
-    #reference_docx: template.docx
+  # bookdown::word_document2:
+  #  reference_docx: template.docx
 bibliography: ../references/swim.bib

### 2020-03-18 scripts/4-figure1.R
leave jdates for now

@@ -25,10 +25,6 @@ edges[, (outmean) := lapply(.SD, mean),
       by = .(island, ANIMAL_ID), .SDcols = tomean]
 
 
-lowjdate <- 149#90
-caribou[, season := NULL]
-caribou[, season := ifelse(JDate > lowjdate & JDate < 365, 'icefree', 'ice')]
-
 (gnet <- ggplot(data = edges[season == 'icefree']) +
     geom_sf(data = islands, fill = '#c7c0bd') +
     geom_edges(

### 2020-03-18 scripts/4-figure1.R
todo drop

@@ -24,6 +24,7 @@ outmean <- c(x = 'meanX', y = 'meanY', xend = 'endmeanX', yend = 'endmeanY')
 edges[, (outmean) := lapply(.SD, mean), 
       by = .(island, ANIMAL_ID), .SDcols = tomean]
 
+# TODO: drop mid island
 
 (gnet <- ggplot(data = edges[season == 'icefree']) +
     geom_sf(data = islands, fill = '#c7c0bd') +
@@ -51,7 +52,7 @@ gcol <- ggplot(data = edges[, .N, by = .(JDate, ANIMAL_ID)]) +
     geom_histogram(aes(JDate, fill = ANIMAL_ID)) +
     guides(fill = FALSE) +
     scale_fill_viridis_d() +
-    geom_vline(aes(xintercept = lowjdate)) +
+    geom_vline(aes(xintercept = 90)) +
     geom_vline(aes(xintercept = 365))+ 
     labs(x = 'Julian Day', y = NULL))
 
@@ -66,7 +67,7 @@ edges[, region := ifelse(meanY < median(meanY) + 6500, 'South', 'North')]
     geom_histogram(aes(JDate, fill = ANIMAL_ID)) +
     guides(fill = FALSE) +
     scale_fill_viridis_d() +
-    geom_vline(aes(xintercept = lowjdate)) +
+    geom_vline(aes(xintercept = 90)) +
     geom_vline(aes(xintercept = 365))+ 
     labs(x = 'Julian Day', y = NULL) + 
     facet_wrap(~region))

### 2020-03-18 scripts/4-figure1.R
save png

@@ -56,8 +56,10 @@ gcol <- ggplot(data = edges[, .N, by = .(JDate, ANIMAL_ID)]) +
     geom_vline(aes(xintercept = 365))+ 
     labs(x = 'Julian Day', y = NULL))
 
+png("graphics/Fig2.png", width = 6000, height = 6000, units = "px", res = 600)
 gnet / ghist + 
   plot_layout(heights = c(3, 1))
+dev.off()
 
 ### Other figs ----
 # south/west

### 2020-03-18 paper/manuscript.Rmd
align
render doc

--- 2020-03-18 paper/manuscript.Rmd
@@
   #    toc: false
-   #bookdown::word_document2:
-    #reference_docx: template.docx
+   bookdown::word_document2:
+    reference_docx: template.docx
 bibliography: ../references/swim.bib
@@
 (range = `r neventidMin`-`r neventidMax`) per year and remained on those islands
-for approximately `r dur` (range = `r durMin`-`r durMax`) days before swimming again. Caribou may swim between
-islands in the ocean for similar reasons they swim in freshwater [@Leblond2016].
-We propose that one such explanation, the forage limitation hypothesis, is a
-likely explanation for oceanic swimming for caribou that live on islands in the
-Fogo Island archipelago.
+for approximately `r dur` (range = `r durMin`-`r durMax`) days before swimming
+again. Caribou may swim between islands in the ocean for similar reasons they
+swim in freshwater [@Leblond2016]. We propose that one such explanation, the
+forage limitation hypothesis, is a likely explanation for oceanic swimming for
+caribou that live on islands in the Fogo Island archipelago.
 <!--Is it better
@@
 In total, 12 of 29 collared female caribou swam between islands in the Fogo
-archipelago. Only `r n10` of these 12 individuals had more than 10 swimming
+Island archipelago. Only `r n10` of these 12 individuals had more than 10 swimming
 events, suggesting that for ~10% of narjed individuals swimming is relatively
@@
 
-<!--![**Figure 2**: Map of the Fogo Island archipelago. Fogo Island is the largest
+![**Figure 2**: Map of the Fogo Island archipelago. Fogo Island is the largest
 island in the archipelago (237.71 km^2^), but there are at least three other
@@
 (*Betula papyrifera*) as well as bogs, lakes, lichen and rocky
-barrens.]()<!--(../graphics/IMG_1562.JPG)-->
+barrens.](../graphics/Fig2.png)
+<!-- edit caption -->
 

### 2020-03-19 paper/manuscript.Rmd
few minor edits to sentences

@@
 running, <!--[@Fish1993]--> and drowning is also possible [@Miller1985]. Despite
-a large number of coastal and island caribou herds, only a handful of examples
+the abundance of coastal and island caribou herds, only a handful of examples
 have documented caribou swimming in the ocean [e.g. @Miller1995; @Miller2002;
@@
 adult females; 1240g, GPS 4400M; Lotek Wireless Inc., Newmarket, Ontario,
-Canada). We identified swimming events as two consecutive GPS (2-hour relocation
-rates) locations from an individual occurred on different islands. Northeastern
-Newfoundland typically experiences pack ice during winter and caribou are known
-locally to travel between islands by walking over the ice. We restricted our GPS
-data to the ice-free period of the year (April 1 - December 31) and we
-identified `r nids` adult female caribou that collectively had `r nevents` swimming events
-over three years (Figure 2). In addition to our own visual and remotely sensed
-observations, residents of Fogo Island have also reported observing caribou
-swimming between islands on numerous occasions.
+Canada). We identified swimming events as two consecutive GPS locations (2-hour
+relocation rates)  from an individual occurred on different islands.
+Northeastern Newfoundland typically experiences pack ice during winter and
+caribou are known locally to travel between islands by walking over the ice. We
+restricted our GPS data to the ice-free period of the year (April 1 - December
+31) and we identified `r nids` adult female caribou that collectively had `r
+nevents` swimming events over three years (Figure 2). In addition to our own
+visual and remotely sensed observations, residents of Fogo Island have also
+reported observing caribou swimming between islands on numerous occasions.
 
@@
 with a peak in November. On average, caribou swam approximately `r neventidMean`
-(range = `r neventidMin`-`r neventidMax`) per year and remained on those islands
-for approximately `r dur` (range = `r durMin`-`r durMax`) days before swimming
-again. Caribou may swim between islands in the ocean for similar reasons they
-swim in freshwater [@Leblond2016]. We propose that one such explanation, the
-forage limitation hypothesis, is a likely explanation for oceanic swimming for
-caribou that live on islands in the Fogo Island archipelago.
+(range = `r neventidMin`-`r neventidMax`) times per year and remained on each
+island for approximately `r dur` (range = `r durMin`-`r durMax`) days before
+swimming again. Caribou may swim between islands in the ocean for similar
+reasons they swim in freshwater [@Leblond2016]. We propose that one such
+explanation, the forage limitation hypothesis, is a likely explanation for
+oceanic swimming for caribou that live on islands in the Fogo Island
+archipelago.
 <!--Is it better
@@
 
-Forage limitation and over-grazing is a major concern for caribou populations
-[@Schaefer2016] and could explain movement between islands. Conspecific
-competition for resources driven by reduced foraging opportunities can decrease
-reproductive success of adult females [@Schaefer2016]. Forage scarcity has been
-proposed as a potential reason for caribou moving between arctic islands on the
-sea-ice in winter [@Miller1977]. Caribou were introduced to Fogo Island (n = 26
-animals between 1964–67) and Change Island (n = 5 animals in 1964) from
-Newfoundland as part of a series of translocations and introductions throughout
-the province (Bergerud & Mercer 1989). It is unknown when caribou began
-colonizing nearby islands, but given that some are small, over-grazing by newly
-arrived caribou may rapidly deplete forage on an annual cycle [@Bergerud1990]. Although
-unconfirmed, it is also possible that Fogo Island has experienced forage
-depletion since the introduction of caribou in the 1960s. More likely, however,
-is fine-scale competition for foraging resources that drives caribou to swim to
-new islands.
+Forage limitation and over-grazing is a major concern for caribou populations:
+it can reduce female reproductive success [@Schaefer2016] and thus could lead to
+movement between islands. Forage scarcity has been proposed as a potential
+reason for caribou moving between arctic islands on the sea-ice in winter
+[@Miller1977]. Caribou were introduced to Fogo Island (n = 26 animals between
+1964–67) and Change Island (n = 5 animals in 1964) from Newfoundland as part of
+a series of translocations and introductions throughout the province (Bergerud &
+Mercer 1989). It is unknown when caribou began colonizing nearby islands, but
+given that some are small, over-grazing by newly arrived caribou may rapidly
+deplete forage on an annual cycle [@Bergerud1990]. Although unconfirmed, it is
+also possible that Fogo Island has experienced forage depletion since the
+introduction of caribou in the 1960s. More likely, however, it is fine-scale
+competition for foraging resources that drives caribou to swim to new islands.
 
@@
 Island archipelago. Only `r n10` of these 12 individuals had more than 10 swimming
-events, suggesting that for ~10% of narjed individuals swimming is relatively
+events, suggesting that for ~10% of collared individuals swimming is relatively
 common, on average occurring approximately every `r indMean` days during

### 2020-03-20 scripts/4-figure1.R
edit themes

@@ -26,6 +26,18 @@ edges[, (outmean) := lapply(.SD, mean),
 
 # TODO: drop mid island
 
+themeMap <- theme(panel.border = element_rect(size = 1, fill = NA),
+                  panel.background = element_rect(fill = "white"), #"#e3ebf9"),
+                  panel.grid = element_line(color = "black", size = 0.2),
+                  axis.text = element_text(size = 12))
+
+themeHist <- theme(panel.border = element_rect(size = 1, fill = NA),
+                  panel.background = element_rect(fill = "white"), #"#e3ebf9"),
+                  panel.grid = element_line(color = "black", size = 0.2),
+                  axis.text = element_text(size = 12))
+
+
+
 (gnet <- ggplot(data = edges[season == 'icefree']) +
     geom_sf(data = islands, fill = '#c7c0bd') +
     geom_edges(
@@ -41,7 +53,8 @@ edges[, (outmean) := lapply(.SD, mean),
     # geom_point(aes(median(meanX), median(meanY) + 6500), size = 4) + 
     guides(color = FALSE) +
     scale_color_viridis_d() + 
-    labs(x = NULL, y = NULL))
+    labs(x = NULL, y = NULL) + 
+    themeMap)
 
 gcol <- ggplot(data = edges[, .N, by = .(JDate, ANIMAL_ID)]) + 
   geom_col(aes(JDate, N, color = ANIMAL_ID)) + 
@@ -54,7 +67,8 @@ gcol <- ggplot(data = edges[, .N, by = .(JDate, ANIMAL_ID)]) +
     scale_fill_viridis_d() +
     geom_vline(aes(xintercept = 90)) +
     geom_vline(aes(xintercept = 365))+ 
-    labs(x = 'Julian Day', y = NULL))
+    labs(x = 'Julian Day', y = 'Frequency') + 
+    themeHist)
 
 png("graphics/Fig2.png", width = 6000, height = 6000, units = "px", res = 600)
 gnet / ghist + 

### 2020-03-20 scripts/4-figure1.R
axis size

@@ -29,12 +29,12 @@ edges[, (outmean) := lapply(.SD, mean),
 themeMap <- theme(panel.border = element_rect(size = 1, fill = NA),
                   panel.background = element_rect(fill = "white"), #"#e3ebf9"),
                   panel.grid = element_line(color = "black", size = 0.2),
-                  axis.text = element_text(size = 12))
+                  axis.text = element_text(size = 14, color = "black"))
 
 themeHist <- theme(panel.border = element_rect(size = 1, fill = NA),
                   panel.background = element_rect(fill = "white"), #"#e3ebf9"),
-                  panel.grid = element_line(color = "black", size = 0.2),
-                  axis.text = element_text(size = 12))
+                  axis.text = element_text(size = 12, color = "black"),
+                  axis.title = element_text(size = 14, color = "black"))
 
 
 
@@ -50,6 +50,7 @@ themeHist <- theme(panel.border = element_rect(size = 1, fill = NA),
       ),
       size = 2
     ) +
+    ggtitle('A)') +
     # geom_point(aes(median(meanX), median(meanY) + 6500), size = 4) + 
     guides(color = FALSE) +
     scale_color_viridis_d() + 
@@ -64,10 +65,11 @@ gcol <- ggplot(data = edges[, .N, by = .(JDate, ANIMAL_ID)]) +
 (ghist <- ggplot(data = edges) +
     geom_histogram(aes(JDate, fill = ANIMAL_ID)) +
     guides(fill = FALSE) +
+    ggtitle('B)') +
     scale_fill_viridis_d() +
     geom_vline(aes(xintercept = 90)) +
     geom_vline(aes(xintercept = 365))+ 
-    labs(x = 'Julian Day', y = 'Frequency') + 
+    labs(x = 'Julian Day', y = NULL) + 
     themeHist)
 
 png("graphics/Fig2.png", width = 6000, height = 6000, units = "px", res = 600)

### 2020-03-20 paper/manuscript.Rmd
Update manuscript.Rmd and manuscript.docx

@@
 
-![**Figure 2**: Map of the Fogo Island archipelago. Fogo Island is the largest
+![**Figure 2**: A) Map of the Fogo Island archipelago with swimming events between islands.
+B) Histogram displaying the distribution of swimming events throughout the year.
+Note, colours correspond to individual caribou.](../graphics/Fig2.png)
+
+<!-- edit caption Fogo Island is the largest
 island in the archipelago (237.71 km^2^), but there are at least three other
@@
 (*Betula papyrifera*) as well as bogs, lakes, lichen and rocky
-barrens.](../graphics/Fig2.png)
-<!-- edit caption -->
+barrens.-->
 

### 2020-03-20 paper/manuscript.Rmd
Update manuscript.Rmd and manuscript.md

@@
   #    toc: false
-   bookdown::word_document2:
-    reference_docx: template.docx
+  # bookdown::word_document2:
+  #  reference_docx: template.docx
 bibliography: ../references/swim.bib

### 2020-03-20 paper/manuscript.Rmd
r nevents
minor edits

@@
 restricted our GPS data to the ice-free period of the year (April 1 - December
-31) and we identified `r nids` adult female caribou that collectively had `r
-nevents` swimming events over three years (Figure 2). In addition to our own
+31) and we identified `r nids` adult female caribou that collectively had `r nevents` 
+swimming events over three years (Figure 2). In addition to our own
 visual and remotely sensed observations, residents of Fogo Island have also
@@
 also possible that Fogo Island has experienced forage depletion since the
-introduction of caribou in the 1960s. More likely, however, it is fine-scale
-competition for foraging resources that drives caribou to swim to new islands.
+introduction of caribou in the 1960s. More likely, however, is that fine-scale
+competition for foraging resources drives caribou to swim to new islands.
 
@@
 between patches. For caribou in the Fogo Island archipelago, the role of forage
-limitation in animal habitat selection patterns remains unknown, but swimming
+limitation in habitat selection patterns remains unknown, but swimming
 between islands may be at least partially governed by density-dependent habitat selection.
@@
 Island archipelago. Only `r n10` of these 12 individuals had more than 10 swimming
-events, suggesting that for ~10% of collared individuals swimming is relatively
-common, on average occurring approximately every `r indMean` days during
+events, suggesting that for ~10% of collared individuals swim relatively
+frequently, on average occurring approximately every `r indMean` days during
 ice-free seasons. We did not detect any swimming events for the remaining 17
 individuals. It remains unclear whether these individuals do not swim, or
-whether swimming is rare for most caribou. Overall, there appears to be three
+whether swimming is rare, and thus undetected, for the majority of individuals Overall, there appears to be three
 groups of individuals: those that swim often (n = 3, every ~`r indMean` days), those that

### 2020-03-20 Untitled.R
ditch random script

@@ -1,4 +1,0 @@
-
-
-### Packages ----
-library(knitcitations)

### 2020-03-20 paper/manuscript.Rmd
minor edits

@@
 In total, 12 of 29 collared female caribou swam between islands in the Fogo
-Island archipelago. Only `r n10` of these 12 individuals had more than 10 swimming
-events, suggesting that for ~10% of collared individuals swim relatively
-frequently, on average occurring approximately every `r indMean` days during
-ice-free seasons. We did not detect any swimming events for the remaining 17
-individuals. It remains unclear whether these individuals do not swim, or
-whether swimming is rare, and thus undetected, for the majority of individuals Overall, there appears to be three
-groups of individuals: those that swim often (n = 3, every ~`r indMean` days), those that
-swim occasionally (n = 9, every ~`r ind2Mean` days), and those that never swam (n = 17). We are also
-unable to draw conclusions about the prevalence of swimming in the ocean by male
-caribou: our initial visual observation was of an adult male, but our GPS data
-only included females.
+Island archipelago. Only `r n10` of these 12 individuals had more than 10
+swimming events, suggesting that for ~10% of collared individuals swim
+relatively frequently, on average occurring approximately every `r indMean` days
+during ice-free seasons. We did not detect any swimming events for the remaining
+17 individuals. It remains unclear whether these individuals do not swim, or
+whether swimming is rare, and thus undetected, for the majority of individuals
+Overall, there appears to be three groups of individuals: those that swim often
+(n = 3, every ~`r indMean` days), those that swim occasionally (n = 9, every ~`r
+ind2Mean` days), and those that never swam (n = 17). We are also unable to draw
+conclusions about the prevalence of swimming in the ocean by male caribou: our
+initial visual observation was of an adult male, but our GPS data only included
+females.
 
@@
 
-Regardless, oceanic swimming is relatively common among a minority of
+Oceanic swimming is relatively common among a minority of
 marked caribou. Of the three collared individuals that swam the most, two were collared on
@@
 around them, are shorter than those between Fogo Island and others (Figure 2).
-Individuals that were born on or spend more time on these smaller islands may
-swim between them more frequently than animals that spend most of their time on
+It is possible that caribou display natal philopatry [@Larue2019] to some of the smaller islands and thus are more likely to
+swim between them more frequently than animals that were born, ospend most of their time, on
 Fogo Island. The forage limitation hypothesis is generally thought to apply

### 2020-03-20 paper/manuscript.Rmd
Update manuscript.Rmd and manuscript.md

@@
 Overall, there appears to be three groups of individuals: those that swim often
-(n = 3, every ~`r indMean` days), those that swim occasionally (n = 9, every ~`r
-ind2Mean` days), and those that never swam (n = 17). We are also unable to draw
+(n = 3, every ~`r indMean` days), those that swim occasionally (n = 9, every ~
+`r ind2Mean` days), and those that never swam (n = 17). We are also unable to draw
 conclusions about the prevalence of swimming in the ocean by male caribou: our
@@
 
-Oceanic swimming is relatively common among a minority of
-marked caribou. Of the three collared individuals that swam the most, two were collared on
-Western Indian Island and the third was collared on Fogo Island. Inter-island
-distances between Western and Eastern Indian Islands, and the smaller islands
-around them, are shorter than those between Fogo Island and others (Figure 2).
-It is possible that caribou display natal philopatry [@Larue2019] to some of the smaller islands and thus are more likely to
-swim between them more frequently than animals that were born, ospend most of their time, on
-Fogo Island. The forage limitation hypothesis is generally thought to apply
-uniformly across populations, but individual caribou could evaluate the costs
-and benefits of swimming to another island differently. It is possible that some
-individuals, in our case ~10%, considered swimming to be less costly or
-perceived competition or predation more acutely than their conspecifics.
+Oceanic swimming is relatively common among a minority of marked caribou. Of the
+three collared individuals that swam the most, two were collared on Western
+Indian Island and the third was collared on Fogo Island. Inter-island distances
+between Western and Eastern Indian Islands, and the smaller islands around them,
+are shorter than those between Fogo Island and others (Figure 2). It is possible
+that caribou display natal philopatry [e.g. @Larue2018] to some of the smaller
+islands and thus are more likely to swim between them more frequently than
+animals that were born, ospend most of their time, on Fogo Island. The forage
+limitation hypothesis is generally thought to apply uniformly across
+populations, but individual caribou could evaluate the costs and benefits of
+swimming to another island differently. It is possible that some individuals, in
+our case ~10%, considered swimming to be less costly or perceived competition or
+predation more acutely than their conspecifics.
 
@@
 density reached 300 animals and it is unknown when caribou began to occupy other
-islands, anecdotal evidence suggests it was during the 1990s. We surmise that as
+islands, although anecdotal evidence suggests it was during the 1990s. We surmise that as
 competition among conspecifics increased along with population size,

### 2020-03-20 paper/citations.R
delete old citations script

@@ -1,62 +1,0 @@
-### Packages ----
-library(knitcitations)
-
-## only 15 citations
-
-## Bastille-Rousseau et al 2016
-#citep('10.1139/cjz-2015-0127')
-
-## Bergerud 1985
-#citep('10.1139/z85-199')
-
-## Bergerud and Page 1987
-#citep('10.1139/z87-249')
-
-## Bergerud and Mercer 1989 - get actual doi
-
-## Bergerud 1990
-#citep('10.1016/S0003-3472(05)80882-6')
-
-## Bradbury et al 2015
-#citep('10.1093/beheco/arv078')
-
-## Fish 1993
-#citep('10.1071/ZO9940079')
-
-## Jeffrey et al 2007
-#citep('10.7557/2.27.4.320')
-
-## Leblond et al 2016
-#citep('10.1186/s40462-016-0079-4')
-
-## Miller et al 1977
-#citep('10.1139/z77-131')
-
-## Miller and Gunn 1986 
-#citep('10.14430/arctic2052')
-
-## Miller 1995 -- get actual doi
-# yes
-## Miller 2002 -- get actual doi
-#yes
-## Morris 1987
-#citep('10.2307/2937087')
-
-## Schaefer 2016
-#citep('10.1093/jmammal/gyv184')
-
-## Rayl et al 2014
-#citep('10.1644/13-MAMM-A-133.1')
-
-## Ricca et al 2012
-#citep('10.1007/s10530-012-0195-z')
-
-## Webber and Vander Wal 2018
-#citep('10.1111/1365-2656.12773')
-
-## van Beest et al 2014
-citep('10.1111/1365-2656.12115')
-
-### Write out bib ----
-write.bibtex(file = "references/knitcite.bib")
-

### 2020-03-20 paper/manuscript.Rmd
ecology.csl in YAML

@@
 bibliography: ../references/swim.bib
-#csl: csl.csl
+csl: ecology.csl
 ---

### 2020-04-02 paper/manuscript.Rmd
minor edits

--- 2020-03-20 paper/manuscript.Rmd
@@
   #    toc: false
-  # bookdown::word_document2:
-  #  reference_docx: template.docx
+   bookdown::word_document2:
+    reference_docx: template.docx
 bibliography: ../references/swim.bib
-#csl: csl.csl
+csl: ecology.csl
 ---
@@
 islands and thus are more likely to swim between them more frequently than
-animals that were born, ospend most of their time, on Fogo Island. The forage
+animals that were born, or spend most of their time, on Fogo Island. The forage
 limitation hypothesis is generally thought to apply uniformly across
@@
 Labrador Wildlife Division, unpublished data]. During the 1990s, population
-density reached 300 animals and it is unknown when caribou began to occupy other
+density reached ~300 animals and it is unknown when caribou began to occupy other
 islands, although anecdotal evidence suggests it was during the 1990s. We surmise that as

### 2020-04-02 paper/manuscript.Rmd
fig2 caption
rm intro
rm comment

@@
 
-**Introduction**
-
 Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
@@
 
-<!-- add numbers in here -->
 In total, `r nevents` swimming events occurred throughout the ice-free season,
@@
 B) Histogram displaying the distribution of swimming events throughout the year.
-Note, colours correspond to individual caribou.](../graphics/Fig2.png)
+Note, colours correspond to individual caribou.
+C) Inset of swimming events between small islands on the northern coast of Fogo island.
+D) Inset of swimming events between islands on the southern coast of Fogo island.
+](../graphics/Fig2.png)
 

### 2020-04-02 paper/manuscript.Rmd
add wg acknowledge
small text edits

@@
 conclusions about the prevalence of swimming in the ocean by male caribou: our
-initial visual observation was of an adult male, but our GPS data only included
+initial visual observation was of an adult male, but our GPS data only included adult
 females.
@@
 likely, however, is that fine-scale competition, density-dependent habitat
-selection, and forage depletion drives individuals to peridiodically swim
+selection, and forage depletion drives individuals to periodically swim
 between islands. Given the presumed low density of coyotes on Fogo Island and
@@
 **Acknowledgements**
-
 We thank M. Laforge, M. Bonar, C. Hart, and R. Huang for help in the field.
-Logistical support was provided by L. Bixby. We also thank members of the
-Newfoundland Wildlife Division including S. Moores, B. Adams, C. Doucet, W.
-Barney, and J. Neville for logistical support in the field as well as help with
-data data collection and management. We thank T. Bergerud and S. Mahoney for
-their vision in initiating much of the work on caribou in Newfoundland. Funding
-for this study was provided by the National Sciences and Engineering Research
-Council (QMRW, JGH, EVW). We respectfully acknowledge the territory in which
-data were collected and analyzed as the ancestral homelands of the Beothuk, and
-the island of Newfoundland as the ancestral homelands of the Mi’kmaq and
-Beothuk.
+Logistical support was provided by L. Bixby. We thank all members of the
+Wildlife Evolutionary Ecology Lab, including C. Hart, C. Prokopenko, J. Kennah,
+J.W. Turner, and S. Boyle for their comments on previous versions of this
+manuscript. Newfoundland Wildlife Division including S. Moores, B. Adams, C.
+Doucet, W. Barney, and J. Neville for logistical support in the field as well as
+help with data data collection and management. We thank T. Bergerud and S.
+Mahoney for their vision in initiating much of the work on caribou in
+Newfoundland. Funding for this study was provided by the National Sciences and
+Engineering Research Council (QMRW, JGH, EVW). We respectfully acknowledge the
+territory in which data were collected and analyzed as the ancestral homelands
+of the Beothuk, and the island of Newfoundland as the ancestral homelands of the
+Mi’kmaq and Beothuk.
 <!-- Simplified the funding statement, do we need to name the specific

### 2020-04-02 paper/manuscript.Rmd
adjust to use text ref, fig1

@@
 
-\captionsetup[figure]{labelformat=empty}
+(ref:fig1) Image of swimming unmarked adult male caribou (*Rangifer tarandus*) from Western to Eastern Indian Island taken on 30 May 2017.
 
-![**Figure 1**: Image of swimming unmarked adult male caribou (*Rangifer tarandus*) from Western to Eastern Indian Island taken on 30 May 2017.](../graphics/IMG_1562.JPG)
+```{r, fig.cap = '(ref:fig1)'}
+knitr::include_graphics('../graphics/IMG_1562.JPG')
+```
 

### 2020-04-02 paper/manuscript.Rmd
adjust to use text ref, fig2
stretch width fig 1 in doc

@@
 
-```{r, fig.cap = '(ref:fig1)'}
-knitr::include_graphics('../graphics/IMG_1562.JPG')
+```{r, fig.cap = '(ref:fig1)', fig.width=4.5}
+knitr::include_graphics('../graphics/IMG_1562_crop.JPG')
 ```
@@
 
-\captionsetup[figure]{labelformat=empty}
-
-![**Figure 2**: A) Map of the Fogo Island archipelago with swimming events between islands.
+(ref:fig2) A) Map of the Fogo Island archipelago with swimming events between islands.
 B) Histogram displaying the distribution of swimming events throughout the year.
@@
 D) Inset of swimming events between islands on the southern coast of Fogo island.
-](../graphics/Fig2.png)
+
+```{r, fig.cap = '(ref:fig2)'}
+knitr::include_graphics('../graphics/Fig2.png')
+```
 

### 2020-04-02 paper/manuscript.Rmd
stretch width fig 2 in doc

@@
 
-```{r, fig.cap = '(ref:fig1)', fig.width=4.5}
+```{r, fig.cap = '(ref:fig1)', fig.width=4}
 knitr::include_graphics('../graphics/IMG_1562_crop.JPG')
@@
 
-```{r, fig.cap = '(ref:fig2)'}
+```{r, fig.cap = '(ref:fig2)', fig.width=4}
 knitr::include_graphics('../graphics/Fig2.png')

### 2020-04-02 paper/manuscript.Rmd
photograph

@@
 
-(ref:fig1) Image of swimming unmarked adult male caribou (*Rangifer tarandus*) from Western to Eastern Indian Island taken on 30 May 2017.
+(ref:fig1) Photograph of swimming unmarked adult male caribou (*Rangifer tarandus*) from Western to Eastern Indian Island taken on 30 May 2017.
 
 ```{r, fig.cap = '(ref:fig1)', fig.width=4}
-knitr::include_graphics('../graphics/IMG_1562_crop.JPG')
+knitr::include_graphics('../graphics/IMG_1562_crop_edit.jpg')
 ```

### 2020-04-02 paper/manuscript.Rmd
render docx

@@
    # github_document
-  # bookdown::pdf_document2:
-  #    number_sections: true
-  #    toc: false
    bookdown::word_document2:

### 2020-04-02 scripts/2-extract-island-locs.R
tidy scripts

@@ -29,36 +29,18 @@ coords <- c('EASTING', 'NORTHING')
 
 caribou[, (coords) := as.data.table(project(cbind(X_COORD, Y_COORD), utm$proj4string))]
 
-# Sub by bounding box
-# caribou <- caribou[EASTING %between% c(690000, 800000) &
-#                      NORTHING %between% c(5470000, 5520000)]
 
 # Sub by date 
 caribou[, season := ifelse(JDate > 90 & JDate < 365, 'icefree', 'ice')]
 
-# Sub by animals that swam
-# TODO: why explicitly selecting?
-# caribou <- caribou[ANIMAL_ID == "FO2016011" |
-#                       ANIMAL_ID == "FO2017001" |
-#                       ANIMAL_ID == "FO2017013"]
-
 
 ### Extract islands ----
-# Extract points on different islands
-# caribou[, island := 
-#   st_join(
-#     st_as_sf(.SD, coords = coords, crs = utm),
-#     st_buffer(islands, 25), 
-#     join = st_intersects)$id, 
-#   .SDcols = coords]
-
 caribou[, island := 
           st_nearest_feature(
             st_as_sf(.SD, coords = coords, crs = utm),
             islands),
         .SDcols = coords]
 
-# 
 # Count locs by island
 caribou[, .N, island]
 

### 2020-04-02 scripts/3-generate-edges.R

@@ -25,9 +25,6 @@ edges[, edgeID := .I]
 
 edges <- edges[!(between(i, 946, 947, incbounds = TRUE) & ANIMAL_ID == 'FO2017012') &
                  !(between(i, 3771, 3772, incbounds = TRUE) & ANIMAL_ID == 'FO2017013')]
-# NORTHING < 5497000 &
-# endislandNORTH < 5497000 &
-#ANIMAL_ID != 'FO2016001']
 
 
 ### Variables ----

### 2020-04-02 scripts/4-figure1.R

@@ -22,7 +22,6 @@ utm <- st_crs('+proj=utm +zone=21 ellps=WGS84')
 
 
 ### Drop erroneous edges ----
-# TODO: move elsewhere?
 edges <- edges[!i %in% c(59, 11156, 4254, 4859, 11157)]
 
 ### Figure 1 ----
@@ -114,7 +113,6 @@ edgesize <- 1.75
   theme(axis.ticks = element_blank())
   )
 
-# TODO: split scales N/S
 (gnetS <- gfogo +
     geom_edges(data = S,
                aes(
@@ -154,7 +152,6 @@ edgesize <- 1.75
 )
     
  
-# TODO: zoom out Fogo for NL context?
 layout <- "AAACCCCDDDD
            AAACCCCDDDD
            BBBCCCCDDDD"

### 2020-04-02 paper/manuscript.Rmd
add R citation

@@
 adult females; 1240g, GPS 4400M; Lotek Wireless Inc., Newmarket, Ontario,
-Canada). We identified swimming events as two consecutive GPS locations (2-hour
+Canada). Using `r substr(version$version.string, 1, 15)` [@R], we identified swimming events as two consecutive GPS locations (2-hour
 relocation rates)  from an individual occurred on different islands.

### 2020-04-03 paper/manuscript.Rmd
del Fish ref

@@
 of swimming for quadrupedal mammals is significantly higher than walking or
-running, <!--[@Fish1993]--> and drowning is also possible [@Miller1985]. Despite
+running, and drowning is also possible [@Miller1985]. Despite
 the abundance of coastal and island caribou herds, only a handful of examples

### 2020-04-03 paper/manuscript.Rmd
peaks in may/nov
Update manuscript.Rmd

@@
 In total, `r nevents` swimming events occurred throughout the ice-free season,
-with a peak in November. On average, caribou swam approximately `r neventidMean`
+with peaks in May and November. On average, caribou swam approximately `r neventidMean`
 (range = `r neventidMin`-`r neventidMax`) times per year and remained on each
@@
 17 individuals. It remains unclear whether these individuals do not swim, or
-whether swimming is rare, and thus undetected, for the majority of individuals
-Overall, there appears to be three groups of individuals: those that swim often
-(n = 3, every ~`r indMean` days), those that swim occasionally (n = 9, every ~
-`r ind2Mean` days), and those that never swam (n = 17). We are also unable to draw
+whether swimming is rare, and thus undetected, for the majority of individuals.
+Overall, there appears to be three groups of individuals: those that swam often
+(n = 3, every ~`r indMean` days), those that swam occasionally (n = 9, every ~
+`r ind2Mean` days), and those that never swam (n = 17). It is therefore possible that individuals vary in their tendency to swim. We are also unable to draw
 conclusions about the prevalence of swimming in the ocean by male caribou: our

### 2020-04-03 paper/manuscript.Rmd
del old comments

@@
 (n = 3, every ~`r indMean` days), those that swam occasionally (n = 9, every ~
-`r ind2Mean` days), and those that never swam (n = 17). It is therefore possible that individuals vary in their tendency to swim. We are also unable to draw
+`r ind2Mean` days), and those that never swam (n = 17). It is therefore possible
+that individuals vary in their tendency to swim. We are also unable to draw
 conclusions about the prevalence of swimming in the ocean by male caribou: our
-initial visual observation was of an adult male, but our GPS data only included adult
-females.
-
-<!-- "During the winter, when individuals could theoretically move between islands more easily by walking across ice, XX individuals were observed moving between islands.
-YY [(a) majority, (b) minority] of these were not detected swimming during the ice-free season and only moved between islands overwinter.
-This suggests that [(a) moving between islands is a relatively consistent behaviour within individuals, where some caribou refrain from both swimming and walking over ice] / [(b) caribou who avoided swimming between islands would still take advantage of the benefits provided by these patches, but only when travelling to them was less risky and costly] -->
-
-<!-- QW: I'm not sure we need to say the below.. 
-JGH - yeah, it's pretty extra I guess. I'm not really a fan of the above para either, it's not very convincing and doesn't add much
-It is theoretically possible that these other
-females moved rapidly back and forth between islands within the 2 hour fix rate
-of our GPS collars, which we would be unable to detect. We consider this highly
-unlikely given the range of residency times for caribou on any given island (min
-- max: XX - XX). <!--double check-->
-<!-- What is the typical residency time? Is this a true statement or am I just making this up? QW: this is something we will look into, I think. -->
+initial visual observation was of an adult male, but our GPS data only included
+adult females.
 

### 2020-04-03 paper/manuscript.Rmd
del more old comments

@@
 
-<!-- QW: changed it a bit - how is this? 
-JGH: simplified these two para's a bit,
-took out most of the predator stuff. I'm a bit confused by the sentence above
-citing your 2018 paper, I'm not really sure what it's saying & how it relates to
-the rest of the section. there were comments to that effect as well, maybe
-expand the end a bit to make it clearer how that relates? -->
-
 Our observations add to the evidence that caribou can, and occasionally do, swim

### 2020-04-03 paper/manuscript.Rmd
del old comments

@@
 archipelago.
-<!--Is it better
-to just state our results here, or give this sort of teaser? Initially I had "86
-events by 3 individuals" here, but then we ignore that for a couple paragraphs
-of context before getting back into it... which seemed weird
-QW: I agree. I've added some Fogo context-->
 
@@
 Mi’kmaq and Beothuk.
-<!-- Simplified the funding statement, do we need to name the specific
-scholarships/grants? Eric opinion??-->
 

### 2020-04-03 scripts/4-figure1.R
add observed loc

@@ -124,6 +124,9 @@ edgesize <- 1.75
                ),
                size = edgesize
     ) +
+    geom_point(aes(east, north), size = 4, fill = '#8a8fd4', shape = 23,
+               data = data.table(east = 699057.51, 
+                                 north = 5491849.29)) + 
     ylim(Sbox[['ymin']], Sbox[['ymax']]) +
     xlim(Sbox[['xmin']], Sbox[['xmax']]) +
   guides(color = FALSE) +

### 2020-04-03 scripts/4-figure1.R
regen fig2

@@ -21,7 +21,7 @@ net <- readRDS('output/island-network.Rds')
 utm <- st_crs('+proj=utm +zone=21 ellps=WGS84')
 
 
-### Drop erroneous edges ----
+### Drop some edges ----
 edges <- edges[!i %in% c(59, 11156, 4254, 4859, 11157)]
 
 ### Figure 1 ----
@@ -126,7 +126,7 @@ edgesize <- 1.75
     ) +
     geom_point(aes(east, north), size = 4, fill = '#8a8fd4', shape = 23,
                data = data.table(east = 699057.51, 
-                                 north = 5491849.29)) + 
+                                 north = 5491600.29)) + 
     ylim(Sbox[['ymin']], Sbox[['ymax']]) +
     xlim(Sbox[['xmin']], Sbox[['xmax']]) +
   guides(color = FALSE) +

### 2020-04-03 paper/manuscript.Rmd
reflow
fig2 caption
minor edits

@@
 observed an unmarked adult male caribou swim between Western and Eastern Indian
-Islands (Figure 1), a distance of at least 470 m which took ~9 minutes (~52 m
+Islands (Figure 1, see Figure 2 for location), a distance of at least 470 m which took ~9 minutes (~52 m
 per minute). This observation prompted us to investigate whether swimming
@@
 adult females; 1240g, GPS 4400M; Lotek Wireless Inc., Newmarket, Ontario,
-Canada). Using `r substr(version$version.string, 1, 15)` [@R], we identified swimming events as two consecutive GPS locations (2-hour
-relocation rates)  from an individual occurred on different islands.
-Northeastern Newfoundland typically experiences pack ice during winter and
-caribou are known locally to travel between islands by walking over the ice. We
-restricted our GPS data to the ice-free period of the year (April 1 - December
-31) and we identified `r nids` adult female caribou that collectively had `r nevents` 
-swimming events over three years (Figure 2). In addition to our own
-visual and remotely sensed observations, residents of Fogo Island have also
-reported observing caribou swimming between islands on numerous occasions.
+Canada). Using `r substr(version$version.string, 1, 15)` [@R], we identified
+swimming events as two consecutive GPS locations (2-hour relocation rates)  from
+an individual occurred on different islands. Northeastern Newfoundland typically
+experiences pack ice during winter and caribou are known locally to travel
+between islands by walking over the ice. We restricted our GPS data to the
+ice-free period of the year (April 1 to December 31) and we identified `r nids`
+adult female caribou that collectively had `r nevents` swimming events over
+three years (Figure 2). In addition to our own visual and remotely sensed
+observations, residents of Fogo Island have also reported observing caribou
+swimming between islands on numerous occasions.
 
@@
 D) Inset of swimming events between islands on the southern coast of Fogo island.
+Note, the grey diamond represents the location we observed an adult male caribou swimming (see Figure 1).
 

### 2020-05-11 paper/manuscript.Rmd
move new P1
two small isl

@@
 
-Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
-that caribou swim in streams, rivers, and lakes during migration [@Leblond2016],
-to avoid predators [@Bergerud1985], and access islands during calving
-[@Bergerud1990]. Even for adept swimmers like caribou, the energetic expenditure
-of swimming for quadrupedal mammals is significantly higher than walking or
-running, and drowning is also possible [@Miller1985]. Despite
-the abundance of coastal and island caribou herds, only a handful of examples
-have documented caribou swimming in the ocean [e.g. @Miller1995; @Miller2002;
-@Jeffery2007; @Ricca2012].
-
 The Fogo Island archipelago, off the coast of Newfoundland, Canada, is home to
 approximately 300 woodland caribou. During routine fieldwork on May 30, 2017, we
-observed an unmarked adult male caribou swim between Western and Eastern Indian
-Islands (Figure 1, see Figure 2 for location), a distance of at least 470 m which took ~9 minutes (~52 m
+observed an unmarked adult male caribou swim between two small islands (Figure 1, see Figure 2 for location), a distance of at least 470 m which took ~9 minutes (~52 m
 per minute). This observation prompted us to investigate whether swimming
@@
 swimming between islands on numerous occasions.
+
+Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
+that caribou swim in streams, rivers, and lakes during migration [@Leblond2016],
+to avoid predators [@Bergerud1985], and access islands during calving
+[@Bergerud1990]. Even for adept swimmers like caribou, the energetic expenditure
+of swimming for quadrupedal mammals is significantly higher than walking or
+running, and drowning is also possible [@Miller1985]. Despite
+the abundance of coastal and island caribou herds, only a handful of examples
+have documented caribou swimming in the ocean [e.g. @Miller1995; @Miller2002;
+@Jeffery2007; @Ricca2012].
 

### 2020-05-11 paper/manuscript.Rmd
collar details

@@
 per minute). This observation prompted us to investigate whether swimming
-behaviour was common among GPS radio-collared caribou in the population (n = 29
-adult females; 1240g, GPS 4400M; Lotek Wireless Inc., Newmarket, Ontario,
-Canada). Using `r substr(version$version.string, 1, 15)` [@R], we identified
+behaviour was common among GPS radio-collared caribou in the population (n = 29, for collaring details see [@Peignier_2019]). Using `r substr(version$version.string, 1, 15)` [@R], we identified
 swimming events as two consecutive GPS locations (2-hour relocation rates)  from

### 2020-05-11 paper/manuscript.Rmd
P1

@@
 approximately 300 woodland caribou. During routine fieldwork on May 30, 2017, we
-observed an unmarked adult male caribou swim between two small islands (Figure 1, see Figure 2 for location), a distance of at least 470 m which took ~9 minutes (~52 m
-per minute). This observation prompted us to investigate whether swimming
-behaviour was common among GPS radio-collared caribou in the population (n = 29, for collaring details see [@Peignier_2019]). Using `r substr(version$version.string, 1, 15)` [@R], we identified
-swimming events as two consecutive GPS locations (2-hour relocation rates)  from
-an individual occurred on different islands. Northeastern Newfoundland typically
+observed an unmarked adult male caribou swim between two small islands (Figure
+1, see Figure 2 for location), a distance of at least 470 m which took ~9
+minutes (~52 m per minute). This observation prompted us to investigate whether
+swimming behaviour was common among GPS radio-collared caribou in the population
+(n = 29, for collaring details see [@Peignier_2019]). We identified swimming
+events as two consecutive GPS locations (2-hour relocation rates)  from an
+individual occurred on different islands. Northeastern Newfoundland typically
 experiences pack ice during winter and caribou are known locally to travel
 between islands by walking over the ice. We restricted our GPS data to the
-ice-free period of the year (April 1 to December 31) and we identified `r nids`
-adult female caribou that collectively had `r nevents` swimming events over
-three years (Figure 2). In addition to our own visual and remotely sensed
-observations, residents of Fogo Island have also reported observing caribou
-swimming between islands on numerous occasions.
+ice-free period of the year (April 1 to December 31) and identified `r nevents`
+swimming events over three years (Figure 2). In addition to our own visual and
+remotely sensed observations, residents of Fogo Island have also reported
+observing caribou swimming between islands on numerous occasions.
 

### 2020-05-11 paper/manuscript.Rmd
P2 move to P3
edits to P2

@@
 
+In total, 12 of 29 collared female caribou swam between islands in the Fogo
+Island archipelago. On average, caribou swam approximately `r neventidMean`
+(range = `r neventidMin`-`r neventidMax`) times per year and remained on each
+island for approximately `r dur` (range = `r durMin`-`r durMax`) days before
+swimming again. Caribou may swim between islands in the ocean for similar
+reasons they swim in freshwater [@Leblond2016]. We propose that one such
+explanation, the forage limitation hypothesis, is a likely explanation for
+oceanic swimming for caribou that live on islands in the Fogo Island
+archipelago.
+
 Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
@@
 @Jeffery2007; @Ricca2012].
-
-In total, `r nevents` swimming events occurred throughout the ice-free season,
-with peaks in May and November. On average, caribou swam approximately `r neventidMean`
-(range = `r neventidMin`-`r neventidMax`) times per year and remained on each
-island for approximately `r dur` (range = `r durMin`-`r durMax`) days before
-swimming again. Caribou may swim between islands in the ocean for similar
-reasons they swim in freshwater [@Leblond2016]. We propose that one such
-explanation, the forage limitation hypothesis, is a likely explanation for
-oceanic swimming for caribou that live on islands in the Fogo Island
-archipelago.
-
 Forage limitation and over-grazing is a major concern for caribou populations:
@@
 
-In total, 12 of 29 collared female caribou swam between islands in the Fogo
-Island archipelago. Only `r n10` of these 12 individuals had more than 10
+ Only `r n10` of these 12 individuals had more than 10
 swimming events, suggesting that for ~10% of collared individuals swim
@@
 17 individuals. It remains unclear whether these individuals do not swim, or
-whether swimming is rare, and thus undetected, for the majority of individuals.
-Overall, there appears to be three groups of individuals: those that swam often
-(n = 3, every ~`r indMean` days), those that swam occasionally (n = 9, every ~
-`r ind2Mean` days), and those that never swam (n = 17). It is therefore possible
-that individuals vary in their tendency to swim. We are also unable to draw
-conclusions about the prevalence of swimming in the ocean by male caribou: our
-initial visual observation was of an adult male, but our GPS data only included
-adult females.
 

### 2020-05-11 paper/manuscript.Rmd
Update manuscript.Rmd

--- 2020-05-11 paper/manuscript.Rmd
@@
 
+In total, 12 of 29 collared female caribou swam between islands in the Fogo
+Island archipelago. On average, caribou swam approximately `r neventidMean`
+(range = `r neventidMin`-`r neventidMax`) times per year and remained on each
+island for approximately `r dur` (range = `r durMin`-`r durMax`) days before
+swimming again. We can separate the population into three apparent groups of individuals: those that swam regularly between islands
+(n = 3, every ~`r indMean` days); those that rarely engaged in swimming events (n = 9, every ~
+`r ind2Mean` days); and those that never swam (n = 17).
+
 Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
@@
 whether swimming is rare, and thus undetected, for the majority of individuals.
-Overall, there appears to be three groups of individuals: those that swam often
-(n = 3, every ~`r indMean` days), those that swam occasionally (n = 9, every ~
-`r ind2Mean` days), and those that never swam (n = 17). It is therefore possible
+
+It is therefore possible
 that individuals vary in their tendency to swim. We are also unable to draw
@@
 adult females.
+
+Caribou may swim between islands in the ocean for similar
+reasons they swim in freshwater [@Leblond2016]. We propose that one such
+explanation, the forage limitation hypothesis, is a likely explanation for
+oceanic swimming for caribou that live on islands in the Fogo Island
+archipelago.
 

### 2020-05-11 paper/manuscript.Rmd
P4

@@
 island for approximately `r dur` (range = `r durMin`-`r durMax`) days before
-swimming again. We can separate the population into three apparent groups of individuals: those that swam regularly between islands
-(n = 3, every ~`r indMean` days); those that rarely engaged in swimming events (n = 9, every ~
-`r ind2Mean` days); and those that never swam (n = 17).
+swimming again. We can separate the population into three apparent groups of
+individuals: those that swam regularly between islands (n = 3, every ~`r
+indMean` days); those that rarely engaged in swimming events (n = 9, every ~ `r
+ind2Mean` days); and those that never swam (n = 17).
 
@@
 @Jeffery2007; @Ricca2012].
+
+Caribou may swim between islands in the ocean for similar
+reasons they swim in freshwater [@Leblond2016], mainly concerning predator avoidance or movement between habitats. We propose that one such
+explanation, the forage limitation hypothesis, is the most likely explanation for
+oceanic swimming for caribou that live on islands in the Fogo Island
+archipelago.
+
 Forage limitation and over-grazing is a major concern for caribou populations:
@@
 adult females.
-
-Caribou may swim between islands in the ocean for similar
-reasons they swim in freshwater [@Leblond2016]. We propose that one such
-explanation, the forage limitation hypothesis, is a likely explanation for
-oceanic swimming for caribou that live on islands in the Fogo Island
-archipelago.
 

### 2020-05-11 paper/manuscript.Rmd
P5

@@
 
-Caribou may swim between islands in the ocean for similar
-reasons they swim in freshwater [@Leblond2016], mainly concerning predator avoidance or movement between habitats. We propose that one such
-explanation, the forage limitation hypothesis, is the most likely explanation for
-oceanic swimming for caribou that live on islands in the Fogo Island
-archipelago.
+Caribou may swim between islands in the ocean for similar reasons they swim in
+freshwater [@Leblond2016], mainly concerning predator avoidance or movement
+between habitats. We propose that one such explanation, the forage limitation
+hypothesis, is the most likely explanation for oceanic swimming for caribou that
+live on islands in the Fogo Island archipelago.
+
+Caribou were introduced to Fogo Island (n = 26 animals between 1964–67) and
+Change Island (n = 5 animals in 1964) from Newfoundland as part of a series of
+translocations and introductions throughout the province [@Bergerud1989]. Typical caribou predators such as black bears (*Ursus americanus*) and
+wolves (*Canis lupus*) are absent from Fogo Island; the dominant predator is
+coyote (*Canis latrans*), which predate calves but it remains unclear what threat
+they pose, if any, to adult caribou. Given the minimal risk of predation to adults,
+we propose that forage limitation is more likely to drive movement between
+islands than predator avoidance. For several decades after their introduction to
+Fogo Island, the caribou population did not exceed ~100 individuals (Bergerud
+and Mercer 1989, Newfoundland and Labrador Wildlife Division, unpublished data).
+During the 1990s, population density reached ~300 animals; anecdotal evidence
+suggests caribou were first seen on nearby islands during this same period.
+Increased competition for resources on Fogo due to this increase in density
+could have precipitated movement to other islands.
 
@@
 reason for caribou moving between arctic islands on the sea-ice in winter
-[@Miller1977]. Caribou were introduced to Fogo Island (n = 26 animals between
-1964–67) and Change Island (n = 5 animals in 1964) from Newfoundland as part of
-a series of translocations and introductions throughout the province (Bergerud &
-Mercer 1989). It is unknown when caribou began colonizing nearby islands, but
+[@Miller1977].
+
+It is unknown when caribou began colonizing nearby islands, but
 given that some are small, over-grazing by newly arrived caribou may rapidly

### 2020-05-11 paper/manuscript.Rmd
bergerud1989 ref
P5

@@
 Change Island (n = 5 animals in 1964) from Newfoundland as part of a series of
-translocations and introductions throughout the province [@Bergerud1989]. Typical caribou predators such as black bears (*Ursus americanus*) and
-wolves (*Canis lupus*) are absent from Fogo Island; the dominant predator is
-coyote (*Canis latrans*), which predate calves but it remains unclear what threat
-they pose, if any, to adult caribou. Given the minimal risk of predation to adults,
+translocations and introductions throughout the province [@Bergerud1989].
+Typical caribou predators such as black bears (*Ursus americanus*) and wolves
+(*Canis lupus*) are absent from Fogo Island; the dominant predator is coyote
+(*Canis latrans*), which predate calves but it remains unclear what threat they
+pose, if any, to adult caribou. Given the minimal risk of predation to adults,
 we propose that forage limitation is more likely to drive movement between
@@
 
-Forage limitation and over-grazing is a major concern for caribou populations:
-it can reduce female reproductive success [@Schaefer2016] and thus could lead to
-movement between islands. Forage scarcity has been proposed as a potential
-reason for caribou moving between arctic islands on the sea-ice in winter
-[@Miller1977].
+Forage limitation and over-grazing is a major concern for caribou populations
+and can lead to reduced female reproductive success [@Schaefer2016]. Forage
+scarcity has been proposed as a potential reason for caribou moving between
+arctic islands on the sea-ice in winter [@Miller1977]. For the Fogo Island
+population, as the population grew and predation pressure was insufficient to
+maintain population density, forage depletion could lead caribou to make
+similar, but riskier and more costly, movements between islands during the
+ice-free season. Given the small size of other islands near Fogo, over-grazing
+by newly arrived caribou could rapidly deplete forage on an annual cycle
+[@Bergerud1990], leading to more frequent movements back and forth between
+islands rather than permanent relocation.
 

### 2020-05-11 paper/manuscript.Rmd
merge p4, p5
trade-offs

@@
 of swimming for quadrupedal mammals is significantly higher than walking or
-running, and drowning is also possible [@Miller1985]. Despite
-the abundance of coastal and island caribou herds, only a handful of examples
-have documented caribou swimming in the ocean [e.g. @Miller1995; @Miller2002;
-@Jeffery2007; @Ricca2012].
-
-Caribou may swim between islands in the ocean for similar reasons they swim in
-freshwater [@Leblond2016], mainly concerning predator avoidance or movement
-between habitats. We propose that one such explanation, the forage limitation
-hypothesis, is the most likely explanation for oceanic swimming for caribou that
-live on islands in the Fogo Island archipelago.
+running, and drowning is also possible [@Miller1985]. Despite the abundance of
+coastal and island caribou herds, only a handful of examples have documented
+caribou swimming in the ocean [e.g. @Miller1995; @Miller2002;
+@Jeffery2007; @Ricca2012]. Caribou may swim between islands in the ocean for
+similar reasons they swim in freshwater [@Leblond2016], mainly concerning
+predator avoidance or movement between habitats. We propose that one such
+explanation, the forage limitation hypothesis, is the most likely explanation
+for oceanic swimming for caribou that live on islands in the Fogo Island
+archipelago.
 
@@
 
-It is unknown when caribou began colonizing nearby islands, but
-given that some are small, over-grazing by newly arrived caribou may rapidly
-deplete forage on an annual cycle [@Bergerud1990]. Although unconfirmed, it is
-also possible that Fogo Island has experienced forage depletion since the
-introduction of caribou in the 1960s. More likely, however, is that fine-scale
-competition for foraging resources drives caribou to swim to new islands.
-
-Movement between islands requires caribou to assess the trade-off associated
-with swimming. This trade-off can be understood as density-dependent habitat
-selection governed by the Ideal Free Distribution [@Morris1987; @Bradbury2015].
-Ideal Free Distribution theory predicts a fitness equilibrium: when the density
-in a given habitat patch has exceeded the optimum for fitness within that patch,
-animals should relocate and settle new habitat patches, so that fitness is equal
-across all patches [@Bradbury2015]. Islands act as discrete habitat patches in
-this case, but the costs associated with swimming create a trade-off when moving
-between patches. For caribou in the Fogo Island archipelago, the role of forage
-limitation in habitat selection patterns remains unknown, but swimming
-between islands may be at least partially governed by density-dependent habitat selection.
+The trade-off associated with swimming between islands can be understood as
+density-dependent habitat selection governed by the Ideal Free Distribution
+[@Morris1987; @Bradbury2015]. Ideal Free Distribution theory predicts a fitness
+equilibrium: when the density in a given habitat patch has exceeded the optimum
+for fitness within that patch, animals should relocate and settle new habitat
+patches, so that fitness is equal across all patches [@Bradbury2015]. Islands
+act as discrete habitat patches in this case, but the costs associated with
+swimming create a trade-off when moving between patches. For caribou in the Fogo
+Island archipelago, the role of forage limitation in habitat selection patterns
+remains unknown, but swimming between islands may be at least partially governed
+by density-dependent habitat selection.
 
@@
 whether swimming is rare, and thus undetected, for the majority of individuals.
+
+It is unknown when caribou began colonizing nearby islands, but
+given that some are small, over-grazing by newly arrived caribou may rapidly
+deplete forage on an annual cycle [@Bergerud1990]. Although unconfirmed, it is
+also possible that Fogo Island has experienced forage depletion since the
+introduction of caribou in the 1960s. More likely, however, is that fine-scale
+competition for foraging resources drives caribou to swim to new islands.
 

### 2020-05-11 paper/manuscript.Rmd
rmd

@@
 
+Forage limitation may be the ultimate explanation for caribou swimming between oceanic islands, but cannot fully explain the variation we observed. Several questions remain:
+
+1) How does the energetic cost or risk of swimming differ from walking across sea ice? How many individuals were observed travelling between islands during the winter, relative to the 12 swimmers?
+
  Only `r n10` of these 12 individuals had more than 10

### 2020-05-11 paper/manuscript.Rmd
question 1
question 2
question 3

@@
 
-Forage limitation may be the ultimate explanation for caribou swimming between oceanic islands, but cannot fully explain the variation we observed. Several questions remain:
+Forage limitation may be the ultimate explanation for caribou swimming between
+oceanic islands, but cannot fully explain the variation we observed. Several
+questions remain:
 
-1) How does the energetic cost or risk of swimming differ from walking across sea ice? How many individuals were observed travelling between islands during the winter, relative to the 12 swimmers?
+1) How does the energetic cost or risk of swimming differ from walking across
+sea ice? How many individuals were observed travelling between islands during
+the winter, relative to the 12 swimmers?
+
+2) At which varying temporal and ecological scale do caribou choose to swim
+between islands? Does the time of day, weather, seasonal trends, or availability
+of food influence swimming?
+
+3) Does reproductive status (presence of a calf-at-heel) or demographics (age,
+sex) influence propensity for oceanic swimming? Our initial observation was of
+an adult male, but the only collared individuals are adult females. Are younger
+animals more likely to swim if they are outcompeted? More likely? Are males more
+likely when in the summer their only concern is foraging, given the relatively
+minimal predation risk they face without calves?
 

### 2020-05-11 paper/manuscript.Rmd
question 4 + 5

@@
 
+4) If swimming is a function of density-dependent habitat selection, what are
+the costs of movement between patches? IFD assumes zero-cost movement, but could
+the risk and energetic costs associated with oceanic swimming be estimated using
+the potential fitness deficit that caribou are willing to tolerate before swimming to a
+new island?
+
+5) In relation to all the previous questions: how do individuals vary in their assessment of the costs and benefits of swimming between islands?
+Why did some caribou regularly engage in swimming events, while more than half of our marked population were never observed swimming in the ocean?
+The forage limitation hypothesis is generally thought to apply uniformly across populations, but individual caribou could evaluate the costs and benefits of swimming to another island differently.
+It is possible that some individuals, in our case ~10%, considered swimming to be less costly or perceived competition or predation more acutely than their conspecifics.
+
  Only `r n10` of these 12 individuals had more than 10

### 2020-05-11 paper/manuscript.Rmd
edits

@@
 
-5) In relation to all the previous questions: how do individuals vary in their assessment of the costs and benefits of swimming between islands?
-Why did some caribou regularly engage in swimming events, while more than half of our marked population were never observed swimming in the ocean?
-The forage limitation hypothesis is generally thought to apply uniformly across populations, but individual caribou could evaluate the costs and benefits of swimming to another island differently.
-It is possible that some individuals, in our case ~10%, considered swimming to be less costly or perceived competition or predation more acutely than their conspecifics.
+5) In relation to all the previous questions: how do individuals vary in their
+assessment of the costs and benefits of swimming between islands? Why did some
+caribou regularly engage in swimming events, while more than half of our marked
+population were never observed swimming in the ocean? The forage limitation
+hypothesis is thought to apply uniformly across populations, but
+individual caribou could evaluate the costs and benefits of swimming to another
+island differently. It is possible that some individuals, in our case ~10%,
+considered swimming to be less costly or perceived competition or predation more
+acutely than their conspecifics.
+
+Oceanic swimming is relatively common among a minority of marked caribou. Of the
+three collared individuals that swam the most, two were collared on Western
+Indian Island and the third was collared on Fogo Island. Inter-island distances
+between Western and Eastern Indian Islands, and the smaller islands around them,
+are shorter than those between Fogo Island and others (Figure 2). It is possible
+that caribou display natal philopatry [e.g. @Larue2018] to some of the smaller
+islands and thus are more likely to swim between them more frequently than
+animals that were born, or spend most of their time, on Fogo Island.
 
@@
 adult females.
-
-Oceanic swimming is relatively common among a minority of marked caribou. Of the
-three collared individuals that swam the most, two were collared on Western
-Indian Island and the third was collared on Fogo Island. Inter-island distances
-between Western and Eastern Indian Islands, and the smaller islands around them,
-are shorter than those between Fogo Island and others (Figure 2). It is possible
-that caribou display natal philopatry [e.g. @Larue2018] to some of the smaller
-islands and thus are more likely to swim between them more frequently than
-animals that were born, or spend most of their time, on Fogo Island. The forage
-limitation hypothesis is generally thought to apply uniformly across
-populations, but individual caribou could evaluate the costs and benefits of
-swimming to another island differently. It is possible that some individuals, in
-our case ~10%, considered swimming to be less costly or perceived competition or
-predation more acutely than their conspecifics.
 

### 2020-05-11 paper/manuscript.Rmd
conclusion

@@
 
-For several decades after their introduction to Fogo Island, the caribou
-population did not exceed ~100 individuals [@Bergerud1989, Newfoundland and
-Labrador Wildlife Division, unpublished data]. During the 1990s, population
-density reached ~300 animals and it is unknown when caribou began to occupy other
-islands, although anecdotal evidence suggests it was during the 1990s. We surmise that as
-competition among conspecifics increased along with population size,
-density-dependent habitat selection resulted in expansion of the population to
-nearby islands. Following Ideal Free Distribution theory, caribou should swim to
-new islands when the average fitness of individuals on the starting island
-exceeds the density-fitness equilibrium [@Morris1987]. While this is an ultimate
-explanation, our data suggest caribou only remain on smaller islands for a few
-days at a time, so the fitness equilibrium remains a theoretical construct. More
-likely, however, is that fine-scale competition, density-dependent habitat
-selection, and forage depletion drives individuals to periodically swim
-between islands. Given the presumed low density of coyotes on Fogo Island and
-the minimal risk they pose, predation risk does not appear to be a major driver
-of swimming.
-
-Our observations add to the evidence that caribou can, and occasionally do, swim
-in the ocean [e.g. @Miller1995; @Miller2002; @Jeffery2007; @Ricca2012]. 
-We suggest that swimming is likely more common than previously thought
-for caribou living on oceanic islands, and that forage limitation and the
-associated density-dependent habitat selection is an ultimate explanation for
-this phenomenon. Although our inference is limited to observations, islands
-appear to represent discrete foraging patches for terrestrial animals that can
-influence fitness via increased foraging opportunities. We also posit that
-individuals can vary in their evaluation of costs and benefits of movement
-between these habitat patches.
+We surmise that as competition among conspecifics increased along with
+population size, density-dependent habitat selection resulted in expansion of
+the population to nearby islands. Following Ideal Free Distribution theory,
+caribou should swim to new islands when the average fitness of individuals on
+the starting island exceeds the density-fitness equilibrium [@Morris1987]. While
+this is an ultimate explanation, our data suggest caribou only remain on smaller
+islands for a few days at a time, so the fitness equilibrium remains a
+theoretical construct. More likely, however, is that fine-scale competition,
+density-dependent habitat selection, and forage depletion drives individuals to
+periodically swim between islands. Given the presumed low density of coyotes on
+Fogo Island and the minimal risk they pose, predation risk does not appear to be
+a major driver of swimming. Our observations add to the evidence that caribou
+can, and occasionally do, swim in the ocean [e.g. @Miller1995; @Miller2002;
+@Jeffery2007; @Ricca2012]. We suggest that swimming is likely more common than
+previously thought for caribou living on oceanic islands, and that forage
+limitation and the associated density-dependent habitat selection is an ultimate
+explanation for this phenomenon. Although our inference is limited to
+observations, islands appear to represent discrete foraging patches for
+terrestrial animals that can influence fitness via increased foraging
+opportunities. We also present several hypotheses related to the nuances and
+further variation within the idea that density-dependent habitat selection
+governs swimming behaviour, and posit that above and beyond these additional
+factors, individuals can vary in their evaluation of costs and benefits of
+movement between these habitat patches.
 

### 2020-05-11 paper/manuscript.Rmd
delete conclusion sentences
delete: it is unknown when caribou...

@@
 
-It is unknown when caribou began colonizing nearby islands, but
-given that some are small, over-grazing by newly arrived caribou may rapidly
-deplete forage on an annual cycle [@Bergerud1990]. Although unconfirmed, it is
-also possible that Fogo Island has experienced forage depletion since the
-introduction of caribou in the 1960s. More likely, however, is that fine-scale
-competition for foraging resources drives caribou to swim to new islands.
-
-It is therefore possible
-that individuals vary in their tendency to swim. We are also unable to draw
-conclusions about the prevalence of swimming in the ocean by male caribou: our
-initial visual observation was of an adult male, but our GPS data only included
-adult females.
-
 We surmise that as competition among conspecifics increased along with

### 2020-05-11 paper/manuscript.Rmd
delete: only n of 12 individuals...

@@
 
- Only `r n10` of these 12 individuals had more than 10
-swimming events, suggesting that for ~10% of collared individuals swim
-relatively frequently, on average occurring approximately every `r indMean` days
-during ice-free seasons. We did not detect any swimming events for the remaining
-17 individuals. It remains unclear whether these individuals do not swim, or
-whether swimming is rare, and thus undetected, for the majority of individuals.
-
 We surmise that as competition among conspecifics increased along with

### 2020-05-11 paper/manuscript.Rmd
delete philopatry comment
question 1

@@
 Forage limitation may be the ultimate explanation for caribou swimming between
-oceanic islands, but cannot fully explain the variation we observed. Several
-questions remain:
+oceanic islands, but a number of proximate mechanisms remain untested. Our
+observations of caribou swimming in the ocean to five questions for future
+research:
 
-1) How does the energetic cost or risk of swimming differ from walking across
-sea ice? How many individuals were observed travelling between islands during
-the winter, relative to the 12 swimmers?
+1) How does the energetic cost or risk of swimming differ from walking on sea
+ice? The cost of swimming may be lower than the risk of falling through sea ice
+and drowning, so what proportion of inter-island movements are swimming events
+compared to walking on ice? Although we are unable to
 
@@
 between Western and Eastern Indian Islands, and the smaller islands around them,
-are shorter than those between Fogo Island and others (Figure 2). It is possible
-that caribou display natal philopatry [e.g. @Larue2018] to some of the smaller
-islands and thus are more likely to swim between them more frequently than
-animals that were born, or spend most of their time, on Fogo Island.
+are shorter than those between Fogo Island and others (Figure 2).
 

### 2020-05-11 paper/manuscript.Rmd
question 2 and 3

@@
 and drowning, so what proportion of inter-island movements are swimming events
-compared to walking on ice? Although we are unable to
+compared to walking on ice? The presence and extent of sea ice varies annually
+in our study area, and although we are unable to determine the exact timing or
+duation of sea ice, we delineate on Figure 2b the timing of typical sea ice
+arrival and departure.
 
-2) At which varying temporal and ecological scale do caribou choose to swim
-between islands? Does the time of day, weather, seasonal trends, or availability
+2) At which temporal and spatial scales do caribou choose to swim
+between islands? Does the time of day, weather, season, or availability
 of food influence swimming?
 
-3) Does reproductive status (presence of a calf-at-heel) or demographics (age,
+3) Does reproductive status (presence of a calf-at-heel) or demographic class (age,
 sex) influence propensity for oceanic swimming? Our initial observation was of
-an adult male, but the only collared individuals are adult females. Are younger
-animals more likely to swim if they are outcompeted? More likely? Are males more
-likely when in the summer their only concern is foraging, given the relatively
-minimal predation risk they face without calves?
+an adult male, but all collared individuals were adult females. Are younger
+caribou more likely to swim due to heightened competition? Are males more
+likely to swim in the summer given the are not restricted based on presence of a calf-at-heel, as females are.
 

### 2020-05-11 paper/manuscript.Rmd
q3
conclusion

@@
 
-3) Does reproductive status (presence of a calf-at-heel) or demographic class (age,
-sex) influence propensity for oceanic swimming? Our initial observation was of
-an adult male, but all collared individuals were adult females. Are younger
-caribou more likely to swim due to heightened competition? Are males more
-likely to swim in the summer given the are not restricted based on presence of a calf-at-heel, as females are.
+3) Does reproductive status (presence of a calf-at-heel) or demographic class
+(age, sex) influence propensity for oceanic swimming? Our initial observation
+was of an adult male, but all collared individuals were adult females. Are
+younger caribou more likely to swim due to heightened competition? Are males
+more likely to swim in the summer given the are not restricted based on presence
+of a calf-at-heel, as females are.
 
@@
 
-We surmise that as competition among conspecifics increased along with
+We surmise that for Fogo Island, as competition among conspecifics increased along with
 population size, density-dependent habitat selection resulted in expansion of
@@
 density-dependent habitat selection, and forage depletion drives individuals to
-periodically swim between islands. Given the presumed low density of coyotes on
-Fogo Island and the minimal risk they pose, predation risk does not appear to be
-a major driver of swimming. Our observations add to the evidence that caribou
+periodically swim between islands. Our focal observations add to the evidence that caribou
 can, and occasionally do, swim in the ocean [e.g. @Miller1995; @Miller2002;

### 2020-05-11 paper/manuscript.Rmd
delete: oceanic swimming is relatively common...

@@
 
-Oceanic swimming is relatively common among a minority of marked caribou. Of the
-three collared individuals that swam the most, two were collared on Western
-Indian Island and the third was collared on Fogo Island. Inter-island distances
-between Western and Eastern Indian Islands, and the smaller islands around them,
-are shorter than those between Fogo Island and others (Figure 2).
-
 We surmise that for Fogo Island, as competition among conspecifics increased along with

### 2020-05-11 paper/manuscript.Rmd
edit ref

@@
 swimming behaviour was common among GPS radio-collared caribou in the population
-(n = 29, for collaring details see [@Peignier_2019]). We identified swimming
+(n = 29, for collaring details see [@Peignier2019]). We identified swimming
 events as two consecutive GPS locations (2-hour relocation rates)  from an

### 2020-05-11 paper/manuscript.Rmd
fix r code in rmd

@@
 swimming again. We can separate the population into three apparent groups of
-individuals: those that swam regularly between islands (n = 3, every ~`r
-indMean` days); those that rarely engaged in swimming events (n = 9, every ~ `r
-ind2Mean` days); and those that never swam (n = 17).
+individuals: those that swam regularly between islands (n = 3, every ~`r indMean days`); those that rarely engaged in swimming events (n = 9, every ~ `r ind2Mean days`); and those that never swam (n = 17).
 

### 2020-05-11 paper/manuscript.Rmd
reference brackets

@@
 swimming behaviour was common among GPS radio-collared caribou in the population
-(n = 29, for collaring details see [@Peignier2019]). We identified swimming
+[n = 29, for collaring details see, @Peignier2019]. We identified swimming
 events as two consecutive GPS locations (2-hour relocation rates)  from an
@@
 swimming again. We can separate the population into three apparent groups of
-individuals: those that swam regularly between islands (n = 3, every ~`r indMean days`); those that rarely engaged in swimming events (n = 9, every ~ `r ind2Mean days`); and those that never swam (n = 17).
+individuals: those that swam regularly between islands (n = 3, every ~`r indMean
+days`); those that rarely engaged in swimming events (n = 9, every ~ `r ind2Mean
+days`); and those that never swam (n = 17).
 

### 2020-05-11 paper/manuscript.Rmd
edits minor

@@
 events as two consecutive GPS locations (2-hour relocation rates)  from an
-individual occurred on different islands. Northeastern Newfoundland typically
+individual that occurred on different islands. Northeastern Newfoundland typically
 experiences pack ice during winter and caribou are known locally to travel
@@
 swimming events over three years (Figure 2). In addition to our own visual and
-remotely sensed observations, residents of Fogo Island have also reported
-observing caribou swimming between islands on numerous occasions.
+remotely sensed observations, residents of Fogo Island have also reported regularly
+observing caribou swimming between islands.
 
@@
 island for approximately `r dur` (range = `r durMin`-`r durMax`) days before
-swimming again. We can separate the population into three apparent groups of
-individuals: those that swam regularly between islands (n = 3, every ~`r indMean
-days`); those that rarely engaged in swimming events (n = 9, every ~ `r ind2Mean
-days`); and those that never swam (n = 17).
+swimming again. Based on swimming data, we separated the population into three
+apparent groups of individuals: those that swam regularly between islands (n =
+3, every ~`r indMean days`); those that rarely engaged in swimming events (n =
+9, every ~ `r ind2Mean days`); and those that never swam (n = 17).
 
@@
 to avoid predators [@Bergerud1985], and access islands during calving
-[@Bergerud1990]. Even for adept swimmers like caribou, the energetic expenditure
+[@Bergerud1990]. Even for adept swimmers like caribou, that have hollow hair shafts to presumably aid in flotation, the energetic expenditure
 of swimming for quadrupedal mammals is significantly higher than walking or

### 2020-05-11 paper/manuscript.Rmd
Update manuscript.Rmd
bergerud ref

@@
 to avoid predators [@Bergerud1985], and access islands during calving
-[@Bergerud1990]. Even for adept swimmers like caribou, that have hollow hair shafts to presumably aid in flotation, the energetic expenditure
-of swimming for quadrupedal mammals is significantly higher than walking or
-running, and drowning is also possible [@Miller1985]. Despite the abundance of
-coastal and island caribou herds, only a handful of examples have documented
-caribou swimming in the ocean [e.g. @Miller1995; @Miller2002;
+[@Bergerud1990]. Even for adept swimmers like caribou, that have hollow hair
+shafts to presumably aid in flotation, the energetic expenditure of swimming for
+quadrupedal mammals is significantly higher than walking or running, and
+drowning is also possible [@Miller1985]. Despite the abundance of coastal and
+island caribou herds, only a handful of examples have documented caribou
+swimming in the ocean [e.g. @Miller1995; @Miller2002;
 @Jeffery2007; @Ricca2012]. Caribou may swim between islands in the ocean for
@@
 islands than predator avoidance. For several decades after their introduction to
-Fogo Island, the caribou population did not exceed ~100 individuals (Bergerud
-and Mercer 1989, Newfoundland and Labrador Wildlife Division, unpublished data).
+Fogo Island, the caribou population did not exceed ~100 individuals
+[@Bergerud1989, Newfoundland and Labrador Wildlife Division, unpublished data].
 During the 1990s, population density reached ~300 animals; anecdotal evidence

### 2020-05-11 paper/manuscript.Rmd
Rayl ref

@@
 (*Canis latrans*), which predate calves but it remains unclear what threat they
-pose, if any, to adult caribou. Given the minimal risk of predation to adults,
+pose, if any, to adult caribou [@Rayl2014]. Given the minimal risk of predation to adults,
 we propose that forage limitation is more likely to drive movement between

### 2020-05-11 paper/manuscript.Rmd
generate doc

@@
 apparent groups of individuals: those that swam regularly between islands (n =
-3, every ~`r indMean days`); those that rarely engaged in swimming events (n =
-9, every ~ `r ind2Mean days`); and those that never swam (n = 17).
+3, every ~`r indMean` days); those that rarely engaged in swimming events (n =
+9, every ~ `r ind2Mean` days); and those that never swam (n = 17).
 

### 2020-05-11 paper/manuscript.Rmd
render both

@@
 output:
-   # github_document
+   github_document: default
    bookdown::word_document2:
      reference_docx: template.docx
+knit: (function(inputFile, encoding) {
+  rmarkdown::render(inputFile, encoding = encoding,
+  output_format = "all") })
 bibliography: ../references/swim.bib

### 2020-05-11 paper/manuscript.Rmd
rm spatsoc
minor edits

@@
 ```{r pkgs}
-library(spatsoc)
 library(data.table)
@@
 that caribou swim in streams, rivers, and lakes during migration [@Leblond2016],
-to avoid predators [@Bergerud1985], and access islands during calving
+to avoid predators [@Bergerud1985], and to access islands during calving
 [@Bergerud1990]. Even for adept swimmers like caribou, that have hollow hair
-shafts to presumably aid in flotation, the energetic expenditure of swimming for
+shafts that aid in flotation, the energetic expenditure of swimming for
 quadrupedal mammals is significantly higher than walking or running, and

### 2020-05-11 paper/manuscript.Rmd
add caribou moving between islands

@@
 drowning is also possible [@Miller1985]. Despite the abundance of coastal and
-island caribou herds, only a handful of examples have documented caribou
-swimming in the ocean [e.g. @Miller1995; @Miller2002;
-@Jeffery2007; @Ricca2012]. Caribou may swim between islands in the ocean for
-similar reasons they swim in freshwater [@Leblond2016], mainly concerning
-predator avoidance or movement between habitats. We propose that one such
-explanation, the forage limitation hypothesis, is the most likely explanation
-for oceanic swimming for caribou that live on islands in the Fogo Island
-archipelago.
+island caribou herds, only a handful of studies have documented caribou swimming
+in the ocean, and most of these lack observation, but rather deduce swimming
+based on presence of caribou on a previously uninhabitated island  [e.g.
+@Miller1995; @Miller2002; @Jeffery2007; @Ricca2012]. Caribou may swim between
+islands in the ocean for similar reasons they swim in freshwater [@Leblond2016],
+mainly concerning predator avoidance or movement between habitats. We propose
+that one such explanation, the forage limitation hypothesis, is the most likely
+explanation for oceanic swimming for caribou that live on islands in the Fogo
+Island archipelago.
 

### 2020-05-11 paper/manuscript.Rmd
minor edits and fig2 caption

@@
 islands in the ocean for similar reasons they swim in freshwater [@Leblond2016],
-mainly concerning predator avoidance or movement between habitats. We propose
+mainly concerning predator avoidance or to access new habitats. We propose
 that one such explanation, the forage limitation hypothesis, is the most likely
@@
 [@Bergerud1989, Newfoundland and Labrador Wildlife Division, unpublished data].
-During the 1990s, population density reached ~300 animals; anecdotal evidence
+During the 1990s, population density reached ~300 animals and anecdotal evidence
 suggests caribou were first seen on nearby islands during this same period.
@@
 oceanic islands, but a number of proximate mechanisms remain untested. Our
-observations of caribou swimming in the ocean to five questions for future
+observations of caribou swimming in the ocean lead to five questions for future
 research:
@@
 (ref:fig2) A) Map of the Fogo Island archipelago with swimming events between islands.
-B) Histogram displaying the distribution of swimming events throughout the year.
-Note, colours correspond to individual caribou.
+B) Histogram displaying the frequency of swimming events throughout the year.
+Note, colours correspond to individual caribou and vertical black bars delineate the ice-free season (April 1 to December 31).
 C) Inset of swimming events between small islands on the northern coast of Fogo island.

### 2020-05-14 paper/manuscript.Rmd
edit to P1
infer swim

@@
 swimming behaviour was common among GPS radio-collared caribou in the population
-[n = 29, for collaring details see, @Peignier2019]. We identified swimming
-events as two consecutive GPS locations (2-hour relocation rates)  from an
-individual that occurred on different islands. Northeastern Newfoundland typically
-experiences pack ice during winter and caribou are known locally to travel
-between islands by walking over the ice. We restricted our GPS data to the
-ice-free period of the year (April 1 to December 31) and identified `r nevents`
-swimming events over three years (Figure 2). In addition to our own visual and
-remotely sensed observations, residents of Fogo Island have also reported regularly
-observing caribou swimming between islands.
+[n = 29, for collaring details see, @Peignier2019]. We identified an
+individual's swimming events as two consecutive GPS locations (2-hour relocation
+rates)  on different islands. Northeastern Newfoundland typically experiences
+pack ice during winter and caribou are known locally to travel between islands
+by walking over the ice. We restricted our GPS data to the ice-free period of
+the year (April 1 to December 31) and identified `r nevents` swimming events
+over three years (Figure 2). In addition to our own visual and remotely sensed
+observations, residents of Fogo Island have also reported regularly observing
+caribou swimming between islands.
 
 In total, 12 of 29 collared female caribou swam between islands in the Fogo
-Island archipelago. On average, caribou swam approximately `r neventidMean`
+Island archipelago. Of the caribou that swam, caribou swam on average approximately `r neventidMean`
 (range = `r neventidMin`-`r neventidMax`) times per year and remained on each
 island for approximately `r dur` (range = `r durMin`-`r durMax`) days before
-swimming again. Based on swimming data, we separated the population into three
+swimming again. Based on GPS data to infer swimming, we separated the population into three
 apparent groups of individuals: those that swam regularly between islands (n =

### 2020-05-14 paper/manuscript.Rmd
update P2
generate paper

@@
 In total, 12 of 29 collared female caribou swam between islands in the Fogo
-Island archipelago. Of the caribou that swam, caribou swam on average approximately `r neventidMean`
-(range = `r neventidMin`-`r neventidMax`) times per year and remained on each
-island for approximately `r dur` (range = `r durMin`-`r durMax`) days before
-swimming again. Based on GPS data to infer swimming, we separated the population into three
-apparent groups of individuals: those that swam regularly between islands (n =
-3, every ~`r indMean` days); those that rarely engaged in swimming events (n =
-9, every ~ `r ind2Mean` days); and those that never swam (n = 17).
+Island archipelago. Of the caribou that swam, caribou swam on average
+approximately `r neventidMean` (range = `r neventidMin`-`r neventidMax`) times
+per year and remained on each island for approximately `r dur` (range = 
+`r durMin`-`r durMax`) days before swimming again. Based on GPS data to infer
+swimming, we separated the population into three apparent groups of individuals:
+those that swam regularly between islands (n = 3, every ~`r indMean` days);
+those that rarely engaged in swimming events (n = 9, every ~ `r ind2Mean` days);
+and those that never swam (n = 17).
 
@@
 islands in the ocean for similar reasons they swim in freshwater [@Leblond2016],
-mainly concerning predator avoidance or to access new habitats. We propose
+mainly concerning predator avoidance or to access new habitat. We propose
 that one such explanation, the forage limitation hypothesis, is the most likely

### 2020-05-14 paper/manuscript.Rmd
minor edits

@@
 islands in the ocean for similar reasons they swim in freshwater [@Leblond2016],
-mainly concerning predator avoidance or to access new habitat. We propose
+mainly concerning predator avoidance or to access new habitats and foraging opportunities. We propose
 that one such explanation, the forage limitation hypothesis, is the most likely
@@
 suggests caribou were first seen on nearby islands during this same period.
-Increased competition for resources on Fogo due to this increase in density
+Increased competition for resources on Fogo Island due to this increase in density
 could have precipitated movement to other islands.
@@
 similar, but riskier and more costly, movements between islands during the
-ice-free season. Given the small size of other islands near Fogo, over-grazing
+ice-free season. Given the small size of other islands near Fogo (Figure 2), over-grazing
 by newly arrived caribou could rapidly deplete forage on an annual cycle

### 2020-05-14 paper/manuscript.Rmd
update Q1

@@
 compared to walking on ice? The presence and extent of sea ice varies annually
-in our study area, and although we are unable to determine the exact timing or
-duation of sea ice, we delineate on Figure 2b the timing of typical sea ice
-arrival and departure.
+in our study area and we are unable to determine the exact timing or duration of
+sea ice from past years. However, we qualitatively delineate the timing of
+typical sea ice arrival and departure (Figure 2B) and deduce that although the
+majority of inter-island movements occur between April 1 and December 31, some
+inter-island movements appear to be caribou walking on sea ice between islands.
+Future research should focus on the trade-off between these types of movement.
 

### 2020-05-14 paper/manuscript.Rmd
update Q3

@@
 3) Does reproductive status (presence of a calf-at-heel) or demographic class
-(age, sex) influence propensity for oceanic swimming? Our initial observation
+(age or sex) influence propensity for oceanic swimming? Our initial observation
 was of an adult male, but all collared individuals were adult females. Are
-younger caribou more likely to swim due to heightened competition? Are males
-more likely to swim in the summer given the are not restricted based on presence
-of a calf-at-heel, as females are.
+younger caribou more likely to swim due to heightened competition? Are males or
+females without a calf-at-heel more likely to swim in the summer given the are
+not restricted based on presence of a calf-at-heel, as females are.
 

### 2020-05-14 paper/manuscript.Rmd
Q5

@@
 population were never observed swimming in the ocean? The forage limitation
-hypothesis is thought to apply uniformly across populations, but
-individual caribou could evaluate the costs and benefits of swimming to another
-island differently. It is possible that some individuals, in our case ~10%,
-considered swimming to be less costly or perceived competition or predation more
-acutely than their conspecifics.
+hypothesis is thought to apply uniformly across populations, but individual
+caribou could evaluate the costs and benefits of swimming differently. It is
+possible that some individuals, in our case ~10%, considered swimming to be less
+costly or perceived competition or predation more acutely than their
+conspecifics.
 

### 2020-05-14 paper/manuscript.Rmd
- moved 'local observations' higher in para before discussing GPS analyses - expanded question (1) swimming vs. ice-walking at end

@@
 1, see Figure 2 for location), a distance of at least 470 m which took ~9
-minutes (~52 m per minute). This observation prompted us to investigate whether
+minutes (~52 m per minute). Residents of Fogo Island confirmed this phenomenon and have also reported regularly observing
+caribou swimming between islands. This observation prompted us to investigate whether
 swimming behaviour was common among GPS radio-collared caribou in the population
@@
 the year (April 1 to December 31) and identified `r nevents` swimming events
-over three years (Figure 2). In addition to our own visual and remotely sensed
-observations, residents of Fogo Island have also reported regularly observing
-caribou swimming between islands.
+over three years (Figure 2).
 
@@
 to avoid predators [@Bergerud1985], and to access islands during calving
-[@Bergerud1990]. Even for adept swimmers like caribou, that have hollow hair
+[@Bergerud1990]. Even for adept swimmers like caribou, which have hollow hair
 shafts that aid in flotation, the energetic expenditure of swimming for
@@
 
-1) How does the energetic cost or risk of swimming differ from walking on sea
-ice? The cost of swimming may be lower than the risk of falling through sea ice
-and drowning, so what proportion of inter-island movements are swimming events
-compared to walking on ice? The presence and extent of sea ice varies annually
-in our study area and we are unable to determine the exact timing or duration of
-sea ice from past years. However, we qualitatively delineate the timing of
-typical sea ice arrival and departure (Figure 2B) and deduce that although the
-majority of inter-island movements occur between April 1 and December 31, some
-inter-island movements appear to be caribou walking on sea ice between islands.
-Future research should focus on the trade-off between these types of movement.
+1) How do the energetic cost and risk of swimming differ from walking on sea
+ice?
+Walking over solid ice is likely less energetically expensive than swimming, but may pose a greater risk of falling through depending on the thickness and stability of ice.
+What proportion of inter-island movements are swimming events compared to walking on ice?
+The presence and extent of sea ice varies annually in our study area and we are unable to determine the exact timing or duration of sea ice from past years.
+However, we qualitatively delineate the timing of typical sea ice arrival and departure (Figure 2B) and deduce that although the majority of inter-island movements occur between April 1 and December 31, some inter-island movements appear to be caribou walking on sea ice between islands.
+Future research on the trade-offs of these types of movement should consider both how the costs associated with each differ, but also how the potential benefits - foraging opportunities - may vary seasonally.
 
@@
 
-3) Does reproductive status (presence of a calf-at-heel) or demographic class
-(age or sex) influence propensity for oceanic swimming? Our initial observation
-was of an adult male, but all collared individuals were adult females. Are
-younger caribou more likely to swim due to heightened competition? Are males or
-females without a calf-at-heel more likely to swim in the summer given the are
-not restricted based on presence of a calf-at-heel, as females are.
+3) Does reproductive status (presence of a calf-at-heel) or demographic class (age or sex) influence propensity for oceanic swimming?
+Our initial observation was of an adult male, but all collared individuals were adult females.
+Are younger caribou more likely to swim if they are unable to compete with older individuals for resources?
+Are males or females without a calf-at-heel more likely to swim in the summer given they are not restricted based on presence of a calf-at-heel, as females are?
 
@@
 the risk and energetic costs associated with oceanic swimming be estimated using
-the potential fitness deficit that caribou are willing to tolerate before swimming to a
-new island?
+the potential fitness deficit that caribou are willing to tolerate before
+swimming to a new island?
 

### 2020-05-14 paper/manuscript.Rmd
re-align paragraphs
minor edits

@@
 1, see Figure 2 for location), a distance of at least 470 m which took ~9
-minutes (~52 m per minute). Residents of Fogo Island confirmed this phenomenon and have also reported regularly observing
-caribou swimming between islands. This observation prompted us to investigate whether
-swimming behaviour was common among GPS radio-collared caribou in the population
-[n = 29, for collaring details see, @Peignier2019]. We identified an
-individual's swimming events as two consecutive GPS locations (2-hour relocation
-rates)  on different islands. Northeastern Newfoundland typically experiences
-pack ice during winter and caribou are known locally to travel between islands
-by walking over the ice. We restricted our GPS data to the ice-free period of
-the year (April 1 to December 31) and identified `r nevents` swimming events
-over three years (Figure 2).
+minutes (~52 m per minute). Residents of Fogo Island confirmed this phenomenon
+and have also reported regularly observing caribou swimming between islands.
+This observation prompted us to investigate whether swimming behaviour was
+common among GPS radio-collared caribou in the population [n = 29, for collaring
+details see, @Peignier2019]. We identified an individual's swimming events as
+two consecutive GPS locations (2-hour relocation rates)  on different islands.
+Northeastern Newfoundland typically experiences pack ice during winter and
+caribou are known locally to travel between islands by walking over the ice. We
+restricted our GPS data to the ice-free period of the year (April 1 to December
+31) and identified `r nevents` swimming events over three years (Figure 2).
 
@@
 drowning is also possible [@Miller1985]. Despite the abundance of coastal and
-island caribou herds, only a handful of studies have documented caribou swimming
+island caribou herds, only a few studies have documented caribou swimming
 in the ocean, and most of these lack observation, but rather deduce swimming
@@
 islands in the ocean for similar reasons they swim in freshwater [@Leblond2016],
-mainly concerning predator avoidance or to access new habitats and foraging opportunities. We propose
-that one such explanation, the forage limitation hypothesis, is the most likely
-explanation for oceanic swimming for caribou that live on islands in the Fogo
-Island archipelago.
+mainly concerning predator avoidance or to access new habitats and foraging
+opportunities. We propose that one such explanation, the forage limitation
+hypothesis, is the most likely explanation for oceanic swimming for caribou that
+live on islands in the Fogo Island archipelago.
 
@@
 (*Canis latrans*), which predate calves but it remains unclear what threat they
-pose, if any, to adult caribou [@Rayl2014]. Given the minimal risk of predation to adults,
-we propose that forage limitation is more likely to drive movement between
-islands than predator avoidance. For several decades after their introduction to
-Fogo Island, the caribou population did not exceed ~100 individuals
-[@Bergerud1989, Newfoundland and Labrador Wildlife Division, unpublished data].
-During the 1990s, population density reached ~300 animals and anecdotal evidence
-suggests caribou were first seen on nearby islands during this same period.
-Increased competition for resources on Fogo Island due to this increase in density
-could have precipitated movement to other islands.
+pose, if any, to adult caribou [@Rayl2014]. Given the minimal risk of predation
+to adults, we propose that forage limitation is more likely to drive movement
+between islands than predator avoidance. For several decades after their
+introduction to Fogo Island, the caribou population did not exceed ~100
+individuals [@Bergerud1989, Newfoundland and Labrador Wildlife Division,
+unpublished data]. During the 1990s, population density reached ~300 animals and
+anecdotal evidence suggests caribou were first seen on nearby islands during
+this same period. Increased competition for resources on Fogo Island due to this
+increase in density could have precipitated movement to other islands.
 
@@
 similar, but riskier and more costly, movements between islands during the
-ice-free season. Given the small size of other islands near Fogo (Figure 2), over-grazing
-by newly arrived caribou could rapidly deplete forage on an annual cycle
-[@Bergerud1990], leading to more frequent movements back and forth between
+ice-free season. Given the small size of other islands near Fogo (Figure 2),
+over-grazing by newly arrived caribou could rapidly deplete forage on an annual
+cycle [@Bergerud1990], leading to more frequent movements back and forth between
 islands rather than permanent relocation.
@@
 1) How do the energetic cost and risk of swimming differ from walking on sea
-ice?
-Walking over solid ice is likely less energetically expensive than swimming, but may pose a greater risk of falling through depending on the thickness and stability of ice.
-What proportion of inter-island movements are swimming events compared to walking on ice?
-The presence and extent of sea ice varies annually in our study area and we are unable to determine the exact timing or duration of sea ice from past years.
-However, we qualitatively delineate the timing of typical sea ice arrival and departure (Figure 2B) and deduce that although the majority of inter-island movements occur between April 1 and December 31, some inter-island movements appear to be caribou walking on sea ice between islands.
-Future research on the trade-offs of these types of movement should consider both how the costs associated with each differ, but also how the potential benefits - foraging opportunities - may vary seasonally.
+ice? Walking over solid ice is likely less energetically expensive than
+swimming, but may pose a greater risk of falling through depending on the
+thickness and stability of ice. What proportion of inter-island movements are
+swimming events compared to walking on ice? Although the presence and extent of sea ice
+varies annually,  we
+deduce that although the majority of inter-island movements occur between April
+1 and December 31, some inter-island movements appear to be caribou walking on
+sea ice between islands (Figure 2B). Future research on the trade-offs of these types of
+movement should consider both how the costs associated with each differ, but
+also how the potential benefits - foraging opportunities - may vary seasonally.
 
@@
 
-3) Does reproductive status (presence of a calf-at-heel) or demographic class (age or sex) influence propensity for oceanic swimming?
-Our initial observation was of an adult male, but all collared individuals were adult females.
-Are younger caribou more likely to swim if they are unable to compete with older individuals for resources?
-Are males or females without a calf-at-heel more likely to swim in the summer given they are not restricted based on presence of a calf-at-heel, as females are?
+3) Does reproductive status (presence of a calf-at-heel) or demographic class
+(age or sex) influence propensity for oceanic swimming? Our initial observation
+was of an adult male, but all collared individuals were adult females. Are
+younger caribou more likely to swim if they are unable to compete with older
+individuals for resources? Are males or females without a calf-at-heel more
+likely to swim in the summer given they are not restricted based on presence of
+a calf-at-heel, as females are?
 
@@
 
-We surmise that for Fogo Island, as competition among conspecifics increased along with
-population size, density-dependent habitat selection resulted in expansion of
-the population to nearby islands. Following Ideal Free Distribution theory,
-caribou should swim to new islands when the average fitness of individuals on
-the starting island exceeds the density-fitness equilibrium [@Morris1987]. While
-this is an ultimate explanation, our data suggest caribou only remain on smaller
-islands for a few days at a time, so the fitness equilibrium remains a
-theoretical construct. More likely, however, is that fine-scale competition,
-density-dependent habitat selection, and forage depletion drives individuals to
-periodically swim between islands. Our focal observations add to the evidence that caribou
-can, and occasionally do, swim in the ocean [e.g. @Miller1995; @Miller2002;
+We surmise that for Fogo Island, as competition among conspecifics increased
+along with population size, density-dependent habitat selection resulted in
+expansion of the population to nearby islands. Following Ideal Free Distribution
+theory, caribou should swim to new islands when the average fitness of
+individuals on the starting island exceeds the density-fitness equilibrium
+[@Morris1987]. While this is an ultimate explanation, our data suggest caribou
+only remain on smaller islands for a few days at a time, so the fitness
+equilibrium remains a theoretical construct. More likely, however, is that
+fine-scale competition, density-dependent habitat selection, and forage
+depletion drives individuals to periodically swim between islands. Our focal
+observations add to the evidence that caribou can, and occasionally do, swim in
+the ocean [e.g. @Miller1995; @Miller2002;
 @Jeffery2007; @Ricca2012]. We suggest that swimming is likely more common than

### 2020-05-15 paper/manuscript.Rmd
EVW comments

@@
 
+<!-- That first paragraph would benefit from a bit of signpost at its end to
+let us know what kind of ideas you're going go talk about near the end. -->
+
 In total, 12 of 29 collared female caribou swam between islands in the Fogo
@@
 research:
+<!-- a) For the questions that arise, maybe try to generalize the language a bit
+so that they are a bit broader: caribou as model, but emphasize the
+ecological/evolutionary process it will comment on. 
+b.) Consider tweaking the
+intro the the questions less as 'someone could do this' as something along the
+lines of : this has ignited in us the idea that we can tailor our data
+collection in our closely studies system to access some of these questions. -->
 

### 2020-05-15 paper/manuscript.Rmd
tweak Q1

@@
 
-1) How do the energetic cost and risk of swimming differ from walking on sea
-ice? Walking over solid ice is likely less energetically expensive than
-swimming, but may pose a greater risk of falling through depending on the
-thickness and stability of ice. What proportion of inter-island movements are
-swimming events compared to walking on ice? Although the presence and extent of sea ice
-varies annually,  we
-deduce that although the majority of inter-island movements occur between April
-1 and December 31, some inter-island movements appear to be caribou walking on
-sea ice between islands (Figure 2B). Future research on the trade-offs of these types of
-movement should consider both how the costs associated with each differ, but
-also how the potential benefits - foraging opportunities - may vary seasonally.
+1) How do energetic costs and risks of swimming differ from walking on sea ice?
+Walking over ice is less energetically expensive than swimming, but may pose a
+greater risk of mortality from falling through the ice. Although the presence
+and extent of sea ice varies annually,  we deduce that although the majority of
+inter-island movements occur between April 1 and December 31, some inter-island
+movements appear to be caribou walking on sea ice between islands (Figure 2B).
+Our aim for future research on the trade-offs of these types of movement
+considers how the costs, e.g. energy expenditure or mortality, and benefits,
+e.g. novel foraging opportunities, may vary seasonally.
 

### 2020-05-15 paper/manuscript.Rmd
Q5 and Morris citation

@@
 density-dependent habitat selection governed by the Ideal Free Distribution
-[@Morris1987; @Bradbury2015]. Ideal Free Distribution theory predicts a fitness
+[@Morris1987]. Ideal Free Distribution theory predicts a fitness
 equilibrium: when the density in a given habitat patch has exceeded the optimum
 for fitness within that patch, animals should relocate and settle new habitat
-patches, so that fitness is equal across all patches [@Bradbury2015]. Islands
+patches, so that fitness is equal across all patches [@Morris1987]. Islands
 act as discrete habitat patches in this case, but the costs associated with
@@
 5) In relation to all the previous questions: how do individuals vary in their
-assessment of the costs and benefits of swimming between islands? Why did some
-caribou regularly engage in swimming events, while more than half of our marked
-population were never observed swimming in the ocean? The forage limitation
-hypothesis is thought to apply uniformly across populations, but individual
-caribou could evaluate the costs and benefits of swimming differently. It is
-possible that some individuals, in our case ~10%, considered swimming to be less
-costly or perceived competition or predation more acutely than their
-conspecifics.
+propensity to swim and their assessment of the costs and benefits of swimming
+between islands? We observed some individuals that regularly engaged in swimming
+events, while others were never observed swimming in the ocean? Our existing research on individual behavioural variation in caribou movement ecology The forage
+limitation hypothesis is thought to apply uniformly across populations, but
+individual caribou could evaluate the costs and benefits of swimming
+differently. It is possible that some individuals, in our case ~10%, considered
+swimming to be less costly or perceived competition or predation more acutely
+than their conspecifics.
 

### 2020-05-15 paper/manuscript.Rmd
Q5

@@
 between islands? We observed some individuals that regularly engaged in swimming
-events, while others were never observed swimming in the ocean? Our existing research on individual behavioural variation in caribou movement ecology The forage
-limitation hypothesis is thought to apply uniformly across populations, but
-individual caribou could evaluate the costs and benefits of swimming
-differently. It is possible that some individuals, in our case ~10%, considered
-swimming to be less costly or perceived competition or predation more acutely
-than their conspecifics.
+events, while others were never observed swimming in the ocean? Our existing
+research on individual behavioural variation in caribou movement ecology
+]@Webber2020] provides a basis to investigate the role of inherent individual
+differences in swimming behaviour. Specifically, the forage limitation
+hypothesis applies uniformly across populations, but individuals could evaluate
+the costs and benefits of swimming differently.
 

### 2020-05-15 paper/manuscript.Rmd
edit paragraph 1

@@
 approximately 300 woodland caribou. During routine fieldwork on May 30, 2017, we
-observed an unmarked adult male caribou swim between two small islands (Figure
-1, see Figure 2 for location), a distance of at least 470 m which took ~9
-minutes (~52 m per minute). Residents of Fogo Island confirmed this phenomenon
-and have also reported regularly observing caribou swimming between islands.
-This observation prompted us to investigate whether swimming behaviour was
-common among GPS radio-collared caribou in the population [n = 29, for collaring
-details see, @Peignier2019]. We identified an individual's swimming events as
-two consecutive GPS locations (2-hour relocation rates)  on different islands.
-Northeastern Newfoundland typically experiences pack ice during winter and
-caribou are known locally to travel between islands by walking over the ice. We
-restricted our GPS data to the ice-free period of the year (April 1 to December
-31) and identified `r nevents` swimming events over three years (Figure 2).
-
-<!-- That first paragraph would benefit from a bit of signpost at its end to
-let us know what kind of ideas you're going go talk about near the end. -->
+observed an unmarked adult male caribou swim between two small islands (Figure 1
+& 2), a distance of at least 470 m which took ~9 minutes (~52 m per minute).
+Residents of Fogo Island confirmed this phenomenon and have also reported
+regularly observing caribou swimming between islands. This observation prompted
+us to investigate whether swimming behaviour was common among GPS radio-collared
+caribou in the population [n = 29, for collaring details see, @Peignier2019]. We
+identified an individual's swimming events as two consecutive GPS locations
+(2-hour relocation rates) on different islands. Newfoundland typically
+experiences pack ice in late winter and caribou ,may travel between islands by
+walking over the ice. To identify swimming events, we restricted our dataset
+between April 1 and December 31 and identified `r nevents` swimming events over
+three years (Figure 2). Given the prevalence of swimming in this population, we
+discuss the causes and consequences of this behaviour and propose future areas
+of research.
 

### 2020-05-15 paper/manuscript.Rmd
add Canada to affiliation

@@
     ^1^Cognitive and Behavioural Ecology Interdisciplinary Program, Memorial
-   University of Newfoundland
+   University of Newfoundland, Canada
     ^2^Department of Biology, Memorial University of Newfoundland, Canada
@@
 
-Caribou were introduced to Fogo Island (n = 26 animals between 1964–67) and
-Change Island (n = 5 animals in 1964) from Newfoundland as part of a series of
-translocations and introductions throughout the province [@Bergerud1989].
-Typical caribou predators such as black bears (*Ursus americanus*) and wolves
-(*Canis lupus*) are absent from Fogo Island; the dominant predator is coyote
-(*Canis latrans*), which predate calves but it remains unclear what threat they
-pose, if any, to adult caribou [@Rayl2014]. Given the minimal risk of predation
-to adults, we propose that forage limitation is more likely to drive movement
-between islands than predator avoidance. For several decades after their
-introduction to Fogo Island, the caribou population did not exceed ~100
-individuals [@Bergerud1989, Newfoundland and Labrador Wildlife Division,
-unpublished data]. During the 1990s, population density reached ~300 animals and
-anecdotal evidence suggests caribou were first seen on nearby islands during
-this same period. Increased competition for resources on Fogo Island due to this
-increase in density could have precipitated movement to other islands.
+Caribou were introduced to Fogo Island between 1964–67 as part of a series of
+translocations throughout the province [@Bergerud1989]. Typical caribou
+predators such as black bears (*Ursus americanus*) and wolves (*Canis lupus*)
+are absent from Fogo Island; the dominant predator is coyote (*Canis latrans*),
+which predate calves but it remains unclear what threat they pose, if any, to
+adult caribou [@Rayl2014]. Given the minimal risk of predation to adults and low
+density of coyote on Fogo Island, we propose that forage limitation is more
+likely to drive movement between islands than predator avoidance. For several
+decades after their introduction to Fogo Island, the caribou population did not
+exceed ~100 individuals [@Bergerud1989, Newfoundland and Labrador Wildlife
+Division, unpublished data]. During the 1990s, population density reached ~300
+animals and anecdotal evidence suggests caribou were first seen on nearby
+islands during this same period. Increased competition for resources on Fogo
+Island due to this increase in density could have precipitated movement to other
+islands.
 
@@
 
-2) At which temporal and spatial scales do caribou choose to swim
+<!--2) At which temporal and spatial scales do caribou choose to swim
 between islands? Does the time of day, weather, season, or availability
-of food influence swimming?
+of food influence swimming?-->
 
-3) Does reproductive status (presence of a calf-at-heel) or demographic class
+<!--3) Does reproductive status (presence of a calf-at-heel) or demographic class
 (age or sex) influence propensity for oceanic swimming? Our initial observation
@@
 likely to swim in the summer given they are not restricted based on presence of
-a calf-at-heel, as females are?
+a calf-at-heel, as females are? -->
 
-4) If swimming is a function of density-dependent habitat selection, what are
+2) If swimming is a function of density-dependent habitat selection, what are
 the costs of movement between patches? IFD assumes zero-cost movement, but could
@@
 
-5) In relation to all the previous questions: how do individuals vary in their
+3) In relation to questions 1 and 2: how do individuals vary in their
 propensity to swim and their assessment of the costs and benefits of swimming

### 2020-07-16 scripts/5-table1.R
add areas, group size and calc cow

@@ -13,10 +13,18 @@ tab1 <- data.table(
   movto = ,
   movfrom = ,
   proplic = ,
-  density = ,
-  grpsize = ,
-  calfcow = 
+  density = cI,
+  grpsize = c('3.05 (2.7, 3.4)',
+              '2.95 (1.89, 4.02)',
+              '3.71 (1.97, 5.46)'),
+  calfcow = c('0.34, (0.28, 0.39) (n = 283 groups)',
+              '0.22, (0.05, 0.39) (n = 23 groups)',
+              '0.67, (0.23, 1.00) (n = 7 groups)'),
+  area = c(255, 7.5, 3.8)
 )
+
+tab1[, density := density / area]
+tab1[, area := NULL]
 
 setnames(tab1,
          c('',
@@ -24,7 +32,7 @@ setnames(tab1,
            'Moves away',
            'Proportion lichen',
            'Estimated density (caribou per km2)',
-           'Average group size',
+           'Average group size (95% CI)',
            'Calf:cow ratio between June and August'))
 
 

### 2020-07-16 scripts/5-table1.R
read islands and lc

@@ -4,16 +4,24 @@
 
 ### Packages ----
 library(data.table)
+library(raster)
+library(sf)
 
-
+### Data ----
+lc <- raster('../nl-landcover/output/fogo_lc.tif')
+islands <- readRDS('output/islandsPoly.Rds')
 
 ### Table 1 ----
+selislands <- islands[islands$id %in% c(120, 124, 128),]
+
+lc == 8
+
 tab1 <- data.table(
   island = c('Fogo Island', 'Western Perry Island', 'Eastern Perry Island'),
   movto = ,
   movfrom = ,
   proplic = ,
-  density = cI,
+  density = c(250, 50, 20),
   grpsize = c('3.05 (2.7, 3.4)',
               '2.95 (1.89, 4.02)',
               '3.71 (1.97, 5.46)'),

### 2020-07-16 scripts/5-table1.R
sample prop lichen

@@ -12,15 +12,15 @@ lc <- raster('../nl-landcover/output/fogo_lc.tif')
 islands <- readRDS('output/islandsPoly.Rds')
 
 ### Table 1 ----
-selislands <- islands[islands$id %in% c(120, 124, 128),]
+selislands <- islands[islands$id %in% c(120, 128, 124),]
 
-lc == 8
+selislands$proplichen <- extract(lc == 8, selislands, fun = mean, na.rm = TRUE)
 
 tab1 <- data.table(
   island = c('Fogo Island', 'Western Perry Island', 'Eastern Perry Island'),
   movto = ,
   movfrom = ,
-  proplic = ,
+  proplic = round(selislands$proplichen[c(1, 3, 2)], 2),
   density = c(250, 50, 20),
   grpsize = c('3.05 (2.7, 3.4)',
               '2.95 (1.89, 4.02)',

### 2020-07-16 scripts/5-table1.R
counts to and from

@@ -10,16 +10,26 @@ library(sf)
 ### Data ----
 lc <- raster('../nl-landcover/output/fogo_lc.tif')
 islands <- readRDS('output/islandsPoly.Rds')
+edges <- readRDS('output/island-edges.Rds')
 
 ### Table 1 ----
-selislands <- islands[islands$id %in% c(120, 128, 124),]
+only <- c(120, 128, 124)
+selislands <- islands[islands$id %in% only,]
 
 selislands$proplichen <- extract(lc == 8, selislands, fun = mean, na.rm = TRUE)
 
+counts <- rbindlist(lapply(only, function(o) {
+  data.table(island = o,
+             from = edges[island == o, .N],
+             to = edges[endisland == o, .N])
+}))
+
+
+
 tab1 <- data.table(
   island = c('Fogo Island', 'Western Perry Island', 'Eastern Perry Island'),
-  movto = ,
-  movfrom = ,
+  movto = counts$to,
+  movfrom = counts$from,
   proplic = round(selislands$proplichen[c(1, 3, 2)], 2),
   density = c(250, 50, 20),
   grpsize = c('3.05 (2.7, 3.4)',

### 2020-09-01 scripts/6-render-md.R
fst 6-render-md.R

@@ -1,0 +1,1 @@
+function() rmarkdown::render('paper/manuscript.Rmd')

### 2020-09-01 scripts/old/3-extractArea.R
rm old/

@@ -1,51 +1,0 @@
-### Island area vs ====
-# Quinn M.R. Webber, Jack G. Hendrix, Alec L. Robitaille,  Eric Vander Wal
-
-### Packages ----
-libs <- c('data.table', 'ggplot2', 'dplyr',
-          'raster', 'prettymapr', 'rosm', 
-          'stars', 'fasterize', 'rgdal')
-lapply(libs, require, character.only = TRUE)
-
-## load files
-duration <- fread("output/duration.csv")
-r <- readRDS('output/islandsPoly.RDS')
-
-## subset duration data
-duration <- duration[, c("ANIMAL_ID","Year","StartIsland","counter")]
-
-
-
-## calculate area
-
-
-areaIsl <- data.table(st_area(r))
-areaIsl$V1 <- areaIsl$V1/100000
-areaIsl$StartIsland <- rownames(areaIsl)
-areaIsl$StartIsland[areaIsl$StartIsland == 43] <- "Fogo"
-#areaIsl$StartIsland[areaIsl$StartIsland == 53] <- "North Long"
-areaIsl$StartIsland[areaIsl$StartIsland == 55] <- "North Long"
-areaIsl$StartIsland[areaIsl$StartIsland == 58] <- "Blundon"
-areaIsl$StartIsland[areaIsl$StartIsland == 67] <- "Brother"
-areaIsl$StartIsland[areaIsl$StartIsland == 68] <- "W. Indian"
-areaIsl$StartIsland[areaIsl$StartIsland == 70] <- "South Long"
-areaIsl$StartIsland[areaIsl$StartIsland == 71] <- "E. Indian"
-areaIsl$StartIsland[areaIsl$StartIsland == 74] <- "Kate"
-
-areaIsl <- areaIsl[StartIsland == "Fogo" | StartIsland == "North Long" |
-                   StartIsland == "Blundon" | StartIsland == "Brother" | 
-                   StartIsland == "W. Indian" | StartIsland == "South Long" |
-                   StartIsland == "E. Indian" | StartIsland == "Kate"]
-areaIsl$
-
-  strsplit(areaIsl$V1, " ")  
-  
-duration2 <- merge(duration, areaIsl, by = "StartIsland")
-
-ggplot(duration) +
-  geom_histogram(aes(log(counter/12), fill = ANIMAL_ID), alpha = 0.5) +
-  facet_wrap(~StartIsland, scale = 'free')
-
-ggplot(duration2) +
[…]

### 2020-09-01 scripts/old/maps.R

@@ -1,59 +1,0 @@
-### Maps ----
-mapview(
-  caribou[ANIMAL_ID %in% edges$ANIMAL_ID],
-  xcol = 'EASTING',
-  ycol = 'NORTHING',
-  zcol = 'island',
-  crs = utm
-)
-
-
-mapview(
-  edges,
-  xcol = 'endislandEAST',
-  ycol = 'endislandNORTH',
-  zcol = 'endisland',
-  crs = utm
-)
-
-mapview(
-  edges,
-  xcol = 'EASTING',
-  ycol = 'NORTHING',
-  zcol = 'island',
-  crs = utm
-)
-
-mapview(
-  caribou[!is.na(island)],
-  xcol = 'EASTING',
-  ycol = 'NORTHING',
-  zcol = 'ANIMAL_ID',
-  crs = utm
-)
-
-mapview(
-  duration,
-  xcol = 'EASTING',
-  ycol = 'NORTHING',
-  zcol = 'ANIMAL_ID',
-  crs = utm
-)
-
-mapview(
-  caribou[island != 32280],
-  xcol = 'EASTING',
-  ycol = 'NORTHING',
-  zcol = 'ANIMAL_ID',
-  crs = utm
-)
-
-
-mapview(
-  caribou[is.na(island)],
-  xcol = 'EASTING',
-  ycol = 'NORTHING',
-  zcol = 'ANIMAL_ID',
-  crs = utm
-)
-

### 2020-09-01 scripts/plan.R
fst plan

@@ -1,0 +1,21 @@
+# Load drake
+library(drake)
+
+# Code to function
+get_islands <- code_to_function('scripts/1-osm-islands.R')
+extract_island_locs <- code_to_function('scripts/2-extract-island-locs.R')
+generate_edges <- code_to_function('scripts/3-generate-edges.R')
+fig_2 <- code_to_function('scripts/4-figure2.R')
+table_1 <- code_to_function('scripts/5-table1.R')
+render_md <- code_to_function('scripts/6-render-md.R')
+
+
+# The plan
+plan <- drake_plan(
+  islands = get_islands(),
+  locs = extract_island_locs(islands),
+  edges = generate_edges(island_locs),
+  fig2 = fig_2(edges),
+  tab1 = table_1(edges),
+  md = render_md(fig2, tab1)
+)

### 2020-09-01 scripts/plan.R
fix plan

@@ -14,7 +14,7 @@ render_md <- code_to_function('scripts/6-render-md.R')
 plan <- drake_plan(
   islands = get_islands(),
   locs = extract_island_locs(islands),
-  edges = generate_edges(island_locs),
+  edges = generate_edges(locs),
   fig2 = fig_2(edges),
   tab1 = table_1(edges),
   md = render_md(fig2, tab1)

### 2020-09-01 paper/manuscript.Rmd
add kable to manuscript
rm double output

@@
 output:
-   github_document: default
    bookdown::word_document2:
      reference_docx: template.docx
-knit: (function(inputFile, encoding) {
-  rmarkdown::render(inputFile, encoding = encoding,
-  output_format = "all") })
 bibliography: ../references/swim.bib
@@
 
+```{r tab1}
+knitr::kable(readRDS('../output/table1.Rds'),
+             caption = '')
+```
+
+\newpage
+
 **Literature Cited**

### 2020-09-01 scripts/6-render-md.R
rm function()

@@ -1,1 +1,1 @@
-function() rmarkdown::render('paper/manuscript.Rmd')
+rmarkdown::render('paper/manuscript.Rmd')

### 2020-09-01 paper/manuscript.Rmd

@@
 title: Interisland movements and oceanic swimming of woodland caribou in Newfoundland
-author: Quinn M.R. Webber^1\*^, Jack G. Hendrix^1^, Alec L. Robitaille^2^,  Eric Vander Wal^1,2^
-date: 
-    ^1^Cognitive and Behavioural Ecology Interdisciplinary Program, Memorial
-   University of Newfoundland, Canada
-    ^2^Department of Biology, Memorial University of Newfoundland, Canada
-    #\*Corresponding author: Quinn M.R. Webber (webber.quinn@gmail.com)
+author:
+  - name: Quinn M.R. Webber
+    institute: cabe
+    correspondence: true
+  - name: Jack G. Hendrix
+    institute: cabe
+  - name: Alec L. Robitaille
+    email: robit.alec@gmail.com
+    institute: bio
+  - name: Eric Vander Wal
+    institute: [bio, cabe]
+institute: 
+  - bio: Department of Biology, Memorial University of Newfoundland, Canada
+  - cabe: Cognitive and Behavioural Ecology Interdisciplinary Program, Memorial University of Newfoundland 
 output:
@@
      reference_docx: template.docx
+     pandoc_args:
+      - '--lua-filter=scholarly-metadata.lua'
+      - '--lua-filter=author-info-blocks.lua'
 bibliography: ../references/swim.bib

### 2020-09-01 scripts/1-osm-islands.R
add island area

@@ -6,6 +6,7 @@
 libs <- c('data.table',
           'sf',
           'osmdata',
+          'lwgeom',
           'fasterize')
 lapply(libs, require, character.only = TRUE)
 
@@ -36,6 +37,8 @@ islands <- st_as_sf(c(st_geometry(polys), castpolys))
 # Basic island id
 islands$id <- seq.int(length.out = nrow(islands))
 
+# Area
+islands$area <- st_area(islands)
 
 ### Reproject islands ----
 # Projections

### 2021-02-12 scripts/5-figure2.R
fig2

@@ -202,12 +202,36 @@ runarea$logIsland <- log10(runarea$islandlen)
 runarea$logArea <- log10(runarea$area/1000000)
 
 
-# run linear model
-a1 <- lm(islandlen~area2, data = runarea)
-summary(a1)
+## Type II curve
+m4 <- drc::drm(area2 ~ islandlen, data = runarea[area2 < 200 & islandlen < 100], fct = MM.2())
+summary(m4)
+plot(m4)
+
+demo.fits <- expand.grid(conc=seq(min(runarea[area2 < 200 & islandlen < 100]$islandlen), max(runarea[area2 < 200 & islandlen < 100]$islandlen), length=100))
+# new data with predictions
+pm <- predict(m4, newdata=demo.fits, interval="confidence") 
+demo.fits$p <- pm[,1]
+demo.fits$pmin <- pm[,2]
+demo.fits$pmax <- pm[,3]
+
+## no log
+aa <- ggplot() +
+  geom_jitter(data = runarea[area2 < 200 & islandlen < 100], 
+              aes(x = islandlen, y = area2),
+              height = 0.2,
+              alpha = 0.5) +
+  geom_ribbon(data=demo.fits, aes(x = conc, y = p, ymin=pmin, ymax=pmax), alpha=0.2) +
+  geom_line(data=demo.fits, aes(x = conc, y = p)) +
+  #xlim(1, 1000) +
+  #ylim(1, 500) +
+  labs(x = 'Residency time (days)',
+       y = expression("Area of island"~km^2)) +
+  ggtitle('A) Not log-transformed') +
+  #coord_trans(x="log", y = "log") +
+  themeHist
 
 # extract CIs
-estMod<-Effect(c( "area2"), partial.residuals = T, a1)
+estMod<-predict(m4) 
 predMod <- data.table(residency = estMod$fit, 
                       area = estMod$x, 
                       lwr = estMod$lower,
@@ -215,31 +239,32 @@ predMod <- data.table(residency = estMod$fit,
 colnames(predMod) <- c("residency","area" ,"lwr", "upr")
 
 labels <- data.frame(
-  x = c(0.001, 0.1, 5, 200),
-  y = c(1, 5, 800, 2500),
+  y = c(0.001, 0.1, 3, 100),
+  x = c(4, 100, 1200, 2000),
   text = c("N. Coastal Islands", "S. Coastal Islands", "Perry Islands", "Fogo Island")
 )
 
 (resTime <- ggplot() +
[…]

### 2021-04-12 paper/manuscript.Rmd
median > mean
new title

@@
 ---
-title: Interisland movements and oceanic swimming of woodland caribou in Newfoundland
+title: On the marginal value of swimming in woodland caribou
 author:
@@
 # Mean per individual
-neventidMean <- round(mean(neventid$N))
+neventidMedian <- round(median(neventid$N))
 

### 2021-04-12 paper/manuscript.Rmd
new P1

@@
 
-The Fogo Island archipelago, off the coast of Newfoundland, Canada, is home to
-approximately 300 woodland caribou. During routine fieldwork on May 30, 2017, we
-observed an unmarked adult male caribou swim between two small islands (Figure 1
-& 2), a distance of at least 470 m which took ~9 minutes (~52 m per minute).
-Residents of Fogo Island confirmed this phenomenon and have also reported
-regularly observing caribou swimming between islands. This observation prompted
-us to investigate whether swimming behaviour was common among GPS radio-collared
-caribou in the population [n = 29, for collaring details see, @Peignier2019]. We
-identified an individual's swimming events as two consecutive GPS locations
-(2-hour relocation rates) on different islands. Newfoundland typically
-experiences pack ice in late winter and caribou ,may travel between islands by
-walking over the ice. To identify swimming events, we restricted our dataset
-between April 1 and December 31 and identified `r nevents` swimming events over
-three years (Figure 2). Given the prevalence of swimming in this population, we
-discuss the causes and consequences of this behaviour and propose future areas
-of research.
+Movement is costly.
+Even for adept swimmers like caribou (Rangifer tarandus), which have hollow hair shafts that aid in flotation, the energetic expenditure of swimming for terrestrial mammals is markedly higher than walking or running (Miller and Gunn 1985).
+The drivers that promote the decision to swim for terrestrial mammals, like caribou, are often related to the profitability of forage (Miller et al.
+1977) and the risks of predation (Jeffery et al.
+2007, Jordan et al.
+2010).
+The Fogo Island archipelago, off the coast of Newfoundland, Canada, is home to approximately 300 woodland caribou, descendants of 26 individuals introduced from the Island of Newfoundland to Fogo Island between 1964–67 (Bergerud and Mercer 1989).
+During fieldwork on May 30, 2017, we observed an unmarked adult male caribou swim between two smaller islands (Figure 1) a distance of 470 m which took approximately 9 minutes.
+Given that swimming is energetically expensive, we ask how often do caribou swim between islands, how long do they occupy islands, and ultimately what motivates swimming?
+Here, we frame swimming and island residency as a model of patch use as described by the marginal value theorem (Charnov 1976) within the context of the forage limitation hypothesis.
+For caribou, the forage limitation hypothesis is the constraints on food availability associated with changes in abundance through time (Klein 1991, Schaefer et al.
+2016).
+Here, we integrate our understanding of forage limitation in caribou populations by using island residency as a proxy for time in patch and island size as a proxy for energy intake in a patch (Figure 2).
 
+`r nevents` 
 In total, 12 of 29 collared female caribou swam between islands in the Fogo

### 2021-04-12 paper/manuscript.Rmd
new P2

@@
 Here, we integrate our understanding of forage limitation in caribou populations by using island residency as a proxy for time in patch and island size as a proxy for energy intake in a patch (Figure 2).
+
+Forage limitation and over-grazing is a concern for caribou populations and can lead to reduced female reproductive success (Klein 1991, Schaefer et al.
+2016).
+Forage scarcity has been proposed as a potential reason for caribou moving between arctic islands on the sea-ice in winter (Miller et al.
+1977).
+We submit that testing the marginal value theorem could improve our understanding of forage limitation.
+The marginal value theorem posits that animals will forage optimally by depleting the resources in a given patch (i.e.
+island) before moving to a new patch (Charnov 1976).
+Depletion of resources on small islands could therefore mirror patterns of over-grazing and forage limitation that occur for caribou herds at higher densities in Newfoundland (Schaefer et al.
+2016) and elsewhere in their range (Klein 1991).
 

### 2021-04-12 paper/manuscript.Rmd
Update manuscript.Rmd
P3

@@
 Movement is costly.
-Even for adept swimmers like caribou (Rangifer tarandus), which have hollow hair shafts that aid in flotation, the energetic expenditure of swimming for terrestrial mammals is markedly higher than walking or running (Miller and Gunn 1985).
+Even for adept swimmers like caribou (*Rangifer tarandus*), which have hollow hair shafts that aid in flotation, the energetic expenditure of swimming for terrestrial mammals is markedly higher than walking or running (Miller and Gunn 1985).
 The drivers that promote the decision to swim for terrestrial mammals, like caribou, are often related to the profitability of forage (Miller et al.
@@
 
-Forage limitation and over-grazing is a concern for caribou populations and can lead to reduced female reproductive success (Klein 1991, Schaefer et al.
-2016).
-Forage scarcity has been proposed as a potential reason for caribou moving between arctic islands on the sea-ice in winter (Miller et al.
-1977).
-We submit that testing the marginal value theorem could improve our understanding of forage limitation.
-The marginal value theorem posits that animals will forage optimally by depleting the resources in a given patch (i.e.
-island) before moving to a new patch (Charnov 1976).
-Depletion of resources on small islands could therefore mirror patterns of over-grazing and forage limitation that occur for caribou herds at higher densities in Newfoundland (Schaefer et al.
-2016) and elsewhere in their range (Klein 1991).
+Forage limitation and over-grazing is a concern for caribou populations and can lead to reduced female
+reproductive success (Klein 1991, Schaefer et al. 2016). Forage scarcity has been proposed as a
+potential reason for caribou moving between arctic islands on the sea-ice in winter (Miller et al.
+1977). We submit that testing the marginal value theorem could improve our understanding of forage
+limitation. The marginal value theorem posits that animals will forage optimally by depleting the
+resources in a given patch (i.e. island) before moving to a new patch (Charnov 1976). Depletion of
+resources on small islands could therefore mirror patterns of over-grazing and forage limitation that
+occur for caribou herds at higher densities in Newfoundland (Schaefer et al. 2016) and elsewhere in
+their range (Klein 1991).
+
+Our focus on forage is due to the limited presence of predators in our system. Throughout their range,
+wolves (*Canis lupus*) are the primary predator of caribou, although there are no wolves on the Island
+of Newfoundland, predation on adult caribou by coyotes (*Canis latrans*) does occur (Lewis and Mahoney
+2014). Within the Fogo Island archipelago, coyotes exist at low density (Huang et al. 2021) and the
+probability of occurrence decreases on the smaller islands. The low density of predators throughout
+the archipelago is therefore an unlikely driver of swimming in our system, though predation may stillinfluence swimming behaviour for caribou elsewhere in their range (e.g. Jeffery et al. 2007).
 

### 2021-04-12 paper/manuscript.Rmd
Update manuscript.Rmd

@@
 
-Movement is costly.
-Even for adept swimmers like caribou (*Rangifer tarandus*), which have hollow hair shafts that aid in flotation, the energetic expenditure of swimming for terrestrial mammals is markedly higher than walking or running (Miller and Gunn 1985).
-The drivers that promote the decision to swim for terrestrial mammals, like caribou, are often related to the profitability of forage (Miller et al.
-1977) and the risks of predation (Jeffery et al.
-2007, Jordan et al.
-2010).
-The Fogo Island archipelago, off the coast of Newfoundland, Canada, is home to approximately 300 woodland caribou, descendants of 26 individuals introduced from the Island of Newfoundland to Fogo Island between 1964–67 (Bergerud and Mercer 1989).
-During fieldwork on May 30, 2017, we observed an unmarked adult male caribou swim between two smaller islands (Figure 1) a distance of 470 m which took approximately 9 minutes.
-Given that swimming is energetically expensive, we ask how often do caribou swim between islands, how long do they occupy islands, and ultimately what motivates swimming?
-Here, we frame swimming and island residency as a model of patch use as described by the marginal value theorem (Charnov 1976) within the context of the forage limitation hypothesis.
-For caribou, the forage limitation hypothesis is the constraints on food availability associated with changes in abundance through time (Klein 1991, Schaefer et al.
-2016).
-Here, we integrate our understanding of forage limitation in caribou populations by using island residency as a proxy for time in patch and island size as a proxy for energy intake in a patch (Figure 2).
+Movement is costly. Even for adept swimmers like caribou (*Rangifer tarandus*), which have hollow hair
+shafts that aid in flotation, the energetic expenditure of swimming for terrestrial mammals is
+markedly higher than walking or running (Miller and Gunn 1985). The drivers that promote the decision
+to swim for terrestrial mammals, like caribou, are often related to the profitability of forage
+(Miller et al. 1977) and the risks of predation (Jeffery et al. 2007, Jordan et al. 2010). The Fogo
+Island archipelago, off the coast of Newfoundland, Canada, is home to approximately 300 woodland
+caribou, descendants of 26 individuals introduced from the Island of Newfoundland to Fogo Island
+between 1964–67 (Bergerud and Mercer 1989). During fieldwork on May 30, 2017, we observed an unmarked
+adult male caribou swim between two smaller islands (Figure 1) a distance of 470 m which took
+approximately 9 minutes. Given that swimming is energetically expensive, we ask how often do caribou
+swim between islands, how long do they occupy islands, and ultimately what motivates swimming? Here,
+we frame swimming and island residency as a model of patch use as described by the marginal value
+theorem (Charnov 1976) within the context of the forage limitation hypothesis. For caribou, the forage
+limitation hypothesis is the constraints on food availability associated with changes in abundance
+through time (Klein 1991, Schaefer et al. 2016). Here, we integrate our understanding of forage
+limitation in caribou populations by using island residency as a proxy for time in patch and island
+size as a proxy for energy intake in a patch (Figure 2).
 
@@
 probability of occurrence decreases on the smaller islands. The low density of predators throughout
-the archipelago is therefore an unlikely driver of swimming in our system, though predation may stillinfluence swimming behaviour for caribou elsewhere in their range (e.g. Jeffery et al. 2007).
+the archipelago is therefore an unlikely driver of swimming in our system, though predation may
+still influence swimming behaviour for caribou elsewhere in their range (e.g. Jeffery et al. 2007).
 

### 2021-04-12 paper/manuscript.Rmd
Update manuscript.Rmd

@@
 still influence swimming behaviour for caribou elsewhere in their range (e.g. Jeffery et al. 2007).
+
+We identified swimming events as two consecutive GPS locations (2-hour relocation rates) on different
+islands. Newfoundland typically experiences pack ice in late winter and caribou may travel between
+islands by walking over the ice. We restricted our dataset between April 1 and December 31 to
+distinguish swimming from walking on ice. We calculated island residency as the duration of time,
+measured on 2-hour relocation rates, an individual caribou was observed on an island. Residency was
+therefore a proxy for when the benefits of foraging no longer exceeded the costs of swimming between
+islands. Our test of the marginal value theorem for the case of oceanic swimming by caribou is based
+on three assumptions: (1) resources are proportional to the size of each island; (2) caribou know
+approximate value of the resources on an island before they swim to and from that island; and (3) the
+cost of swimming to and from an island is less than the value of resources on that island when the
+caribou departs. Should these assumptions hold, we would predict a positive relationship where caribou
+depart smaller islands sooner than larger islands.
 

### 2021-04-12 paper/manuscript.Rmd
P4

@@
 
-`r nevents` 
+Swimming behavior was common among GPS radio-collared caribou in the population (n = 29, for collaring
+details see Peignier et al. 2019). We identified `r nevents` swimming events over three years (Figure
+2). In total, 13 of 29 collared female caribou swam among islands (Figure 2). Of the caribou that
+swam, caribou swam on average approximately `r neventidMean` (range = `r neventidMin`-`r neventidMax`)
+times per year and remained on each island for a median of `r dur` (range = 
+`r durMin`-`r durMax`) days before swimming again. We found that caribou residency increased
+proportionally with the size of the island, such that caribou spent less time on smaller islands and
+more time on larger islands, providing preliminary evidence for links between forage limitation and
+the marginal value theorem (Figure 2B).
+
 In total, 12 of 29 collared female caribou swam between islands in the Fogo

### 2021-04-12 paper/manuscript.Rmd
P5 - questions

@@
 the marginal value theorem (Figure 2B).
+
+The apparent relationship between island size and residency creates a framework to test our 
+assumptions about the marginal value of resource use by caribou. We extend this framework to test
+prospective mechanisms that drive movement by caribou more generally, and swimming specifically. For
+example:
+1)	If caribou assess the relative value of resources in a patch, or on an island, we predict that residency should increase as the value of resources elsewhere becomes depressed.
+Longer relative residency might be an indicator of forage limitation and variation in residency could be a predictor of fitness costs and reduced reproductive success.
+2)	Density-dependent habitat selection (Fretwell and Lucas 1969, Morris 2003) could be invoked to explain what drives caribou to swim and expand their ranges to include smaller islands.
+As caribou density increases on larger islands, competition for resources increases such that animals may have similar fitness benefits by selecting islands that at lower density would not be as profitable.
+At high local density, individual residency may decrease, thus, increasing the frequency of swimming.
+As such, density-dependent habitat selection may represent an ultimate explanation for why caribou swim between islands.
+3)	Cognition is an important proximate mechanism that drives swimming behaviour.
+For example, caribou primarily rely on their sense of hearing and smell thus may not see or perceive an island when they enter the water to swim to it.
+How then do caribou know about the presence, and subsequently the value, of an island?
+Social and spatial cognition are almost certainly involved and tests of the role of cognition could help elucidate why caribou swim.
+For example, a novice individual may trust an experienced individual to lead them to a new island, a result which is driven by aspects of social cognition (Seyfarth and Cheney 2015).
+Meanwhile, once an individual has visited an island, it may become part of the cognitive map used to navigate within their home range (e.g.
+African elephants Loxondonta africana Presotto et al.
+2019).
 

### 2021-04-12 paper/manuscript.Rmd
cut old ms

@@
 2). In total, 13 of 29 collared female caribou swam among islands (Figure 2). Of the caribou that
-swam, caribou swam on average approximately `r neventidMean` (range = `r neventidMin`-`r neventidMax`)
-times per year and remained on each island for a median of `r dur` (range = 
+swam, caribou swam on average approximately `r neventidMedian` (range = `r neventidMin`-` neventidMax`) times per year and remained on each island for a median of `r dur` (range = 
 `r durMin`-`r durMax`) days before swimming again. We found that caribou residency increased
@@
 
-In total, 12 of 29 collared female caribou swam between islands in the Fogo
-Island archipelago. Of the caribou that swam, caribou swam on average
-approximately `r neventidMean` (range = `r neventidMin`-`r neventidMax`) times
-per year and remained on each island for approximately `r dur` (range = 
-`r durMin`-`r durMax`) days before swimming again. Based on GPS data to infer
-swimming, we separated the population into three apparent groups of individuals:
-those that swam regularly between islands (n = 3, every ~`r indMean` days);
-those that rarely engaged in swimming events (n = 9, every ~ `r ind2Mean` days);
-and those that never swam (n = 17).
-
-Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
-that caribou swim in streams, rivers, and lakes during migration [@Leblond2016],
-to avoid predators [@Bergerud1985], and to access islands during calving
-[@Bergerud1990]. Even for adept swimmers like caribou, which have hollow hair
-shafts that aid in flotation, the energetic expenditure of swimming for
-quadrupedal mammals is significantly higher than walking or running, and
-drowning is also possible [@Miller1985]. Despite the abundance of coastal and
-island caribou herds, only a few studies have documented caribou swimming
-in the ocean, and most of these lack observation, but rather deduce swimming
-based on presence of caribou on a previously uninhabitated island  [e.g.
-@Miller1995; @Miller2002; @Jeffery2007; @Ricca2012]. Caribou may swim between
-islands in the ocean for similar reasons they swim in freshwater [@Leblond2016],
-mainly concerning predator avoidance or to access new habitats and foraging
-opportunities. We propose that one such explanation, the forage limitation
-hypothesis, is the most likely explanation for oceanic swimming for caribou that
-live on islands in the Fogo Island archipelago.
-
-Caribou were introduced to Fogo Island between 1964–67 as part of a series of
-translocations throughout the province [@Bergerud1989]. Typical caribou
-predators such as black bears (*Ursus americanus*) and wolves (*Canis lupus*)
-are absent from Fogo Island; the dominant predator is coyote (*Canis latrans*),
-which predate calves but it remains unclear what threat they pose, if any, to
-adult caribou [@Rayl2014]. Given the minimal risk of predation to adults and low
-density of coyote on Fogo Island, we propose that forage limitation is more
-likely to drive movement between islands than predator avoidance. For several
-decades after their introduction to Fogo Island, the caribou population did not
-exceed ~100 individuals [@Bergerud1989, Newfoundland and Labrador Wildlife
-Division, unpublished data]. During the 1990s, population density reached ~300
-animals and anecdotal evidence suggests caribou were first seen on nearby
-islands during this same period. Increased competition for resources on Fogo
-Island due to this increase in density could have precipitated movement to other
-islands.
-
-Forage limitation and over-grazing is a major concern for caribou populations
-and can lead to reduced female reproductive success [@Schaefer2016]. Forage
-scarcity has been proposed as a potential reason for caribou moving between
-arctic islands on the sea-ice in winter [@Miller1977]. For the Fogo Island
-population, as the population grew and predation pressure was insufficient to
-maintain population density, forage depletion could lead caribou to make
-similar, but riskier and more costly, movements between islands during the
-ice-free season. Given the small size of other islands near Fogo (Figure 2),
-over-grazing by newly arrived caribou could rapidly deplete forage on an annual
-cycle [@Bergerud1990], leading to more frequent movements back and forth between
-islands rather than permanent relocation.
-
-The trade-off associated with swimming between islands can be understood as
-density-dependent habitat selection governed by the Ideal Free Distribution
-[@Morris1987]. Ideal Free Distribution theory predicts a fitness
-equilibrium: when the density in a given habitat patch has exceeded the optimum
-for fitness within that patch, animals should relocate and settle new habitat
-patches, so that fitness is equal across all patches [@Morris1987]. Islands
-act as discrete habitat patches in this case, but the costs associated with
-swimming create a trade-off when moving between patches. For caribou in the Fogo
-Island archipelago, the role of forage limitation in habitat selection patterns
-remains unknown, but swimming between islands may be at least partially governed
-by density-dependent habitat selection.
-
-Forage limitation may be the ultimate explanation for caribou swimming between
-oceanic islands, but a number of proximate mechanisms remain untested. Our
-observations of caribou swimming in the ocean lead to five questions for future
-research:
-<!-- a) For the questions that arise, maybe try to generalize the language a bit
-so that they are a bit broader: caribou as model, but emphasize the
-ecological/evolutionary process it will comment on. 
-b.) Consider tweaking the
-intro the the questions less as 'someone could do this' as something along the
-lines of : this has ignited in us the idea that we can tailor our data
-collection in our closely studies system to access some of these questions. -->
-
-1) How do energetic costs and risks of swimming differ from walking on sea ice?
-Walking over ice is less energetically expensive than swimming, but may pose a
-greater risk of mortality from falling through the ice. Although the presence
-and extent of sea ice varies annually,  we deduce that although the majority of
-inter-island movements occur between April 1 and December 31, some inter-island
-movements appear to be caribou walking on sea ice between islands (Figure 2B).
-Our aim for future research on the trade-offs of these types of movement
-considers how the costs, e.g. energy expenditure or mortality, and benefits,
-e.g. novel foraging opportunities, may vary seasonally.
-
-<!--2) At which temporal and spatial scales do caribou choose to swim
-between islands? Does the time of day, weather, season, or availability
-of food influence swimming?-->
-
-<!--3) Does reproductive status (presence of a calf-at-heel) or demographic class
-(age or sex) influence propensity for oceanic swimming? Our initial observation
-was of an adult male, but all collared individuals were adult females. Are
-younger caribou more likely to swim if they are unable to compete with older
-individuals for resources? Are males or females without a calf-at-heel more
-likely to swim in the summer given they are not restricted based on presence of
-a calf-at-heel, as females are? -->
-
-2) If swimming is a function of density-dependent habitat selection, what are
-the costs of movement between patches? IFD assumes zero-cost movement, but could
-the risk and energetic costs associated with oceanic swimming be estimated using
-the potential fitness deficit that caribou are willing to tolerate before
-swimming to a new island?
-
-3) In relation to questions 1 and 2: how do individuals vary in their
-propensity to swim and their assessment of the costs and benefits of swimming
-between islands? We observed some individuals that regularly engaged in swimming
-events, while others were never observed swimming in the ocean? Our existing
-research on individual behavioural variation in caribou movement ecology
-]@Webber2020] provides a basis to investigate the role of inherent individual
-differences in swimming behaviour. Specifically, the forage limitation
-hypothesis applies uniformly across populations, but individuals could evaluate
-the costs and benefits of swimming differently.
-
-We surmise that for Fogo Island, as competition among conspecifics increased
-along with population size, density-dependent habitat selection resulted in
-expansion of the population to nearby islands. Following Ideal Free Distribution
-theory, caribou should swim to new islands when the average fitness of
-individuals on the starting island exceeds the density-fitness equilibrium
-[@Morris1987]. While this is an ultimate explanation, our data suggest caribou
-only remain on smaller islands for a few days at a time, so the fitness
-equilibrium remains a theoretical construct. More likely, however, is that
-fine-scale competition, density-dependent habitat selection, and forage
-depletion drives individuals to periodically swim between islands. Our focal
-observations add to the evidence that caribou can, and occasionally do, swim in
-the ocean [e.g. @Miller1995; @Miller2002;
-@Jeffery2007; @Ricca2012]. We suggest that swimming is likely more common than
-previously thought for caribou living on oceanic islands, and that forage
-limitation and the associated density-dependent habitat selection is an ultimate
-explanation for this phenomenon. Although our inference is limited to
-observations, islands appear to represent discrete foraging patches for
-terrestrial animals that can influence fitness via increased foraging
-opportunities. We also present several hypotheses related to the nuances and
-further variation within the idea that density-dependent habitat selection
-governs swimming behaviour, and posit that above and beyond these additional
-factors, individuals can vary in their evaluation of costs and benefits of
-movement between these habitat patches.
-
 **Acknowledgements**
@@
 ```{r, fig.cap = '(ref:fig1)', fig.width=4}
-knitr::include_graphics('../graphics/IMG_1562_crop_edit.jpg')
+knitr::include_graphics('../graphics/Fig1.jpg')
 ```

### 2021-04-12 paper/manuscript.Rmd
P6

@@
 
+From our natural history observation and proximal movement data on caribou swimming, characteristics
+of destination islands, and residency on the islands, we have presented some evidence for the marginal
+value of resources that may serve as for evidence of forage limitation. Further investigation into the
+nuance of forage profitability, local density, and the role of both social and spatial cognition is
+needed. Here, we have proposed a thought exercise that integrates the marginal value theorem with a
+novel natural history observation to explain swimming behavior for terrestrial animals and opens the
+door for future studies of caribou in our system and other island archipelagos inhabited by caribou.
+
 **Acknowledgements**

### 2021-04-12 paper/manuscript.Rmd
acknowledgements

@@
 **Acknowledgements**
-We thank M. Laforge, M. Bonar, C. Hart, and R. Huang for help in the field.
-Logistical support was provided by L. Bixby. We thank all members of the
-Wildlife Evolutionary Ecology Lab, including C. Hart, C. Prokopenko, J. Kennah,
-J.W. Turner, and S. Boyle for their comments on previous versions of this
-manuscript. Newfoundland Wildlife Division including S. Moores, B. Adams, C.
-Doucet, W. Barney, and J. Neville for logistical support in the field as well as
-help with data data collection and management. We thank T. Bergerud and S.
-Mahoney for their vision in initiating much of the work on caribou in
-Newfoundland. Funding for this study was provided by the National Sciences and
-Engineering Research Council (QMRW, JGH, EVW). We respectfully acknowledge the
-territory in which data were collected and analyzed as the ancestral homelands
-of the Beothuk, and the island of Newfoundland as the ancestral homelands of the
+We thank M. Laforge, M. Bonar, C. Hart, and R. Huang for help in the field. Logistical support was
+provided by L. Bixby. We thank J. Schaefer and one anonymous reviewer as well as all members of the
+Wildlife Evolutionary Ecology Lab for their comments on previous versions of this manuscript.
+Newfoundland Wildlife Division including S. Moores, B. Adams, C. Doucet, W. Barney, and J. Neville for
+logistical support in the field as well as help with data data collection and management. We thank T.
+Bergerud and S. Mahoney for their vision in initiating much of the work on caribou in Newfoundland.
+Funding for this study was provided by the Natural Sciences and Engineering Research Council (QMRW,
+JGH, EVW). We respectfully acknowledge the territory in which data were collected and analyzed as the
+ancestral homelands of the Beothuk, and the island of Newfoundland as the ancestral homelands of the
 Mi’kmaq and Beothuk.

### 2021-04-12 paper/manuscript.Rmd
fig1 caption
Fig2 caption

@@
 
-(ref:fig1) Photograph of swimming unmarked adult male caribou (*Rangifer tarandus*) from Western to Eastern Indian Island taken on 30 May 2017.
+(ref:fig1) Photograph of swimming unmarked adult male caribou (*Rangifer tarandus*) from Western 
+Perry Island to Eastern Perry Island taken on 30 May 2017.
 
@@
 
-(ref:fig2) A) Map of the Fogo Island archipelago with swimming events between islands.
-B) Histogram displaying the frequency of swimming events throughout the year.
-Note, colours correspond to individual caribou and vertical black bars delineate the ice-free season (April 1 to December 31).
-C) Inset of swimming events between small islands on the northern coast of Fogo island.
-D) Inset of swimming events between islands on the southern coast of Fogo island.
-Note, the grey diamond represents the location we observed an adult male caribou swimming (see Figure 1).
+(ref:fig2) A) Map of the Fogo Island archipelago with swimming events from 13 caribou between islands.
+B) Correlation between area of an island (km2) and the residency in days an individual caribou spent
+on that island (linear mixed effects regression with ID as a random effect: ß = 1.08, SE = 0.06; tdf =
+177 = 17.5, p < 0.001). C) Inset of swimming events between small islands on the northern coast of
+Fogo island. D) Inset of swimming events between islands on the southern coast of Fogo island. The
+grey diamond represents the location we observed an adult male caribou swimming (see Figure 1). E)
+Histogram displaying the frequency of swimming events throughout the year. Colors correspond to
+individual caribou, which are consistent in all panels, and vertical black bars delineate the ice-free
+season (April 1 to December 31).
 

### 2021-05-25 scripts/8-extract-lc.R
Update 8-extract-lc.R

@@ -1,0 +1,56 @@
+
+library(sf)
+library(rgdal)
+library(raster)
+library(data.table)
+
+extract_lc <- function(value, lc, poly) {
+  if (is.na(value)) {
+    extract(is.na(lc), poly, fun = mean)
+  } else {
+    extract(lc == value, poly, fun = mean, na.rm = TRUE)
+  }
+}
+
+## load data
+fogopolys <- readRDS('../caribou-swimming/output/islandsPoly.Rds')
+fogolc <- raster('../fogo_coyote_repeat/data/raw-data/Landcover/FogoSDSS_RS.tif')
+
+## extract proportion of habitat type by island
+values <- c(NA, unique(fogolc))
+extracts <- lapply(values, extract_lc, fogolc, fogopolys)
+names(extracts) <- paste0('prop', values)
+DT <- as.data.table(extracts)
+DT[, id := fogopolys$id]
+DT[, area := fogopolys$area]
+saveRDS(DT, '../caribou-swimming/output/fogo-prop-lc.Rds')
+
+## load data
+DT <- readRDS('../caribou-swimming/output/fogo-prop-lc.Rds')
+legend <- fread('../fogo_coyote_repeat/data/raw-data/Landcover/Legend.csv')
+
+setnames(DT, c("prop1", "prop2", "prop3", "prop4",
+               "prop5", "prop6", "prop7", "prop8",
+               "prop9", "propNA"),
+         unique(legend$Cover))
+
+DT <- DT[id == 120 | id == 128 | id == 124 |
+         id == 75 | id == 70 | id == 125  |
+         id == 78 | id == 73 | id == 58 | id == 72]
+
+## change names of islands ids to island names 
+DT$id[DT$id == 58] <- "Island 6"
+DT$id[DT$id == 72] <- "Island 7"
+DT$id[DT$id == 70] <- "Island 1"
+DT$id[DT$id == 73] <- "Island 2"
+DT$id[DT$id == 75] <- "Island 3"
+DT$id[DT$id == 78] <- "Island 4"
+DT$id[DT$id == 125] <- "Island 5"
+DT$id[DT$id == 120] <- "Fogo"
+DT$id[DT$id == 124] <- "E. Perry"
+DT$id[DT$id == 128] <- "W. Perry"
+
+DT$other <- DT$Broadleaf +DT$MixedWood + DT$Anthro + DT$NotAvail
+DT$conifer <- DT$ConiferScrub + DT$ConiferForest
+
+DT$areakm2 <- DT$area/1000000

### 2021-05-25 paper/manuscript.Rmd
P1 and P2

@@
 
-Movement is costly. Even for adept swimmers like caribou (*Rangifer tarandus*), which have hollow hair
-shafts that aid in flotation, the energetic expenditure of swimming for terrestrial mammals is
-markedly higher than walking or running (Miller and Gunn 1985). The drivers that promote the decision
-to swim for terrestrial mammals, like caribou, are often related to the profitability of forage
-(Miller et al. 1977) and the risks of predation (Jeffery et al. 2007, Jordan et al. 2010). The Fogo
-Island archipelago, off the coast of Newfoundland, Canada, is home to approximately 300 woodland
-caribou, descendants of 26 individuals introduced from the Island of Newfoundland to Fogo Island
-between 1964–67 (Bergerud and Mercer 1989). During fieldwork on May 30, 2017, we observed an unmarked
-adult male caribou swim between two smaller islands (Figure 1) a distance of 470 m which took
-approximately 9 minutes. Given that swimming is energetically expensive, we ask how often do caribou
-swim between islands, how long do they occupy islands, and ultimately what motivates swimming? Here,
-we frame swimming and island residency as a model of patch use as described by the marginal value
-theorem (Charnov 1976) within the context of the forage limitation hypothesis. For caribou, the forage
-limitation hypothesis is the constraints on food availability associated with changes in abundance
-through time (Klein 1991, Schaefer et al. 2016). Here, we integrate our understanding of forage
-limitation in caribou populations by using island residency as a proxy for time in patch and island
-size as a proxy for energy intake in a patch (Figure 2).
+Movement is costly. Even for adept swimmers like caribou (*Rangifer tarandus*), which 
+have hollow hair shafts that aid in flotation, the energetic expenditure of swimming 
+for terrestrial mammals is markedly higher than walking or running (Miller and Gunn 1985). 
+The drivers that promote the decision to swim for terrestrial mammals, like caribou, 
+are often related to the profitability of forage (Miller et al. 1977) and the risks 
+of predation (Jeffery et al. 2007, Jordan et al. 2010). The Fogo Island archipelago, 
+off the coast of Newfoundland, Canada, is home to approximately 300 woodland caribou, 
+descendants of 26 individuals introduced from the Island of Newfoundland to Fogo Island 
+between 1964–67 (Bergerud and Mercer 1989). During fieldwork on May 30, 2017, we 
+observed an unmarked adult male caribou swim between two smaller islands (Figure 1), 
+a distance of 470 m which took approximately 9 minutes. Given that swimming is 
+energetically expensive, we ask how often do caribou swim between islands, how 
+long do they occupy islands, and ultimately what motivates swimming? Here, we 
+frame swimming and island residency as a model of patch use as described by the 
+marginal value theorem (Charnov 1976) within the context of the forage limitation 
+hypothesis. For caribou, the forage limitation hypothesis refers to the 
+constraints on food availability associated with changes in abundance (Klein 1991, 
+Schaefer et al. 2016). Here, we integrate our understanding of forage limitation in
+caribou populations by using island residency as a proxy for time in a patch and 
+island size as a proxy for energy intake in a patch (Figure 2).
 
-Forage limitation and over-grazing is a concern for caribou populations and can lead to reduced female
-reproductive success (Klein 1991, Schaefer et al. 2016). Forage scarcity has been proposed as a
-potential reason for caribou moving between arctic islands on the sea-ice in winter (Miller et al.
-1977). We submit that testing the marginal value theorem could improve our understanding of forage
-limitation. The marginal value theorem posits that animals will forage optimally by depleting the
-resources in a given patch (i.e. island) before moving to a new patch (Charnov 1976). Depletion of
-resources on small islands could therefore mirror patterns of over-grazing and forage limitation that
-occur for caribou herds at higher densities in Newfoundland (Schaefer et al. 2016) and elsewhere in
-their range (Klein 1991).
+Forage limitation and over-grazing can reduce reproductive success of female caribou (Klein 1991, Schaefer et al.
+2016).
+Forage scarcity has been proposed as a potential reason for caribou moving between arctic islands on the sea-ice in winter (Miller et al.
+1977).
+We submit that testing the marginal value theorem could improve our understanding of forage limitation.
+The marginal value theorem posits that animals will forage optimally by depleting the resources in a given patch (i.e.
+island) before moving to a new patch (Charnov 1976).
+Depletion of resources on small islands could therefore mirror patterns of over-grazing and forage limitation that occur for caribou herds at higher densities in Newfoundland (Schaefer et al.
+2016) and elsewhere in their range (Klein 1991).
 

### 2021-05-25 scripts/extract-lc.R
Update extract-lc.R

@@ -1,56 +1,0 @@
-
-library(sf)
-library(rgdal)
-library(raster)
-library(data.table)
-
-extract_lc <- function(value, lc, poly) {
-  if (is.na(value)) {
-    extract(is.na(lc), poly, fun = mean)
-  } else {
-    extract(lc == value, poly, fun = mean, na.rm = TRUE)
-  }
-}
-
-## load data
-fogopolys <- readRDS('../caribou-swimming/output/islandsPoly.Rds')
-fogolc <- raster('../fogo_coyote_repeat/data/raw-data/Landcover/FogoSDSS_RS.tif')
-
-## extract proportion of habitat type by island
-values <- c(NA, unique(fogolc))
-extracts <- lapply(values, extract_lc, fogolc, fogopolys)
-names(extracts) <- paste0('prop', values)
-DT <- as.data.table(extracts)
-DT[, id := fogopolys$id]
-DT[, area := fogopolys$area]
-saveRDS(DT, '../caribou-swimming/output/fogo-prop-lc.Rds')
-
-## load data
-DT <- readRDS('../caribou-swimming/output/fogo-prop-lc.Rds')
-legend <- fread('../fogo_coyote_repeat/data/raw-data/Landcover/Legend.csv')
-
-setnames(DT, c("prop1", "prop2", "prop3", "prop4",
-               "prop5", "prop6", "prop7", "prop8",
-               "prop9", "propNA"),
-         unique(legend$Cover))
-
-DT <- DT[id == 120 | id == 128 | id == 124 |
-         id == 75 | id == 70 | id == 125  |
-         id == 78 | id == 73 | id == 58 | id == 72]
-
-## change names of islands ids to island names 
-DT$id[DT$id == 58] <- "Island 6"
-DT$id[DT$id == 72] <- "Island 7"
-DT$id[DT$id == 70] <- "Island 1"
-DT$id[DT$id == 73] <- "Island 2"
-DT$id[DT$id == 75] <- "Island 3"
-DT$id[DT$id == 78] <- "Island 4"
-DT$id[DT$id == 125] <- "Island 5"
-DT$id[DT$id == 120] <- "Fogo"
-DT$id[DT$id == 124] <- "E. Perry"
-DT$id[DT$id == 128] <- "W. Perry"
-
-DT$other <- DT$Broadleaf +DT$MixedWood + DT$Anthro + DT$NotAvail
-DT$conifer <- DT$ConiferScrub + DT$ConiferForest
-
-DT$areakm2 <- DT$area/1000000

### 2021-05-25 paper/manuscript.Rmd
P3

@@
 
-Our focus on forage is due to the limited presence of predators in our system. Throughout their range,
-wolves (*Canis lupus*) are the primary predator of caribou, although there are no wolves on the Island
-of Newfoundland, predation on adult caribou by coyotes (*Canis latrans*) does occur (Lewis and Mahoney
-2014). Within the Fogo Island archipelago, coyotes exist at low density (Huang et al. 2021) and the
-probability of occurrence decreases on the smaller islands. The low density of predators throughout
-the archipelago is therefore an unlikely driver of swimming in our system, though predation may
-still influence swimming behaviour for caribou elsewhere in their range (e.g. Jeffery et al. 2007).
+Our focus on forage is due to the limited presence of predators in our system.
+Throughout their range, wolves (*Canis lupus*) are the primary predator of caribou.
+Because there are no wolves on the Island of Newfoundland, coyotes (*Canis latrans*) and black bears (*Ursus americanus*) are the majors predator of caribou (Lewis and Mahoney 2014).
+Within the Fogo Island archipelago there are no black bears and coyotes exist at low density (Huang et al.
+2021), while the probability of coyote occurrence decreases on the smaller islands.
+The low density of predators throughout the archipelago is therefore an unlikely driver of swimming in our system, though predation may still influence swimming behaviour for caribou elsewhere in their range (e.g.
+Jeffery et al.
+2007).
 

### 2021-05-25 paper/manuscript.Rmd
P4

@@
 
-We identified swimming events as two consecutive GPS locations (2-hour relocation rates) on different
-islands. Newfoundland typically experiences pack ice in late winter and caribou may travel between
-islands by walking over the ice. We restricted our dataset between April 1 and December 31 to
-distinguish swimming from walking on ice. We calculated island residency as the duration of time,
-measured on 2-hour relocation rates, an individual caribou was observed on an island. Residency was
-therefore a proxy for when the benefits of foraging no longer exceeded the costs of swimming between
-islands. Our test of the marginal value theorem for the case of oceanic swimming by caribou is based
-on three assumptions: (1) resources are proportional to the size of each island; (2) caribou know
-approximate value of the resources on an island before they swim to and from that island; and (3) the
-cost of swimming to and from an island is less than the value of resources on that island when the
-caribou departs. Should these assumptions hold, we would predict a positive relationship where caribou
-depart smaller islands sooner than larger islands.
+We identified swimming events as two consecutive GPS locations (2-hour relocation rates) on different islands (n = 29, for collaring details see Peignier et al.
+2019).
+Newfoundland typically experiences pack ice in late winter and caribou may travel between islands by walking over the ice.
+We restricted our dataset between April 1 and December 31 to distinguish swimming from walking on ice.
+We calculated island residency as the duration of time, measured on 2-hour relocation rates, an individual caribou was observed on an island.
+Residency was therefore a proxy for when the benefits of foraging no longer exceeded the costs of swimming between islands.
+Our test of the marginal value theorem for the case of oceanic swimming by caribou is based on three assumptions: (1) resources are proportional to the size of each island above approximately 0.1km2; (2) caribou know the approximate value of the resources on an island before they swim to and from that island; and (3) the cost of swimming to and from an island is less than the value of resources on that island when the caribou departs.
+Should these assumptions hold, we would predict a positive relationship where caribou depart smaller islands sooner than larger islands.
 

### 2021-05-25 paper/manuscript.Rmd
P5

@@
 
-Swimming behavior was common among GPS radio-collared caribou in the population (n = 29, for collaring
-details see Peignier et al. 2019). We identified `r nevents` swimming events over three years (Figure
-2). In total, 13 of 29 collared female caribou swam among islands (Figure 2). Of the caribou that
-swam, caribou swam on average approximately `r neventidMedian` (range = `r neventidMin`-` neventidMax`) times per year and remained on each island for a median of `r dur` (range = 
-`r durMin`-`r durMax`) days before swimming again. We found that caribou residency increased
-proportionally with the size of the island, such that caribou spent less time on smaller islands and
-more time on larger islands, providing preliminary evidence for links between forage limitation and
-the marginal value theorem (Figure 2B).
+Swimming behavior was common among GPS radio-collared caribou in the population (n = 29, for collaring details see Peignier et al.
+2019).
+We identified `r nevents` swimming events over three years (Figure 2).
+In total, 13 of 29 collared female caribou swam among islands (Figure 2).
+Of the caribou that swam, caribou swam on average approximately `r neventidMedian` (range = `r neventidMin`-` neventidMax`) times per year and remained on each island for a median of `r dur` (range =  `r durMin`-`r durMax`) days before swimming again.
+We found that caribou residency increased proportionally with the size of the island for islands larger than approximately 0.1 km^2, such that caribou spent less time on smaller islands and more time on larger islands, providing preliminary evidence for links between forage limitation and the marginal value theorem (Figure 2B).
+In many cases, the smaller islands in the archipelago are forage limited and the larger islands provide greater opportunities to forage.
 

### 2021-05-25 paper/manuscript.Rmd
conclusion

@@
 
-From our natural history observation and proximal movement data on caribou swimming, characteristics
-of destination islands, and residency on the islands, we have presented some evidence for the marginal
-value of resources that may serve as for evidence of forage limitation. Further investigation into the
-nuance of forage profitability, local density, and the role of both social and spatial cognition is
-needed. Here, we have proposed a thought exercise that integrates the marginal value theorem with a
-novel natural history observation to explain swimming behavior for terrestrial animals and opens the
-door for future studies of caribou in our system and other island archipelagos inhabited by caribou.
+From our natural history observations and proximal movement data on caribou swimming, characteristics of destination islands, and residency on the islands, we have presented some evidence for the marginal value of resources that may serve as evidence of forage limitation for caribou swimming between islands.
+Further investigation into the nuance of forage profitability, local density, and the role of both social and spatial cognition is needed.
+Here, we have proposed a thought exercise that integrates the marginal value theorem with novel natural history observations to explain swimming behavior for terrestrial animals and opens the door for future studies of caribou in our system and other island archipelagos inhabited by caribou.
 

### 2021-05-25 paper/manuscript.Rmd
add references

@@
 **Literature Cited**
+
+Bergerud, A. T., and W. E. Mercer. 1989. Caribou introductions in eastern North America. Wildlife Society Bulletin 17:111–120.
+
+Charnov, E. L. 1976. Optimal foraging, the marginal value theorem. Theoretical Population Biology 9:129–136.
+
+Fretwell, S. D., and H. L. J. Lucas. 1969. On territorial behaviour and other factors influencing habitat distribution in birds. Acta Biotheoretica 19:16–36.
+
+Huang, R.
+K.
+K., Q.
+M.
+R.
+Webber, M.
+P.
+Laforge, A.
+L.
+Robitaille, M.
+Bonar, J.
+Balluffi-Fry, S.
+Zabihi-Seissan, and E.
+Vander Wal.
+2021.
+Coyote (*Canis latrans*) diet and spatial co-occurrence with woodland caribou (*Rangifer tarandus caribou*).
+Canadian Jounal of Zoology 99:391–399.
+
+Jeffery, R. A., R. D. Otto, and F. R. Phillips. 2007. George’s Island, Labrador - A high-density predator-free refuge for a woodland caribou subpopulation? Rangifer:51–56.
+
+Jordan, P. A., R. O. Peterson, and K. A. Ledoux. 2010. Swimming Wolves, Canis lupus, attack a swimming Moose, Alces alces. Canadian Field-Naturalist 124:54–56.
+
+Klein, D. R. 1991. Limiting factors in caribou population ecology. Rangifer 11:30.
+
+Lewis, K.
+P., and S.
+P.
+Mahoney.
+2014.
+Caribou survival, fate, and cause of mortality in Newfoundland: a summary and analysis of the patterns and causes of caribou survival and mortality in Newfoundland during a period of rapid population decline (2003-2012).
+Technical Bulletin No.
+009, Sustainable Development and Strategic Science.
+Government of Newfoundland and Labrador, St.
+John’s, NL.
+
+Miller, F. L., and A. Gunn. 1985. Observations of Barren-Ground Caribou Travelling on Thin Ice during Autumn Migration. Arctic 39:85–88.
+
+Miller, F.
+L., R.
+H.
+Russell, and A.
+Gunn.
+1977.
+Interisland movements of Peary caribou (*Rangifer tarandus pearyi*) on western Queen Elizabeth Islands, Arctic Canada.
+Canadian Journal of Zoology 55:1029–1037.
+
+Morris, D. W. 2003. Toward an ecological synthesis: A case for habitat selection. Oecologia 136:1–13.
+
+Peignier, M., Q.
+M.
+R.
+Webber, E.
+L.
+Koen, M.
+P.
+Laforge, A.
+L.
+Robitaille, and E.
+Vander Wal.
+2019.
+Space use and social association in a gregarious ungulate: Testing the conspecific attraction and resource dispersion hypotheses.
+Ecology & Evolution 9:5133–5145.
+
+Presotto, A., R.
+Fayrer-Hosken, C.
+Curry, and M.
+Madden.
+2019.
+Spatial mapping shows that some African elephants use cognitive maps to navigate the core but not the periphery of their home ranges.
+Animal Cognition 22:251–263.
+
+Schaefer, J. A., S. P. Mahoney, J. N. Weir, J. G. Luther, and C. E. Soulliere. 2016. Decades of habitat use reveal food limitation of Newfoundland caribou. Journal of Mammalogy 97:386–393.
+
+Seyfarth, R. M., and D. L. Cheney. 2015. Social cognition. Animal Behaviour 103:191–202.

### 2021-06-04 paper/supplement.Rmd
fst sup

---
title: On the marginal value of swimming in woodland caribou
author:
  - name: Quinn M.R. Webber
    institute: cabe
    correspondence: true
  - name: Jack G. Hendrix
    institute: cabe
  - name: Alec L. Robitaille
    email: robit.alec@gmail.com
    institute: bio
  - name: Eric Vander Wal
    institute: [bio, cabe]
institute: 
  - bio: Department of Biology, Memorial University of Newfoundland, Canada
  - cabe: Cognitive and Behavioural Ecology Interdisciplinary Program, Memorial University of Newfoundland 
output:
   bookdown::word_document2:
     reference_docx: template.docx
     pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: ../references/swim.bib
csl: ecology.csl
---

```{r knitropts, include = FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      eval = TRUE,
                      fig.align = 'center')
```

```{r pkgs}
library(data.table)
```

|  			Habitat 			type 		        |  			Description 		     |  			Common 			species and groups of species 		        |
|-|-|-|
|  			Lichen 		     |  			A 			non-forested shrub-dominated habitat found on hummocky terrain and 			may have thin soils with exposed bedrock, ericaceous shrubs 			dominate having relatively high percent cover. Tree species are 			always stunted and rarely grow above the shrub layer. The herb 			layer is species-rich, is nearly always dominated by ground 			lichen. 		                       |  			Caribou 			lichen (Cladonia sp.)  			  			Sheep 			laurel (Kalmia sp.)  			  			Dwarf 			willow (Salix herbacea) 		                          |
|  			Wetland 		     |  			Characterized 			by open or ponded water. Dominated by aquatic and submergent 			vegetation such as yellow pond lily and rushes. 		           |  			Sedges 			(Carex sp.)  			  			Rushes 			(Juncus sp.)  			  			Mosses 		                       |
|  			Conifer 			forest 		        |  			Dominant 			tree species are typically balsam fir or black spruce with lesser 			amount of larch, ericaceous shrubs are found in this habitat and 			sheep laurel. 		              |  			White 			spruce (Picea glauca)  			  			Black 			spruce (Picea mariana)  			  			Balsam 			fir (Abies balsamea)  			  			Larch 			(Larix laricina) 		                                   |
|  			Rocky 			barrens 		        |  			Windswept 			with thin layer that is comprised of bedrock, exposed soil, stone, 			and boulders. The shrub layer is composed mainly of ericaceous 			shrubs. Ground cover consists of grasses, lichens, mosses, and 			smaller berry plants. 		                 |  			Alder 			(Alnus sp.)  			  			Sheep 			laurel (Kalmia sp.)  			  			Crowberry 			(Empetrum nigrum)  			  			Vaccinium 			sp. 		                                   |

### 2021-06-04 scripts/10-render-sup.R
fst 10 render sup

@@ -1,0 +1,1 @@
+rmarkdown::render('paper/supplement.Rmd')

### 2021-06-04 scripts/plan.R
update plan

@@ -8,7 +8,10 @@ generate_edges <- code_to_function('scripts/3-generate-edges.R')
 tables <- code_to_function('scripts/4-tables.R')
 fig_2 <- code_to_function('scripts/5-figure2.R')
 render_md <- code_to_function('scripts/6-render-md.R')
-
+island_dist <- code_to_function('scripts/7-island-dist.R')
+extract_lc <- code_to_function('scripts/8-extract-lc.R')
+fig_s1 <- code_to_function('scripts/9-figureS1.R')
+render_sup <- code_to_function('scripts/10-render-sup.R')
 
 # The plan
 plan <- drake_plan(
@@ -17,5 +20,9 @@ plan <- drake_plan(
   edges = generate_edges(locs),
   tabs = tables(edges),
   fig2 = fig_2(edges, tabs),
-  md = render_md(fig2, tabs)
+  md = render_md(fig2, tabs),
+  isldist = island_dist(locs, edges),
+  lc = extract_lc(locs),
+  figs1 = fig_s1(locs, edges),
+  sup = render_sup(lc, figs1)
 )

### 2021-06-04 scripts/5-figure2.R
missing ggplot2

@@ -11,7 +11,8 @@ libs <- c(
   'ggnetwork',
   'patchwork',
   'effects',
-  'lme4'
+  'lme4',
+  'ggplot2'
 )
 lapply(libs, require, character.only = TRUE)
 

### 2021-06-04 scripts/9-figureS1.R

@@ -8,7 +8,8 @@ libs <- c(
   'ggnetwork',
   'patchwork',
   'effects',
-  'mapview'
+  'mapview',
+  'ggplot2'
 )
 lapply(libs, require, character.only = TRUE)
 

### 2021-06-07 scripts/8-extract-lc.R
fix output, set names

@@ -1,9 +1,14 @@
+### Extract land cover ====
+# Quinn M.R. Webber, Jack G. Hendrix, Alec L. Robitaille,  Eric Vander Wal
 
+
+### Packages ----
 library(sf)
 library(rgdal)
 library(raster)
 library(data.table)
 
+### Functions ----
 extract_lc <- function(value, lc, poly) {
   if (is.na(value)) {
     extract(is.na(lc), poly, fun = mean)
@@ -12,45 +17,41 @@ extract_lc <- function(value, lc, poly) {
   }
 }
 
-## load data
+
+### Data ----
 fogopolys <- readRDS('output/islandsPoly.Rds')
 fogolc <- raster('../nl-landcover/output/fogo_lc.tif')
+legend <- fread('../nl-landcover/input/FINAL_PRODUCT/FINAL_RC_legend.csv')
 
-## extract proportion of habitat type by island
+### Extract proportion of habitat type by island ----
 values <- c(NA, unique(fogolc))
 extracts <- lapply(values, extract_lc, fogolc, fogopolys)
 names(extracts) <- paste0('prop', values)
+
 DT <- as.data.table(extracts)
+
 DT[, id := fogopolys$id]
 DT[, area := fogopolys$area]
-saveRDS(DT, 'output/fogo-prop-lc.Rds')
-
-## load data
-DT <- readRDS('output/fogo-prop-lc.Rds')
-legend <- fread('../nl-landcover/input/FINAL_PRODUCT/FINAL_RC_legend.csv')
 
 setnames(DT, c("prop1", "prop2", "prop3", "prop4",
                "prop5", "prop6", "prop7", "prop8",
                "prop9", "propNA"),
-         unique(legend$Cover))
+         c(unique(legend$Landcover), "Not Available"))
 
-DT <- DT[id == 120 | id == 128 | id == 124 |
-         id == 75 | id == 70 | id == 125  |
-         id == 78 | id == 73 | id == 58 | id == 72]
+selislands <- data.table(
+  id = c(120, 128, 124, 75, 70, 125, 78, 73, 58, 72),
+  name = c("Fogo", "W. Perry", "E. Perry", "Island 3", 
+           "Island 1", "Island 5", "Island 4", 
+           "Island 2", "Island 6", "Island 7")
+)
[…]

### 2021-06-07 scripts/9-figureS1.R
add title

@@ -1,3 +1,5 @@
+### Figure S1 ====
+# Quinn M.R. Webber, Jack G. Hendrix, Alec L. Robitaille,  Eric Vander Wal
 
 ### Packages ----
 libs <- c(

### 2021-06-07 scripts/9-figureS1.R
no palette necessary

@@ -34,11 +34,6 @@ themeMap <- theme(panel.border = element_rect(size = 1, fill = NA),
                   axis.text = element_blank(),
                   axis.title = element_blank())
 
-# Palette
-pal <- unique(edges, by = 'ANIMAL_ID')[order(region), .(ID = unique(ANIMAL_ID), col = scales::viridis_pal()(.N))]
-cols <- pal[, setNames(col, ID)]
-
-
 # Base islands ------------------------------------------------------------
 labels <- data.table(id = c(58, ## Island 6
                             72, ## Island 7

### 2021-06-07 paper/manuscript.Rmd
minor edits to paper

@@
 Because there are no wolves on the Island of Newfoundland, coyotes (*Canis latrans*) and black bears (*Ursus americanus*) are the majors predator of caribou (Lewis and Mahoney 2014).
-Within the Fogo Island archipelago there are no black bears and coyotes exist at low density (Huang et al.
-2021), while the probability of coyote occurrence decreases on the smaller islands.
+Within the Fogo Island archipelago there are no black bears; the density of coyotes is low (Huang et al.
+2021) and the probability of coyote occurrence decreases on the smaller islands.
 The low density of predators throughout the archipelago is therefore an unlikely driver of swimming in our system, though predation may still influence swimming behaviour for caribou elsewhere in their range (e.g.
@@
 
-Swimming behavior was common among GPS radio-collared caribou in the population (n = 29, for collaring details see Peignier et al.
-2019).
 We identified `r nevents` swimming events over three years (Figure 2).

### 2021-06-08 scripts/5-figure2.R
fix subset only island 5

@@ -27,7 +27,7 @@ runarea <- readRDS('output/runarea.Rds')
 utm <- st_crs('+proj=utm +zone=21 ellps=WGS84')
 
 # Drop some edges
-edges <- edges[!i %in% c(59, 11156, 4254, 4859, 11157)]
+edges <- edges[!(ANIMAL_ID == 'FO2016013' & (island == 5 | endisland == 5))]
 
 
 # summary stats for % swims per island

### 2021-06-08 scripts/5-figure2.R
adjust bbox

@@ -65,19 +65,19 @@ themeRes <- theme(legend.position = 'none',
 tofirst <- c('EASTING', 'NORTHING', 'endislandEAST', 'endislandNORTH')
 outfirst <- c(x = 'firstX', y = 'firstY', xend = 'endfirstX', yend = 'endfirstY')
 
-edges[, (outfirst) := .SD,
+edges[, (outfirst) := .SD, 
       by = .(island, ANIMAL_ID), .SDcols = tofirst]
 
 
 # North/South regions
-edges[, region := ifelse(firstY < median(firstY) + 6500, 'South', 'North')]
+edges[, region := ifelse(firstY < median(firstY) + 6000, 'South', 'North')]
 N <- edges[season == 'icefree' & region == 'North']
 S <- edges[season == 'icefree' & region == 'South']
 
 # Bboxes
-Nbox <- c(ymin = min(N$NORTHING), 
+Nbox <- c(ymin = min(N$NORTHING) + 1250, 
           ymax = max(N$NORTHING),
-          xmin = min(N$EASTING), 
+          xmin = min(N$EASTING) + 5000, 
           xmax = max(N$EASTING))
 Nsfbox <- st_as_sf(st_as_sfc(st_bbox(Nbox, crs = utm)))
 Nsfbox$label <- 'C'

### 2021-06-08 scripts/5-figure2.R
rm old labels

@@ -96,21 +96,12 @@ cols <- pal[, setNames(col, ID)]
 
 
 # Base islands ------------------------------------------------------------
-labels <- data.table(id = c(120, 124, 128),
-                     label = c('Fogo Island', 'E. Perry Island', 'W. Perry Island'))
-islands <- left_join(islands, labels, 'id')
-
 (gfogo <- ggplot(islands) + 
-    geom_sf(fill = islandcol, size = 0.13, color = coastcol) + 
+    geom_sf(fill = islandcol, size = 0.3, color = coastcol) + 
     themeMap +
     theme(axis.text = element_text(size = 11, color = 'black')) +
     scale_y_continuous(label = function(x) sprintf('%.2f°N', x)) +
     scale_x_continuous(label = function(x) sprintf('%.1f°W', -1 * x)))
-
-(gfogothick <- ggplot(islands) + 
-    geom_sf(fill = islandcol, size = 0.3, color = coastcol) + 
-    themeMap +
-    theme(axis.text = element_text(size = 11, color = 'black')))
 
 
 

### 2021-06-08 scripts/5-figure2.R
rm gg thick

@@ -119,7 +119,7 @@ cols <- pal[, setNames(col, ID)]
 
 # Edges -------------------------------------------------------------------
 edgesize <- 1
-(gnetN <- gfogothick +
+(gnetN <- gfogo +
     geom_edges(data = N,
                aes(
                  x = firstX,
@@ -139,7 +139,7 @@ edgesize <- 1
   theme(axis.ticks = element_blank())
   )
 
-(gnetS <- gfogothick +
+(gnetS <- gfogo +
     geom_edges(data = S,
                aes(
                  x = firstX,

### 2021-06-08 scripts/4-tables.R
only 14, switch to merge and replace islands output

@@ -16,14 +16,14 @@ caribou <- readRDS('output/islands-locs.Rds')
 
 ### Set island labels ----
 setDT(islands)
-numbsub <- 15
+numbsub <- 14
 labislands <- islands[id %in% edges[, unique(island)]][order(-area)][seq.int(numbsub)]
 nms <- c('Fogo Island', 'W. Perry', 'E. Perry')
 labislands[, label := c(nms, paste0('Island ', 4:numbsub))]
-labislands <- st_as_sf(labislands)
+islands <- st_as_sf(merge(islands, labislands[, .(id, label)], by = 'id', all.x = TRUE))
 
 ### Table 1 ----
-selislands <- labislands[labislands$label %in% nms, ]
+selislands <- islands[islands$label %in% nms, ]
 
 selislands$proplichen <- extract(lc == 8, selislands, fun = mean, na.rm = TRUE)
 
@@ -78,7 +78,4 @@ runarea[, .(area, islandlen)]
 ### Output ----
 saveRDS(runarea, 'output/runarea.Rds')
 saveRDS(tab1, 'output/table1.Rds')
-saveRDS(labislands, 'output/island-labels.Rds')
-
-
-  
+saveRDS(islands, 'output/islandsPoly.Rds')

### 2021-06-08 scripts/5-figure2.R
islands centroids and repels

@@ -162,6 +162,15 @@ edgesize <- 1
   theme(axis.ticks = element_blank())
 )
 
+islandcents <- st_centroid(
+  islands[
+  !is.na(islands$label) &
+  !grepl('Island ', islands$label) & 
+  islands$area > units::as_units(1e5, 'm^2'),
+])
+islandrepels <- islands[!islands$label %in% islandcents$label &
+                          !is.na(islands$label),]
+
 (withboxes <- gfogo +
     geom_sf(
       data = Nsfbox,

### 2021-06-08 scripts/3-generate-edges.R
drop drop edges in 3

@@ -22,11 +22,6 @@ edges <- caribou[island != endisland]
 # Edge ID
 edges[, edgeID := .I]
 
-
-edges <- edges[!(between(i, 946, 947, incbounds = TRUE) & ANIMAL_ID == 'FO2017012') &
-                 !(between(i, 3771, 3772, incbounds = TRUE) & ANIMAL_ID == 'FO2017013')]
-
-
 ### Variables ----
 # Step length
 difXY <- c('difX', 'difY')

### 2021-06-08 scripts/4-tables.R
fiddle to remove fogo head

@@ -17,9 +17,9 @@ caribou <- readRDS('output/islands-locs.Rds')
 ### Set island labels ----
 setDT(islands)
 numbsub <- 14
-labislands <- islands[id %in% edges[, unique(island)]][order(-area)][seq.int(numbsub)]
+labislands <- islands[id %in% edges[, unique(island)]][order(-area)][seq.int(numbsub)[-4]]
 nms <- c('Fogo Island', 'W. Perry', 'E. Perry')
-labislands[, label := c(nms, paste0('Island ', 4:numbsub))]
+labislands[, label := c(nms, paste0('Island ', 4:(numbsub-1)))]
 islands <- st_as_sf(merge(islands, labislands[, .(id, label)], by = 'id', all.x = TRUE))
 
 ### Table 1 ----
@@ -78,4 +78,4 @@ runarea[, .(area, islandlen)]
 ### Output ----
 saveRDS(runarea, 'output/runarea.Rds')
 saveRDS(tab1, 'output/table1.Rds')
-saveRDS(islands, 'output/islandsPoly.Rds')
+saveRDS(islands, 'output/islandsPoly-with-label.Rds')

### 2021-06-08 scripts/5-figure2.R
read with-label version

@@ -18,7 +18,7 @@ lapply(libs, require, character.only = TRUE)
 
 
 ### Data ----
-islands <- readRDS('output/islandsPoly.Rds')
+islands <- readRDS('output/islandsPoly-with-label.Rds')
 edges <- readRDS('output/island-edges.Rds')
 net <- readRDS('output/island-network.Rds')
 runarea <- readRDS('output/runarea.Rds')

### 2021-06-08 scripts/5-figure2.R
drops island repels from fig2

@@ -168,8 +168,6 @@ islandcents <- st_centroid(
   !grepl('Island ', islands$label) & 
   islands$area > units::as_units(1e5, 'm^2'),
 ])
-islandrepels <- islands[!islands$label %in% islandcents$label &
-                          !is.na(islands$label),]
 
 (withboxes <- gfogo +
     geom_sf(
@@ -186,11 +184,9 @@ islandrepels <- islands[!islands$label %in% islandcents$label &
       size = 0.3
     ) +
     geom_sf_label(data = Ssfbox, aes(label = label)) +
-    geom_sf_label_repel(aes(label = label), 
-                        data = islands[islands$label != 'Fogo Island', ], 
-                        nudge_x = 1.3e4, size = 2) +
-    geom_sf_label(aes(label = label), 
-                        data = islands[islands$label == 'Fogo Island', ], size = 2) +
+    geom_sf_label(aes(label = label),
+                  data = islandcents,
+                  size = 2) +
     theme(axis.title = element_blank())
 )
 

### 2021-06-08 scripts/5-figure2.R
fiddle fiddle

@@ -76,7 +76,7 @@ S <- edges[season == 'icefree' & region == 'South']
 
 # Bboxes
 Nbox <- c(ymin = min(N$NORTHING) + 1250, 
-          ymax = max(N$NORTHING),
+          ymax = max(N$NORTHING) + 100,
           xmin = min(N$EASTING) + 5000, 
           xmax = max(N$EASTING))
 Nsfbox <- st_as_sf(st_as_sfc(st_bbox(Nbox, crs = utm)))
@@ -176,17 +176,19 @@ islandcents <- st_centroid(
       color = 'black',
       size = 0.3
     ) +
-    geom_sf_label(data = Nsfbox, aes(label = label)) +
+    geom_sf_label(data = Nsfbox, aes(label = label), segment.alpha = 0) +
     geom_sf(
       data = Ssfbox,
       fill = NA,
       color = 'black',
       size = 0.3
     ) +
-    geom_sf_label(data = Ssfbox, aes(label = label)) +
-    geom_sf_label(aes(label = label),
-                  data = islandcents,
-                  size = 2) +
+    geom_sf_label(aes(label = label), data = islandcents[grepl('Fogo', islandcents$label), ],
+                        size = 2) +
+    geom_sf_label_repel(aes(label = label), data = islandcents[!grepl('Fogo', islandcents$label),],
+                        size = 2, nudge_x = 7000) +
+    geom_sf_label(data = Ssfbox, aes(label = label), segment.alpha = 0) +
+    
     theme(axis.title = element_blank())
 )
 

### 2021-06-08 scripts/8-extract-lc.R
use with labels for extract lc

@@ -19,31 +19,26 @@ extract_lc <- function(value, lc, poly) {
 
 
 ### Data ----
-fogopolys <- readRDS('output/islandsPoly.Rds')
+islands <- readRDS('output/islandsPoly-with-label.Rds')
 fogolc <- raster('../nl-landcover/output/fogo_lc.tif')
 legend <- fread('../nl-landcover/input/FINAL_PRODUCT/FINAL_RC_legend.csv')
 
 ### Extract proportion of habitat type by island ----
 values <- c(NA, unique(fogolc))
-extracts <- lapply(values, extract_lc, fogolc, fogopolys)
+extracts <- lapply(values, extract_lc, fogolc, islands)
 names(extracts) <- paste0('prop', values)
 
 DT <- as.data.table(extracts)
 
-DT[, id := fogopolys$id]
-DT[, area := fogopolys$area]
+DT[, id := islands$id]
+DT[, area := islands$area]
 
 setnames(DT, c("prop1", "prop2", "prop3", "prop4",
                "prop5", "prop6", "prop7", "prop8",
                "prop9", "propNA"),
          c(unique(legend$Landcover), "Not Available"))
 
-selislands <- data.table(
-  id = c(120, 128, 124, 75, 70, 125, 78, 73, 58, 72),
-  name = c("Fogo", "W. Perry", "E. Perry", "Island 3", 
-           "Island 1", "Island 5", "Island 4", 
-           "Island 2", "Island 6", "Island 7")
-)
+selislands <- islands[!is.na(islands$label), ]
 DT <- DT[selislands, on = 'id']
 
 DT[, Other := Broadleaf + `Mixed Wood`  + `Anthropogenic and disturbance` +

### 2021-06-08 scripts/9-figureS1.R
use with labels for fig s1

@@ -4,7 +4,6 @@
 ### Packages ----
 libs <- c(
   'data.table',
-  'dplyr',
   'sf',
   'ggsflabel',
   'ggnetwork',
@@ -35,20 +34,13 @@ themeMap <- theme(panel.border = element_rect(size = 1, fill = NA),
                   axis.title = element_blank())
 
 # Base islands ------------------------------------------------------------
-labels <- data.table(id = c(58, ## Island 6
-                            72, ## Island 7
-                            70, ## Island 1
-                            73, ## Island 2
-                            75, ## Island 3
-                            78, ## Island 4
-                            120, ## Fogo
-                            124, ## E. Perry
-                            125, ## Island 5
-                            128), ## W. Perry
-                     label = c('Island 6', 'Island 7',
-                               'Island 1', 'Island 2', 'Island 3', 'Island 4',
-                               'Fogo Island', 'E. Perry Island','Island 5','W. Perry Island'))
-islands <- left_join(islands, labels, 'id')
+labels <- data.table(
+  id = c(118, 126, 122, 75, 70, 125, 78, 73, 58, 72),
+  label = c("Fogo", "W. Perry", "E. Perry", "Island 3", 
+           "Island 1", "Island 5", "Island 4", 
+           "Island 2", "Island 6", "Island 7")
+)
+selislands <- merge(islands, labels, on = 'id')
 
 
 png('graphics/FigS1.png',
@@ -56,15 +48,15 @@ png('graphics/FigS1.png',
   height = 3000,
   units = 'px',
   res = 600)
-ggplot(islands) + 
+ggplot(selislands) + 
   geom_sf(fill = islandcol, size = 0.3, color = coastcol) + 
   guides(color = FALSE) +
   labs(x = NULL, y = NULL) + 
   geom_sf_label_repel(aes(label = label), 
-                      data = islands[islands$label != 'Fogo Island', ], 
+                      data = selislands[selislands$label != 'Fogo Island', ], 
                       nudge_x = 2.2e3, size = 2) +
   geom_sf_label(aes(label = label), 
-                data = islands[islands$label == 'Fogo Island', ], size = 2) +
+                data = selislands[selislands$label == 'Fogo Island', ], size = 2) +
   themeMap + 
   theme(axis.text = element_text(size = 11, color = 'black'),
         axis.ticks = element_blank())

### 2021-06-08 scripts/9-figureS1.R
more with labels fig s1

@@ -15,7 +15,7 @@ lapply(libs, require, character.only = TRUE)
 
 
 ### Data ----
-islands <- readRDS('output/islandsPoly.Rds')
+islands <- readRDS('output/islandsPoly-with-label.Rds')
 edges <- readRDS('output/island-edges.Rds')
 
 
@@ -34,14 +34,8 @@ themeMap <- theme(panel.border = element_rect(size = 1, fill = NA),
                   axis.title = element_blank())
 
 # Base islands ------------------------------------------------------------
-labels <- data.table(
-  id = c(118, 126, 122, 75, 70, 125, 78, 73, 58, 72),
-  label = c("Fogo", "W. Perry", "E. Perry", "Island 3", 
-           "Island 1", "Island 5", "Island 4", 
-           "Island 2", "Island 6", "Island 7")
-)
-selislands <- merge(islands, labels, on = 'id')
-
+selislands <- islands[!is.na(islands$label),]
+centislands <- st_centroid(selislands)
 
 png('graphics/FigS1.png',
   width = 3000,
@@ -53,10 +47,10 @@ ggplot(selislands) +
   guides(color = FALSE) +
   labs(x = NULL, y = NULL) + 
   geom_sf_label_repel(aes(label = label), 
-                      data = selislands[selislands$label != 'Fogo Island', ], 
+                      data = centislands, 
                       nudge_x = 2.2e3, size = 2) +
-  geom_sf_label(aes(label = label), 
-                data = selislands[selislands$label == 'Fogo Island', ], size = 2) +
+  # geom_sf_label(aes(label = label), 
+  #               data = selislands[selislands$label == 'Fogo Island', ], size = 2) +
   themeMap + 
   theme(axis.text = element_text(size = 11, color = 'black'),
         axis.ticks = element_blank())

### 2021-06-08 paper/manuscript.Rmd
render manuscript

@@
       - '--lua-filter=author-info-blocks.lua'
+  bookdown::github_document2:
+    toc: true
+knit: (function(inputFile, encoding) {
+  rmarkdown::render(inputFile, encoding = encoding,
+  output_dir = "paper", output_format = "all") })
 bibliography: ../references/swim.bib

### 2021-06-08 scripts/6-render-md.R

@@ -1,1 +1,1 @@
-rmarkdown::render('paper/manuscript.Rmd')
+rmarkdown::render('paper/manuscript.Rmd', 'all')

### 2021-06-08 scripts/10-render-sup.R
non merci

@@ -1,1 +1,1 @@
-rmarkdown::render('paper/supplement.Rmd')
+rmarkdown::render('paper/supplement.Rmd', 'all')

### 2021-06-08 paper/supplement.Rmd
render supplement

--- 2021-06-04 paper/supplement.Rmd
@@
     reference_docx: template.docx
+    fig_caption: no
     pandoc_args:
@@
       - '--lua-filter=author-info-blocks.lua'
+  bookdown::github_document2:
+    toc: true
 bibliography: ../references/swim.bib
@@
 library(data.table)
+DT <- readRDS('../output/islandsPoly-with-label.Rds')
+setDT(DT)
 ```
 
-|  			Habitat 			type 		        |  			Description 		     |  			Common 			species and groups of species 		        |
-|-|-|-|
-|  			Lichen 		     |  			A 			non-forested shrub-dominated habitat found on hummocky terrain and 			may have thin soils with exposed bedrock, ericaceous shrubs 			dominate having relatively high percent cover.
-Tree species are 			always stunted and rarely grow above the shrub layer.
-The herb 			layer is species-rich, is nearly always dominated by ground 			lichen.
-		                       |  			Caribou 			lichen (Cladonia sp.)  			  			Sheep 			laurel (Kalmia sp.)  			  			Dwarf 			willow (Salix herbacea) 		                          |
-|  			Wetland 		     |  			Characterized 			by open or ponded water.
-Dominated by aquatic and submergent 			vegetation such as yellow pond lily and rushes.
-		           |  			Sedges 			(Carex sp.)  			  			Rushes 			(Juncus sp.)  			  			Mosses 		                       |
-|  			Conifer 			forest 		        |  			Dominant 			tree species are typically balsam fir or black spruce with lesser 			amount of larch, ericaceous shrubs are found in this habitat and 			sheep laurel.
-		              |  			White 			spruce (Picea glauca)  			  			Black 			spruce (Picea mariana)  			  			Balsam 			fir (Abies balsamea)  			  			Larch 			(Larix laricina) 		                                   |
-|  			Rocky 			barrens 		        |  			Windswept 			with thin layer that is comprised of bedrock, exposed soil, stone, 			and boulders.
-The shrub layer is composed mainly of ericaceous 			shrubs.
-Ground cover consists of grasses, lichens, mosses, and 			smaller berry plants.
-		                 |  			Alder 			(Alnus sp.)  			  			Sheep 			laurel (Kalmia sp.)  			  			Crowberry 			(Empetrum nigrum)  			  			Vaccinium 			sp.
-		                                   |
+\newpage
+
+```{r}
+knitr::include_graphics('../graphics/FigS1.png')
+```
+
+Figure S1: Map of the Fogo Island archipelago.
+`r DT[!is.na(label), .N]` islands are identified and the dominant habitats on these islands are presented in Table S1.
+Note, islands listed as Island 1–5 correspond to those presented in Panel D in Figure 2 and Islands 6 and 7 correspond to those presented in Panel C in Figure 2.
+These numbered islands do not have official names and for ease of interpretation we identified them using numbers.
+
+\newpage
+
+```{r}
+prop <- readRDS('../output/fogo-prop-lc.Rds')[, .(
+  Island = label, 
+  `Area (km2)` = as.numeric(round(area / 1e+6, digits = 3)),
+  Lichen = `Lichen and Heath`,
+  Wetland,
+  `Conifer forest` = `Conifer Forest`,
+  `Rocky barrens` = `Rocky Barren`,
+  Freshwater = Water,
+  Other = Other
+)]
+mkpercent <- colnames(prop)[-c(1, 2)]
+prop[, (mkpercent) := lapply(.SD, function(x) scales::label_percent(0.01)(x)), .SDcols = mkpercent]
+knitr::kable(prop[order(-`Area (km2)`)], align = 'l',
+             caption = 'Table S1: Summary of primary habitat types for eight islands in the Fogo Island Archipelago.
+Habitat types were determined using supervised classification of Landsat 7 imagery (Integrated Informatics Inc., 2014).
+Habitats pooled in the ‘other’ category include broadleaf forest, mixed-wood forest, and anthropogenic habitats and we pooled conifer forest and conifer scrub together as these habitats are similar in their species composition.
+Descriptions of each habitat type and dominant plant species are presented in Table S2 and the location of each island is presented in Figure S1.')
+```
+
+\newpage
+
+```{r}
+knitr::kable(fread('../paper/tabs2.csv'), align = 'l',
+             caption = 'Table S2: Summary of common plant species and groups of species in four habitat types on Fogo Island, Newfoundland.
+Habitat types were determined using supervised classification of Landsat 7 imagery with 30×30m pixels to generate land cover classifications (Integrated Informatics Inc., 2014).')
+```
+
+**References**
+
+Integrated Informatics Inc.
+2014.
+Sustainable Development and Strategic Science Branch Land cover classification.
+Sustainable Development and Strategic Science, Government of Newfoundland and Labrador, St.
+John’s, NL.

### 2021-06-08 paper/supplement.Rmd
edit caption Fig S1

@@
 `r DT[!is.na(label), .N]` islands are identified and the dominant habitats on these islands are presented in Table S1.
-Note, islands listed as Island 1–5 correspond to those presented in Panel D in Figure 2 and Islands 6 and 7 correspond to those presented in Panel C in Figure 2.
-These numbered islands do not have official names and for ease of interpretation we identified them using numbers.
+Note, numbered islands do not have official names and for ease of interpretation we identified them using numbers.
 

### 2021-06-08 paper/AppendixS1.Rmd
chunk opts, subtitle

--- 2021-06-08 paper/supplement.Rmd
@@
 title: On the marginal value of swimming in woodland caribou
+subtitle: Appendix S1
 author:
@@
 ```{r knitropts, include = FALSE}
-knitr::opts_chunk$set(echo = FALSE,
+knitr::opts_chunk$set(
+  echo = FALSE,
   eval = TRUE,
-                      fig.align = 'center')
+  fig.align = 'center'
+)
 ```
@@
 library(data.table)
+library(ftExtra)
 DT <- readRDS('../output/islandsPoly-with-label.Rds')

### 2021-06-08 paper/AppendixS1.Rmd
switch to flextable

@@
 prop[, (mkpercent) := lapply(.SD, function(x) scales::label_percent(0.01)(x)), .SDcols = mkpercent]
-knitr::kable(prop[order(-`Area (km2)`)], align = 'l',
-             caption = 'Table S1: Summary of primary habitat types for eight islands in the Fogo Island Archipelago.
+```
+
+```{r}
+library(flextable)
+ft <- qflextable(prop[order(-`Area (km2)`)])
+set_caption(ft,
+            'Summary of primary habitat types for eight islands in the Fogo Island Archipelago.
 Habitat types were determined using supervised classification of Landsat 7 imagery (Integrated Informatics Inc., 2014).
@@
 ```{r}
-knitr::kable(fread('../paper/tabs2.csv'), align = 'l',
-             caption = 'Table S2: Summary of common plant species and groups of species in four habitat types on Fogo Island, Newfoundland.
-Habitat types were determined using supervised classification of Landsat 7 imagery with 30×30m pixels to generate land cover classifications (Integrated Informatics Inc., 2014).')
+ft2 <- flextable(fread('../paper/tabs2.csv'))
+
+set_caption(autofit(
+  colformat_md(ft2)),
+  caption = 'Summary of common plant species and groups of species in four habitat types on Fogo Island, Newfoundland.
+Habitat types were determined using supervised classification of Landsat 7 imagery with 30×30m pixels to generate land cover classifications (Integrated Informatics Inc., 2014).'
+)
 ```
+
+\newpage
 
